### 2015-05-01 Notes/N_eff_prior_paper.tex
add folder structure and paper skeleton, lensing noise is now there

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.0
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}

% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

\begin{document}

% Use the \preprint command to place your local institutional report
% number in the upper righthand corner of the title page in preprint mode.
% Multiple \preprint commands are allowed.
% Use the 'preprintnumbers' class option to override journal defaults
% to display numbers if necessary
%\preprint{}

%Title of paper
\title{$N_{\text{eff}}$ priors}

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current
% author. Explanatory text should go in the []'s, actual e-mail
% address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro foreach each type of information

% \affiliation command applies to all authors since the last
% \affiliation command. The \affiliation command should follow the
% other information
% \affiliation can be followed by \email, \homepage, \thanks as well.
\author{}
%\email[]{Your e-mail address}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{}

%Collaboration name if desired (requires use of superscriptaddress
%option in \documentclass). \noaffiliation is required (may also be
%used with the \author command).
%\collaboration can be followed by \email, \homepage, \thanks as well.
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
Try to understand of prior on the $n_{eff}$ parameter. With a Fisher matrix approach.

\end{abstract}

% insert suggested PACS numbers in braces on next line
\pacs{}s
% insert suggested keywords - APS authors don't need to do this
%\keywords{}

%\maketitle must follow title, authors, abstract, \pacs, and \keywords
\maketitle

% body of paper here - Use proper section commands
% References should be done using the \cite, \ref, and \label commands
\section{Introduction}

% Put \label in argument of \section for cross-referencing
%\section{\label{}}
\subsection{}
\subsubsection{}

\section{Theory}

\section{Data}
\section{Results}
\section{Conclusions}

% If in two-column mode, this environment will change to single-column
% format so that long equations can be displayed. Use
% sparingly.
%\begin{widetext}
% put long equation here
%\end{widetext}

% figures should be put into the text as floats.
% Use the graphics or graphicx packages (distributed with LaTeX2e)
% and the \includegraphics macro defined in those packages.
% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz,
% and Frank Mittelbach for instance.
%
% Here is an example of the general form of a figure:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Use the figure* environment if the figure should span across the
% entire page. There is no need to do explicit centering.

% \begin{figure}
% \includegraphics{}%
% \caption{\label{}}
% \end{figure}

% Surround figure environment with turnpage environment for landscape
% figure
% \begin{turnpage}
% \begin{figure}
% \includegraphics{}%
% \caption{\label{}}
% \end{figure}
% \end{turnpage}

% tables should appear as floats within the text
%
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
% The ruledtabular enviroment adds doubled rules to table and sets a
% reasonable default table settings.
% Use the table* environment to get a full-width table in two-column
% Add \usepackage{longtable} and the longtable (or longtable*}
% environment for nicely formatted long tables. Or use the the [H]
% placement option to break a long table (with less control than 
% in longtable).
% \begin{table}%[H] add [H] placement to break table across pages
% \caption{\label{}}
% \begin{ruledtabular}
% \begin{tabular}{}
% Lines of table here ending with \\
% \end{tabular}
% \end{ruledtabular}
% \end{table}

% Surround table environment with turnpage environment for landscape
% table
% \begin{turnpage}
% \begin{table}
% \caption{\label{}}
% \begin{ruledtabular}
% \begin{tabular}{}
% \end{tabular}
% \end{ruledtabular}
% \end{table}
% \end{turnpage}

% Specify following sections are appendices. Use \appendix* if there
% only one appendix.
%\appendix
%\section{}

% If you have acknowledgments, this puts in the proper section head.
%\begin{acknowledgments}
% put your acknowledgments here.
%\end{acknowledgments}

% Create the reference section using BibTeX:
\bibliography{basename of .bib file}

\end{document}

%
% ****** End of file apstemplate.tex ******

### 2015-05-01 code/compute_noisebias.f90

@@ -1,0 +1,73 @@
+PROGRAM compute_noisebias
+! A for computing "noisebias," Nl, of the reconstructed lensing potential:
+!
+! 1/Nl = 1/(2l+1) * sum_{l2=l}^{lmax} * sum_{l3=l2}^{lmax}
+!                 * A_{l1l2l3}^2/[C(l2)C(l3)Delta(l1,l2,l3)].
+!
+! REF: Lewis, Challinor & Hanson, JCAP 03, 018 (2011), arXiv:1101.2234
+! April 25, 2012: E.Komatsu
+  IMPLICIT none
+  character(len=128) :: filename
+  integer :: lmax=4900 ! maximum multipole
+  integer :: lmin=2    ! minimum multipole
+  integer :: i,l,l1,l2,l3,info,dum
+  double precision :: l3min,l3max,gaunt,fourpi,twopi,delta
+  double precision :: Al1l2l3,noisebias
+  double precision, allocatable, dimension(:) :: cl,wigner3j
+  print*,'lmin=',lmin
+  print*,'lmax=',lmax
+  twopi=2d0*3.1415926535d0
+  fourpi=4d0*3.1415926535d0
+! read in lensed Cl
+  allocate(cl(lmax))
+  filename='/Users/alessandromanzotti/Work/Astrophysics/Neff_prior/code/fiducial_lensedcls.dat'
+  print*,'cl file (lensed)= ',trim(filename)
+  open(2,file=filename, status='old')
+  do l=2,lmax
+     read(2,*)dum,cl(l)
+     cl(l)=cl(l)/dble(l)/(dble(l)+1d0)*twopi/(2.726d6)**2d0
+  enddo
+  close(2)
+! begin computing noisebias
+  allocate(wigner3j(2*lmax+1))
+  open(3,file='multipole_noisebias_TT.txt',status='unknown')
+  do l1=lmin,lmax
+     noisebias=0d0
+     do l2=l1,lmax
+        call DRC3JJ(dble(l1),dble(l2),0d0,0d0, &
+             l3min,l3max,wigner3j,2*lmax+1,info)
+        if ((int(l3max).ge.l2)) then
+           do l3=l2,int(l3max)
+              if ((l3.le.lmax).and.(mod(l1+l2+l3,2).eq.0))then
+                 gaunt=wigner3j(l3-int(l3min)+1)/dsqrt(fourpi) &
[…]

### 2015-05-01 code/fisher.py

@@ -1,78 +1,145 @@
+'''
+
+Simple fisher code to test prior effect on N effective neutrino estimates
+
+TODO:
+HIGH: lensing noise
+
+low importance: ini file, automate the data creation process and derivative.
+
+what are this data? documents better
+
+create classes or at least external utilities. Maybe a better way to interface with cosmosis
+
+
+The matrix is going to run an 4 parameters:
+
+CONVENTIONS:
+
+PARAMETER ORDER = h0, ns, As, Neff
+
+'''
+
+__author__ = "Y. Park"
+
+
 import numpy as np
 import math
 import matplotlib.pyplot as plt
+import util
 
-def C(iell,ell,parbin):
+# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
 
-	s = 350*math.sqrt(20626*60*60)/math.sqrt(10**6*0.25*5*365*24*60*60)
-	t = 2/60/180*math.pi
 
-	fac = ell*(ell+1)/2/math.pi
-	fac2 = ell**4/ell/(ell+1)
+def years2sec(years):
+    return years * 365 * 24. * 60. * 60.
 
-	N = s**2 * math.exp( ell * (ell+1) * t**2 / 8 / math.log(2) )
+lmax = 3000
 
-	C = np.array([[dats[iell,1,parbin]/fac+N,dats[iell,2,parbin]/fac,0],\
-		[dats[iell,2,parbin]/fac,dats[iell,3,parbin]/fac+N*2**0.5,0],\
-		[0,0,dats[iell,4,parbin]]])
-	return C
+N_phi_l = np.loadtxt('multipole_noisebias.txt')
 
+
+def C(iell, ell, parbin):
+    '''
+    Given CMB data dats it normalize them anf create a 3x3 matrix
+
+    ell is the multiple
+    iell is the index in the data ell corresponds to
+
+    '''
+
+    # noise definition from the number of observations and time
+    # eq 1 of W.hu et al snowmass paper 10^6 detectors
+    Y = 0.25  # 25%yeld
+    N_det = 10 ** 6  # 1 milion of detectors
+    s = 350 * math.sqrt(20626 * 60 * 60) / math.sqrt(N_det * Y * years2sec(5))  # half sky in arcmin^2
[…]

### 2015-05-01 code/generate_data.py

@@ -1,0 +1,48 @@
+'''
+Script to produced the different Cls with different parameters needed to compute derivative for the Fisher matrices.
+
+It chenge the ini file of CAMB at each step with configparser
+and it then run camb.
+
+f' = -f(x+2h) + 8f(x+h) -8f(x-h)+f(x-2h)
+      ---------------------------------
+                    12h
+For better accuracy the derivative  should be computed with a 5 point stencil algorithms
+
+CONVENTIONS:
+
+PARAMETER ORDER = h0, ns, As, Neff
+
+
+'''
+
+
+import configparser
+import subprocess
+import numpy as np
+
+# deltax to be used with different parameters
+
+# gaps beween  x1 x_-1 these three are used to get the value of the derivative in the middle
+pargaps = np.array([0.002, 0.02, 2e-11, 0.02])  # h0, ns, As, Neff
+
+
+
+config = configparser.ConfigParser()
+configfile='./fiducial.ini'
+
+# run fiducial
+subprocess.call(['/Users/alessandromanzotti/Work/Software/camb2013/camb',configfile])
+
+
+
+
+# config.read(configfile)
+# # config.set('camb','massless_neutrinos',2.046)
+# config.set('camb','output_root','N_eff_{:f}'.format(2.046))
+
+# with open(configfile, 'w') as confile:
+#     config.write(confile)
+
+
+# subprocess.call(['/Users/alessandromanzotti/Work/Software/camb2013/camb',configfile])

### 2015-05-04 code/generate_data.py
data generation seems to work now. To be tested anyway

@@ -20,6 +20,7 @@ PARAMETER ORDER = h0, ns, As, Neff
 import configparser
 import subprocess
 import numpy as np
+import sys
 
 # deltax to be used with different parameters
 
@@ -30,19 +31,47 @@ pargaps = np.array([0.002, 0.02, 2e-11, 0.02])  # h0, ns, As, Neff
 
 config = configparser.ConfigParser()
 configfile='./fiducial.ini'
+config.read(configfile)
 
 # run fiducial
-subprocess.call(['/Users/alessandromanzotti/Work/Software/camb2013/camb',configfile])
-
-
-
-
-# config.read(configfile)
-# # config.set('camb','massless_neutrinos',2.046)
-# config.set('camb','output_root','N_eff_{:f}'.format(2.046))
-
-# with open(configfile, 'w') as confile:
-#     config.write(confile)
-
-
 # subprocess.call(['/Users/alessandromanzotti/Work/Software/camb2013/camb',configfile])
+
+# get fiducial values to figure out where to compute next
+
+h = config.getfloat('camb','hubble')
+As = config.getfloat('camb','scalar_amp(1)')
+ns = config.getfloat('camb','scalar_spectral_index(1)')
+N_eff = config.getfloat('camb','massless_neutrinos') # is it true? what do we want to keep fixed?
+
+
+# generate values to compute Cls
+values = {}
+
+values['hubble'] = pargaps[0] * np.array([-2,-1,1,2]) + h
+values['scalar_amp(1)'] = pargaps[1] * np.array([-2,-1,1,2]) + As
+values['scalar_spectral_index(1)'] = pargaps[2] * np.array([-2,-1,1,2]) + ns
+values['massless_neutrinos'] = pargaps[3] * np.array([-2,-1,1,2]) + N_eff
+
+# start loop on them and generate.
+
+for key, value in values.iteritems():
+    for i in np.arange(0,4):
+        config.set('camb',key,values[key][i])
+        config.set('camb','output_root','./data/run1/'+key+'_{:f}'.format(values[key][i]))
+        configfile_temp = key+'_{:f}'.format(values[key][i]) +'.ini'
+        with open(configfile_temp, 'w') as confile:
+            config.write(confile)
+        # subprocess.call(['/Users/alessandromanzotti/Work/Software/camb2013/camb',configfile_temp])
+
+
+
+
+
+sys.exit()
+config.set('camb','hubble',h)
[…]

### 2015-05-05 code/generate_data.py
add digitize noise curves from figure Fig 1 1402.4108

@@ -64,7 +64,7 @@ for key, value in values.iteritems():
         # subprocess.call(['/Users/alessandromanzotti/Work/Software/camb2013/camb',configfile_temp])
 
 
-
+np.savetxt('wu_cdd_noise_4.txt',np.vstack((np.arange(2,4800),noise_6_fun( np.arange(2,4800))).T))
 
 
 sys.exit()

### 2015-05-10 Notes/N_eff_prior_paper.tex
data generation seems to work now. To be tested anyway
still working on code, plotting routines added, automatic generation of data in place

@@
-%% ****** Start of file apstemplate.tex ****** %
-%%
-%%
-%%   This file is part of the APS files in the REVTeX 4 distribution.
-%%   Version 4.1r of REVTeX, August 2010
-%%
-%%
-%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
-%%
-%%   See the REVTeX 4 README file for restrictions and more information.
-%%
-%
-% This is a template for producing manuscripts for use with REVTEX 4.0
-% Copy this file to another name and then work on that file.
-% That way, you always have this original template file to use.
-%
 % Group addresses by affiliation; use superscriptaddress for long
@@
 %\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
-
+\usepackage{bm}
+\usepackage{amsmath}
 % You should use BibTeX and apsrev.bst for references
@@
 \begin{abstract}
-Try to understand of prior on the $n_{eff}$ parameter. With a Fisher matrix approach.
+Try to understand the effect of prior on the $N_{\text{eff}}$ parameter with a Fisher matrix approach. Where should you invest time and money if you really care about the value of $N_{\text{eff}}$?
 
@@
 % insert suggested PACS numbers in braces on next line
-\pacs{}s
+\pacs{}
 % insert suggested keywords - APS authors don't need to do this
@@
 \section{Theory}
+
+\begin{eqnarray}
+\centering
+	s\ [\;\mu {\rm K.arcmin}\;] \equiv \frac{ {\rm NET}\ [\; \mu{\rm K.}\sqrt{s}\;] \times \sqrt{f_{sky} \ [\;{\rm arcmin}^2\;] }}{ \sqrt {N_{\rm det} \times Y \times \Delta T\ [\;{\rm s }\;]}}.
+	\label{eq:sensitivity_definition}
+\end{eqnarray}
+
+\begin{equation}
+	\centering
+		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle
+	\label{eq:Fij_def}
+\end{equation}
+
+ \begin{eqnarray}
+ 	\centering
+		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
+	\label{eq:cov_definition}
+\end{eqnarray}
+
+\begin{equation}
+\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}
+\end{equation}
+
+\begin{equation}
+F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,fid})^2}, 
+\end{equation}
+
+\begin{eqnarray}
+	T_{\nu} = \left( \frac{4}{11} \right)^{1/3} T_{\gamma} 
+	\label{eq:tnu_propto_tgamma}
+\end{eqnarray}
 

### 2015-05-10 code/fisher.py

@@ -116,7 +116,7 @@ for iell, ell in enumerate(range(2, 5000)):
             print j, cj[1,1]
             print ''
         # sys.exit()
-
+print fisher
 
 d = []
 d2 = []

### 2015-05-10 code/fisher_2.py

@@ -102,7 +102,8 @@ def C(iell, ell, parbin, data):
 # =============================
 lmax = 5000
 N_phi_l = np.loadtxt('multipole_noisebias.txt')
-run_idx = 1
+run_idx = 3
+fsky = 0.5
 # =============================
 
 
@@ -144,10 +145,10 @@ for iell, ell in enumerate(range(2, lmax)):
 
         for j in range(0, 4):
             # computing derivatives.
-            ci = (C(iell, ell, i * 4 + 3,dats) - C(iell, ell, i * 4 + 2,dats)) /2./ pargaps[values.keys()[i]]
-            cj = (C(iell, ell, j * 4 + 3,dats) - C(iell, ell, j * 4 + 2,dats)) /2./ pargaps[values.keys()[j]]
+            # ci = (C(iell, ell, i * 4 + 3,dats) - C(iell, ell, i * 4 + 2,dats)) /2./ pargaps[values.keys()[i]]
+            # cj = (C(iell, ell, j * 4 + 3,dats) - C(iell, ell, j * 4 + 2,dats)) /2./ pargaps[values.keys()[j]]
 
-            print j, cj[1,1]
+            # print j, cj[1,1]
 
             # f' = -f(x+2h) + 8f(x+h) -8f(x-h)+f(x-2h)
                   # ---------------------------------
@@ -177,7 +178,7 @@ d3 = []
 for i in np.arange(-3, -1, 0.1):
     fisher1 = fisher.copy()
     # Cicle on H0 priors
-    fisher1[1, 1] += 1 / (10 ** i * 0.673) ** 2
+    fisher1[1, 1] += 1 / (10 ** i * 67.04346) ** 2
     # Invert and get Neff error with these priors
 
     d.append(math.sqrt(np.linalg.inv(fisher1)[0, 0]))
@@ -185,7 +186,7 @@ for i in np.arange(-3, -1, 0.1):
     fisher2 = fisher.copy()
     # Cicle on H0 priors
 
-    fisher2[1, 1] += 1 / (10 ** i * 0.673) ** 2
+    fisher2[1, 1] += 1 / (10 ** i * 67.04346) ** 2
 
     # add 1% prior on ns
     fisher2[2, 2] += 1 / (0.01 * 0.96) ** 2
@@ -197,7 +198,7 @@ for i in np.arange(-3, -1, 0.1):
     fisher3 = fisher.copy()[[0, 1], :][:, [0, 1]]
[…]

### 2015-05-10 code/generate_data.py

@@ -43,7 +43,7 @@ As = config.getfloat('camb', 'scalar_amp(1)')
 ns = config.getfloat('camb', 'scalar_spectral_index(1)')
 N_eff = config.getfloat('camb', 'massless_neutrinos')  # is it true? what do we want to keep fixed?
 
-np.savetxt("./data/run3/fiducial_pars.txt", np.array([h, ns, As, N_eff]))
+np.savetxt("./data/run2/fiducial_pars.txt", np.array([h, ns, As, N_eff]))
 
 # generate values to compute Cls
 values = {}
@@ -58,12 +58,12 @@ pargaps_dict['scalar_spectral_index(1)'] = pargaps[1]
 pargaps_dict['scalar_amp(1)'] = pargaps[2]
 pargaps_dict['massless_neutrinos'] = pargaps[3]
 # save a pickle of data values to be re-used
-with open("./data/run3/grid_values.p", "wb") as output_file:
+with open("./data/run2/grid_values.p", "wb") as output_file:
     pickle.dump(values, output_file)
 
 # save datagaps
 
-with open("./data/run3/par_gaps.p", "wb") as output_file:
+with open("./data/run2/par_gaps.p", "wb") as output_file:
     pickle.dump(pargaps_dict, output_file)
 
 

### 2015-05-10 code/plotting/plot_deriv.py

@@ -182,6 +182,8 @@ for i, key in enumerate(par_gaps.keys()):
         np.nan_to_num(dats[:, 1, 4 * i + 3] - dats[:, 1, 4 * i + 2]) / (2 * par_gaps[key])), linewidth=1, color='b')
     plot3 = plt.loglog(data_fid[:, 0], np.abs(
         np.nan_to_num(dats[:, 1, 4 * i + 4] - dats[:, 1, 4 * i + 1]) / (4 * par_gaps[key])), linewidth=1, color='r')
+    plot3 = plt.loglog(data_fid[:, 0], np.abs(
+        np.nan_to_num(-dats[:, 1, 4 * i + 4] + 8. * dats[:, 1, 4 * i + 3] - 8. * dats[:, 1, 4 * i + 2] + dats[:, 1, 4 * i + 1]) / (12 * par_gaps[key])), linewidth=1 ,marker='o')
     fg.tight_layout(pad=0.1)
 
     legend = ax1.legend()
@@ -200,6 +202,9 @@ for i, key in enumerate(par_gaps.keys()):
         np.nan_to_num(dats[:, 2, 4 * i + 3] - dats[:, 2, 4 * i + 2]) / (2 * par_gaps[key])), linewidth=1, color='b')
     plot3 = plt.loglog(data_fid[:, 0], np.abs(
         np.nan_to_num(dats[:, 2, 4 * i + 4] - dats[:, 2, 4 * i + 1]) / (4 * par_gaps[key])), linewidth=1, color='r')
+    plot3 = plt.loglog(data_fid[:, 0], np.abs(
+        np.nan_to_num( -dats[:, 2, 4 * i + 4] + 8. * dats[:, 2, 4 * i + 3] - 8. * dats[:,2 , 4 * i + 2] + dats[:, 2, 4 * i + 1]) / (12 * par_gaps[key])) , linewidth=1, marker='o')
+
     fg.tight_layout(pad=0.1)
 
     legend = ax1.legend()
@@ -219,6 +224,12 @@ for i, key in enumerate(par_gaps.keys()):
     plot3 = plt.loglog(data_fid[:, 0], np.abs(
         np.nan_to_num(dats[:, 5, 4 * i + 4] - dats[:, 5, 4 * i + 1]) / (4 * par_gaps[key])), linewidth=1, color='r')
     fg.tight_layout(pad=0.1)
[…]

### 2015-05-24 code/compute_lens_reconstruction_noise_levels.py
now plotting also against tau prior

@@ -8,6 +8,7 @@
 # full-sky and flat-sky limits for comparison.
 
 # TODO load details from ini
+#  we can include recursive techniques from hirata seljack the idea is that flbb is now clbb-clbb lens obtained from the ebphi estimator.
 
 import numpy as np
 import pylab as pl
@@ -152,6 +153,7 @@ you need a 5x5 wit all the estimator and cross correlation between them
     # invert at a given l
     for ell in np.arange(1, ells):
         # invert at a given l
+        print ell,temp[:, :, ell],np.linalg.inv(temp[:, :, ell]).flatten()
         nlpp_mv[ell] = 1. / np.sum(np.linalg.inv(temp[:, :, ell]).flatten())
 
     return nlpp_mv

### 2015-05-24 Notes/N_eff_prior_paper.tex
new plots and Notes

@@
 \usepackage{amsmath}
+\usepackage{graphicx}
+
 % You should use BibTeX and apsrev.bst for references
@@
 \begin{document}
+\graphicspath{{../images/}}
 
@@
 \begin{abstract}
-Try to understand the effect of prior on the $N_{\text{eff}}$ parameter with a Fisher matrix approach. Where should you invest time and money if you really care about the value of $N_{\text{eff}}$?
+Try to understand the effect of priors on different parameters on the $N_{\text{eff}}$ errors with a Fisher matrix approach.
+Where should you invest time and money if you really care about the value of $N_{\text{eff}}$?
 
@@
 \section{Theory}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[scale=0.6]{PS_phi_with_noise.pdf}
+\caption{Lensing potential power spectrum for our fiducial cosmology together with the lensing reconstruction noise $N^{\phi}$ used in this work.}
+\label{fig:phi-cl-noise}
+\end{center}
+\end{figure}
 
@@
 \section{Results}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[scale=0.6]{h0_fisher.pdf}
+\caption{}
+\label{fig:phi-cl-noise}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[scale=0.6]{tau_fisher.pdf}
+\caption{}
+\label{fig:phi-cl-noise}
+\end{center}
+\end{figure}
+
 \section{Conclusions}

### 2015-05-24 code/plotting/plot_sigmaN_prior_H0.py

@@ -60,25 +60,11 @@ def noise(Y, N_det, beam_arc, ell):
 
 # =============================
 
-run_idx = 3
 
+sigma1per = np.loadtxt('../output/sigma_H0_1percent.txt')
+sigma_no = np.loadtxt('../output/sigma_H0_noPrior.txt')
+sigma_perfect = np.loadtxt('../output/sigma_H0_perfect_prior.txt')
 
-# READ PARAMS
-dats = np.genfromtxt('../data/run{}/fiducial_lenspotentialcls.dat'.format(run_idx))
-fid = np.genfromtxt('../data/run{}/fiducial_pars.txt'.format(run_idx))
-# load parameter grid dictionary. The format is a pickle
-values = pickle.load(open('../data/run{}/grid_values.p'.format(run_idx), "rb"))
-par_gaps = pickle.load(open('../data/run{}/par_gaps.p'.format(run_idx), "rb"))
-fac = (dats[:, 0] * (dats[:, 0] + 1.) / 2. / np.pi) / (7.4311 * 10 ** 12)
-noise_l = noise(0.25, 10e6, 2., dats[:, 0])
-noise1 = np.loadtxt('../data/noise/wu_cdd_noise_6.txt')
-noise2 = np.loadtxt('../data/noise/wu_cdd_noise_5.txt')
-noise3 = np.loadtxt('../data/noise/wu_cdd_noise_4.txt')
-noise_tt = np.loadtxt('../TT_lensing_noise.txt')
-noise_ee = np.loadtxt('../EE_lensing_noise.txt')
-noise_eb = np.loadtxt('../EB_lensing_noise.txt')
-noise_te = np.loadtxt('../TE_lensing_noise.txt')
-noise_tb = np.loadtxt('../TB_lensing_noise.txt')
 
 # ============================================
 # ============================================
@@ -194,29 +180,21 @@ plt.rcParams['legend.handletextpad'] = 0.3
 
 # plot2 = plt.semilogx(data_fid[:,0] , 10.*np.nan_to_num((dats[:,1,]- dats[:,1,] )/data_fid[:,1]),linewidth=1, color='b',label=r'$C^{T}$')
 
-plot_type = plt.semilogy
-plot2 = plot_type(dats[:, 0], dats[:, 5], linewidth=1, label=r'$\phi$')
[…]

### 2015-05-24 code/plotting/plot_sigmaN_prior_tau.py

@@ -60,25 +60,11 @@ def noise(Y, N_det, beam_arc, ell):
 
 # =============================
 
-run_idx = 3
 
+sigma1per = np.loadtxt('../output/sigma_tau_1percent.txt')
+sigma_no = np.loadtxt('../output/sigma_tau_noPrior.txt')
+sigma_perfect = np.loadtxt('../output/sigma_tau_perfect_prior.txt')
 
-# READ PARAMS
-dats = np.genfromtxt('../data/run{}/fiducial_lenspotentialcls.dat'.format(run_idx))
-fid = np.genfromtxt('../data/run{}/fiducial_pars.txt'.format(run_idx))
-# load parameter grid dictionary. The format is a pickle
-values = pickle.load(open('../data/run{}/grid_values.p'.format(run_idx), "rb"))
-par_gaps = pickle.load(open('../data/run{}/par_gaps.p'.format(run_idx), "rb"))
-fac = (dats[:, 0] * (dats[:, 0] + 1.) / 2. / np.pi) / (7.4311 * 10 ** 12)
-noise_l = noise(0.25, 10e6, 2., dats[:, 0])
-noise1 = np.loadtxt('../data/noise/wu_cdd_noise_6.txt')
-noise2 = np.loadtxt('../data/noise/wu_cdd_noise_5.txt')
-noise3 = np.loadtxt('../data/noise/wu_cdd_noise_4.txt')
-noise_tt = np.loadtxt('../TT_lensing_noise.txt')
-noise_ee = np.loadtxt('../EE_lensing_noise.txt')
-noise_eb = np.loadtxt('../EB_lensing_noise.txt')
-noise_te = np.loadtxt('../TE_lensing_noise.txt')
-noise_tb = np.loadtxt('../TB_lensing_noise.txt')
 
 # ============================================
 # ============================================
@@ -194,29 +180,21 @@ plt.rcParams['legend.handletextpad'] = 0.3
 
 # plot2 = plt.semilogx(data_fid[:,0] , 10.*np.nan_to_num((dats[:,1,]- dats[:,1,] )/data_fid[:,1]),linewidth=1, color='b',label=r'$C^{T}$')
 
-plot_type = plt.semilogy
-plot2 = plot_type(dats[:, 0], dats[:, 5], linewidth=1, label=r'$\phi$')
[…]

### 2015-05-25 Notes/N_eff_prior_paper.tex
change notes

@@
 
+\begin{itemize}
+\item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet.
+\item We are varying $\tau ~ n_{s} ~ A_{s} ~ N_{\rm eff} ~ H_{0}$. The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, $\Omega_{\nu}h^{2}=0.0009$
+\item Next TODO: Check for bugs and errors code in its early stages. PCA? what prior is more important? Full MCMC should not be extremely hard with cosmosis.
+\end{itemize}
+
 \end{abstract}
@@
 \label{fig:phi-cl-noise}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[scale=0.6]{PS_with_noise.pdf}
+\caption{CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
+\label{fig:cmb-cl-noise}
 \end{center}

### 2015-05-25 code/plotting/plot_sigmaN_prior_H0.py

@@ -184,6 +184,7 @@ plot_type = plt.semilogx
 plot2 = plot_type(10 ** np.arange(-3, -1, 0.1), np.array(sigma_no) * 100., linewidth=1,label='No Priors')
 plot2 = plot_type(10 ** np.arange(-3, -1, 0.1), np.array(sigma_perfect) * 100., linewidth=1,label='Perfect Priors')
 plot2 = plot_type(10 ** np.arange(-3, -1, 0.1), np.array(sigma1per) * 100., linewidth=1,label=r'1$\%$ Priors')
+plot2=plt.axvline(0.01,linewidth=1, ls= '--')
 
 
 legend = ax1.legend()

### 2015-05-27 code/fisher_2.py
adding output folder

@@ -205,17 +205,18 @@ for i in np.arange(-3, -1, 0.1):
     # Invert and get Neff error with these priors
     d2.append(math.sqrt(np.linalg.inv(fisher2)[1, 1]))
 
+    fisher3 = fisher.copy()[[0, 1], :][:, [0, 1]]
+
     fisher3[0, 0] += 1 / (10 ** i * 67.04346) ** 2
 
-    fisher3 = fisher.copy()[[0, 1], :][:, [0, 1]]
     # Cicle on H0 priors
 
     # Invert and get Neff error with these priors
     d3.append(math.sqrt(np.linalg.inv(fisher3)[1, 1]))
 
-np.savetxt('sigma_H0_1percent.txt',d)
-np.savetxt('sigma_H0_noPrior.txt',d2)
-np.savetxt('sigma_H0_perfect_prior.txt',d3)
+np.savetxt('output/sigma_H0_1percent.txt',d)
+np.savetxt('output/sigma_H0_noPrior.txt',d2)
+np.savetxt('output/sigma_H0_perfect_prior.txt',d3)
 
 plt.clf()
 plt.plot(10 ** np.arange(-3, -1, 0.1), np.array(d) * 100., label='No Priors')
@@ -272,9 +273,9 @@ for i in np.arange(-3, -1, 0.1):
     # Invert and get Neff error with these priors
     d3.append(math.sqrt(np.linalg.inv(fisher3)[1, 1]))
 
-np.savetxt('sigma_tau_1percent.txt',d)
-np.savetxt('sigma_tau_noPrior.txt',d2)
-np.savetxt('sigma_tau_perfect_prior.txt',d3)
+np.savetxt('output/sigma_tau_1percent.txt',d)
+np.savetxt('output/sigma_tau_noPrior.txt',d2)
+np.savetxt('output/sigma_tau_perfect_prior.txt',d3)
 
 plt.clf()
 plt.plot(10 ** np.arange(-3, -1, 0.1), np.array(d) * 100., label='No Priors')

### 2015-05-27 code/fisher_2.py
removing junk

@@ -323,9 +323,9 @@ for i in np.arange(-3, -1, 0.1):
     # Invert and get Neff error with these priors
     d3.append(math.sqrt(np.linalg.inv(fisher3)[1, 1]))
 
-np.savetxt('sigma_ns_1percent.txt',d)
-np.savetxt('sigma_ns_noPrior.txt',d2)
-np.savetxt('sigma_ns_perfect_prior.txt',d3)
+np.savetxt('output/sigma_ns_1percent.txt',d)
+np.savetxt('output/sigma_ns_noPrior.txt',d2)
+np.savetxt('output/sigma_ns_perfect_prior.txt',d3)
 
 print 'finally how much constraint on parameters without prior?'
 print ''

### 2015-05-28 code/generate_data.py
split commit to add omeganu2

@@ -36,7 +36,7 @@ configfile = './fiducial.ini'
 config.read(configfile)
 
 # run fiducial
-subprocess.call(['/Users/alessandromanzotti/Work/Software/camb2013/camb', configfile])
+subprocess.call(['/home/manzotti/local/camb2013/camb', configfile])
 
 # get fiducial values to figure out where to compute next
 
@@ -45,6 +45,11 @@ tau = config.getfloat('camb', 're_optical_depth')
 As = config.getfloat('camb', 'scalar_amp(1)')
 ns = config.getfloat('camb', 'scalar_spectral_index(1)')
 N_eff = config.getfloat('camb', 'massless_neutrinos')  # is it true? what do we want to keep fixed?
+omnuh2 = config.getfloat('camb', 'omnuh2')
+# Load it for later
+omch2 = config.getfloat('camb', 'omch2')
+
+pargaps = np.array([1., 0.05, 5e-11, 0.08, 0.00462259,omnuh2*0.40])
 
 
 fid = {}
@@ -53,14 +58,16 @@ fid['scalar_spectral_index(1)'] =  ns
 fid['scalar_amp(1)'] = As
 fid['massless_neutrinos'] =  N_eff
 fid['re_optical_depth'] = tau
+fid['omnuh2'] = omnuh2
+
 fid = collections.OrderedDict(sorted(fid.items(), key=lambda t: t[0]))
 
-with open("./data/run3/fid_values.p", "wb") as output_file:
+with open("./data/run2/fid_values.p", "wb") as output_file:
     pickle.dump(fid, output_file)
 
 # KEEP THIS IN THE ALPABETHICAL ORDER OR USE THE DICT
 
-np.savetxt("./data/run3/fiducial_pars.txt", np.array([h, N_eff, tau, As, ns]))
+np.savetxt("./data/run2/fiducial_pars.txt", np.array([h, N_eff, tau, As, ns]))
 print np.array([h, ns, As, N_eff, tau])
 
 
@@ -81,6 +88,8 @@ values['scalar_spectral_index(1)'] = pargaps[1] * np.array([-2, -1, 1, 2]) + ns
 values['scalar_amp(1)'] = pargaps[2] * np.array([-2, -1, 1, 2]) + As
 values['massless_neutrinos'] = pargaps[3] * np.array([-2, -1, 1, 2]) + N_eff
[…]

### 2015-05-28 code/plotting/plot_deriv.py

@@ -23,19 +23,8 @@ from matplotlib.ticker import MaxNLocator  # needed for tick
 
 # READ DATA
 
-data0 = np.genfromtxt('../data/dat0.txt', dtype=float)
-data1 = np.genfromtxt('../data/dat1.txt', dtype=float)
-data2 = np.genfromtxt('../data/dat2.txt', dtype=float)
-data3 = np.genfromtxt('../data/dat3.txt', dtype=float)
-data4 = np.genfromtxt('../data/dat4.txt', dtype=float)
-data5 = np.genfromtxt('../data/dat5.txt', dtype=float)
-data6 = np.genfromtxt('../data/dat6.txt', dtype=float)
-data7 = np.genfromtxt('../data/dat7.txt', dtype=float)
-data8 = np.genfromtxt('../data/dat8.txt', dtype=float)
 
-# =============================
-
-run_idx = 3
+run_idx = 2
 
 
 # READ PARAMS
@@ -54,7 +43,6 @@ for key, value in values.iteritems():
         dats = np.dstack((dats, newdat))
 
 
-pargaps = np.array([0.002, 0.02, 2e-11, 0.02])  # h0, ns, As, Neff
 
 
 # ============================================
@@ -62,7 +50,7 @@ pargaps = np.array([0.002, 0.02, 2e-11, 0.02])  # h0, ns, As, Neff
 # ============================================
 
 
-print "read an array of size=", data1.shape
+print "read an array of size=", dats.shape
 
 
 # ============================================
@@ -171,19 +159,25 @@ label['massless_neutrinos'] = 'N_{eff}'
 label['hubble'] = 'H_{0}'
 label['scalar_amp(1)'] = 'A_{s}'
 label['scalar_spectral_index(1)'] = 'n_{s}'
+label['omnuh2'] = '\Omega'
+label['re_optical_depth'] = '\tau'
 
 
 for i, key in enumerate(par_gaps.keys()):
     print key, i
-    plot2 = plt.loglog(data_fid[:, 0], np.abs(
+    plot2 = plt.loglog(dats[:, 0, 0], np.abs(
[…]

### 2015-06-22 Notes/N_eff_prior_paper.tex
adding new marginalization technique

@@
 \item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet.
-\item We are varying $\tau ~ n_{s} ~ A_{s} ~ N_{\rm eff} ~ H_{0}$. The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, $\Omega_{\nu}h^{2}=0.0009$
+\item We are varying $\tau, ~ n_{s}, ~ A_{s}, ~ N_{\rm eff}, ~ H_{0}, w,~\Omega_{\nu}h^{2},~\Omega_{c}h^{2},~\Omega_{b}h^{2}$.
+The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, $\Omega_{\nu}h^{2}=0.0009$
 \item Next TODO: Check for bugs and errors code in its early stages. PCA? what prior is more important? Full MCMC should not be extremely hard with cosmosis.
@@
 
+\subsection{Fisher Matrix formalism}
+Copied from paper change it for pubblication:
+
+To avoid any possible numerical instability in the marginalising procedure, we calculate the the Fisher matrix marginalised over the standard LCDM parameters following (Albrecht et al. 2009)
+\begin{equation}
+G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
+\end{equation}
+where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
+marginal\}$; therefore, $F^{\phi\phi}$ is the block of the total Fisher matrix containing the parameters we want to constrain, whilst $F^{\psi\psi}$ is the nuisance-parameter Fisher sub-matrix.
+Here, $\Lambda$ is the diagonal matrix whose elements are the eigenvalues of $F^{\psi\psi}$, whilst U is the orthogonal matrix diagonalising $F^{\psi\psi}$.
+By using Eq., our marginalising procedure is more stable, since degeneracies in $F^{\phi\phi}$ are properly propagated to G with no instabilities, and we do not even worry about a possibly ill-conditioned $F^{\phi\phi}$ sub-matrix, since we check its stability on the fly by the diagonalisation.
+
 \begin{figure}[htbp]
@@
 \includegraphics[scale=0.6]{PS_with_noise.pdf}
-\caption{CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
+\caption{CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.
+They are basically equivalent to a cosmic variance limited experiment with $l_{max}\sim 3000$}
 \label{fig:cmb-cl-noise}
@@
 \section{Data}
+
+\begin{table}[htdp]
+\caption{Fiducial values used.}
+\begin{center}
+\begin{tabular}{|c|c|}
+\hline
+$H_{0}$ &\\
+\hline
+\hline
+$\tau$ &  \\
+\hline
+
+\hline
+$A_{s}$ & \\
+\hline
+
+\hline
+$n_{s}$ & \\
+\hline
+
+\hline
+$N_{eff}$ & \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
+
 \section{Results}
+
+\begin{table}[htdp]
+\caption{How well we do constrain separate parameters with this data without any external prior?}
+\begin{center}
+\begin{tabular}{|c|c|}
+\hline
+$H_{0}$ & 0.32\% \\
+\hline
+\hline
+$\tau$ & 2.26 \%\\
+\hline
+
+\hline
+$A_{s}$ & 0.37 \%\\
+\hline
+
+\hline
+$n_{s}$ & 0.25 \%\\
+\hline
+
+\hline
+$N_{eff}$ & 0.9 \%\\
+\hline
+
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
 

### 2015-06-22 code/plotting/plot_deriv.py
imgages and plotting

@@ -1,3 +1,15 @@
+'''
+Plot scipt to plot the derivatives of CMB specctra respect to all the parameters of the Fisher matrix.
+
+Different lines corresponde to different gaps and techniques.
+
+
+
+'''
+
+
+
+
 import numpy as np
 import pickle as pickle
 #import sys
@@ -108,10 +120,7 @@ plt.rcParams['ytick.major.size'] = 5
 plt.rcParams['ytick.minor.width'] = 0.13
 plt.rcParams['ytick.minor.size'] = 2.8
 
-# fg = plt.figure(figsize=fig_dims)
-fg = plt.figure(figsize=(8, 8))
 
-ax1 = plt.subplot2grid((1, 1), (0, 0))  # , rowspan=2)
 #ax2 = plt.subplot2grid((1,2), (0, 1))
 #ax3 = plt.subplot2grid((2,2), (1, 1))
 
@@ -119,7 +128,7 @@ ax1 = plt.subplot2grid((1, 1), (0, 0))  # , rowspan=2)
 # set colors and lines
 lines = ["-", "--", "-.", ":"]
 linecycler = cycle(lines)
-ax1.set_color_cycle(['#38197A', '#FC7D00', '#318B10', '#B61911'])
+# ax1.set_color_cycle(['#38197A', '#FC7D00', '#318B10', '#B61911'])
 colorcycler = cycle(['#38197A', '#FC7D00', '#318B10', '#B61911'])
 
 # ============================================
@@ -160,18 +169,26 @@ label['massless_neutrinos'] = 'N_{eff}'
 label['hubble'] = 'H_{0}'
 label['scalar_amp(1)'] = 'A_{s}'
 label['scalar_spectral_index(1)'] = 'n_{s}'
-label['omnuh2'] = '\Omega'
-label['re_optical_depth'] = '\tau'
+label['omnuh2'] = r'\Omega_{\nu}'
+label['re_optical_depth'] = r'~\tau'
+label['ombh2'] = '\Omega_{b}'
+label['omch2'] = '\Omega_{c}'
+label['w'] = 'w'
+
+
+
+# fg = plt.figure(figsize=fig_dims)
 
 
 for i, key in enumerate(par_gaps.keys()):
     print key, i
-    plot2 = plt.loglog(dats[:, 0, 0], np.abs(
[…]

### 2015-06-22 code/util.py
util module fixed

@@ -1,57 +1,0 @@
-'''
-
-Simple fisher code to test prior effect on N effective neutrino estimates
-
-TODO:
-
-what are this data? documents better
-
-create classes or at least external utilities. Maybe a better way to interface with cosmosis
-
-
-Noise inspired by W. Hu et al 1402.4108 SNOWMASS paper
-
-TODO:
-
-noise calculatiions routines
-
-'''
-
-__author__ = "Y. Park"
-
-
-import numpy as np
-import math
-import matplotlib.pyplot as plt
-
-import os, copy, glob
-
-# Simple noise generation
-
-def bl(fwhm_arcmin, lmax):
-    """ returns the map-level transfer function for a symmetric Gaussian beam.
-         * fwhm_arcmin      - beam full-width-at-half-maximum (fwhm) in arcmin.
-         * lmax             - maximum multipole.
-    """
-    ls = np.arange(0, lmax+1)
-    return np.exp( -(fwhm_arcmin * np.pi/180./60.)**2 / (16.*np.log(2.)) * ls*(ls+1.) )
-
-def nl(noise_uK_arcmin, fwhm_arcmin, lmax):
-    """ returns the beam-deconvolved noise power spectrum in units of uK^2 for
-          * noise_uK_arcmin - map noise level in uK.arcmin
-          * fwhm_arcmin     - beam full-width-at-half-maximum (fwhm) in arcmin.
-          * lmax            - maximum multipole.
-    """
-    return (noise_uK_arcmin * np.pi/180./60.)**2 / bl(fwhm_arcmin, lmax)**2
-
-def noise_uK_arcmin(noise_uK_arcmin, fwhm_arcmin, lmax):
-    """ returns noise_uK_arcmin given the number of detectors and the time of obsrvation
-    """
-    return (noise_uK_arcmin * np.pi/180./60.)**2 / bl(fwhm_arcmin, lmax)**2
-
-
-
-
-
-
-#

### 2015-06-22 code/utils.py

@@ -6,6 +6,27 @@ import pickle
 import sys
 
 
+def bl(fwhm_arcmin, lmax):
+    """ returns the map-level transfer function for a symmetric Gaussian beam.
+         * fwhm_arcmin      - beam full-width-at-half-maximum (fwhm) in arcmin.
+         * lmax             - maximum multipole.
+    """
+    ls = np.arange(0, lmax+1)
+    return np.exp( -(fwhm_arcmin * np.pi/180./60.)**2 / (16.*np.log(2.)) * ls*(ls+1.) )
+
+def nl(noise_uK_arcmin, fwhm_arcmin, lmax):
+    """ returns the beam-deconvolved noise power spectrum in units of uK^2 for
+          * noise_uK_arcmin - map noise level in uK.arcmin
+          * fwhm_arcmin     - beam full-width-at-half-maximum (fwhm) in arcmin.
+          * lmax            - maximum multipole.
+    """
+    return (noise_uK_arcmin * np.pi/180./60.)**2 / bl(fwhm_arcmin, lmax)**2
+
+def noise_uK_arcmin(noise_uK_arcmin, fwhm_arcmin, lmax):
+    """ returns noise_uK_arcmin given the number of detectors and the time of obsrvation
+    """
+    return (noise_uK_arcmin * np.pi/180./60.)**2 / bl(fwhm_arcmin, lmax)**2
+
 def load_data(run_idx, lensed, values):
     '''
     build the dats matrix of data used in the fisher code. If lensed it is composed by lesned CMB cls + CMB lensing cls like cldd clde cldt
@@ -157,8 +178,10 @@ def fisher_marginalize(fisher, marginalize_parameters_list, fid):
     # s array of eigenvalues
     Fqr = fisher[indeces_to_keep, :][:, indeces_to_marg]
     U, s, V = linalg.svd(Frr, full_matrices=True)
+    print 'marginalize matrix eig', s
     s_inv = np.diag(1/s)
     G = Fqq - np.dot( (np.dot(Fqr,U)) , np.dot(s_inv,((np.dot(Fqr,U)).T)))
+    return G
 
 class Fisher_matrix(object):
 

### 2015-08-05 Notes/N_eff_prior_paper.tex
working on Notes

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
+\documentclass[aps,prd,preprint,groupedaddress]{revtex4-1}
 %\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
@@
 % below if necessary.
-%\bibliographystyle{apsrev4-1}
+\bibliographystyle{apsrev4-1}
 
@@
 %Title of paper
-\title{$N_{\text{eff}}$ priors}
+\title{The importance of external priors for the S4 generation of CMB experiments.}
 
-% repeat the \author .. \affiliation  etc. as needed
-% \email, \thanks, \homepage, \altaffiliation all apply to the current
-% author. Explanatory text should go in the []'s, actual e-mail
-% address or url should go in the {}'s for \email and \homepage.
-% Please use the appropriate macro foreach each type of information
-
-% \affiliation command applies to all authors since the last
-% \affiliation command. The \affiliation command should follow the
-% other information
-% \affiliation can be followed by \email, \homepage, \thanks as well.
-\author{}
-%\email[]{Your e-mail address}
-%\homepage[]{Your web page}
-%\thanks{}
-%\altaffiliation{}
-\affiliation{}
-
-%Collaboration name if desired (requires use of superscriptaddress
-%option in \documentclass). \noaffiliation is required (may also be
-%used with the \author command).
-%\collaboration can be followed by \email, \homepage, \thanks as well.
-%\collaboration{}
-%\noaffiliation
+\author{A. Manzotti}
+%\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+\author{S. Dodelson}
+%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
 
@@
 \begin{abstract}
-Try to understand the effect of priors on different parameters on the $N_{\text{eff}}$ errors with a Fisher matrix approach.
-Where should you invest time and money if you really care about the value of $N_{\text{eff}}$?
+The next generation of CMB experiments is expected to drastically improve our understanding of neutrino physics, inflation, Dark Matter and Dark Energy.
+Indeed they will measure with cosmic variance limited precision the temperature and polarization mode together with a lensing potential reconstruction with signal to noise bigger than one on several scales.
+Detailed forecasts have been done and will tremendously help to plan and motivate this next generation of CMB experiments.
+However we think that theoretical priors coming from external non-CMB experiments were not fully analyzed.
+For this reason in this work we use a fisher-matrix technique to test how improving on external priors can lead to improvement in S4 CMB parameters constraint.
+We find that..
 
 \begin{itemize}
-\item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet.
+\item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet. Needed?
 \item We are varying $\tau, ~ n_{s}, ~ A_{s}, ~ N_{\rm eff}, ~ H_{0}, w,~\Omega_{\nu}h^{2},~\Omega_{c}h^{2},~\Omega_{b}h^{2}$.
-The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, $\Omega_{\nu}h^{2}=0.0009$
-\item Next TODO: Check for bugs and errors code in its early stages. PCA? what prior is more important? Full MCMC should not be extremely hard with cosmosis.
+The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, which corresponds to $\Omega_{\nu}h^{2}=0.0009$.
+\item Next TODO: Check for bugs and errors code (DONE look at Zhen test).
+PCA?
+what prior is more important?
+Full MCMC should not be extremely hard with cosmosis (I implemented a toy likelihood analysis to be improved).
 \end{itemize}
@@
 % References should be done using the \cite, \ref, and \label commands
-\section{Introduction}
+\section{Introduction \label{sec:intro}}
 
-% Put \label in argument of \section for cross-referencing
-%\section{\label{}}
-\subsection{}
-\subsubsection{}
+Since the first results coming from the COBE experiments great results.
 
-\section{Theory}
+Far from finished S4 experiment will measure
+
+This will give us amazing information on..
+
+Forecast is important \cite{wu:2014} ... we think that also a detailed study of priors will be important in planning new experiment.
+
+For this reason we...
+
+This paper is organized as follow:
+
+\section{Assumption and methods \label{sec:intro} }
+We start this section by introducing the Fisher matrix formalism a simple but powerful technique widely used to predict the constrains on the parameters of the assumed cosmological model with future experiments.
+We then specify the model we used in this work and the fiducial parameters we assumed.
+We adopt a standard $\Lambda$CDM model with basic possible extensions with parameters taken from recent CMB+BAO+SN fit.
 
 \subsection{Fisher Matrix formalism}
-Copied from paper change it for pubblication:
+The starting point to estimate errors on cosmological parameters  is the Bayes theorem that relates the likelihood of measuring a set of data given the parameters of the model to what we want: the posterior probability of those parameters given the data.
+These are related by the priors probability of the parameters $P(\theta)$ through:
+\begin{equation}
+\mathcal{L}(\theta|d)\propto \mathcal{L}(d|\theta)P(\theta),
+\end{equation}
 
-To avoid any possible numerical instability in the marginalising procedure, we calculate the the Fisher matrix marginalised over the standard LCDM parameters following (Albrecht et al. 2009)
+where,as usual we have neglect the probability of the data itself.
+These general framework can be simplified in the context of cosmology by focusing on small perturbations around the maximum of a Gaussian likelihood. 
+Indeed we will assume that the likelihood can be approximated as Gaussian.
+While this approximate the true likelihood most of the time the opposite as been found in other cases \cite{2012JCAP...09..009W}.
+Even if other methods of analysis have been proposed \cite{2006JCAP...10..013P,2006astro.ph..9591A} this is still the standard assumption when future constrains have been forecasted \cite{wu:2014}.
+Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$.
+A consequence is that we are then able to estimate the uncertainties on those parameters by a simple analysis of the likelihood curvature around the fiducial values.
+As usual we define the fisher matrix elements as:
+\begin{equation}
+	\centering
+		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle,
+	\label{eq:Fij_def}
+\end{equation}
+where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters array that, by definition, maximize the likelihood.
+We refer the reader for a detailed explanation of the fact that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$ with in our case $X,Y\in\{T,E,\phi\}$ the Fisher matrix can be rewritten as:
+\begin{equation}
+ F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right)
+ \end{equation}
+ with $\boldsymbol{C}_\ell$ encapsulating the usual power spectra in a matrix structure:
+ \begin{eqnarray}
+ 	\centering
+		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
+	\label{eq:cov_definition}
+\end{eqnarray}
+
+The power of this method come from the fact that, given the Cramer-Rao inequality we can get the errors on parameter $i$ marginalized over all the other parameters simply inverting the Fisher matrix as:
+\begin{equation}
+\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}.
+\label{eq:cramer-rao}
+\end{equation}
+Approaching the forecast problem using a Fisher matrix introduce some technical difficulties together with the previously mentioned assumption that the likelihood is gaussian.
+Indeed it is known that the more the paramters used in the analysis the easier it is to face numerical isseus.
+Fisher matrix are can be ill-conditioned: a small change in the fisher matrix led to a big change in its inverse.
+Given \refeq{cramer-rao} this can be a problem for error estimation.
+While we understand this possible concern we carefully compare our results to the literature  \footnote{we thank the authors of \cite{pan:2015} for their collaboration.}
+Furthermore we use the same technique used in \cite{2006astro.ph..9591A} to avoid any possible numerical instability in the marginalising procedure.
+This allow not to invert the entire matrix when we want to marginalize over a set of parameters minimizin numerical issue and conserving parameters degeneracies as much as possible.
+
+Priors on cosmological parameters coming from external experiments, the main focus or our work can be easily treated in this contest.
+For example a $1\%$ prior on $H_{0}$ can be obtained by:
+\begin{equation}
+F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,\text{fid}})^2}, 
+\end{equation}
+
+PUT in an appendix?
 \begin{equation}
@@
 
-\begin{eqnarray}
-\centering
-	s\ [\;\mu {\rm K.arcmin}\;] \equiv \frac{ {\rm NET}\ [\; \mu{\rm K.}\sqrt{s}\;] \times \sqrt{f_{sky} \ [\;{\rm arcmin}^2\;] }}{ \sqrt {N_{\rm det} \times Y \times \Delta T\ [\;{\rm s }\;]}}.
-	\label{eq:sensitivity_definition}
-\end{eqnarray}
+\subsection{CMB S4 experiments: level of noise.}
 
+The next generation of CMB experiments defined as S4 will have an extraordinary resolution and depth.\
+We will also assume that for the measurement we use, the temperature and E-mode polarization of the CMB large scale foreground like dust are under control and negligible.
+We will deal with the poisson noise in the temperature signal by discarding all the small scales modes with $\ell>\ell_{T,max}=3000$.
+The remaining source of noise, the instrumental noise is added to the power spectrum in the usual way:
  \begin{equation}
  	\centering
-		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle
-	\label{eq:Fij_def}
+		N^{ XX' }_\ell = s^{\, 2} \exp \left(\ell(\ell+1) \frac{\theta^{\ 2}_{\textsc{fwhm}}}{8\log2}\right),
+	\label{eq:beamnoise}
 \end{equation}
+where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represent instrumental the white noise.
+We decide to use a $1.5$ muK and a beam of 1 arcmin.
+Note that  $s \rightarrow s\times \sqrt{2}$ in the case of $ XX' = \{ EE, BB \}$.
 
- \begin{eqnarray}
- 	\centering
-		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
-	\label{eq:cov_definition}
-\end{eqnarray}
+Together with E and T we will also consider the information contained in the  
+lensing potential $phi$ reconstructed from CMB experiments.
+The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending CMB photons.
+This introduce non gaussianities that couples different modes in the otherwise independent CMB modes and using a quadrature estimator technique it can be reconstructed \cite{okamoto:2003,hu:2002}.
+For the noise associated to the recosntructed $\phi$ we follow \cite{okamoto:2003,hu:2002} without using more advanced iterative techniques.
 
-\begin{equation}
-\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}
-\end{equation}
+We parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We also allow different Dark Energy framework by introducing the equations of state parameter w as a free parameter.
 
-\begin{equation}
-F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,fid})^2}, 
-\end{equation}
+We use the $\Lambda$CDM parameter values from Table 2 of \textit{Planck} best fit value \cite{planck-collaboration:2014g}, i.e.
+$\Omega_c h^2 =  0.12029$,  $\Omega_b h^2  = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc.
+Regarding the $\Lambda$CDM extension we chose a neutrino energy densities$\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$  $\simeq$ 85\ meV and we also set $w=1$.
+
+%\begin{eqnarray}
+%\centering
+%	s\ [\;\mu {\rm K.arcmin}\;] \equiv \frac{ {\rm NET}\ [\; \mu{\rm K.}\sqrt{s}\;] \times \sqrt{f_{sky} \ [\;{\rm arcmin}^2\;] }}{ \sqrt {N_{\rm det} \times Y \times \Delta T\ [\;{\rm s }\;]}}.
+%	\label{eq:sensitivity_definition}
+%\end{eqnarray}
+%
 
@@
 
-\section{Data}
+\section{Results: \label{sec:results}}
 
@@
 
-\section{Results}
-
 \begin{table}[htdp]
@@
 \begin{center}
-\begin{tabular}{|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-$H_{0}$ & 0.32\% \\
-\hline
-\hline
-$\tau$ & 2.26 \%\\
-\hline
+$H_{0}$ &$ M_{\nu}$ &$\Omega_{bc}h^{2}$&$\Omega_{b}h^{2}$&$\tau$&$A_{s}$&$n_{s}$ \\
+$0.80 \%$&$40.92\%$&$0.44\%$&$0.11\%$&$3.075\%$&$0.53\%$&$0.18\%$\\
 
-\hline
-$A_{s}$ & 0.37 \%\\
-\hline
-
-\hline
-$n_{s}$ & 0.25 \%\\
-\hline
-
-\hline
-$N_{eff}$ & 0.9 \%\\
 \hline
@@
 \label{default}
+\caption{Things to notice: this is done with the Zhen test parameters ($\ell<3000$ and no error on lensing).
+Now if we trust it CMB alone can get a $0.8\%$ error on $H_{0}$ thus I am not surprised if a prior on H would not help.
+However what do we think about it?
+is it really CMB better than SN.
+People will not agree on that.
+$\tau$ will probably improve and also M$\nu$ from BAO may help improving parameters.}
 \end{table}%
@@
 
-\section{Conclusions}
+\section{Conclusions \label{sec:conclusions}}
+We find that...
+
+This means that supernova experiments CMB spatial distortions and large scale structure can/will/should
+
+Future works
 
@@
 
-% If you have acknowledgments, this puts in the proper section head.
-%\begin{acknowledgments}
-% put your acknowledgments here.
-%\end{acknowledgments}
+\begin{acknowledgments}
+We thank Youngsoo Park for..
+AM want to thank Zhen Pan who allow a careful comparison of our results.
+This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.
+%%%=================================================================
+The work of SD is supported by the U.S. Department of Energy, including grant DE-FG02-95ER40896.
+\end{acknowledgments}
 
 % Create the reference section using BibTeX:
-\bibliography{basename of .bib file}
+\bibliography{N_eff_prior_paper}
 

### 2015-08-05 Notes/N_eff_prior_paper.tex
uploading notes

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,prd,preprint,groupedaddress]{revtex4-1}
+\documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
 %\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
@@
 \usepackage{graphicx}
-
+\input{papers_packages_command._include_tex}
 % You should use BibTeX and apsrev.bst for references
@@
 \begin{document}
-\graphicspath{{../images/}}
+\graphicspath{{images/}}
 
@@
 The next generation of CMB experiments is expected to drastically improve our understanding of neutrino physics, inflation, Dark Matter and Dark Energy.
-Indeed they will measure with cosmic variance limited precision the temperature and polarization mode together with a lensing potential reconstruction with signal to noise bigger than one on several scales.
-Detailed forecasts have been done and will tremendously help to plan and motivate this next generation of CMB experiments.
-However we think that theoretical priors coming from external non-CMB experiments were not fully analyzed.
-For this reason in this work we use a fisher-matrix technique to test how improving on external priors can lead to improvement in S4 CMB parameters constraint.
-We find that..
+Indeed they will measure with cosmic variance limited precision the temperature and E mode polarization that will also allow the reconstruction of lensing potential with a signal to noise bigger than one on several scales.
+Detailed forecasts have been done and will tremendously help to plan and motivate the next generation of CMB experiments.
+However we think that the dependence of this results on the assumed external priors need to be explored.
+For this reason in this work we use a fisher-matrix technique to test how external priors can lead to improvement in S4 CMB parameters constraint.
+We find that quite surprisingly CMB S4 are very powerful even on
 
@@
 
-Far from finished S4 experiment will measure
+Far from finished S4 experiment will reveal several unclear aspect of the universe like the neutrino sector, Dark Energy and Dark matter properties.
 
-This will give us amazing information on..
+The number of relativistic species,like neutrinos, present in the early influence alter the expansion history of the universe.
+This can be probe by CMB by comparing the sound horizon scale with the silk damping.
+Indeed their ration scales differently with the expansion rate giving $r_{d}/r_{s}\propto\sqrt{H}$ (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
+Their mass on the other hand has very little influence on the CMB physics because for the range of masses allowed by recent constraints neutrinos are relativist at the last scattering surface.
+As a lot of other low redshift effect however they modify the CMB through the lensing effect of large scale structure on CMB photons.
+Different neutrino masses led to different growth of structure which consequently lead to different CMB lensing.
+The lensing potential can be reconstructed from the CMB itself opening the possibilities of constraining the neutrino mass with CMB.
+In the future according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a consistent factor of two improvement when BAO are added.
+Even if these constraints depends on the fiducial values used the recent future will be a big step from the neutrino sector.
+
+CMB is also sensitive to Dark energy (see \cite{2010MNRAS.405.2639J}) mainly from its effect on the Hubble function that modifies the distance to the last scattering surface.
+Furthermore the influence of Dark Energy in the low redshift universe modify the energy of CMB photons through the Integrated sachs wolfe Effect.
+Also as it was true in the neutrino case, CMB can thorugh lensing probe the growth of structure that depends on the dark energy properties and in particular its equation of state.
+When trying to constrain dark energy properties however we have to remember that they are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
+To get competitive constraint, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
+
+The importance of external priors in the dark energy case lead us to a more general concept. It is certainly true that CMB S4 will open an all new window
+but the synergy with other experiments like large scale structure clustering and weak lensing, BAO targeted experiments, SN, and proposed experiments like SPIRE.
+Indeed forecasted constrains very often assume external realistic prior.
 
@@
 
-This paper is organized as follow:
+This paper is organized as follow: in \refsec{method} we in \refsec{results} and we conclude with a discussion of our results in \refsec{conclusions}
 
-\section{Assumption and methods \label{sec:intro} }
-We start this section by introducing the Fisher matrix formalism a simple but powerful technique widely used to predict the constrains on the parameters of the assumed cosmological model with future experiments.
-We then specify the model we used in this work and the fiducial parameters we assumed.
-We adopt a standard $\Lambda$CDM model with basic possible extensions with parameters taken from recent CMB+BAO+SN fit.
+\section{Assumption and methods \label{sec:methods} }
+We start this section by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints on the parameters of the assumed cosmological model.
+Then, we describe in details the $\nu\Lambda$CDM cosmological model we use in this work and the fiducial parameters we assume (from recent BAO+CMB).
 
 \subsection{Fisher Matrix formalism}
-The starting point to estimate errors on cosmological parameters  is the Bayes theorem that relates the likelihood of measuring a set of data given the parameters of the model to what we want: the posterior probability of those parameters given the data.
-These are related by the priors probability of the parameters $P(\theta)$ through:
+The starting point to estimate errors on cosmological parameters is the Bayes theorem that relates the likelihood of measuring a set of data given the parameters of the model ($\mathcal{L}(d|\theta)$) to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
+These are related by the prior probability of the parameters $P(\theta)$ through:
 \begin{equation}
@@
 \end{equation}
-
-where,as usual we have neglect the probability of the data itself.
-These general framework can be simplified in the context of cosmology by focusing on small perturbations around the maximum of a Gaussian likelihood. 
+where, as usual, we have neglect the probability of the data itself.
+These general framework can be simplified in the context of cosmology by focusing not on the entire posterior distribution but on small perturbations around the maximum of an assumed Gaussian likelihood.
 Indeed we will assume that the likelihood can be approximated as Gaussian.
-While this approximate the true likelihood most of the time the opposite as been found in other cases \cite{2012JCAP...09..009W}.
+While this approximate the true likelihood most of the time, the opposite has been found in other cases \cite{2012JCAP...09..009W}.
 Even if other methods of analysis have been proposed \cite{2006JCAP...10..013P,2006astro.ph..9591A} this is still the standard assumption when future constrains have been forecasted \cite{wu:2014}.
+Moreover the gaussian approximation gets better at smaller scales which, a part for the exception of the optical depth parameters $\tau$, is where most of the information is coming from in our case.
 Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$.
-A consequence is that we are then able to estimate the uncertainties on those parameters by a simple analysis of the likelihood curvature around the fiducial values.
+Given that we are then able to estimate the uncertainties on those parameters by a simple analysis of the likelihood curvature around the fiducial values.
 As usual we define the fisher matrix elements as:
@@
 where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters array that, by definition, maximize the likelihood.
-We refer the reader for a detailed explanation of the fact that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$ with in our case $X,Y\in\{T,E,\phi\}$ the Fisher matrix can be rewritten as:
+We refer the reader to \cite{} for a detailed explanation of the fact that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$ with $X,Y\in\{T,E,\phi\}$ the Fisher matrix can be rewritten as:
 \begin{equation}
@@
 \end{eqnarray}
-
-The power of this method come from the fact that, given the Cramer-Rao inequality we can get the errors on parameter $i$ marginalized over all the other parameters simply inverting the Fisher matrix as:
+The terms $N_\ell^{X}$ represents the instrumental noise of the experiment and will be discussed in \refsec{noise}.
+The power of this method descend from Cramer-Rao inequality that relates the errors on parameter $i$ marginalized over all the other parameters $\sigma_i$ to the Fisher matrix as:
 \begin{equation}
 \sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}.
-\label{eq:cramer-rao}
+\label{eqn:cramer-rao}
 \end{equation}
-Approaching the forecast problem using a Fisher matrix introduce some technical difficulties together with the previously mentioned assumption that the likelihood is gaussian.
-Indeed it is known that the more the paramters used in the analysis the easier it is to face numerical isseus.
-Fisher matrix are can be ill-conditioned: a small change in the fisher matrix led to a big change in its inverse.
+
+This powerful and simple approach however introduce some technical difficulties together with the previously mentioned assumption that the likelihood is gaussian.
+Indeed it is known that increasing the number of parameters used in the analysis can lead to numerical issues.
+Fisher matrix  can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse.
 Given \refeq{cramer-rao} this can be a problem for error estimation.
-While we understand this possible concern we carefully compare our results to the literature  \footnote{we thank the authors of \cite{pan:2015} for their collaboration.}
-Furthermore we use the same technique used in \cite{2006astro.ph..9591A} to avoid any possible numerical instability in the marginalising procedure.
-This allow not to invert the entire matrix when we want to marginalize over a set of parameters minimizin numerical issue and conserving parameters degeneracies as much as possible.
+While we understand this as possible concern we carefully compare our results to the literature  \footnote{we thank the authors of \cite{pan:2015} for their collaboration.}
+Furthermore we use the same technique used in \cite{2006astro.ph..9591A} to avoid possible numerical instability in the marginalizing procedure.
+This allow not to invert the entire matrix when we want to marginalize over a set of parameters minimizing numerical issue and conserving parameters degeneracies as much as possible.
 
@@
 \end{equation}
-
-PUT in an appendix?
-\begin{equation}
-G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
-\end{equation}
-where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
-marginal\}$; therefore, $F^{\phi\phi}$ is the block of the total Fisher matrix containing the parameters we want to constrain, whilst $F^{\psi\psi}$ is the nuisance-parameter Fisher sub-matrix.
-Here, $\Lambda$ is the diagonal matrix whose elements are the eigenvalues of $F^{\psi\psi}$, whilst U is the orthogonal matrix diagonalising $F^{\psi\psi}$.
-By using Eq., our marginalising procedure is more stable, since degeneracies in $F^{\phi\phi}$ are properly propagated to G with no instabilities, and we do not even worry about a possibly ill-conditioned $F^{\phi\phi}$ sub-matrix, since we check its stability on the fly by the diagonalisation.
 
@@
 
-\subsection{CMB S4 experiments: level of noise.}
+\subsection{CMB S4 experiments: level of noise.\label{sec:noise}}
 
@@
 
-\begin{eqnarray}
-	T_{\nu} = \left( \frac{4}{11} \right)^{1/3} T_{\gamma} 
-	\label{eq:tnu_propto_tgamma}
-\end{eqnarray}
+%\begin{eqnarray}
+%	T_{\nu} = \left( \frac{4}{11} \right)^{1/3} T_{\gamma} 
+%	\label{eq:tnu_propto_tgamma}
+%\end{eqnarray}
 
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[scale=0.6]{h0_fisher.pdf}
-\caption{}
-\label{fig:phi-cl-noise}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[scale=0.6]{tau_fisher.pdf}
-\caption{}
-\label{fig:phi-cl-noise}
-\end{center}
-\end{figure}
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[scale=0.6]{h0_fisher.pdf}
+%\caption{}
+%\label{fig:phi-cl-noise}
+%\end{center}
+%\end{figure}
+%
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[scale=0.6]{tau_fisher.pdf}
+%\caption{}
+%\label{fig:phi-cl-noise}
+%\end{center}
+%\end{figure}
 
@@
 % only one appendix.
-%\appendix
+\appendix
+\section{Appendix I}
+\begin{equation}
+G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
+\end{equation}
+where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
+marginal\}$; therefore, $F^{\phi\phi}$ is the block of the total Fisher matrix containing the parameters we want to constrain, whilst $F^{\psi\psi}$ is the nuisance-parameter Fisher sub-matrix.
+Here, $\Lambda$ is the diagonal matrix whose elements are the eigenvalues of $F^{\psi\psi}$, whilst U is the orthogonal matrix diagonalising $F^{\psi\psi}$.
+By using Eq., our marginalizing procedure is more stable, since degeneracies in $F^{\phi\phi}$ are properly propagated to G with no instabilities, and we do not even worry about a possibly ill-conditioned $F^{\phi\phi}$ sub-matrix, since we check its stability on the fly by the diagonalisation.
+
 %\section{}
@@
 \begin{acknowledgments}
-We thank Youngsoo Park for..
+We thank Youngsoo Park for his contribution in the early stage of this work.
 AM want to thank Zhen Pan who allow a careful comparison of our results.

### 2015-08-14 n_priors_code/fisher_joint.py
new prior plot

@@ -76,7 +76,7 @@ def C(iell, ell, parbin, data):
     # eq 1 of W.hu et al snowmass paper 10^6 detectors
     Y = 0.25  # 25% yeld
     N_det = 10 ** 6  # 1 milion of detectors
-    s = 350. * np.sqrt(fsky2arcmin(0.75)) / np.sqrt(N_det * Y * years2sec(5))  # half sky in arcmin^2
+    s = 350. * np.sqrt(fsky2arcmin(fsky)) / np.sqrt(N_det * Y * years2sec(5))  # half sky in arcmin^2
     # s = 0.48 as in table from paper so it is ok.
     t = 1. / 60. / 180. * np.pi  # 2arcmin to rads beam
     fac = (ell * (ell + 1.) / 2. / np.pi) / (7.4311 * 10 ** 12)
@@ -109,10 +109,10 @@ def C(iell, ell, parbin, data):
 # =============================
 l_t_max = 3000  # this is the multipole you want to cut the temperature Cl at, to simulate the effect of foregrounds
 lmax = 3000
-lmin = 2
+lmin = 30
 N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_6.txt')
-data_folder = 'varying_lambda/run1'
-output_folder = 'varying_lambda/run1/output'
+data_folder = 'varying_lambda/run2'
+output_folder = 'varying_lambda/run2/output'
 fsky = 0.5
 lensed= False
 exclude = None

### 2015-08-14 n_priors_code/plotting/plot_prior_all.py

@@ -1,7 +1,9 @@
 '''
-Plot scipt to plot the derivatives of CMB specctra respect to all the parameters of the Fisher matrix.
+This plotting script is used to plot the effect of external priors on all the parameters involved.
 
-Different lines corresponde to different gaps and techniques.
+To put all the parameters in the same plot the prior is defined as the improvement from the CMB S4 results we get.
+
+the same is true for the improvement on the y axis parameters
 
 
 
@@ -35,10 +37,11 @@ from palettable.colorbrewer.qualitative import Set1_9
 
 # READ DATA
 
-
+# DEFINE YOUR FOLDER HERE
 base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_lambda'
 run_idx = 1
+# ======
 fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type, str(run_idx)), "rb"))
 values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
 par_gaps = pickle.load(open(base_dir + 'data/{}/run{}/par_gaps.p'.format(data_type, str(run_idx)), "rb"))
@@ -46,7 +49,7 @@ fisher_mat = np.loadtxt(base_dir + 'data/{}/run{}/output/fisher_mat.txt'.format(
 
 
 # ============================================
-# ============================================
+# PLOTTING DEFINITION SKIP TO ~155
 # ============================================
 
 
@@ -148,6 +151,7 @@ plt.rcParams['legend.handletextpad'] = 0.3
 # REAL PLOT HERE
 # ============================================
 
+# DEFINE LABELS
 
 label = {}
 
@@ -157,7 +161,7 @@ label['mnu'] = 'M'
 label['scalar_amp(1)'] = 'A_{s}'
 label['scalar_spectral_index(1)'] = 'n_{s}'
 label['omnuh2'] = r'\Omega_{\nu}'
-label['re_optical_depth'] = r'~\tau'
+label['re_optical_depth'] = r'\tau'
 label['ombh2'] = '\Omega_{b}h^{2}'
 label['ombch2'] = '\Omega_{m}h^{2}'
 label['omch2'] = '\Omega_{c}h^{2}'
@@ -165,47 +169,53 @@ label['w'] = 'w'
 fisher_inv = np.linalg.inv(fisher_mat)
[…]

### 2015-08-14 Notes/N_eff_prior_paper.tex

--- 2015-06-22 Notes/N_eff_prior_paper.tex
@@
 \usepackage{amsmath}
+\usepackage{lineno}
 \usepackage{graphicx}
+\input{papers_packages_command._include_tex}
 
-% You should use BibTeX and apsrev.bst for references
-% Choosing a journal automatically selects the correct APS
-% BibTeX style file (bst file), so only uncomment the line
-% below if necessary.
-%\bibliographystyle{apsrev4-1}
+\bibliographystyle{apsrev4-1}
 
 \begin{document}
-\graphicspath{{../images/}}
-
-% Use the \preprint command to place your local institutional report
-% number in the upper righthand corner of the title page in preprint mode.
-% Multiple \preprint commands are allowed.
-% Use the 'preprintnumbers' class option to override journal defaults
-% to display numbers if necessary
-%\preprint{}
+\graphicspath{{images/}}
 
 %Title of paper
-\title{$N_{\text{eff}}$ priors}
+\title{The importance of external priors for the next generation of CMB experiments.}
 
-% repeat the \author .. \affiliation  etc. as needed
-% \email, \thanks, \homepage, \altaffiliation all apply to the current
-% author. Explanatory text should go in the []'s, actual e-mail
-% address or url should go in the {}'s for \email and \homepage.
-% Please use the appropriate macro foreach each type of information
-
-% \affiliation command applies to all authors since the last
-% \affiliation command. The \affiliation command should follow the
-% other information
-% \affiliation can be followed by \email, \homepage, \thanks as well.
-\author{}
-%\email[]{Your e-mail address}
-%\homepage[]{Your web page}
-%\thanks{}
-%\altaffiliation{}
-\affiliation{}
-
-%Collaboration name if desired (requires use of superscriptaddress
-%option in \documentclass). \noaffiliation is required (may also be
-%used with the \author command).
-%\collaboration can be followed by \email, \homepage, \thanks as well.
-%\collaboration{}
-%\noaffiliation
+\author{A. Manzotti}
+%\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+\author{S. Dodelson}
+%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
 
 \date{\today}
+\linenumbers
+\begin{abstract}
+The next generation of cosmic microwave background (CMB) experiments can potentially improve drastically what we know about neutrino physics, inflation, Dark Matter and Dark Energy.
+Indeed they will likely exploit the great sensitivity of the CMB to the details of physical processes in the early Universe (through temperature and polarization) as well as in the late universe (through CMB secondary effects like lensing).
+They will be possible because the temperature anisotropies and E mode polarization of the CMB will be measured with cosmic variance limited precision.
+Such a low level of noise, together with the improved resolution, will also allow CMB experiment to reconstruct the lensing potential of high redshift large scale structure with a signal to noise bigger than one on several scales.
+Indeed recently, forecasts papers validate these possible outcomes and help to plan and motivate the next generation of CMB experiments.
+However these results also indirectly disclose the needed synergy between CMB and other experiments, for example the need of imposing external prior on some of the cosmological parameters.
+We think that the extent to which CMB constraints depend on external priors needs to be explored, for instance to guide the combined effort of different experimental fields.
+For this reason, in this work we adopt a fisher-matrix technique to test how external priors can lead to improvement in the S4 generation of CMB parameters constraint. 
+We find that quite surprisingly CMB S4 are very powerful even on .. There are few exceptions.. like $w$.
 
-\begin{abstract}
-Try to understand the effect of priors on different parameters on the $N_{\text{eff}}$ errors with a Fisher matrix approach.
-Where should you invest time and money if you really care about the value of $N_{\text{eff}}$?
-
-\begin{itemize}
-\item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet.
-\item We are varying $\tau, ~ n_{s}, ~ A_{s}, ~ N_{\rm eff}, ~ H_{0}, w,~\Omega_{\nu}h^{2},~\Omega_{c}h^{2},~\Omega_{b}h^{2}$.
-The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, $\Omega_{\nu}h^{2}=0.0009$
-\item Next TODO: Check for bugs and errors code in its early stages. PCA? what prior is more important? Full MCMC should not be extremely hard with cosmosis.
-\end{itemize}
+%\begin{itemize}
+%\item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet. Needed?
+%\item We are varying $\tau, ~ n_{s}, ~ A_{s}, ~ N_{\rm eff}, ~ H_{0}, w,~\Omega_{\nu}h^{2},~\Omega_{c}h^{2},~\Omega_{b}h^{2}$.
+The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, which corresponds to $\Omega_{\nu}h^{2}=0.0009$.
+%\item Next TODO: Check for bugs and errors code (DONE look at Zhen test).
+PCA?
+what prior is more important?
+Full MCMC should not be extremely hard with cosmosis (I implemented a toy likelihood analysis to be improved).
+%\end{itemize}
+%
 
@@
 
-% insert suggested PACS numbers in braces on next line
 \pacs{}
 % insert suggested keywords - APS authors don't need to do this
-%\keywords{}
-
-%\maketitle must follow title, authors, abstract, \pacs, and \keywords
+\keywords{CMB, neutrinos}
 \maketitle
 
-% body of paper here - Use proper section commands
-% References should be done using the \cite, \ref, and \label commands
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
-% Put \label in argument of \section for cross-referencing
-%\section{\label{}}
-\subsection{}
-\subsubsection{}
+Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
+Moreover they also represent an amazing success in instrumental physics.
+Indeed every new generation of satellite experiments improved its sensitivity gaining almost a factor of ten in sensitivity compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
+Experimentally we went from observing the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
+With the improved sensitivity we menaged to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
+The same is true for ground experiment where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
+Ground experiments have already reached the maximum possible sensitivity per detector but will significantly increase the number of detectors in future generation.
+These series of successes is far from its end.
+Indeed the next generation of CMB experiments (S4) is now in its planning stage and it promises to measure with cosmic variance limited precision the E-mode polarization together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+As happened in the past this new sensitivity will improve our understanding of several areas like neutrinos, inflation, Dark Energy and Dark matter.
+In this work we will mainly focus on two of these: neutrinos and Dark Energy.
 
-\section{Theory}
+The main aspects of neutrinos physics that CMB can access are the number of relativistic degree of freedom in the early universe $N_{\rm eff}$ and the total mass of the three neutrino species $M_{\nu}$.
+Indeed the number of relativistic species, like neutrinos, in the early Universe alter the expansion history.
+This effect can be probed with CMB measurement by comparing the sound horizon scale, obtained from the CMB peaks position, with the silk damping scale.
+Indeed because they scale differently with the expansion rate $H(z)$ their ratio $r_{d}/r_{s}\propto\sqrt{H}$ (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
+This allow us to constrain $N_{\rm eff}$ and any deviation from the predicted value $N_{\rm eff}=3.046$ will be a sign of possible extensions of the Standard Model.
+The mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
+As other low redshift effects however, massive neutrinos modify the CMB through the lensing effect of large scale structure on the CMB photons.
+Different neutrino masses result into different growth of structure which consequently lead to different CMB lensing.
+The lensing potential can be reconstructed from the CMB itself, opening the possibility of constraining the neutrino mass.
+In the future, according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement when BAO data are also used.
+The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-\subsection{Fisher Matrix formalism}
-Copied from paper change it for pubblication:
+As previously mentioned, CMB is also sensitive to Dark energy (see \cite{2010MNRAS.405.2639J}).
+The current generation of experiments measured with percent accuracy the energy density of Dark Energy.
+The next big challenge is to identify the nature of this mysterious component.
+The first step will be to measure any deviation from the equation of state predicted by a cosmological constant model.
+Furthermore CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
+Indeed dark energy mainly affect the CMB because of its effect on the universe's expansion that modifies the distance to the last scattering surface.
+Furthermore the influence of Dark Energy on the low redshift universe modify the energy of CMB photons through the Integrated Sachs Wolfe Effect.
+Also, as it was true in the neutrino case, CMB lensing can probe the growth of structure that depends on the dark energy properties and in particular on its equation of state.
+However dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
+To get competitive constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
 
-To avoid any possible numerical instability in the marginalising procedure, we calculate the the Fisher matrix marginalised over the standard LCDM parameters following (Albrecht et al. 2009)
+The importance of external priors in the dark energy case is actually more general issue.
+Even if it is certainly true that CMB S4 will open an all new window on several area of physics and cosmology the synergy with other experiments is crucial.
+In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae.
+For this reason forecasted constrains very often assume external realistic priors.
+To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
+This can tell us for example which CMB parameter constraints will benefit more from improvement on external priors or if proposed experiments like PIXIE \cite{2011JCAP...07..025K} can indirectly also benefit CMB constraints.
+
+This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
+In \refsec{results} we will describe our findings and we then conclude with a discussion of our results in \refsec{conclusions}.
+
+\section{Assumption and methods \label{sec:methods}}
+The main goal of this paper is to measure the effect of external priors on the statistical errors of cosmological parameters derived from future CMB experiments.
+We start in \refssec{fm-formalism} by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints on the parameters of the assumed cosmological model.
+To apply this technique, we need to choose a model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the details of the CMB experiments we have in mind.
+These will be presented in \refssec{cosmo-noise}.
+
+\subsection{Fisher Matrix formalism \label{subsec:fm-formalism}}
+The natural starting point to estimate errors on cosmological parameters is the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
+These two are related by the prior probability of the parameters $P(\theta)$ through:
 \begin{equation}
-G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
+\mathcal{L}(\theta|d)\propto \mathcal{L}(d|\theta)P(\theta),
 \end{equation}
-where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
-marginal\}$; therefore, $F^{\phi\phi}$ is the block of the total Fisher matrix containing the parameters we want to constrain, whilst $F^{\psi\psi}$ is the nuisance-parameter Fisher sub-matrix.
-Here, $\Lambda$ is the diagonal matrix whose elements are the eigenvalues of $F^{\psi\psi}$, whilst U is the orthogonal matrix diagonalising $F^{\psi\psi}$.
-By using Eq., our marginalising procedure is more stable, since degeneracies in $F^{\phi\phi}$ are properly propagated to G with no instabilities, and we do not even worry about a possibly ill-conditioned $F^{\phi\phi}$ sub-matrix, since we check its stability on the fly by the diagonalisation.
+where, as usual, we have neglected the probability of the data itself.
+These general framework can be simplified in our case. We will focus not on an entirely general posterior distribution, but we will assume that the likelihoods are gaussian.
+Furthermore we will not derive the global shape of the likelihood but we will obtain parameters errors by studying small perturbations around its maximum.
+These are the two basic assumptions of the Fisher approach.
+
+Regarding the first assumption, we recognize the fact that, even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}.
+Despite this, we do not need the level of accuracy that a careful modeling of the likelihoods will give us, because we are looking for the general behavior of parameters errors as a function of external priors.
+Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, a part for the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
+
+Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$ and we will get the errors on those by a simple analysis of the likelihood curvature around the fiducial values.
+Indeed, as usual, we define the fisher matrix elements as the curvature:
+\begin{equation}
+	\centering
+		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle,
+	\label{eqn:Fij_def}
+\end{equation}
+where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters array that, by definition, maximize the likelihood.
+We refer the reader to \cite{} for a detailed proof that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, the Fisher matrix of a CMB experiment can be rewritten as:
+\begin{equation}
+ F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
+ \label{eqn:Fij_def2}
+ \end{equation}
+ In this work we will use the CMB temperature and E mode polarization together with lensing reconstruction to constrain parameters.
+For this reason, $\boldsymbol{C}_\ell$ that encapsulate the used power spectra in a matrix structure is:
+ \begin{eqnarray}
+ 	\centering
+		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
+	\label{eqn:cov_definition}
+\end{eqnarray}
+Note that we are neglecting the term $C_\ell^{E\phi}$.
+As also noticed in previous literature like (\cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
+Furthermore the terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
+The power of the Fisher approach descend from the Cramer-Rao inequality that relates the errors on parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
+\begin{equation}
+\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}.
+\label{eqn:cramer-rao}
+\end{equation}
+Once $F_{ij}$ is computed following \refeq{Fij_def2} it is straightforwards to get the error $\sigma_i$ from \refeq{cramer-rao}.
+Furthermore in this context it is easy to introduce external priors on cosmological parameters, which is the main focus or our work.
+Indeed to add priors we simply need to add external experiments Fisher matrixes before performing the matrix inversion of \refeq{cramer-rao}, i.e.:
+\begin{equation}
+F_{total}=F_{CMB}+F_{external}.
+\end{equation}
+In the same way we can add priors on a single cosmological parameter just by adding the prior to the matrix element.
+For example, a $1\%$ prior on $H_{0}$ can be obtained by:
+\begin{equation}
+F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,\text{fid}})^2}.
+\end{equation}
+
+We chose a Fisher matrix approach because of its ability to rapidly forecast future experiments performances without generating mock data together with the ease of including external priors. 
+This technique however introduce some technical difficulties together the previously mentioned assumption that the likelihood is gaussian. 
+Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves contest were a lot of parameters are needed).
+Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
+Even if other methods of analysis have been proposed \cite{2006JCAP...10..013P,2006astro.ph..9591A} this is still the standard method used to forecast future constrains \cite{wu:2014}.
+While we how these issues can be a possible concern we carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
+We compute the derivatives in \refeq{Fij_def2} using a 5 points formula:
+\begin{equation}
+\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim \frac{-C(\theta_{0}+2h)+8C(\theta_{0}+h)-8C(\theta_{0}-h)+C(\theta_{0}-2h)}{12h}.
+\label{eqn:deriv}
+\end{equation}
+This high order definition allow us to use a bigger gap h around the fiducial parameters $\theta_{0}$.
+As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing C negligible.
+We also test the robustness of this calculation by changing the gap h in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
+Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
+Lastly we implement the same technique of \cite{2006astro.ph..9591A} to avoid possible numerical instability in the marginalizing procedure.
+This allow us not to invert the entire matrix when we want to marginalize over a set of parameters, in order to minimize numerical issues and conserve parameters degeneracies as much as possible.
 
@@
 \includegraphics[scale=0.6]{PS_phi_with_noise.pdf}
-\caption{Lensing potential power spectrum for our fiducial cosmology together with the lensing reconstruction noise $N^{\phi}$ used in this work.}
+\caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
+In the figure: the deflection power spectrum for our fiducial cosmology together with the two examples of lensing reconstruction noise $N^{\phi}$ used in this work.}
 \label{fig:phi-cl-noise}
@@
 \includegraphics[scale=0.6]{PS_with_noise.pdf}
-\caption{CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.
-They are basically equivalent to a cosmic variance limited experiment with $l_{max}\sim 3000$}
+\caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
+In the figure: CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
 \label{fig:cmb-cl-noise}
@@
 
-\begin{eqnarray}
-\centering
-	s\ [\;\mu {\rm K.arcmin}\;] \equiv \frac{ {\rm NET}\ [\; \mu{\rm K.}\sqrt{s}\;] \times \sqrt{f_{sky} \ [\;{\rm arcmin}^2\;] }}{ \sqrt {N_{\rm det} \times Y \times \Delta T\ [\;{\rm s }\;]}}.
-	\label{eq:sensitivity_definition}
-\end{eqnarray}
+\subsection{CMB S4 experiments: level of noise.\label{subsec:cosmo-noise}}
+In this subsection we will describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra C and the noise power N.
+The first step is the cosmological model we used to compute $C^{X,Y}$ with $X,Y \in \{T,E,\phi\}$.
+We parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We also allow different Dark Energy model by introducing the equations of state parameter w as a varying parameter.
+We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
+$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc.
+Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
+Given the model, we use the CAMB software to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
+Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
 
+We chose the second element of \refeq{Fij_def2}, the noise power in the CMB $N_{\ell}^{T,E}$ and in the lensing reconstruction $N_{\ell}^{\phi}$ with the next generation of CMB experiments (S4) in mind.
+For the temperature and E-mode polarization of the CMB, together with the an improved depth and resolution we will also assume that large scale foregrounds, like dust, are under control or negligible.
+We will deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
+The remaining source of noise, the instrumental noise is added to the power spectrum in the usual way:
  \begin{equation}
  	\centering
-		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle
-	\label{eq:Fij_def}
+		N^{X}_\ell = s^{\, 2} \exp \left(\ell(\ell+1) \frac{\theta^{\ 2}_{\textsc{fwhm}}}{8\log2}\right),
+	\label{eq:beamnoise}
 \end{equation}
+where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represent the instrumental white noise.
+We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of 1 arcmin (PRELIMINARY).
+Note that $1.5$ $\mu$K-arcmin is the noise in temperature and we need $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
 
- \begin{eqnarray}
- 	\centering
-		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
-	\label{eq:cov_definition}
-\end{eqnarray}
+Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from CMB experiments.
+The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons.
+This introduce non gaussianities that couple different modes in the otherwise independent CMB modes and it can be reconstructed using a quadrature estimator technique \cite{okamoto:2003,hu:2002}.
+For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow \cite{okamoto:2003,hu:2002} without using more advanced iterative techniques.
 
-\begin{equation}
-\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}
-\end{equation}
+%\begin{eqnarray}
+%\centering
+%	s\ [\;\mu {\rm K.arcmin}\;] \equiv \frac{ {\rm NET}\ [\; \mu{\rm K.}\sqrt{s}\;] \times \sqrt{f_{sky} \ [\;{\rm arcmin}^2\;] }}{ \sqrt {N_{\rm det} \times Y \times \Delta T\ [\;{\rm s }\;]}}.
+%	\label{eq:sensitivity_definition}
+%\end{eqnarray}
+%
 
-\begin{equation}
-F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,fid})^2}, 
-\end{equation}
+%\begin{eqnarray}
+%	T_{\nu} = \left( \frac{4}{11} \right)^{1/3} T_{\gamma} 
+%	\label{eq:tnu_propto_tgamma}
+%\end{eqnarray}
 
-\begin{eqnarray}
-	T_{\nu} = \left( \frac{4}{11} \right)^{1/3} T_{\gamma} 
-	\label{eq:tnu_propto_tgamma}
-\end{eqnarray}
-
-\section{Data}
+\section{Results \label{sec:results}}
 
 \begin{table}[htdp]
-\caption{Fiducial values used.}
 \begin{center}
@@
 \hline
-$H_{0}$ &\\
+$H_{0}$ & $67.11$ km/s/Mpc\\
 \hline
 \hline
-$\tau$ &  \\
+$\tau$ & $0.0925$ \\
 \hline
@@
 \hline
-$A_{s}$ & \\
+$A_{s}$ &$2.215 \times 10^{-9}$ \\
 \hline
@@
 \hline
-$n_{s}$ & \\
+$n_{s}$ & $0.9624$ \\
 \hline
@@
 \hline
-$N_{eff}$ & \\
+$N_{eff}$ & 3.046\\
 \hline
@@
 \end{center}
+\caption{Fiducial values of cosmological parameters used in this work.}
 \label{default}
@@
 
-\section{Results}
+\begin{table}[htdp]
 
-\begin{table}[htdp]
-\caption{How well we do constrain separate parameters with this data without any external prior?}
 \begin{center}
-\begin{tabular}{|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-$H_{0}$ & 0.32\% \\
-\hline
-\hline
-$\tau$ & 2.26 \%\\
-\hline
+$H_{0}$ &$ M_{\nu}$ &$\Omega_{bc}h^{2}$&$\Omega_{b}h^{2}$&$\tau$&$A_{s}$&$n_{s}$ \\
+$0.80 \%$&$40.92\%$&$0.44\%$&$0.11\%$&$3.075\%$&$0.53\%$&$0.18\%$\\
 
 \hline
-$A_{s}$ & 0.37 \%\\
-\hline
-
-\hline
-$n_{s}$ & 0.25 \%\\
-\hline
-
-\hline
-$N_{eff}$ & 0.9 \%\\
-\hline
-
 \end{tabular}
+\caption{How well we do constrain separate parameters with this data without any external prior?
+Things to notice: this is done with the Zhen test parameters ($\ell<3000$ and no error on lensing).
+Now if we trust it CMB alone can get a $0.8\%$ error on $H_{0}$ thus I am not surprised if a prior on H would not help.
+However what do we think about it?
+is it really CMB better than SN.
+People will not agree on that.
+$\tau$ will probably improve and also M$\nu$ from BAO may help improving parameters.}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.6]{h0_fisher.pdf}
+\includegraphics[scale=0.6]{prior_w.pdf}
 \caption{}
-\label{fig:phi-cl-noise}
+\label{fig:prior_w}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.6]{tau_fisher.pdf}
+\includegraphics[scale=0.6]{prior_massless_neutrinos.pdf}
 \caption{}
-\label{fig:phi-cl-noise}
+\label{fig:prior_massless_neutrinos}
 \end{center}
@@
 
-\section{Conclusions}
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[scale=0.6]{prior_re_optical_depth.pdf}
+\caption{}
+\label{fig:prior_re_optical_depth}
+\end{center}
+\end{figure}
 
-% If in two-column mode, this environment will change to single-column
-% format so that long equations can be displayed. Use
-% sparingly.
-%\begin{widetext}
-% put long equation here
-%\end{widetext}
+\section{Conclusions \label{sec:conclusions}}
+--We find that...
 
-% figures should be put into the text as floats.
-% Use the graphics or graphicx packages (distributed with LaTeX2e)
-% and the \includegraphics macro defined in those packages.
-% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz,
-% and Frank Mittelbach for instance.
-%
-% Here is an example of the general form of a figure:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Use the figure* environment if the figure should span across the
-% entire page. There is no need to do explicit centering.
+--This means that supernova experiments CMB spatial distortions and large scale structure can/will/should
 
-% \begin{figure}
-% \includegraphics{}%
-% \caption{\label{}}
-% \end{figure}
-
-% Surround figure environment with turnpage environment for landscape
-% figure
-% \begin{turnpage}
-% \begin{figure}
-% \includegraphics{}%
-% \caption{\label{}}
-% \end{figure}
-% \end{turnpage}
+--Future works
 
@@
 % only one appendix.
-%\appendix
-%\section{}
+\appendix
+\section{Marginalization}
+\begin{equation}
+G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
+\end{equation}
+where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
+marginal\}$; therefore, $F^{\phi\phi}$ is the block of the total Fisher matrix containing the parameters we want to constrain, whilst $F^{\psi\psi}$ is the nuisance-parameter Fisher sub-matrix.
+Here, $\Lambda$ is the diagonal matrix whose elements are the eigenvalues of $F^{\psi\psi}$, whilst U is the orthogonal matrix diagonalising $F^{\psi\psi}$.
+By using Eq., our marginalizing procedure is more stable, since degeneracies in $F^{\phi\phi}$ are properly propagated to G with no instabilities, and we do not even worry about a possibly ill-conditioned $F^{\phi\phi}$ sub-matrix, since we check its stability on the fly by the diagonalisation.
 
-% If you have acknowledgments, this puts in the proper section head.
-%\begin{acknowledgments}
-% put your acknowledgments here.
-%\end{acknowledgments}
+\begin{acknowledgments}
+We thank Youngsoo Park for his contribution in the early stage of this work.
+AM want to thank Zhen Pan who allows a careful cross-check of our results.
+This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.
+%%%=================================================================
+The work of SD is supported by the U.S. Department of Energy, including grant DE-FG02-95ER40896.
+\end{acknowledgments}
 
 % Create the reference section using BibTeX:
-\bibliography{basename of .bib file}
+\bibliography{N_eff_prior_paper}
 
 \end{document}
-
-%
-% ****** End of file apstemplate.tex ******

### 2015-09-04 Notes/N_eff_prior_paper.tex
minor changes

@@
 \usepackage{graphicx}
-\input{papers_packages_command._include_tex}
-
+%\input{papers_packages_command._include_tex}
+\newcommand\refsec[1]{\S\ref{sec:#1}}
+\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
 \bibliographystyle{apsrev4-1}
@@
 %Title of paper
-\title{The importance of external priors for the next generation of CMB experiments.}
+\title{External priors for the next generation of CMB experiments.}
 
-\author{A. Manzotti}
+\author{Alessandro Manzotti}
 %\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
@@
 %\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-\author{S. Dodelson}
+\author{Scott Dodelson}
 %\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
@@
 \begin{abstract}
-The next generation of cosmic microwave background (CMB) experiments can potentially improve drastically what we know about neutrino physics, inflation, Dark Matter and Dark Energy.
-Indeed they will likely exploit the great sensitivity of the CMB to the details of physical processes in the early Universe (through temperature and polarization) as well as in the late universe (through CMB secondary effects like lensing).
-They will be possible because the temperature anisotropies and E mode polarization of the CMB will be measured with cosmic variance limited precision.
-Such a low level of noise, together with the improved resolution, will also allow CMB experiment to reconstruct the lensing potential of high redshift large scale structure with a signal to noise bigger than one on several scales.
-Indeed recently, forecasts papers validate these possible outcomes and help to plan and motivate the next generation of CMB experiments.
+The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy.
+%Indeed these experiments will likely exploit the sensitivity of the CMB to the details of physical processes in the early Universe (through temperature and polarization) as well as in the late universe (through CMB secondary effects like lensing).
+%They will be possible because the temperature anisotropies and E mode polarization of the CMB will be measured with cosmic variance limited precision. 
+The low level of noise, together with improved angular resolution, will allow CMB experiments to reconstruct the lensing potential of high redshift large scale structure with a signal to noise bigger than one over a wide range of scales.
+%Indeed recently, forecast papers validate these possible outcomes and help to plan and motivate the next generation of CMB experiments.
 However these results also indirectly disclose the needed synergy between CMB and other experiments, for example the need of imposing external prior on some of the cosmological parameters.
 We think that the extent to which CMB constraints depend on external priors needs to be explored, for instance to guide the combined effort of different experimental fields.
-For this reason, in this work we adopt a fisher-matrix technique to test how external priors can lead to improvement in the S4 generation of CMB parameters constraint. 
+Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
+Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
 We find that quite surprisingly CMB S4 are very powerful even on .. There are few exceptions.. like $w$.
@@
 where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represent the instrumental white noise.
-We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of 1 arcmin (PRELIMINARY).
+We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin (PRELIMINARY).
 Note that $1.5$ $\mu$K-arcmin is the noise in temperature and we need $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
@@
 The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons.
-This introduce non gaussianities that couple different modes in the otherwise independent CMB modes and it can be reconstructed using a quadrature estimator technique \cite{okamoto:2003,hu:2002}.
+This introduce non gaussianities that couple different modes in the otherwise independent CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
 For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow \cite{okamoto:2003,hu:2002} without using more advanced iterative techniques.
@@
 % Create the reference section using BibTeX:
-\bibliography{N_eff_prior_paper}
+\bibliography{N_eff_prior_paperNotes}
 

### 2015-09-08 Notes/N_eff_prior_paper.tex
adding commands to get rid of the additional tex include
working on the draft

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,prd,preprint,groupedaddress]{revtex4-1}
+\documentclass[aps,prd,reprint,superscriptaddress]{revtex4-1}
 %\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
@@
 \newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
+\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
+
+% For various journals
+%\newcommand{\apj}{ApJ}
+\newcommand{\physrep}{Physics Reports}
+\newcommand{\aapr}{A\&A Rev.}
+\newcommand{\apjl}{ApJL}
+\newcommand{\apjs}{ApJS}
+\newcommand{\aap}{A\&A}
+\newcommand{\mnras}{MNRAS}
+%\newcommand{\prd}{Phys. Rev. D}
+%\newcommand{\physrev}{Phys. Rev.}
+\newcommand{\physrevlett}{Phys. Rev. Lett.}
+\newcommand{\jcap}{{\em JCAP }}
+\newcommand{\aj}{AJ }
+
 \bibliographystyle{apsrev4-1}
@@
 \author{Alessandro Manzotti}
-%\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
-%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
+\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
 \author{Scott Dodelson}
-%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
-%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+
+\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
 
 \date{\today}
-\linenumbers
 \begin{abstract}
 The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
-%Indeed these experiments will likely exploit the sensitivity of the CMB to the details of physical processes in the early Universe (through temperature and polarization) as well as in the late universe (through CMB secondary effects like lensing).
-%They will be possible because the temperature anisotropies and E mode polarization of the CMB will be measured with cosmic variance limited precision. 
-The low level of noise, together with improved angular resolution, will allow CMB experiments to reconstruct the lensing potential of high redshift large scale structure with a signal to noise bigger than one over a wide range of scales.
+Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
 %Indeed recently, forecast papers validate these possible outcomes and help to plan and motivate the next generation of CMB experiments.
@@
 Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
-We find that quite surprisingly CMB S4 are very powerful even on .. There are few exceptions.. like $w$.
-
-%\begin{itemize}
-%\item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet. Needed?
-%\item We are varying $\tau, ~ n_{s}, ~ A_{s}, ~ N_{\rm eff}, ~ H_{0}, w,~\Omega_{\nu}h^{2},~\Omega_{c}h^{2},~\Omega_{b}h^{2}$.
-The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, which corresponds to $\Omega_{\nu}h^{2}=0.0009$.
-%\item Next TODO: Check for bugs and errors code (DONE look at Zhen test).
-PCA?
-what prior is more important?
-Full MCMC should not be extremely hard with cosmosis (I implemented a toy likelihood analysis to be improved).
-%\end{itemize}
-%
-
+We find that, quite surprisingly, CMB S4 are very powerful even on .. There are few exceptions.. like $w$.
 \end{abstract}
@@
 \section{Introduction}\label{sec:intro}
+Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
+They will maintain their role also in the near future thanks to the unprecedented level of noise and accuracy of their next generation (S4).
+The recent past is indeed reassuring. 
+Every new generation of satellite experiments improved the sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
+This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
+With the improved sensitivity we managed to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
+The same success characterized ground experiments where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
+These series of successes is far from its end.
+Indeed the next generation of CMB experiments (S4) is now in its planning stage and it will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+As it happened in the past, this new sensitivity will improve our understanding of several areas of astrophysics like neutrinos, inflation, Dark Energy and Dark matter.
 
-Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
-Moreover they also represent an amazing success in instrumental physics.
-Indeed every new generation of satellite experiments improved its sensitivity gaining almost a factor of ten in sensitivity compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
-Experimentally we went from observing the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
-With the improved sensitivity we menaged to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
-The same is true for ground experiment where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
-Ground experiments have already reached the maximum possible sensitivity per detector but will significantly increase the number of detectors in future generation.
-These series of successes is far from its end.
-Indeed the next generation of CMB experiments (S4) is now in its planning stage and it promises to measure with cosmic variance limited precision the E-mode polarization together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As happened in the past this new sensitivity will improve our understanding of several areas like neutrinos, inflation, Dark Energy and Dark matter.
-In this work we will mainly focus on two of these: neutrinos and Dark Energy.
-
-The main aspects of neutrinos physics that CMB can access are the number of relativistic degree of freedom in the early universe $N_{\rm eff}$ and the total mass of the three neutrino species $M_{\nu}$.
+%In this work we will mainly focus on two of these: neutrinos and Dark Energy.
+The main effects of neutrinos in the CMB can be parametrize by the number of relativistic degree of freedom in the early universe $N_{\rm eff}$ and the total mass of the three neutrino species $M_{\nu}$.
 Indeed the number of relativistic species, like neutrinos, in the early Universe alter the expansion history.
-This effect can be probed with CMB measurement by comparing the sound horizon scale, obtained from the CMB peaks position, with the silk damping scale.
-Indeed because they scale differently with the expansion rate $H(z)$ their ratio $r_{d}/r_{s}\propto\sqrt{H}$ (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
-This allow us to constrain $N_{\rm eff}$ and any deviation from the predicted value $N_{\rm eff}=3.046$ will be a sign of possible extensions of the Standard Model.
-The mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
-As other low redshift effects however, massive neutrinos modify the CMB through the lensing effect of large scale structure on the CMB photons.
-Different neutrino masses result into different growth of structure which consequently lead to different CMB lensing.
+This can be probed using the CMB by comparing the sound horizon scale, obtained from the CMB peaks positions, with the silk damping scale taking advantage of the fact that they scale differently with the expansion rate $H(z)$
+%their ratio $r_{d}/r_{s}\propto\sqrt{H}$ 
+(see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
+This allow CMB experiment to constrain $N_{\rm eff}$. Any deviation from the predicted value $N_{\rm eff}=3.046$ will be a sign of possible extensions of the Standard Model.
+The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
+As other low redshift effects however, massive neutrinos modify the CMB through the lensing of the CMB photons by large scale structures.
+Different neutrino masses result into a different growth of structure which consequently lead to different CMB lensing.
 The lensing potential can be reconstructed from the CMB itself, opening the possibility of constraining the neutrino mass.
-In the future, according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement when BAO data are also used.
+In the future, according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when baryon acoustic oscillations (BAO) data are also used.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-As previously mentioned, CMB is also sensitive to Dark energy (see \cite{2010MNRAS.405.2639J}).
+As previously mentioned, CMB is also sensitive to  the Dark energy properties (see \cite{2010MNRAS.405.2639J}).
 The current generation of experiments measured with percent accuracy the energy density of Dark Energy.
-The next big challenge is to identify the nature of this mysterious component.
-The first step will be to measure any deviation from the equation of state predicted by a cosmological constant model.
+The challenge for future experiments is to identify the nature of this mysterious component.
+The first step will be to measure any deviation in the equation of state from the one predicted by a cosmological constant model.
 Furthermore CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
-Indeed dark energy mainly affect the CMB because of its effect on the universe's expansion that modifies the distance to the last scattering surface.
-Furthermore the influence of Dark Energy on the low redshift universe modify the energy of CMB photons through the Integrated Sachs Wolfe Effect.
-Also, as it was true in the neutrino case, CMB lensing can probe the growth of structure that depends on the dark energy properties and in particular on its equation of state.
+Indeed dark energy affect the CMB because due to its effect on the universe's expansion that modifies the distance to the last scattering surface together with a modification  of CMB photons energy through the Integrated Sachs Wolfe Effect.
+Furthermore different DE models lead to a different growth of large scale structure.
 However dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
@@
 
-The importance of external priors in the dark energy case is actually more general issue.
-Even if it is certainly true that CMB S4 will open an all new window on several area of physics and cosmology the synergy with other experiments is crucial.
+In general, external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
+For this reason the synergy with other experiments is crucial.
 In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae.
-For this reason forecasted constrains very often assume external realistic priors.
 To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
-This can tell us for example which CMB parameter constraints will benefit more from improvement on external priors or if proposed experiments like PIXIE \cite{2011JCAP...07..025K} can indirectly also benefit CMB constraints.
+%This can tell us for example which CMB parameter constraints will benefit more from improvement on external priors or if proposed experiments like PIXIE \cite{2011JCAP...07..025K} can indirectly also benefit CMB constraints.
 
 This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
-In \refsec{results} we will describe our findings and we then conclude with a discussion of our results in \refsec{conclusions}.
+In \refsec{results} we will describe our results and we then conclude with a discussion in \refsec{conclusions}.
 
-\section{Assumption and methods \label{sec:methods}}
-The main goal of this paper is to measure the effect of external priors on the statistical errors of cosmological parameters derived from future CMB experiments.
-We start in \refssec{fm-formalism} by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints on the parameters of the assumed cosmological model.
-To apply this technique, we need to choose a model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the details of the CMB experiments we have in mind.
+\section{Assumptions and methods \label{sec:methods}}
+In this paper we want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
+We start this section in \refssec{fm-formalism} by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints.
+To apply this technique, we need to specify our cosmological model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the specifications of S4 CMB experiments.
 These will be presented in \refssec{cosmo-noise}.
@@
 Furthermore we will not derive the global shape of the likelihood but we will obtain parameters errors by studying small perturbations around its maximum.
-These are the two basic assumptions of the Fisher approach.
+These are the two basic assumptions of the Fisher-matrix approach.
 
 Regarding the first assumption, we recognize the fact that, even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}.
-Despite this, we do not need the level of accuracy that a careful modeling of the likelihoods will give us, because we are looking for the general behavior of parameters errors as a function of external priors.
+Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
 Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, a part for the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
 
-Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$ and we will get the errors on those by a simple analysis of the likelihood curvature around the fiducial values.
+Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$ and we will get the errors on those from the likelihood curvature around the fiducial values.
 Indeed, as usual, we define the fisher matrix elements as the curvature:
@@
 \end{equation}
-where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters array that, by definition, maximize the likelihood.
-We refer the reader to \cite{} for a detailed proof that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, the Fisher matrix of a CMB experiment can be rewritten as:
+where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximize the likelihood.
+We refer the reader to \cite{} for a detailed proof that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
 \begin{equation}
@@
  \end{equation}
- In this work we will use the CMB temperature and E mode polarization together with lensing reconstruction to constrain parameters.
-For this reason, $\boldsymbol{C}_\ell$ that encapsulate the used power spectra in a matrix structure is:
+ In this work we will use the CMB temperature and E mode polarization together with the reconstructed lensing potential to constrain parameters.
+For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
  \begin{eqnarray}
@@
 Note that we are neglecting the term $C_\ell^{E\phi}$.
-As also noticed in previous literature like (\cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
-Furthermore the terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
+As also noticed in previous literature (like \cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
+The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
 The power of the Fisher approach descend from the Cramer-Rao inequality that relates the errors on parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
@@
 Once $F_{ij}$ is computed following \refeq{Fij_def2} it is straightforwards to get the error $\sigma_i$ from \refeq{cramer-rao}.
-Furthermore in this context it is easy to introduce external priors on cosmological parameters, which is the main focus or our work.
-Indeed to add priors we simply need to add external experiments Fisher matrixes before performing the matrix inversion of \refeq{cramer-rao}, i.e.:
+Furthermore in this context it is easy to introduce external priors on cosmological parameters.
+Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiements before performing the matrix inversion of \refeq{cramer-rao}, i.e.:
 \begin{equation}
-F_{total}=F_{CMB}+F_{external}.
+F_{total}=F_{CMB}+\sum F_{external}.
 \end{equation}
-In the same way we can add priors on a single cosmological parameter just by adding the prior to the matrix element.
+In the same way we can add priors on a single cosmological parameter just by modyfing the correspondent matrix element.
 For example, a $1\%$ prior on $H_{0}$ can be obtained by:
@@
 
-We chose a Fisher matrix approach because of its ability to rapidly forecast future experiments performances without generating mock data together with the ease of including external priors. 
+We chose a Fisher matrix approach because of its ability to rapidly forecast future experiments performances without generating mock data, together with the ease of including external priors. 
 This technique however introduce some technical difficulties together the previously mentioned assumption that the likelihood is gaussian. 
-Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves contest were a lot of parameters are needed).
+Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves contest were a lot of parameters are used).
 Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
 Even if other methods of analysis have been proposed \cite{2006JCAP...10..013P,2006astro.ph..9591A} this is still the standard method used to forecast future constrains \cite{wu:2014}.
-While we how these issues can be a possible concern we carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
+We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
 We compute the derivatives in \refeq{Fij_def2} using a 5 points formula:
@@
 \end{equation}
-This high order definition allow us to use a bigger gap h around the fiducial parameters $\theta_{0}$.
+This high order definition allow us to use a bigger gap $h$ around the fiducial parameters $\theta_{0}$.
 As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing C negligible.
-We also test the robustness of this calculation by changing the gap h in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
+We also test the robustness of this calculation by changing the gap $h$ in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
 Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
@@
 \begin{center}
-\includegraphics[scale=0.6]{PS_phi_with_noise.pdf}
+\includegraphics[scale=0.4]{PS_phi_with_noise.pdf}
 \caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
@@
 \begin{center}
-\includegraphics[scale=0.6]{PS_with_noise.pdf}
+\includegraphics[scale=0.4]{PS_with_noise.pdf}
 \caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
@@
 The first step is the cosmological model we used to compute $C^{X,Y}$ with $X,Y \in \{T,E,\phi\}$.
-We parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We also allow different Dark Energy model by introducing the equations of state parameter w as a varying parameter.
+We parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy model by introducing the equations of state parameter w as a varying parameter.
 We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
@@
 Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
-Given the model, we use the CAMB software to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
+Given the model, we use the CAMB \cite{Lewis:1999bs} software to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
 Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
@@
 The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons.
-This introduce non gaussianities that couple different modes in the otherwise independent CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
-For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow \cite{okamoto:2003,hu:2002} without using more advanced iterative techniques.
+This introduce non gaussianities that couple different otherwise independent CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
+For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow the non-iterative technique of \cite{okamoto:2003,hu:2002}.
 
@@
 
-\begin{table}[htdp]
-\begin{center}
-\begin{tabular}{|c|c|}
-\hline
-$H_{0}$ & $67.11$ km/s/Mpc\\
-\hline
-\hline
-$\tau$ & $0.0925$ \\
-\hline
-
-\hline
-$A_{s}$ &$2.215 \times 10^{-9}$ \\
-\hline
-
-\hline
-$n_{s}$ & $0.9624$ \\
-\hline
-
-\hline
-$N_{eff}$ & 3.046\\
-\hline
-\end{tabular}
-\end{center}
-\caption{Fiducial values of cosmological parameters used in this work.}
-\label{default}
-\end{table}%
+%\begin{table}[htdp]
+%\begin{center}
+%\begin{tabular}{|c|c|}
+%\hline
+%$H_{0}$ & $67.11$ km/s/Mpc\\
+%\hline
+%\hline
+%$\tau$ & $0.0925$ \\
+%\hline
+%
+%\hline
+%$A_{s}$ &$2.215 \times 10^{-9}$ \\
+%\hline
+%
+%\hline
+%$n_{s}$ & $0.9624$ \\
+%\hline
+%
+%\hline
+%$N_{eff}$ & 3.046\\
+%\hline
+%\end{tabular}
+%\end{center}
+%\caption{Fiducial values of cosmological parameters used in this work.}
+%\label{default}
+%\end{table}%
 
@@
 \begin{center}
-\includegraphics[scale=0.6]{prior_w.pdf}
+\includegraphics{prior_w_snow_mass.pdf}
 \caption{}
@@
 \begin{center}
-\includegraphics[scale=0.6]{prior_massless_neutrinos.pdf}
+\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
 \caption{}
@@
 \begin{center}
-\includegraphics[scale=0.6]{prior_re_optical_depth.pdf}
+\includegraphics{prior_re_optical_depth_snow_mass.pdf}
 \caption{}
@@
 \appendix
-\section{Marginalization}
+\section{Marginalization , not needed probably}
 \begin{equation}
@@
 We thank Youngsoo Park for his contribution in the early stage of this work.
-AM want to thank Zhen Pan who allows a careful cross-check of our results.
+AM wants to thank Zhen Pan who allows a careful cross-check of our results.
 This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.
@@
 % Create the reference section using BibTeX:
-\bibliography{N_eff_prior_paperNotes}
+\bibliography{N_eff_prior_paper}
 

### 2015-09-08 Notes/N_eff_prior_paper.tex
working on the draft 2B

@@
 Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
-%Indeed recently, forecast papers validate these possible outcomes and help to plan and motivate the next generation of CMB experiments.
-However these results also indirectly disclose the needed synergy between CMB and other experiments, for example the need of imposing external prior on some of the cosmological parameters.
-We think that the extent to which CMB constraints depend on external priors needs to be explored, for instance to guide the combined effort of different experimental fields.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
@@
 
-\subsection{CMB S4 experiments: level of noise.\label{subsec:cosmo-noise}}
-In this subsection we will describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra C and the noise power N.
-The first step is the cosmological model we used to compute $C^{X,Y}$ with $X,Y \in \{T,E,\phi\}$.
-We parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy model by introducing the equations of state parameter w as a varying parameter.
+\subsection{CMB S4 experiments specifications.\label{subsec:cosmo-noise}}
+In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
+%The first step is the cosmological model we used to compute $C^{X,Y}$ with $X,Y \in \{T,E,\phi\}$.
+Firstly we parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy models by introducing the equations of state parameter $w$ as a varying parameter.
 We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
 $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc.
-Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
-Given the model, we use the CAMB \cite{Lewis:1999bs} software to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
+Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
+Given the model, we use CAMB \cite{Lewis:1999bs} to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
 Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
 
-We chose the second element of \refeq{Fij_def2}, the noise power in the CMB $N_{\ell}^{T,E}$ and in the lensing reconstruction $N_{\ell}^{\phi}$ with the next generation of CMB experiments (S4) in mind.
-For the temperature and E-mode polarization of the CMB, together with the an improved depth and resolution we will also assume that large scale foregrounds, like dust, are under control or negligible.
-We will deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
-The remaining source of noise, the instrumental noise is added to the power spectrum in the usual way:
+The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the expected level for the next generation of CMB experiments (S4).
+For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
+We deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
+The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
  \begin{equation}
@@
 \end{equation}
-where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represent the instrumental white noise.
+where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
 We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin (PRELIMINARY).
@@
 
-Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from CMB experiments.
+Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from the CMB.
 The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons.
-This introduce non gaussianities that couple different otherwise independent CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
+This introduces non gaussianities that couple different, otherwise independent, CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
 For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow the non-iterative technique of \cite{okamoto:2003,hu:2002}.

### 2015-09-08 Notes/N_eff_prior_paper.tex
working on the draft 2B

@@
 \subsection{Fisher Matrix formalism \label{subsec:fm-formalism}}
-The natural starting point to estimate errors on cosmological parameters is the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
+To estimate errors on cosmological parameters we follow the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
 These two are related by the prior probability of the parameters $P(\theta)$ through:
@@
 \end{equation}
-where, as usual, we have neglected the probability of the data itself.
+where, as usual, we have neglected the probability of the data $P(d)$.
 These general framework can be simplified in our case. We will focus not on an entirely general posterior distribution, but we will assume that the likelihoods are gaussian.
@@
 Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
-Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, a part for the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
+Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
 
@@
 \end{equation}
-where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximize the likelihood.
-We refer the reader to \cite{} for a detailed proof that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
+where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximizes the likelihood.
+ With the usual definition \cite{} $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
 \begin{equation}
@@
  \end{equation}
- In this work we will use the CMB temperature and E mode polarization together with the reconstructed lensing potential to constrain parameters.
+ In this work we will constrain cosmological parameters with the CMB temperature and E mode polarization together with the reconstructed lensing potential.
 For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
@@
 As also noticed in previous literature (like \cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
-The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
-The power of the Fisher approach descend from the Cramer-Rao inequality that relates the errors on parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
+The term $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
+The power of the Fisher approach descend from the Cramer-Rao inequality that relates the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
 \begin{equation}
@@
 Furthermore in this context it is easy to introduce external priors on cosmological parameters.
-Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiements before performing the matrix inversion of \refeq{cramer-rao}, i.e.:
+Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
 \begin{equation}
-F_{total}=F_{CMB}+\sum F_{external}.
+F_{\rm total}=F_{\rm CMB}+\sum F_{\rm external}.
 \end{equation}
-In the same way we can add priors on a single cosmological parameter just by modyfing the correspondent matrix element.
+In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
 For example, a $1\%$ prior on $H_{0}$ can be obtained by:
@@
 
-We chose a Fisher matrix approach because of its ability to rapidly forecast future experiments performances without generating mock data, together with the ease of including external priors. 
-This technique however introduce some technical difficulties together the previously mentioned assumption that the likelihood is gaussian. 
-Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves contest were a lot of parameters are used).
+We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data, together with the ease of including external priors. 
+This technique however introduces also some technical difficulties and assumptions. 
+Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
 Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
@@
 \end{equation}
+
 This high order definition allow us to use a bigger gap $h$ around the fiducial parameters $\theta_{0}$.

### 2015-09-08 Notes/N_eff_prior_paper.tex
working on the draft

@@
 \newcommand{\refssec}[1]{section~\ref{subsec:#1}}
-
+\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
 % For various journals
@@
 Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
-We find that, quite surprisingly, CMB S4 are very powerful even on .. There are few exceptions.. like $w$.
+We find that CMB S4 are quite powerful even on .. There are few exceptions.. like $w$.
 \end{abstract}
@@
 We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data, together with the ease of including external priors. 
-This technique however introduces also some technical difficulties and assumptions. 
+This technique however introduces also some technical difficulties. 
 Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
 Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
-Even if other methods of analysis have been proposed \cite{2006JCAP...10..013P,2006astro.ph..9591A} this is still the standard method used to forecast future constrains \cite{wu:2014}.
+Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard method used to forecast future constrains \cite{wu:2014}.
 We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
@@
 \section{Results \label{sec:results}}
+Our main results are shown in \reffig{prior_massless_neutrinos} \reffig{prior_w}, \reffig{prior_re_optical_depth}
 
@@
 \includegraphics{prior_w_snow_mass.pdf}
-\caption{}
+\caption{$w$ costraints will be dramatically improved by $H_{0}$ external priors. 
+\footnotesize In the figure: on the y axis we have the relative error on $w$ as a function of the prior we impose on the other parameters.
+These priors on the x-axis are measured relative to the error on that parameters from CMB Stage 4 experiments alone.}
 \label{fig:prior_w}
@@
 \includegraphics{prior_massless_neutrinos_snow_mass.pdf}
-\caption{}
+\caption{$N_{eff}$ costraints mainly benefit from $\Omega_{b}$ and $n_{s}$ external priors.}
 \label{fig:prior_massless_neutrinos}
@@
 \includegraphics{prior_re_optical_depth_snow_mass.pdf}
-\caption{}
+\caption{$\tau$ constraints are limited by the well known degeneracies with $A_{s}$. An external prior on $A_{s}$ already tightly constrained from CMB alone can still be beneficial. }
 \label{fig:prior_re_optical_depth}

### 2015-09-10 Notes/N_eff_prior_paper.tex
working on the draft:intro

@@
 These series of successes is far from its end.
-Indeed the next generation of CMB experiments (S4) is now in its planning stage and it will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As it happened in the past, this new sensitivity will improve our understanding of several areas of astrophysics like neutrinos, inflation, Dark Energy and Dark matter.
+Indeed the next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+As it happened in the past, this new sensitivity will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.
 
-%In this work we will mainly focus on two of these: neutrinos and Dark Energy.
-The main effects of neutrinos in the CMB can be parametrize by the number of relativistic degree of freedom in the early universe $N_{\rm eff}$ and the total mass of the three neutrino species $M_{\nu}$.
-Indeed the number of relativistic species, like neutrinos, in the early Universe alter the expansion history.
-This can be probed using the CMB by comparing the sound horizon scale, obtained from the CMB peaks positions, with the silk damping scale taking advantage of the fact that they scale differently with the expansion rate $H(z)$
-%their ratio $r_{d}/r_{s}\propto\sqrt{H}$ 
-(see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
-This allow CMB experiment to constrain $N_{\rm eff}$. Any deviation from the predicted value $N_{\rm eff}=3.046$ will be a sign of possible extensions of the Standard Model.
+The cosmological dependence to neutrinos is primarily due to their contribution to the number of relativistic species $N_{\rm eff}$ in the early phase of the universe  together with the role they play, given their non zero mass $M_{\nu}$, into the late growth of cosmic structures.
+Because relativistic species, like neutrinos, present in the early Universe are the main drivers of the cosmic expansion their number is encapsulated in the expansion rate $H(z)$.
+This rate can be powerfully tested using the CMB.
+Mainly by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
+The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale and will unequivocally test with data the predicted value of the standard model $N_{\rm eff}=3.046$.
 The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
-As other low redshift effects however, massive neutrinos modify the CMB through the lensing of the CMB photons by large scale structures.
-Different neutrino masses result into a different growth of structure which consequently lead to different CMB lensing.
-The lensing potential can be reconstructed from the CMB itself, opening the possibility of constraining the neutrino mass.
-In the future, according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when baryon acoustic oscillations (BAO) data are also used.
+As other low redshift effects however, massive neutrinos modify the CMB by altering the growth of the large scale structure responsible for the the lensing of the CMB photons.
+Different neutrino masses consequently lead to different CMB lensing.
+Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise drastically improving the lensing reconstruction.
+This will turn in a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
+To put this progress in prospective, CMB alone will constrain the relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when we introduce priors from baryon acoustic oscillations (BAO) data \cite{wu:2014}.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-As previously mentioned, CMB is also sensitive to  the Dark energy properties (see \cite{2010MNRAS.405.2639J}).
+As previously mentioned, CMB is also sensitive to the properties of Dark energy (see \cite{2010MNRAS.405.2639J}).
 The current generation of experiments measured with percent accuracy the energy density of Dark Energy.
-The challenge for future experiments is to identify the nature of this mysterious component.
-The first step will be to measure any deviation in the equation of state from the one predicted by a cosmological constant model.
+The challenge for future experiments is to reveal the nature of this mysterious component.
+For example a crucial step will be to investigate any possible deviations in the equation of state from the value $w=-1$ predicted by a cosmological constant model.
 Furthermore CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
-Indeed dark energy affect the CMB because due to its effect on the universe's expansion that modifies the distance to the last scattering surface together with a modification  of CMB photons energy through the Integrated Sachs Wolfe Effect.
-Furthermore different DE models lead to a different growth of large scale structure.
-However dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
-To get competitive constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
+Indeed dark energy affects the CMB because it alters the universe's expansion consequently changing the distance to the last scattering surface.
+Furthermore different DE models lead to a different growth of large scale structure that can be inferred from the CMB lensing.
+However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
+Similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
 
@@
 In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae.
-To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
-%This can tell us for example which CMB parameter constraints will benefit more from improvement on external priors or if proposed experiments like PIXIE \cite{2011JCAP...07..025K} can indirectly also benefit CMB constraints.
+To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study in detail the dependance of CMB future constrains on the external priors assumed.
 

### 2015-09-10 Notes/N_eff_prior_paper.tex
working on the draft:intro

@@
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
-Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
+Here, we examine quantitatively with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
 We find that CMB S4 are quite powerful even on .. There are few exceptions.. like $w$.
@@
 \section{Introduction}\label{sec:intro}
-Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
-They will maintain their role also in the near future thanks to the unprecedented level of noise and accuracy of their next generation (S4).
+Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe and they will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation.
 The recent past is indeed reassuring. 
-Every new generation of satellite experiments improved the sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
+Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
 This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
-With the improved sensitivity we managed to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
-The same success characterized ground experiments where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
+With the improved sensitivity we extend our understanding of the universe scientific from a proof of its almost flat geometry to a well tested model $\Lambda$CDM with a percent level constraint on its parameters.
+The same success characterized ground CMB experiments, where the evolution from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
 These series of successes is far from its end.
 Indeed the next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As it happened in the past, this new sensitivity will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.
+As it happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.
 
 The cosmological dependence to neutrinos is primarily due to their contribution to the number of relativistic species $N_{\rm eff}$ in the early phase of the universe  together with the role they play, given their non zero mass $M_{\nu}$, into the late growth of cosmic structures.
-Because relativistic species, like neutrinos, present in the early Universe are the main drivers of the cosmic expansion their number is encapsulated in the expansion rate $H(z)$.
-This rate can be powerfully tested using the CMB.
-Mainly by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
-The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale and will unequivocally test with data the predicted value of the standard model $N_{\rm eff}=3.046$.
+Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$.
+This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
+The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale of the CMB power spectrum and will unequivocally test the value $N_{\rm eff}=3.046$ predicted by the standard model .
 The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
@@
 Different neutrino masses consequently lead to different CMB lensing.
-Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise drastically improving the lensing reconstruction.
-This will turn in a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
-To put this progress in prospective, CMB alone will constrain the relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when we introduce priors from baryon acoustic oscillations (BAO) data \cite{wu:2014}.
+Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction.
+This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
+Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-As previously mentioned, CMB is also sensitive to the properties of Dark energy (see \cite{2010MNRAS.405.2639J}).
-The current generation of experiments measured with percent accuracy the energy density of Dark Energy.
-The challenge for future experiments is to reveal the nature of this mysterious component.
-For example a crucial step will be to investigate any possible deviations in the equation of state from the value $w=-1$ predicted by a cosmological constant model.
-Furthermore CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
-Indeed dark energy affects the CMB because it alters the universe's expansion consequently changing the distance to the last scattering surface.
-Furthermore different DE models lead to a different growth of large scale structure that can be inferred from the CMB lensing.
+As previously mentioned, CMB is also sensitive to the properties of dark energy (see \cite{2010MNRAS.405.2639J}).
+Thanks to the current generation of experiments we know with extraordinary accuracy its energy density.
+The challenge for the next generation is to reveal the nature of this mysterious component.
+For example, a crucial step to rule out part of the proposed models will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant.
+Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface.
+Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing.
+Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
 However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
@@
 
-In general, external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
-For this reason the synergy with other experiments is crucial.
+The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector. 
+The future of cosmology will hinge on the ability of combining different probes.
+In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
 In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
 To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study in detail the dependance of CMB future constrains on the external priors assumed.
+For all the parameters of the standard $\Lambda$CDM with massive neutrinos and $w\neq-1$ dark energy model we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
 
 This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
-In \refsec{results} we will describe our results and we then conclude with a discussion in \refsec{conclusions}.
+In \refsec{results} we will describe our results and we then conclude with a discussion of them in \refsec{conclusions}.
 
@@
 In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
-%The first step is the cosmological model we used to compute $C^{X,Y}$ with $X,Y \in \{T,E,\phi\}$.
 Firstly we parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy models by introducing the equations of state parameter $w$ as a varying parameter.
@@
 % only one appendix.
-\appendix
-\section{Marginalization , not needed probably}
-\begin{equation}
-G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
-\end{equation}
-where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
+
+%\appendix
+%\section{Marginalization , not needed probably}
+%\begin{equation}
+%G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
+%\end{equation}
+%where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
 marginal\}$; therefore, $F^{\phi\phi}$ is the block of the total Fisher matrix containing the parameters we want to constrain, whilst $F^{\psi\psi}$ is the nuisance-parameter Fisher sub-matrix.

### 2015-09-11 Notes/N_eff_prior_paper.tex
small chnges

--- 2015-09-08 Notes/N_eff_prior_paper.tex
@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,prd,reprint,superscriptaddress]{revtex4-1}
+\documentclass[aps,prd,preprint,groupedaddress]{revtex4-1}
 %\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
@@
 \usepackage{graphicx}
-%\input{papers_packages_command._include_tex}
-\newcommand\refsec[1]{\S\ref{sec:#1}}
-\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
-\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
-\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
-% For various journals
-%\newcommand{\apj}{ApJ}
-\newcommand{\physrep}{Physics Reports}
-\newcommand{\aapr}{A\&A Rev.}
-\newcommand{\apjl}{ApJL}
-\newcommand{\apjs}{ApJS}
-\newcommand{\aap}{A\&A}
-\newcommand{\mnras}{MNRAS}
-%\newcommand{\prd}{Phys. Rev. D}
-%\newcommand{\physrev}{Phys. Rev.}
-\newcommand{\physrevlett}{Phys. Rev. Lett.}
-\newcommand{\jcap}{{\em JCAP }}
-\newcommand{\aj}{AJ }
-
+\input{papers_packages_command._include_tex}
+%\newcommand\refsec[1]{\S\ref{sec:#1}}
+%\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
 \bibliographystyle{apsrev4-1}
@@
 \author{Alessandro Manzotti}
-\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
-\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+%\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
 \author{Scott Dodelson}
-
-\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
 
 \date{\today}
+\linenumbers
 \begin{abstract}
 The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy.
-Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
+%Indeed these experiments will likely exploit the sensitivity of the CMB to the details of physical processes in the early Universe (through temperature and polarization) as well as in the late universe (through CMB secondary effects like lensing).
+%They will be possible because the temperature anisotropies and E mode polarization of the CMB will be measured with cosmic variance limited precision. 
+The low level of noise, together with improved angular resolution, will allow CMB experiments to reconstruct the lensing potential of high redshift large scale structure with a signal to noise bigger than one over a wide range of scales.
+%Indeed recently, forecast papers validate these possible outcomes and help to plan and motivate the next generation of CMB experiments.
+However these results also indirectly disclose the needed synergy between CMB and other experiments, for example the need of imposing external prior on some of the cosmological parameters.
+We think that the extent to which CMB constraints depend on external priors needs to be explored, for instance to guide the combined effort of different experimental fields.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
 Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
-We find that CMB S4 are quite powerful even on .. There are few exceptions.. like $w$.
+We find that quite surprisingly CMB S4 are very powerful even on .. There are few exceptions.. like $w$.
+
+%\begin{itemize}
+%\item We are now using a simplistic lensing noise but I have lensing noise with the usual Hu Okamoto formula. Iterative Seljack not there yet. Needed?
+%\item We are varying $\tau, ~ n_{s}, ~ A_{s}, ~ N_{\rm eff}, ~ H_{0}, w,~\Omega_{\nu}h^{2},~\Omega_{c}h^{2},~\Omega_{b}h^{2}$.
+The neutrino sector consists of one massive neutrino of $m_{\nu}=0.083$ eV, which corresponds to $\Omega_{\nu}h^{2}=0.0009$.
+%\item Next TODO: Check for bugs and errors code (DONE look at Zhen test).
+PCA?
+what prior is more important?
+Full MCMC should not be extremely hard with cosmosis (I implemented a toy likelihood analysis to be improved).
+%\end{itemize}
+%
+
 \end{abstract}
@@
 \section{Introduction}\label{sec:intro}
+
 Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
-They will maintain their role also in the near future thanks to the unprecedented level of noise and accuracy of their next generation (S4).
-The recent past is indeed reassuring. 
-Every new generation of satellite experiments improved the sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
-This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
-With the improved sensitivity we managed to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
-The same success characterized ground experiments where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
+Moreover they also represent an amazing success in instrumental physics.
+Indeed every new generation of satellite experiments improved its sensitivity gaining almost a factor of ten in sensitivity compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
+Experimentally we went from observing the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
+With the improved sensitivity we menaged to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
+The same is true for ground experiment where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
+Ground experiments have already reached the maximum possible sensitivity per detector but will significantly increase the number of detectors in future generation.
 These series of successes is far from its end.
-Indeed the next generation of CMB experiments (S4) is now in its planning stage and it will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As it happened in the past, this new sensitivity will improve our understanding of several areas of astrophysics like neutrinos, inflation, Dark Energy and Dark matter.
+Indeed the next generation of CMB experiments (S4) is now in its planning stage and it promises to measure with cosmic variance limited precision the E-mode polarization together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+As happened in the past this new sensitivity will improve our understanding of several areas like neutrinos, inflation, Dark Energy and Dark matter.
+In this work we will mainly focus on two of these: neutrinos and Dark Energy.
 
-%In this work we will mainly focus on two of these: neutrinos and Dark Energy.
-The main effects of neutrinos in the CMB can be parametrize by the number of relativistic degree of freedom in the early universe $N_{\rm eff}$ and the total mass of the three neutrino species $M_{\nu}$.
+The main aspects of neutrinos physics that CMB can access are the number of relativistic degree of freedom in the early universe $N_{\rm eff}$ and the total mass of the three neutrino species $M_{\nu}$.
 Indeed the number of relativistic species, like neutrinos, in the early Universe alter the expansion history.
-This can be probed using the CMB by comparing the sound horizon scale, obtained from the CMB peaks positions, with the silk damping scale taking advantage of the fact that they scale differently with the expansion rate $H(z)$
-%their ratio $r_{d}/r_{s}\propto\sqrt{H}$ 
-(see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
-This allow CMB experiment to constrain $N_{\rm eff}$. Any deviation from the predicted value $N_{\rm eff}=3.046$ will be a sign of possible extensions of the Standard Model.
-The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
-As other low redshift effects however, massive neutrinos modify the CMB through the lensing of the CMB photons by large scale structures.
-Different neutrino masses result into a different growth of structure which consequently lead to different CMB lensing.
+This effect can be probed with CMB measurement by comparing the sound horizon scale, obtained from the CMB peaks position, with the silk damping scale.
+Indeed because they scale differently with the expansion rate $H(z)$ their ratio $r_{d}/r_{s}\propto\sqrt{H}$ (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
+This allow us to constrain $N_{\rm eff}$ and any deviation from the predicted value $N_{\rm eff}=3.046$ will be a sign of possible extensions of the Standard Model.
+The mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
+As other low redshift effects however, massive neutrinos modify the CMB through the lensing effect of large scale structure on the CMB photons.
+Different neutrino masses result into different growth of structure which consequently lead to different CMB lensing.
 The lensing potential can be reconstructed from the CMB itself, opening the possibility of constraining the neutrino mass.
-In the future, according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when baryon acoustic oscillations (BAO) data are also used.
+In the future, according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement when BAO data are also used.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-As previously mentioned, CMB is also sensitive to  the Dark energy properties (see \cite{2010MNRAS.405.2639J}).
+As previously mentioned, CMB is also sensitive to Dark energy (see \cite{2010MNRAS.405.2639J}).
 The current generation of experiments measured with percent accuracy the energy density of Dark Energy.
-The challenge for future experiments is to identify the nature of this mysterious component.
-The first step will be to measure any deviation in the equation of state from the one predicted by a cosmological constant model.
+The next big challenge is to identify the nature of this mysterious component.
+The first step will be to measure any deviation from the equation of state predicted by a cosmological constant model.
 Furthermore CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
-Indeed dark energy affect the CMB because due to its effect on the universe's expansion that modifies the distance to the last scattering surface together with a modification  of CMB photons energy through the Integrated Sachs Wolfe Effect.
-Furthermore different DE models lead to a different growth of large scale structure.
+Indeed dark energy mainly affect the CMB because of its effect on the universe's expansion that modifies the distance to the last scattering surface.
+Furthermore the influence of Dark Energy on the low redshift universe modify the energy of CMB photons through the Integrated Sachs Wolfe Effect.
+Also, as it was true in the neutrino case, CMB lensing can probe the growth of structure that depends on the dark energy properties and in particular on its equation of state.
 However dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
@@
 
-In general, external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
-For this reason the synergy with other experiments is crucial.
+The importance of external priors in the dark energy case is actually more general issue.
+Even if it is certainly true that CMB S4 will open an all new window on several area of physics and cosmology the synergy with other experiments is crucial.
 In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae.
+For this reason forecasted constrains very often assume external realistic priors.
 To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
-%This can tell us for example which CMB parameter constraints will benefit more from improvement on external priors or if proposed experiments like PIXIE \cite{2011JCAP...07..025K} can indirectly also benefit CMB constraints.
+This can tell us for example which CMB parameter constraints will benefit more from improvement on external priors or if proposed experiments like PIXIE \cite{2011JCAP...07..025K} can indirectly also benefit CMB constraints.
 
 This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
-In \refsec{results} we will describe our results and we then conclude with a discussion in \refsec{conclusions}.
+In \refsec{results} we will describe our findings and we then conclude with a discussion of our results in \refsec{conclusions}.
 
-\section{Assumptions and methods \label{sec:methods}}
-In this paper we want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
-We start this section in \refssec{fm-formalism} by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints.
-To apply this technique, we need to specify our cosmological model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the specifications of S4 CMB experiments.
+\section{Assumption and methods \label{sec:methods}}
+The main goal of this paper is to measure the effect of external priors on the statistical errors of cosmological parameters derived from future CMB experiments.
+We start in \refssec{fm-formalism} by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints on the parameters of the assumed cosmological model.
+To apply this technique, we need to choose a model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the details of the CMB experiments we have in mind.
 These will be presented in \refssec{cosmo-noise}.
@@
 \subsection{Fisher Matrix formalism \label{subsec:fm-formalism}}
-To estimate errors on cosmological parameters we follow the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
+The natural starting point to estimate errors on cosmological parameters is the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
 These two are related by the prior probability of the parameters $P(\theta)$ through:
@@
 \end{equation}
-where, as usual, we have neglected the probability of the data $P(d)$.
+where, as usual, we have neglected the probability of the data itself.
 These general framework can be simplified in our case. We will focus not on an entirely general posterior distribution, but we will assume that the likelihoods are gaussian.
 Furthermore we will not derive the global shape of the likelihood but we will obtain parameters errors by studying small perturbations around its maximum.
-These are the two basic assumptions of the Fisher-matrix approach.
+These are the two basic assumptions of the Fisher approach.
 
 Regarding the first assumption, we recognize the fact that, even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}.
-Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
-Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
+Despite this, we do not need the level of accuracy that a careful modeling of the likelihoods will give us, because we are looking for the general behavior of parameters errors as a function of external priors.
+Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, a part for the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
 
-Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$ and we will get the errors on those from the likelihood curvature around the fiducial values.
+Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$ and we will get the errors on those by a simple analysis of the likelihood curvature around the fiducial values.
 Indeed, as usual, we define the fisher matrix elements as the curvature:
@@
 \end{equation}
-where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximizes the likelihood.
- With the usual definition \cite{} $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
+where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters array that, by definition, maximize the likelihood.
+We refer the reader to \cite{} for a detailed proof that with the usual definition of power spectrum $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, the Fisher matrix of a CMB experiment can be rewritten as:
 \begin{equation}
@@
  \end{equation}
- In this work we will constrain cosmological parameters with the CMB temperature and E mode polarization together with the reconstructed lensing potential.
-For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
+ In this work we will use the CMB temperature and E mode polarization together with lensing reconstruction to constrain parameters.
+For this reason, $\boldsymbol{C}_\ell$ that encapsulate the used power spectra in a matrix structure is:
  \begin{eqnarray}
@@
 Note that we are neglecting the term $C_\ell^{E\phi}$.
-As also noticed in previous literature (like \cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
-The term $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
-The power of the Fisher approach descend from the Cramer-Rao inequality that relates the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
+As also noticed in previous literature like (\cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
+Furthermore the terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
+The power of the Fisher approach descend from the Cramer-Rao inequality that relates the errors on parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
 \begin{equation}
@@
 Once $F_{ij}$ is computed following \refeq{Fij_def2} it is straightforwards to get the error $\sigma_i$ from \refeq{cramer-rao}.
-Furthermore in this context it is easy to introduce external priors on cosmological parameters.
-Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
+Furthermore in this context it is easy to introduce external priors on cosmological parameters, which is the main focus or our work.
+Indeed to add priors we simply need to add external experiments Fisher matrixes before performing the matrix inversion of \refeq{cramer-rao}, i.e.:
 \begin{equation}
-F_{\rm total}=F_{\rm CMB}+\sum F_{\rm external}.
+F_{total}=F_{CMB}+F_{external}.
 \end{equation}
-In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
+In the same way we can add priors on a single cosmological parameter just by adding the prior to the matrix element.
 For example, a $1\%$ prior on $H_{0}$ can be obtained by:
@@
 
-We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data, together with the ease of including external priors. 
-This technique however introduces also some technical difficulties. 
-Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
+We chose a Fisher matrix approach because of its ability to rapidly forecast future experiments performances without generating mock data together with the ease of including external priors. 
+This technique however introduce some technical difficulties together the previously mentioned assumption that the likelihood is gaussian. 
+Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves contest were a lot of parameters are needed).
 Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
-Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard method used to forecast future constrains \cite{wu:2014}.
-We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
+Even if other methods of analysis have been proposed \cite{2006JCAP...10..013P,2006astro.ph..9591A} this is still the standard method used to forecast future constrains \cite{wu:2014}.
+While we how these issues can be a possible concern we carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
 We compute the derivatives in \refeq{Fij_def2} using a 5 points formula:
@@
 \end{equation}
-
-This high order definition allow us to use a bigger gap $h$ around the fiducial parameters $\theta_{0}$.
+This high order definition allow us to use a bigger gap h around the fiducial parameters $\theta_{0}$.
 As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing C negligible.
-We also test the robustness of this calculation by changing the gap $h$ in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
+We also test the robustness of this calculation by changing the gap h in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
 Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
@@
 \begin{center}
-\includegraphics[scale=0.4]{PS_phi_with_noise.pdf}
+\includegraphics[scale=0.6]{PS_phi_with_noise.pdf}
 \caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
@@
 \begin{center}
-\includegraphics[scale=0.4]{PS_with_noise.pdf}
+\includegraphics[scale=0.6]{PS_with_noise.pdf}
 \caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
@@
 
-\subsection{CMB S4 experiments specifications.\label{subsec:cosmo-noise}}
-In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
-%The first step is the cosmological model we used to compute $C^{X,Y}$ with $X,Y \in \{T,E,\phi\}$.
-Firstly we parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy models by introducing the equations of state parameter $w$ as a varying parameter.
+\subsection{CMB S4 experiments: level of noise.\label{subsec:cosmo-noise}}
+In this subsection we will describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra C and the noise power N.
+The first step is the cosmological model we used to compute $C^{X,Y}$ with $X,Y \in \{T,E,\phi\}$.
+We parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We also allow different Dark Energy model by introducing the equations of state parameter w as a varying parameter.
 We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
 $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc.
-Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
-Given the model, we use CAMB \cite{Lewis:1999bs} to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
+Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
+Given the model, we use the CAMB software to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
 Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
 
-The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the expected level for the next generation of CMB experiments (S4).
-For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
-We deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
-The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
+We chose the second element of \refeq{Fij_def2}, the noise power in the CMB $N_{\ell}^{T,E}$ and in the lensing reconstruction $N_{\ell}^{\phi}$ with the next generation of CMB experiments (S4) in mind.
+For the temperature and E-mode polarization of the CMB, together with the an improved depth and resolution we will also assume that large scale foregrounds, like dust, are under control or negligible.
+We will deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
+The remaining source of noise, the instrumental noise is added to the power spectrum in the usual way:
  \begin{equation}
@@
 \end{equation}
-where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
+where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represent the instrumental white noise.
 We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin (PRELIMINARY).
@@
 
-Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from the CMB.
+Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from CMB experiments.
 The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons.
-This introduces non gaussianities that couple different, otherwise independent, CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
-For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow the non-iterative technique of \cite{okamoto:2003,hu:2002}.
+This introduce non gaussianities that couple different modes in the otherwise independent CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
+For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow \cite{okamoto:2003,hu:2002} without using more advanced iterative techniques.
 
@@
 \section{Results \label{sec:results}}
-Our main results are shown in \reffig{prior_massless_neutrinos} \reffig{prior_w}, \reffig{prior_re_optical_depth}
 
-%\begin{table}[htdp]
-%\begin{center}
-%\begin{tabular}{|c|c|}
-%\hline
-%$H_{0}$ & $67.11$ km/s/Mpc\\
-%\hline
-%\hline
-%$\tau$ & $0.0925$ \\
-%\hline
-%
-%\hline
-%$A_{s}$ &$2.215 \times 10^{-9}$ \\
-%\hline
-%
-%\hline
-%$n_{s}$ & $0.9624$ \\
-%\hline
-%
-%\hline
-%$N_{eff}$ & 3.046\\
-%\hline
-%\end{tabular}
-%\end{center}
-%\caption{Fiducial values of cosmological parameters used in this work.}
-%\label{default}
-%\end{table}%
+\begin{table}[htdp]
+\begin{center}
+\begin{tabular}{|c|c|}
+\hline
+$H_{0}$ & $67.11$ km/s/Mpc\\
+\hline
+\hline
+$\tau$ & $0.0925$ \\
+\hline
+
+\hline
+$A_{s}$ &$2.215 \times 10^{-9}$ \\
+\hline
+
+\hline
+$n_{s}$ & $0.9624$ \\
+\hline
+
+\hline
+$N_{eff}$ & 3.046\\
+\hline
+\end{tabular}
+\end{center}
+\caption{Fiducial values of cosmological parameters used in this work.}
+\label{default}
+\end{table}%
 
@@
 \begin{center}
-\includegraphics{prior_w_snow_mass.pdf}
-\caption{$w$ costraints will be dramatically improved by $H_{0}$ external priors. 
-\footnotesize In the figure: on the y axis we have the relative error on $w$ as a function of the prior we impose on the other parameters.
-These priors on the x-axis are measured relative to the error on that parameters from CMB Stage 4 experiments alone.}
+\includegraphics[scale=0.6]{prior_w.pdf}
+\caption{}
 \label{fig:prior_w}
@@
 \begin{center}
-\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
-\caption{$N_{eff}$ costraints mainly benefit from $\Omega_{b}$ and $n_{s}$ external priors.}
+\includegraphics[scale=0.6]{prior_massless_neutrinos.pdf}
+\caption{}
 \label{fig:prior_massless_neutrinos}
@@
 \begin{center}
-\includegraphics{prior_re_optical_depth_snow_mass.pdf}
-\caption{$\tau$ constraints are limited by the well known degeneracies with $A_{s}$. An external prior on $A_{s}$ already tightly constrained from CMB alone can still be beneficial. }
+\includegraphics[scale=0.6]{prior_re_optical_depth.pdf}
+\caption{}
 \label{fig:prior_re_optical_depth}
@@
 \appendix
-\section{Marginalization , not needed probably}
+\section{Marginalization}
 \begin{equation}
@@
 We thank Youngsoo Park for his contribution in the early stage of this work.
-AM wants to thank Zhen Pan who allows a careful cross-check of our results.
+AM want to thank Zhen Pan who allows a careful cross-check of our results.
 This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.

### 2015-09-11 Notes/N_eff_prior_paper.tex
last section with new suggestions for the figures

--- 2015-09-10 Notes/N_eff_prior_paper.tex
@@
 %\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
-\usepackage{bm}
+%\usepackage{bm}
 \usepackage{amsmath}
@@
 \usepackage{graphicx}
+%<<<<<<< HEAD
+%\input{papers_packages_command._include_tex}
+
+%\newcommand\refsec[1]{\S\ref{sec:#1}}
+%\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
 %\input{papers_packages_command._include_tex}
@@
 \newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
-% For various journals
 %\newcommand{\apj}{ApJ}
@@
 %\newcommand{\prd}{Phys. Rev. D}
-%\newcommand{\physrev}{Phys. Rev.}
+\newcommand{\physrev}{Phys. Rev.}
 \newcommand{\physrevlett}{Phys. Rev. Lett.}
@@
 
+%>>>>>>> 683231e77bc036bad411dd2244ca99280568d723
 \bibliographystyle{apsrev4-1}
@@
 \section{Assumptions and methods \label{sec:methods}}
+
+{\bf A lot of this is standard and need not be included, which will make the paper much shorter: a good thing!}
+
 In this paper we want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
@@
 \section{Results \label{sec:results}}
-Our main results are shown in \reffig{prior_massless_neutrinos} \reffig{prior_w}, \reffig{prior_re_optical_depth}
+
+{\bf Let's open up one parameter at a time: first standard LCDM with neff, and then with mnu, and then with w.
+Also, let's show a curve in each plot with constraints on *all*  parameters, not just one of them, as that will give people a sense of what is gained if multiple priors can be obtained.}
+
+\subsection{Relativistic Degrees of Freedom}
+
+In the standard cosmology, there are three active neutrinos thermally produced in the early universe.
+Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation.
+Decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$.
+The additional fraction depends not only on well-known neutrino scattering rates but also finite temperature quantum corrections.
+Were we to measure this deviation of $N_{\rm eff}$ from 3, it would be an indication that we understand well physics when the Universe was about a second old, so this is a natural target for upcoming experiments.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
+\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the 6 cosmological parameters in the standard model.
+Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{old}=4.1\%$.
+The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.
+The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
+\label{fig:prior_massless_neutrinos}
+\end{center}
+\end{figure}
+
+\reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other 6 parameters in the standard cosmological model ($H_0, \Omega_mh^2,\Omega_bh^2, n_s, A_s,$ and $\tau$).
+With no priors on the other parameters, the projected 1-sigma error is  about 0.02, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$.
+The sensitivity comes from the effect of extra species on the damping tail of the CMB anisotropies, both in temperature and polarization, so parameters that also affect the shape of the damping tail, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are most degenerate with $N_{\rm eff}$.
+As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction.
+The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
+
+\subsection{Neutrino Masses}
+
+Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget.
+This effect, probed by CMB lensing, will allow us to determine how much of the matter density is made up of neutrinos.
+Since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$.
+This is particularly exciting because there is a lower limit on $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{prior_omnuh2_snow_mass.pdf}
+\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
+Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
+The lower limit on the sum of the neutrino masses is 50 meV.}
+\label{fig:prior_omeganuh2}
+\end{center}
+\end{figure}
+
+\reffig{prior_omeganuh2} shows that, without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
+Adding in external priors helps this significantly though, as discussed in \citet{2013arXiv1309.5383A}.
+There, the prior was deescribed as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
+Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$, which indeed determines these low-redshift distances.
+As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior.
+Adding in other priors would help further.
 
@@
 
+\subsection{Dark Energy Equation of State}
+
+While the primordial CMB has limited information about the late-time dark energy equation of state, CMB lensing probes the growth of structure at late times and therefore is sensitive to $w$.
+Here we are using only the information from the lensing power spectrum as inferred from the CMB, but there is even more information not considered here contained in the many galaxy clusters that upcoming missions will detect via the Sunyaev-Zel'dovich effect.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{prior_w_snow_mass.pdf}
+\caption{$w$ costraints will be dramatically improved by $H_{0}$ external priors. 
+\footnotesize In the figure: on the y axis we have the relative error on $w$ as a function of the prior we impose on the other parameters.
+These priors on the x-axis are measured relative to the error on that parameters from CMB Stage 4 experiments alone.}
+\label{fig:prior_w}
+\end{center}
+\end{figure}
+
+\reffig{prior_w} shows that without any priors, CMB experiments will not do much better than current constraints, which hover around 10\%.
+However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably.
+{\bf We should try to understand why.}
+
+{\bf I don't really understand this table.}
 \begin{table}[htdp]
@@
 \end{table}%
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{prior_w_snow_mass.pdf}
-\caption{$w$ costraints will be dramatically improved by $H_{0}$ external priors. 
-\footnotesize In the figure: on the y axis we have the relative error on $w$ as a function of the prior we impose on the other parameters.
-These priors on the x-axis are measured relative to the error on that parameters from CMB Stage 4 experiments alone.}
-\label{fig:prior_w}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
-\caption{$N_{eff}$ costraints mainly benefit from $\Omega_{b}$ and $n_{s}$ external priors.}
-\label{fig:prior_massless_neutrinos}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{prior_re_optical_depth_snow_mass.pdf}
-\caption{$\tau$ constraints are limited by the well known degeneracies with $A_{s}$. An external prior on $A_{s}$ already tightly constrained from CMB alone can still be beneficial. }
-\label{fig:prior_re_optical_depth}
-\end{center}
-\end{figure}
-
-\section{Conclusions \label{sec:conclusions}}
---We find that...
-
---This means that supernova experiments CMB spatial distortions and large scale structure can/will/should
-
---Future works
 

### 2015-09-14 n_priors_code/fisher_joint_test.py
speed 3x improve in fisher

@@ -57,48 +57,27 @@ def fsky2arcmin(fsky):
     return 41253. * fsky * 60. * 60.
 
 
-def calc_deriv_vectorial(data):
-    '''
+# def calc_deriv_vectorial(iell, ell, fisher_index ,parbin, data,pargaps,values):
+#     '''
+#     Given CMB data dats it normalize them anf create a 3x3 matrix
 
-    Work in progress vectorial derivative calcualtion this can improve a lot the speed
+#     ell is the multiple
+#     iell is the index in the data ell corresponds to
 
-    remember the order from CAMB
-     l CTT CEE CBB CTE Cdd CdT CdE
-
-    ell: array of l values
-    should be easy you gest have to deal with the slight difference in ell starting finishing point
-    '''
-    # noise definition from the number of observations and time
-    # eq 1 of W.hu et al snowmass paper 10^6 detectors
-    # Y = 0.25  # 25% yeld
-    # N_det = 10 ** 6  # 1 milion of detectors
-    # these are taken from global
-    ell = data[:, 0, 0]
-
-    global Y, sec_of_obs, arcmin_from_fsky
-    # PREPARE NOISE LEVEL
-
-    s = 350. * np.sqrt(arcmin_from_fsky) / np.sqrt(N_det * Y * sec_of_obs)  # half sky in arcmin^2
-    t = 1. / 60. / 180. * np.pi  # 2arcmin to rads beam
-    fac = (ell * (ell + 1.) / 2. / np.pi) / (7.4311 * 10 ** 12)
-    fac2 = (ell * (ell + 1.))
-    # Final CMB noise definition
-    N = (s * np.pi / 180. / 60.) ** 2 * np.exp(ell * (ell + 1.) * t ** 2 / 8. / np.log(2))
-    # this noise is in mu_K so check CMB accordingly
-    # N_phi = 0. * N_phi_l[iell, 1] * ell ** 2
-    # is it a 3x3 matrix? with
-    # TT,TE,Tphi
-    # TE,EE,Ephi
-    # phiT,phiE,phiphi
-    C = np.array([[data[iell, 1, parbin] + fac * N, data[iell, 4, parbin], data[iell, 6, parbin]],
-                  [data[iell, 4, parbin], data[iell, 2, parbin] + fac * N * 2.,               0.],
-                  [data[iell, 6, parbin],          0.,         data[iell, 5, parbin] + N_phi_l[iell, 1]]]
[…]

### 2015-09-14 n_priors_code/fisher_old.py

@@ -1,162 +1,0 @@
-'''
-
-Simple fisher code to test prior effect on N effective neutrino estimates
-
-TODO:
-HIGH: lensing noise (almost there: Komatsu perfect is there other from quicklens almost)
-
-low importance: ini file, automate the data creation process and derivative. Think about PCA to understand what is more important for N_eff
-
-what are this data? documents better
-
-create classes or at least external utilities. Maybe a better way to interface with cosmosis
-
-
-The matrix is going to run an 4 parameters:
-
-CONVENTIONS:
-
-PARAMETER ORDER = h0, ns, As, Neff
-
-'''
-
-__author__ = "Y. Park"
-
-
-import numpy as np
-import math
-import matplotlib.pyplot as plt
-import util
-
-# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
-
-
-def years2sec(years):
-    return years * 365 * 24. * 60. * 60.
-
-lmax = 3000
-
-N_phi_l = np.loadtxt('multipole_noisebias.txt')
-
-
-def C(iell, ell, parbin):
-    '''
-    Given CMB data dats it normalize them anf create a 3x3 matrix
-
-    ell is the multiple
-    iell is the index in the data ell corresponds to
-
-    '''
-
-    # noise definition from the number of observations and time
-    # eq 1 of W.hu et al snowmass paper 10^6 detectors
-    Y = 0.25  # 25%yeld
-    N_det = 10 ** 6  # 1 milion of detectors
-    s = 350 * math.sqrt(20626 * 60 * 60) / math.sqrt(N_det * Y * years2sec(5))  # half sky in arcmin^2
-    # print 'noise is ',s
-    # s = 0.48 as in table from paper so it is ok.
-    t = 2 / 60 / 180 * math.pi  # 2arcmin to rads beam
-    fac = ell * (ell + 1.) / 2. / math.pi
-    fac2 = ell ** 4 / ell / (ell + 1)
-
-    # Final CMB noise definition
-    N = 0#s ** 2 * math.exp(ell * (ell + 1) * t ** 2 / 8 / math.log(2))
-    # N_phi = 0. * N_phi_l[iell, 1] * ell ** 2
-
-    # Check again in particular cosmosis ouptut lensing
[…]

### 2015-09-14 n_priors_code/utils.py

@@ -92,7 +92,7 @@ def study_prior_H0_on_N_eff():
         # fid['re_optical_depth']) ** 2)
 
         d.append(
-            math.sqrt(np.linalg.inv(fisher1)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
+            np.sqrt(np.linalg.inv(fisher1)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
 
         fisher2 = fisher.copy()
         # Cicle on H0 priors
@@ -110,7 +110,7 @@ def study_prior_H0_on_N_eff():
 
         # Invert and get Neff error with these priors
         d2.append(
-            math.sqrt(np.linalg.inv(fisher2)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
+            np.sqrt(np.linalg.inv(fisher2)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
 
         fisher3 = fisher.copy()[[fid.keys().index('re_optical_depth'), fid.keys().index('massless_neutrinos')], :][
             :, [fid.keys().index('re_optical_depth'), fid.keys().index('massless_neutrinos')]]
@@ -120,7 +120,7 @@ def study_prior_H0_on_N_eff():
 
         # Invert and get Neff error with these priors
         d3.append(
-            math.sqrt(np.linalg.inv(fisher3)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
+            np.sqrt(np.linalg.inv(fisher3)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
 
         np.savetxt('output_cmb/sigma_tau_1percent.txt', d2)
         np.savetxt('output_cmb/sigma_tau_noPrior.txt', d)
@@ -155,7 +155,7 @@ def study_prior_tau_on_N_eff(fid, fisher, output_folder, header):
         # fid['re_optical_depth']) ** 2)
 
         d.append(
-            math.sqrt(np.linalg.inv(fisher1)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
+            np.sqrt(np.linalg.inv(fisher1)[fid.keys().index('massless_neutrinos'), fid.keys().index('massless_neutrinos')]))
 
         fisher2 = fisher.copy()
         # Cicle on H0 priors
[…]

### 2015-09-14 Notes/N_eff_prior_paper.tex

@@
 Here, we examine quantitatively with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
-We find that CMB S4 are quite powerful even on .. There are few exceptions.. like $w$.
+We find that CMB S4 
 \end{abstract}
@@
 \begin{equation}
-\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim \frac{-C(\theta_{0}+2h)+8C(\theta_{0}+h)-8C(\theta_{0}-h)+C(\theta_{0}-2h)}{12h}.
+\begin{split}
+\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim & \frac{1}{12 h} [ -C(\theta_{0}+2h)+ 8C(\theta_{0}+h) \\ &-8C(\theta_{0}-h)+C(\theta_{0}-2h)].
+\end{split}
 \label{eqn:deriv}
@@
 \section{Results \label{sec:results}}
+In this section we present our main results. We will start from
 
@@
 
-\subsection{Relativistic Degrees of Freedom}
+\subsection{Relativistic Degrees of Freedom, $N_{\rm eff}$}
 
-In the standard cosmology, there are three active neutrinos thermally produced in the early universe.
+In the standard cosmology, three active neutrinos are thermally produced in the early universe.
 Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation.
-Decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$.
+However decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$.
 The additional fraction depends not only on well-known neutrino scattering rates but also finite temperature quantum corrections.
-Were we to measure this deviation of $N_{\rm eff}$ from 3, it would be an indication that we understand well physics when the Universe was about a second old, so this is a natural target for upcoming experiments.
-
+Upcoming experiment have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
+Furthermore any significant deviation from this value could be a hint of a different scenario not predicted by the standard model. 
+%==========
 \begin{figure}[htbp]
@@
 \includegraphics{prior_massless_neutrinos_snow_mass.pdf}
-\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the 6 cosmological parameters in the standard model.
-Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{old}=4.1\%$.
+\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model.
+Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm old}=4.1\%$.
 The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.
-The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
+The bottom-most (grey) curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
 \label{fig:prior_massless_neutrinos}
@@
 \end{figure}
-
+%==========
 \reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other 6 parameters in the standard cosmological model ($H_0, \Omega_mh^2,\Omega_bh^2, n_s, A_s,$ and $\tau$).
-With no priors on the other parameters, the projected 1-sigma error is  about 0.02, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$.
+With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$.
 The sensitivity comes from the effect of extra species on the damping tail of the CMB anisotropies, both in temperature and polarization, so parameters that also affect the shape of the damping tail, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are most degenerate with $N_{\rm eff}$.
@@
 \includegraphics{prior_w_snow_mass.pdf}
-\caption{$w$ costraints will be dramatically improved by $H_{0}$ external priors. 
-\footnotesize In the figure: on the y axis we have the relative error on $w$ as a function of the prior we impose on the other parameters.
-These priors on the x-axis are measured relative to the error on that parameters from CMB Stage 4 experiments alone.}
+\caption{Constraints on the dark energy equation of state $w$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
+An external prior on $H_{0}$ will be crucial to improve the constraint.}
 \label{fig:prior_w}
@@
 
-{\bf I don't really understand this table.}
-\begin{table}[htdp]
-
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|}
-\hline
-$H_{0}$ &$ M_{\nu}$ &$\Omega_{bc}h^{2}$&$\Omega_{b}h^{2}$&$\tau$&$A_{s}$&$n_{s}$ \\
-$0.80 \%$&$40.92\%$&$0.44\%$&$0.11\%$&$3.075\%$&$0.53\%$&$0.18\%$\\
-
-\hline
-\end{tabular}
-\caption{How well we do constrain separate parameters with this data without any external prior?
+%{\bf I don't really understand this table.}
+%\begin{table}[htdp]
+%
+%\begin{center}
+%\begin{tabular}{|c|c|c|c|c|c|c|}
+%\hline
+%$H_{0}$ &$ M_{\nu}$ &$\Omega_{bc}h^{2}$&$\Omega_{b}h^{2}$&$\tau$&$A_{s}$&$n_{s}$ \\
+%$0.80 \%$&$40.92\%$&$0.44\%$&$0.11\%$&$3.075\%$&$0.53\%$&$0.18\%$\\
+%
+%\hline
+%\end{tabular}
+%\caption{How well we do constrain separate parameters with this data without any external prior?
 Things to notice: this is done with the Zhen test parameters ($\ell<3000$ and no error on lensing).
@@
 $\tau$ will probably improve and also M$\nu$ from BAO may help improving parameters.}
-\end{center}
-\label{default}
-\end{table}%
+%\end{center}
+%\label{default}
+%\end{table}%
 

### 2015-09-14 n_priors_code/fisher_CMB.py
update fisher CMB only to work with new format

@@ -1,393 +1,0 @@
-'''
-
-SECOND VERSION TO SUBSTITUTE THE first one A. Manzotti
-
-Simple fisher code to test prior effect on N effective neutrino estimates
-
-TODO:
-HIGH: lensing noise (almost there: Komatsu and quicklens. MV not yet)
-
-low importance: ini file. Think about PCA to understand what is more important for N_eff
-
-
-create classes or at least external utilities. Maybe a better way to interface with cosmosis
-
-# DATA FROM CAMB
-
-we use the _lenspotentialcls so
-
- l CTT CEE CBB CTE Cdd CdT CdE
- 0  1  2    3   4   5   6  7
-
-  CX are l(l+1)Cl/2pi and Cdd=[l(l+1)]^2 Clphi/2pi, CdT=[l(l+1)]^(3/2) ClphiT/2pi, CdE=[l(l+1)]^(3/2)ClphiE/2pi
-
-
-The matrix is going to run an 4 parameters:
-
-CONVENTIONS:
-alphabetical in CAMB description
-hubble,massless_neutrinos,omnuh2,re_optical_depth,scalar_amp(1),scalar_spectral_index(1)
-PARAMETER ORDER = H0,Neff,Omega_nuh^2,tau,As,ns
-                    0  1      2        3   4  5
-
-
-
-GOAL
-
-1) reproduce Wu paper
-
-'''
-
-
-import numpy as np
-import math
-import matplotlib.pyplot as plt
-import utils
-import pickle
-import sys
-
-# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
-
-
-def years2sec(years):
-    ''' years to sec '''
-    return years * 365 * 24. * 60. * 60.
-
-
-def fsky2arcmin(fsky):
-    '''convert fsky in fraction of unity to arcmin^2'''
-    # 41253 square degrees in all sky
-    return 41253. * fsky * 60. * 60.
-
-
-def C(iell, ell, parbin, data):
-    '''
-    Given CMB data dats it normalize them and create a 3x3 matrix
-
-    ell is the multiple
-    iell is the index in the data ell corresponds to
-
-    remember the order from CAMB
-     l CTT CEE CBB CTE Cdd CdT CdE
-
-    '''
-
-    # noise definition from the number of observations and time
-    # eq 1 of W.hu et al snowmass paper 10^6 detectors
-    Y = 0.25  # 25%yeld
[…]

### 2015-09-14 n_priors_code/fisher_CMB_test.py

@@ -1,0 +1,311 @@
+'''
+
+SECOND VERSION TO SUBSTITUTE THE first one (AM)
+
+Simple fisher code to test prior effect on N effective neutrino estimates
+
+TODO:
+HIGH: lensing noise (almost there: Komatsu and quicklens.)
+
+low importance: ini file. Think about PCA to understand what is more important for N_eff
+
+
+create classes or at least external utilities. Maybe a better way to interface with cosmosis
+
+# DATA FROM CAMB
+
+we use the _lenspotentialcls so
+
+ l CTT CEE CBB CTE Cdd CdT CdE
+ 0  1  2    3   4   5   6  7
+
+  CX are l(l+1)Cl/2pi and Cdd=[l(l+1)]^2 Clphi/2pi, CdT=[l(l+1)]^(3/2) ClphiT/2pi, CdE=[l(l+1)]^(3/2)ClphiE/2pi
+
+
+The matrix is going to run an 4 parameters:
+
+CONVENTIONS:
+alphabetical in CAMB description. Use the ordered dict tools to explore it.
+hubble,massless_neutrinos,omnuh2,re_optical_depth,scalar_amp(1),scalar_spectral_index(1)
+
+
+GOAL
+
+1) reproduce Wu paper and beyond/ DONE. Cross checked with Zhen Pan UC Davis
+
+'''
+
+
+import numpy as np
+import math
+import matplotlib.pyplot as plt
+import utils
+import pickle
+import sys
+
+# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
+
+
+def years2sec(years):
+    ''' years to sec '''
+    return years * 365 * 24. * 60. * 60.
+
+
+def fsky2arcmin(fsky):
+    '''convert fsky in fraction of unity to arcmin^2'''
+    # 41253 square degrees in all sky
+    return 41253. * fsky * 60. * 60.
+
+
+# def calc_deriv_vectorial(iell, ell, fisher_index ,parbin, data,pargaps,values):
+#     '''
+#     Given CMB data dats it normalize them anf create a 3x3 matrix
+
+#     ell is the multiple
+#     iell is the index in the data ell corresponds to
+
+#     remember the order from CAMB
+#      l CTT CEE CBB CTE Cdd CdT CdE
+
+
+#     '''
+#     calc_c_general(data,parabin)
+
+#     ci = (-calc_c_general(dats,fisher_index * 4 + 4) + 8. * calc_c_general(dats, fisher_index * 4 + 3) - 8. *
[…]

### 2015-09-14 n_priors_code/fisher_CMB_test.py
update fisher CMB only to work with new format

@@ -201,7 +201,7 @@ arcmin_from_fsky = fsky2arcmin(fsky)
 sec_of_obs = years2sec(5)
 Y = 0.25  # 25% yeld
 # ===================
-header = 'Joint fisher CMB T E + phi lensing used \n'
+header = 'Joint fisher CMB T E. NO LENSING used \n'
 header += 'lmax={} \n lmin={} \n l_t_max={} \n fsky={} \n lensed={} \n data_folder={} \n N_det={} \n'.format(
     lmax, lmin, l_t_max, fsky, lensed, data_folder, N_det)
 

### 2015-09-17 n_priors_code/plotting/plot_cov.py
new plots added

@@ -1,0 +1,60 @@
+#!/usr/bin/env python
+'''
+This script plot the covariance matrix with usual Planck colours for the given set of data.
+
+
+'''
+
+
+__author__ = "A.Manzotti"
+__license__ = "GPL"
+__version__ = "2.0"
+__maintainer__ = "A.Manzotti"
+__email__ = "manzotti.alessandro@gmail.com"
+__status__ = "Production"
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+# READ DATA
+# DEFINE YOUR FOLDER HERE
+base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
+data_type = 'varying_lambda'
+run_idx = 2
+# ======
+
+dat = np.genfromtxt(base_dir + 'data/{}/run{}/output/param_cov.txt'.format(data_type, str(run_idx)))
+fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type, str(run_idx)), "rb"))
+label = {}
+
+
+
+label['massless_neutrinos'] = r'$N_{eff}$'
+label['hubble'] = r'$H_{0}$'
+label['scalar_amp(1)'] = r'$A_{s}$'
+label['scalar_spectral_index(1)'] = r'$n_{s}$'
+label['omnuh2'] = r'$\Omega$'
+label['re_optical_depth'] = r'$\tau$'
+label['massless_neutrinos'] =  r'$N_{eff}$'
+label['w'] = r'$w$'
+label['ombh2'] = r'$\Omega_{b}$'
+label['omch2'] = r'$\Omega_{c}$'
+label['omnuh2'] = r'$\Omega_{\nu}$'
+
+tick_array = np.arange(np.size(fid.keys()))
+
+plt.imshow(dat,cmap=plt.get_cmap('bwr'),interpolation='nearest',vmin=-1,vmax=1)
+
+plt.xticks(tick_array,[label[key] for key in fid.keys() ])
+plt.yticks(tick_array,[label[key] for key in fid.keys() ])
+
+
+plt.gca().xaxis.set_ticks_position('top')
+
+plt.colorbar()
+
+plt.savefig('../../images/parcov_cmb.pdf')
+
+plt.savefig(base_dir + 'data/{}/run{}/output/parcov_cmb.pdf'.format(data_type, str(run_idx) ), dpi=400, papertype='Letter',
+            format='pdf', bbox_inches='tight')
+

### 2015-09-17 n_priors_code/plotting/plot_triangle_ellipses_plot.py

@@ -1,11 +1,21 @@
+#!/usr/bin/env python
 '''
-Plot scipt to plot the derivatives of CMB specctra respect to all the parameters of the Fisher matrix.
+This script plot the usual tringle ellispes plot for the given dataset.
+Remember this is a Fisher matrix code so everything is approximated to be gaussian.
 
-Different lines corresponde to different gaps and techniques.
-
-
-
+This was written to compare our results with the one given by Zhen
 '''
+
+
+__author__ = "A.Manzotti"
+__license__ = "GPL"
+__version__ = "2.0"
+__maintainer__ = "A.Manzotti"
+__email__ = "manzotti.alessandro@gmail.com"
+__status__ = "Production"
+
+
+
 
 
 

### 2015-09-17 n_priors_code/plotting/plotcov.py

@@ -1,35 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-import pickle
-
-run_idx = 4
-dat = np.genfromtxt('../output_cmb/param_cov.txt')
-fid = pickle.load(open('../data/run{}/fid_values.p'.format(run_idx), "rb"))
-label = {}
-
-label['massless_neutrinos'] = r'$N_{eff}$'
-label['hubble'] = r'$H_{0}$'
-label['scalar_amp(1)'] = r'$A_{s}$'
-label['scalar_spectral_index(1)'] = r'$n_{s}$'
-label['omnuh2'] = r'$\Omega$'
-label['re_optical_depth'] = r'$\tau$'
-label['massless_neutrinos'] =  r'$N_{eff}$'
-label['w'] = r'$w$'
-label['ombh2'] = r'$\Omega_{b}$'
-label['omch2'] = r'$\Omega_{c}$'
-label['omnuh2'] = r'$\Omega_{\nu}$'
-
-tick_array = np.arange(np.size(fid.keys()))
-
-plt.imshow(dat,cmap=plt.get_cmap('bwr'),interpolation='nearest',vmin=-1,vmax=1)
-
-plt.xticks(tick_array,[label[key] for key in fid.keys() ])
-plt.yticks(tick_array,[label[key] for key in fid.keys() ])
-
-
-plt.gca().xaxis.set_ticks_position('top')
-
-plt.colorbar()
-
-plt.savefig('../../images/parcov_cmb.pdf')
-plt.show()

### 2015-09-17 n_priors_code/utils.py
minor changes code

@@ -207,7 +207,7 @@ def return_simgax_y_prior(fid, fisher, x, y, prior_val):
         fisher1 = fisher.copy()
         # Cicle on H0 priors
         fisher1[fid.keys().index(y), fid.keys().index(y)] += 1 / \
-            (prior_val * fid[y] ) ** 2
+            (prior_val * fid[y]) ** 2
         print np.diag(fisher1)
 
         return np.sqrt(np.linalg.inv(fisher1)[fid.keys().index(x), fid.keys().index(x)])
@@ -221,12 +221,13 @@ def return_simgax_y_prior(fid, fisher, x, y, prior_val):
             # Cicle on H0 priors
 
             fisher1[fid.keys().index(y), fid.keys().index(y)] += 1 / \
-                (prior * fid[y] ) ** 2
+                (prior * fid[y]) ** 2
 
             sigma_x_prior[i] = np.sqrt(np.linalg.inv(fisher1)[fid.keys().index(x), fid.keys().index(x)])
         return sigma_x_prior
     else:
         sys.exit('type not recognized')
+
 
 def return_simgax_y_prior(fid, fisher, x, y, prior_val):
     '''
@@ -245,7 +246,7 @@ def return_simgax_y_prior(fid, fisher, x, y, prior_val):
         fisher1 = fisher.copy()
         # Cicle on H0 priors
         fisher1[fid.keys().index(y), fid.keys().index(y)] += 1 / \
-            (prior_val * fid[y] ) ** 2
+            (prior_val * fid[y]) ** 2
         print np.diag(fisher1)
 
         return np.sqrt(np.linalg.inv(fisher1)[fid.keys().index(x), fid.keys().index(x)])
@@ -259,7 +260,7 @@ def return_simgax_y_prior(fid, fisher, x, y, prior_val):
             # Cicle on H0 priors
 
             fisher1[fid.keys().index(y), fid.keys().index(y)] += 1 / \
-                (prior * fid[y] ) ** 2
+                (prior * fid[y]) ** 2
 
             sigma_x_prior[i] = np.sqrt(np.linalg.inv(fisher1)[fid.keys().index(x), fid.keys().index(x)])
         return sigma_x_prior
@@ -268,7 +269,6 @@ def return_simgax_y_prior(fid, fisher, x, y, prior_val):
 
 
 def apply_simgax_y_prior(fid, fisher, x, y, prior_val):
-
     '''
     Concept return sigma x given a possible prior on y
 
[…]

### 2015-09-17 Notes/N_eff_prior_paper.tex
minor edit notes

@@
 \usepackage{graphicx}
-%<<<<<<< HEAD
-%\input{papers_packages_command._include_tex}
 
@@
 \newcommand{\aj}{AJ }
-
-%>>>>>>> 683231e77bc036bad411dd2244ca99280568d723
 \bibliographystyle{apsrev4-1}
@@
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
-Here, we examine quantitatively with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
-We find that CMB S4 
+Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
+We find that CMB S4... 
 \end{abstract}
@@
 
-{\bf A lot of this is standard and need not be included, which will make the paper much shorter: a good thing!}
-
 In this paper we want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
@@
 
-\subsection{Fisher Matrix formalism \label{subsec:fm-formalism}}
 To estimate errors on cosmological parameters we follow the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
@@
 
-\subsection{CMB S4 experiments specifications.\label{subsec:cosmo-noise}}
 In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
@@
 \section{Results \label{sec:results}}
-In this section we present our main results. We will start from
-
-{\bf Let's open up one parameter at a time: first standard LCDM with neff, and then with mnu, and then with w.
-Also, let's show a curve in each plot with constraints on *all*  parameters, not just one of them, as that will give people a sense of what is gained if multiple priors can be obtained.}
+In this section we present our main results. We will start from..
 
@@
 Upcoming experiment have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
-Furthermore any significant deviation from this value could be a hint of a different scenario not predicted by the standard model. 
+Furthermore any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model. 
 %==========
@@
 %==========
+
 \reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other 6 parameters in the standard cosmological model ($H_0, \Omega_mh^2,\Omega_bh^2, n_s, A_s,$ and $\tau$).
@@
 
-\subsection{Neutrino Masses}
+\subsection{Neutrino Masses, $\sum m_\nu$}
 
@@
 
-\subsection{Dark Energy Equation of State}
+\subsection{Dark Energy Equation of State, $w$}
 
@@
 However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably.
-{\bf We should try to understand why.}
+{\bf We should try to understand why.
+Probably just a geometrical degeneracy in the way $w$ and $H_{0}$ enter the definition of the distance to the LSS $D(z_{\rm LSS})$}
 
-%{\bf I don't really understand this table.}
-%\begin{table}[htdp]
-%
-%\begin{center}
-%\begin{tabular}{|c|c|c|c|c|c|c|}
-%\hline
-%$H_{0}$ &$ M_{\nu}$ &$\Omega_{bc}h^{2}$&$\Omega_{b}h^{2}$&$\tau$&$A_{s}$&$n_{s}$ \\
-%$0.80 \%$&$40.92\%$&$0.44\%$&$0.11\%$&$3.075\%$&$0.53\%$&$0.18\%$\\
-%
-%\hline
-%\end{tabular}
-%\caption{How well we do constrain separate parameters with this data without any external prior?
-Things to notice: this is done with the Zhen test parameters ($\ell<3000$ and no error on lensing).
-Now if we trust it CMB alone can get a $0.8\%$ error on $H_{0}$ thus I am not surprised if a prior on H would not help.
-However what do we think about it?
-is it really CMB better than SN.
-People will not agree on that.
-$\tau$ will probably improve and also M$\nu$ from BAO may help improving parameters.}
-%\end{center}
-%\label{default}
-%\end{table}%
+\section{Conclusions  \label{sec:conclusions}}
 

### 2015-10-12 Notes/N_eff_prior_paper.tex
new paper text label v2

@@
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
-Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
-We find that CMB S4... 
+Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
+We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$  
+if the largest scale at reach is not enough to constraint $\tau$ internally.
+An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
 \end{abstract}
@@
 
-In this paper we want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
-We start this section in \refssec{fm-formalism} by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints.
+We want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
+We approach the problem with a Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints.
 To apply this technique, we need to specify our cosmological model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the specifications of S4 CMB experiments.
-These will be presented in \refssec{cosmo-noise}.
 
-To estimate errors on cosmological parameters we follow the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
-These two are related by the prior probability of the parameters $P(\theta)$ through:
-\begin{equation}
-\mathcal{L}(\theta|d)\propto \mathcal{L}(d|\theta)P(\theta),
-\end{equation}
-where, as usual, we have neglected the probability of the data $P(d)$.
-These general framework can be simplified in our case. We will focus not on an entirely general posterior distribution, but we will assume that the likelihoods are gaussian.
-Furthermore we will not derive the global shape of the likelihood but we will obtain parameters errors by studying small perturbations around its maximum.
-These are the two basic assumptions of the Fisher-matrix approach.
-
-Regarding the first assumption, we recognize the fact that, even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}.
-Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
-Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
-
-Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$ and we will get the errors on those from the likelihood curvature around the fiducial values.
-Indeed, as usual, we define the fisher matrix elements as the curvature:
+As usual, we define the fisher matrix elements as the curvature:
 \begin{equation}
@@
 where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximizes the likelihood.
- With the usual definition \cite{} $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
+It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian.
+Even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}.
+Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
+Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
+
+With the usual definition $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
 \begin{equation}
@@
 \begin{center}
-\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
+\includegraphics{prior_massless_neutrinos_snow_mass_lmin=4_lmax=4499.pdf}
 \caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model.
@@
 
+%==========
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics{prior_omnuh2_snow_mass.pdf}
+\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
 \caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
@@
 \end{figure}
+%==========
 
@@
 
+%==========
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics{prior_w_snow_mass.pdf}
+\includegraphics{prior_w_snow_mass_lmin=4_lmax=4499.pdf}
 \caption{Constraints on the dark energy equation of state $w$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.

### 2015-10-12 Notes/N_eff_prior_paper_v2.tex
new paper text label v2

--- 2015-10-12 Notes/N_eff_prior_paper.tex
@@
 The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
-Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
+Indeed the low level of noise, together with improved angular resolution, will allow a precise measurement of the CMB polarization anisotropies as well as an improved reconstruction of the lensing potential of high redshift large scale structure.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
 Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
-We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$  
-if the largest scale at reach is not enough to constraint $\tau$ internally.
+We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$ unless the largest scale at reach for CMB S4 will be large enough to constraint $\tau$ internally thought the reionization bump.
 An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
@@
 \section{Introduction}\label{sec:intro}
-Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe and they will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation.
-The recent past is indeed reassuring. 
-Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
-This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
-With the improved sensitivity we extend our understanding of the universe scientific from a proof of its almost flat geometry to a well tested model $\Lambda$CDM with a percent level constraint on its parameters.
-The same success characterized ground CMB experiments, where the evolution from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
-These series of successes is far from its end.
-Indeed the next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As it happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.
+Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
+They have the potential to maintain this important role in the near future thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation.
+Indeed, the next generation of CMB experiments (S4), now in its planning stage, will measure the E-mode polarization with almost cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+This new sensitivity, in particular if supported with the essential synergy with other cosmological probes, will be a crucial step in our understanding of dark matter, inflation, Dark Energy and neutrinos.
 
-The cosmological dependence to neutrinos is primarily due to their contribution to the number of relativistic species $N_{\rm eff}$ in the early phase of the universe  together with the role they play, given their non zero mass $M_{\nu}$, into the late growth of cosmic structures.
-Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$.
-This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
-The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale of the CMB power spectrum and will unequivocally test the value $N_{\rm eff}=3.046$ predicted by the standard model .
+Neutrinos properties can be effectively constrained with CMB measurements because they influence both the early history of the universe and the later stage of structure formation. 
+Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their effectivenumber $N_{\rm eff}$ affects the expansion rate.
+This can be powerfully tested using the CMB, by comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
+The future experiments, with their high resolution, will be able to measure the CMB power spectrum, at small scales, deeply into the damping tale, testing any possible deviations from $N_{\rm eff}=3.046$, the value predicted by the standard model.
 The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
-As other low redshift effects however, massive neutrinos modify the CMB by altering the growth of the large scale structure responsible for the the lensing of the CMB photons.
+However massive neutrinos influence the growth of structure at scales shorter than their free streaming.
 Different neutrino masses consequently lead to different CMB lensing.
 Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction.
-This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
-Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
+This will turn into such a precise constraints of the sum of neutrino masses to get a possible hint of the hierarchies of the individual masses.
+Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $45\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-As previously mentioned, CMB is also sensitive to the properties of dark energy (see \cite{2010MNRAS.405.2639J}).
-Thanks to the current generation of experiments we know with extraordinary accuracy its energy density.
-The challenge for the next generation is to reveal the nature of this mysterious component.
-For example, a crucial step to rule out part of the proposed models will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant.
-Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface.
+As previously mentioned, CMB is also a powerful probe of the dark energy properties (see \cite{2010MNRAS.405.2639J}).
+The challenge for future experiments is to go beyond the now well measured energy density and to determine other dark energy properties.
+Among these, a crucial step would be to constraint the equation of state, the ratio of pressure and energy density, looking for possible deviations from the value $w=-1$ expected in a cosmological constant scenario.
+CMB is sensitive to Dark energy because this component alters the universe's expansion and it consequently changes the distance to the last scattering surface.
 Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing.
-Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
 However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
-Similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
+As a consequence, similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
 
-The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector. 
-The future of cosmology will hinge on the ability of combining different probes.
+The crucial importance of the feedback from external experiments is true not only for the dark energy and the neutrino sector. 
+In the future, cosmology will hinge on the ability of combining different probes.
 In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
 In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
-To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study in detail the dependance of CMB future constrains on the external priors assumed.
-For all the parameters of the standard $\Lambda$CDM with massive neutrinos and $w\neq-1$ dark energy model we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
+To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
+For all the parameters of the standard $\Lambda$CDM plus massive neutrinos ($\Omega_c h^2,\Omega_b h^2,A_s,n_s,\tau,H_0,M_{\nu}$) together with possibly testable non-standard extension ($N_{\rm eff},w,w_{a}$) we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
 
@@
 
-We want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
-We approach the problem with a Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints.
-To apply this technique, we need to specify our cosmological model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the specifications of S4 CMB experiments.
+To measure quantitatively the impact of external priors on the CMB ability to constrain cosmological parameters we use a Fisher matrix formalism.
+In this section we will quickly review  the technique and we will then present our fiducial cosmological model together with the experimental specifications of S4 CMB we used.
 
-As usual, we define the fisher matrix elements as the curvature:
+As usual, we define the Fisher matrix elements as the curvature of the likelihood:
 \begin{equation}
@@
 \end{equation}
-where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximizes the likelihood.
+where $\theta_{i,j}$ represents two of the cosmological parameters and $\boldsymbol{\theta_0}$ is the fiducial values array that, by definition, maximizes the likelihood.
 It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian.
-Even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}.
-Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
+Even if the gaussian approximation has been shown to be appropriate most of the time, exceptions exist in the literature \cite{2012JCAP...09..009W}.
+However, the level of accuracy needed in this work does not require a more careful modeling of the likelihoods.
 Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
 
-With the usual definition $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
+For CMB experiments the Fisher matrix can be related to the power spectrum $\boldsymbol{C}_\ell$ by:
 \begin{equation}
- F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
+ F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
  \label{eqn:Fij_def2}
  \end{equation}
- In this work we will constrain cosmological parameters with the CMB temperature and E mode polarization together with the reconstructed lensing potential.
+In this work we constrain cosmological parameters with CMB temperature and E mode polarization together with the reconstructed lensing potential of large scale structure.
 For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
@@
  	\centering
-		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
+		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{T\phi} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{T\phi} & 0 & C_\ell^{\phi} + N_\ell^{\phi}\end{array}\right).
 	\label{eqn:cov_definition}
@@
 Note that we are neglecting the term $C_\ell^{E\phi}$.
-As also noticed in previous literature (like \cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
-The term $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
-The power of the Fisher approach descend from the Cramer-Rao inequality that relates the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
+As also noticed in previous literature \cite{wu:2014,2013PhRvD..87h3008H} this term contains very little information while adding possible numerical issues.
+The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be described at the end of this section.
+
+A lower bound on the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, can be easily obtained as:
 \begin{equation}
-\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}.
+\sigma_i \equiv \sigma (\theta_i) \geq \sqrt{(\mathbf{ F^{-1}})_{ii}}.
 \label{eqn:cramer-rao}
 \end{equation}
-Once $F_{ij}$ is computed following \refeq{Fij_def2} it is straightforwards to get the error $\sigma_i$ from \refeq{cramer-rao}.
-Furthermore in this context it is easy to introduce external priors on cosmological parameters.
-Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
+Furthermore it is easy to introduce external priors on cosmological parameters.
+Indeed we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
 \begin{equation}
@@
 In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
-For example, a $1\%$ prior on $H_{0}$ can be obtained by:
+For example, a $1\%$ prior on $H_{0}$ can be added by:
 \begin{equation}
@@
 
-We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data, together with the ease of including external priors. 
-This technique however introduces also some technical difficulties. 
+We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data and easily including external priors. 
+This technique however introduces also technical difficulties. 
 Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
-Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
-Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard method used to forecast future constrains \cite{wu:2014}.
+Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix lead to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
+Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard procedure used to forecast future constrains \cite{wu:2014}.
 We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
-We compute the derivatives in \refeq{Fij_def2} using a 5 points formula:
-\begin{equation}
-\begin{split}
-\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim & \frac{1}{12 h} [ -C(\theta_{0}+2h)+ 8C(\theta_{0}+h) \\ &-8C(\theta_{0}-h)+C(\theta_{0}-2h)].
-\end{split}
-\label{eqn:deriv}
-\end{equation}
-
-This high order definition allow us to use a bigger gap $h$ around the fiducial parameters $\theta_{0}$.
-As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing C negligible.
-We also test the robustness of this calculation by changing the gap $h$ in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
-Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
-Lastly we implement the same technique of \cite{2006astro.ph..9591A} to avoid possible numerical instability in the marginalizing procedure.
+We compute the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
+%\begin{equation}
+%\begin{split}
+%\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim & \frac{1}{12 h} [ -C(\theta_{0}+2h)+ 8C(\theta_{0}+h) \\ &-8C(\theta_{0}-h)+C(\theta_{0}-2h)].
+%\end{split}
+%\label{eqn:deriv}
+%\end{equation}
+this high order approach allows us to use bigger steps around the fiducial parameters to compute derivatives.
+As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing $\boldsymbol{C}_\ell$ negligible.
+We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
+%Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
+%Lastly we implement the same technique of \cite{2006astro.ph..9591A} to avoid possible numerical instability in the marginalizing procedure.
 This allow us not to invert the entire matrix when we want to marginalize over a set of parameters, in order to minimize numerical issues and conserve parameters degeneracies as much as possible.
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[scale=0.4]{PS_phi_with_noise.pdf}
-\caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
-In the figure: the deflection power spectrum for our fiducial cosmology together with the two examples of lensing reconstruction noise $N^{\phi}$ used in this work.}
-\label{fig:phi-cl-noise}
-\end{center}
-\end{figure}
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[scale=0.4]{PS_phi_with_noise.pdf}
+%\caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
+%In the figure: the deflection power spectrum for our fiducial cosmology together with the two examples of lensing reconstruction noise $N^{\phi}$ used in this work.}
+%\label{fig:phi-cl-noise}
+%\end{center}
+%\end{figure}
+%
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[scale=0.4]{PS_with_noise.pdf}
+%\caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
+%In the figure: CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
+%\label{fig:cmb-cl-noise}
+%\end{center}
+%\end{figure}
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[scale=0.4]{PS_with_noise.pdf}
-\caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
-In the figure: CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
-\label{fig:cmb-cl-noise}
-\end{center}
-\end{figure}
-
-In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
-Firstly we parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy models by introducing the equations of state parameter $w$ as a varying parameter.
+We compute the power specctra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB.
+Our base cosmology is a flat $\nu \Lambda$CDM. 
+We then extend the parameter space by introducing $N_{\rm eff}$ and $w,w_{a}$ as free parameters.
 We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
-$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc.
-Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
-Given the model, we use CAMB \cite{Lewis:1999bs} to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}.
-Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
+$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc and a neutrino total mass $M_{\nu}$ $\simeq$ 85\ meV.
+Regarding the $\Lambda$CDM extensions, we chose, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1,w_{a}=0$.
+It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
 
@@
 For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
+This allow us to use all the polarization power spectrum multipoles all the way up to $\ell_{\rm E,max}=5000$.
 We deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
@@
 We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin (PRELIMINARY).
-Note that $1.5$ $\mu$K-arcmin is the noise in temperature and we need $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
+Note that $1.5$ $\mu$K-arcmin is the noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
 
-Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from the CMB.
+%Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from the CMB.
 The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons.
 This introduces non gaussianities that couple different, otherwise independent, CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
-For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow the non-iterative technique of \cite{okamoto:2003,hu:2002}.
-
-%\begin{eqnarray}
-%\centering
-%	s\ [\;\mu {\rm K.arcmin}\;] \equiv \frac{ {\rm NET}\ [\; \mu{\rm K.}\sqrt{s}\;] \times \sqrt{f_{sky} \ [\;{\rm arcmin}^2\;] }}{ \sqrt {N_{\rm det} \times Y \times \Delta T\ [\;{\rm s }\;]}}.
-%	\label{eq:sensitivity_definition}
-%\end{eqnarray}
-%
-
-%\begin{eqnarray}
-%	T_{\nu} = \left( \frac{4}{11} \right)^{1/3} T_{\gamma} 
-%	\label{eq:tnu_propto_tgamma}
-%\end{eqnarray}
+The noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ is modeled assuming an iterative reconstruction technique.
+Indeed it has been shown \cite{seljak:2004} that the low level of noise expected in CMB S4 experiment will be in the regime  where this technique improve our constraints compared to the quadratic estimator introduced in \cite{okamoto:2003,hu:2002}.
 
 \section{Results \label{sec:results}}
-In this section we present our main results. We will start from..
+In this section we present our main results. We will start from our base $\nu \Lambda$CDM and we then explore possible extensions.
+
+\subsection{Neutrino Masses, $\sum m_\nu$}
+
+Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget.
+This effect, probed by CMB lensing, will allow us to determine the contribution of neutrinos to the total matter density.
+Since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$.
+This is particularly exciting because we now know that neutrinos are massive with a lower limit on $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
+
+%==========
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
+\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
+Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
+The lower limit on the sum of the neutrino masses is 50 meV.}
+\label{fig:prior_omeganuh2}
+\end{center}
+\end{figure}
+%==========
+%==========
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{prior_omnuh2_snow_mass_lmin=50_lmax=3000_ndet=10000_fsky=0.75}
+\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
+Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
+The lower limit on the sum of the neutrino masses is 50 meV.}
+\label{fig:prior_omeganuh2_tau}
+\end{center}
+\end{figure}
+%==========
+
+\reffig{prior_omeganuh2} shows that, without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
+Adding in external priors helps this significantly though, as discussed in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
+There, the prior was deescribed as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
+Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$, which indeed determines these low-redshift distances.
+As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior.
+Adding in other priors would help further.
 
@@
 
-\subsection{Neutrino Masses, $\sum m_\nu$}
-
-Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget.
-This effect, probed by CMB lensing, will allow us to determine how much of the matter density is made up of neutrinos.
-Since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$.
-This is particularly exciting because there is a lower limit on $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
-
-%==========
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
-\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
-Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
-The lower limit on the sum of the neutrino masses is 50 meV.}
-\label{fig:prior_omeganuh2}
-\end{center}
-\end{figure}
-%==========
-
-\reffig{prior_omeganuh2} shows that, without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
-Adding in external priors helps this significantly though, as discussed in \citet{2013arXiv1309.5383A}.
-There, the prior was deescribed as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
-Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$, which indeed determines these low-redshift distances.
-As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior.
-Adding in other priors would help further.
-
-%\begin{table}[htdp]
-%\begin{center}
-%\begin{tabular}{|c|c|}
-%\hline
-%$H_{0}$ & $67.11$ km/s/Mpc\\
-%\hline
-%\hline
-%$\tau$ & $0.0925$ \\
-%\hline
-%
-%\hline
-%$A_{s}$ &$2.215 \times 10^{-9}$ \\
-%\hline
-%
-%\hline
-%$n_{s}$ & $0.9624$ \\
-%\hline
-%
-%\hline
-%$N_{eff}$ & 3.046\\
-%\hline
-%\end{tabular}
-%\end{center}
-%\caption{Fiducial values of cosmological parameters used in this work.}
-%\label{default}
-%\end{table}%
-
 \subsection{Dark Energy Equation of State, $w$}
@@
 \section{Conclusions  \label{sec:conclusions}}
+The planning of the next CMB Stage IV experiments represents an exciting challenge.
+The possible reward is remarkable.
+We can finally put a tight constraint on the total mass of neutrinos with deeply implications on particle physics like, for example, the ability to distinguish between normal and inverted hierarchy.
+We will measure with great accuracy the number of relativistic species, testing both standard model prediction and, eventually, probing new physics.
+The same is true for dark energy where we will tighten the constraint on $w$,$w_{a}$ getting closer to test the nature of the accelerated expansion of the universe.
+These however will not be possible without an optimal synergy with other experiments observing a variety of others cosmological probes.
+Indeed this should be taken into account in this CMB Stage IV planning phase.
+Understanding CMB weaknesses and the impact of external data can have a deep influence in planning the experiment specifications.
+For this reason in this work we explored the impact of external priors on CMB cosmological constraints and in particular on the neutrino sector and dark energy parameters.
+We find:
+\begin{description}
+\item[$N_{\rm eff}$] 
+\item[$M_{\nu}$] 
+\item[$w$,$w_{a}$] 
+This results suggest...
+
+Comment on issue and future.
+This will be even more relevant when we open the parameter space.
+Will CMB + others be able to discover somethign CMB can't alone?
+$N_{\rm eff}$ extensions, mass splitting etc
+
+\end{description}
 
@@
 %\appendix
-%\section{Marginalization , not needed probably}
-%\begin{equation}
-%G = F^{\phi\phi} - F^{\phi\psi}U\Lambda^{-1}U^{T}F^{\phi\psi},
-%\end{equation}
-%where we define $\phi = \{N_{eff},...\}$ and $\psi = \{\Omega_{m} ...
-marginal\}$; therefore, $F^{\phi\phi}$ is the block of the total Fisher matrix containing the parameters we want to constrain, whilst $F^{\psi\psi}$ is the nuisance-parameter Fisher sub-matrix.
-Here, $\Lambda$ is the diagonal matrix whose elements are the eigenvalues of $F^{\psi\psi}$, whilst U is the orthogonal matrix diagonalising $F^{\psi\psi}$.
-By using Eq., our marginalizing procedure is more stable, since degeneracies in $F^{\phi\phi}$ are properly propagated to G with no instabilities, and we do not even worry about a possibly ill-conditioned $F^{\phi\phi}$ sub-matrix, since we check its stability on the fly by the diagonalisation.
 
@@
 We thank Youngsoo Park for his contribution in the early stage of this work.
-AM wants to thank Zhen Pan who allows a careful cross-check of our results.
+AM wants to thank Zhen Pan who allowed a careful cross-check of our results.
 This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.

### 2015-10-12 n_priors_code/fisher_joint_test.py
remove data add images to notes

@@ -50,6 +50,7 @@ import matplotlib.pyplot as plt
 import utils
 import pickle
 import sys
+import collections
 
 # util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
 
@@ -65,7 +66,7 @@ def fsky2arcmin(fsky):
     return 41253. * fsky * 60. * 60.
 
 
-def calc_deriv_vectorial(fisher_index, dats, pargaps, values):
+def calc_deriv_vectorial(fisher_index, dats, pargaps, values, order=5):
     '''
     Given CMB data dats it normalize them anf create a 3x3 matrix
 
@@ -75,8 +76,23 @@ def calc_deriv_vectorial(fisher_index, dats, pargaps, values):
     remember the order from CAMB
      l CTT CEE CBB CTE Cdd CdT CdE
     '''
-    ci = (-calc_c_general(dats, fisher_index * 4 + 4) + 8. * calc_c_general(dats, fisher_index * 4 + 3) - 8. *
-          calc_c_general(dats, fisher_index * 4 + 2) + calc_c_general(dats, fisher_index * 4 + 1)) / (12. * pargaps[values.keys()[fisher_index]])
+    if order == 5 and np.shape(values[values.keys()[0]])[0] == 4:
+        ci = (-calc_c_general(dats, fisher_index * 4 + 4) + 8. * calc_c_general(dats, fisher_index * 4 + 3) - 8. *
+              calc_c_general(dats, fisher_index * 4 + 2) + calc_c_general(dats, fisher_index * 4 + 1)) / (12. * pargaps[values.keys()[fisher_index]])
+
+    elif order == 7 and np.shape(values[values.keys()[0]])[0] == 6:
+        ci = (calc_c_general(dats, fisher_index * 6 + 6) - 9. * calc_c_general(dats, fisher_index * 6 + 5) + 45. *
+              calc_c_general(dats, fisher_index * 6 + 4) - 45. * calc_c_general(dats, fisher_index * 6 + 3) + 9. * calc_c_general(dats, fisher_index * 6 + 2) - calc_c_general(dats, fisher_index * 6 + 1)) / (60. * pargaps[values.keys()[fisher_index]])
+
+    elif order == 9 and np.shape(values[values.keys()[0]])[0] == 8:
+        ci = (-3. * calc_c_general(dats, fisher_index * 8 + 8) + 32. * calc_c_general(dats, fisher_index * 8 + 7) - 168. *
[…]

### 2015-10-12 n_priors_code/generate_data.py

@@ -37,7 +37,7 @@ import os
 import warnings
 
 output_folder = 'varying_all'
-output_folder_2 = 'run4'
+output_folder_2 = 'run2'
 camb_location = '/home/manzotti/local/camb2013/camb'
 
 config = configparser.ConfigParser()
@@ -95,21 +95,21 @@ with open("./data/{}/{}/fid_values.p".format(output_folder, output_folder_2), "w
 # ================================================
 
 pargaps_dict = {}
-pargaps_dict['hubble'] = fid['hubble'] * 0.05
-pargaps_dict['helium_fraction'] = fid['helium_fraction'] * 0.05
+pargaps_dict['hubble'] = fid['hubble'] * 0.02
+pargaps_dict['helium_fraction'] = fid['helium_fraction'] * 0.02
 # for the value with fiducial = 0 we can not take a percentage. So fraction of 1509.07471 Table III
-pargaps_dict['omk'] = 0.02
+pargaps_dict['omk'] = 0.01
 pargaps_dict['scalar_nrun(1)'] = 5e-2  # for the value with fiducial = 0 we can not take a percentage
-pargaps_dict['scalar_spectral_index(1)'] = fid['scalar_spectral_index(1)'] * 0.04
-pargaps_dict['scalar_amp(1)'] = fid['scalar_amp(1)'] * 0.05
-pargaps_dict['re_optical_depth'] = fid['re_optical_depth'] * 0.16
-pargaps_dict['omnuh2'] = fid['omnuh2'] * 0.15
-pargaps_dict['ombh2'] = fid['ombh2'] * 0.05
-pargaps_dict['omch2'] = fid['omch2'] * 0.05
-pargaps_dict['massless_neutrinos'] = fid['massless_neutrinos'] * 0.05
-pargaps_dict['w'] = fid['w'] * 0.15
+pargaps_dict['scalar_spectral_index(1)'] = fid['scalar_spectral_index(1)'] * 0.02
+pargaps_dict['scalar_amp(1)'] = fid['scalar_amp(1)'] * 0.02
+pargaps_dict['re_optical_depth'] = fid['re_optical_depth'] * 0.03
+pargaps_dict['omnuh2'] = fid['omnuh2'] * 0.04
+pargaps_dict['ombh2'] = fid['ombh2'] * 0.02
+pargaps_dict['omch2'] = fid['omch2'] * 0.02
[…]

### 2015-10-12 n_priors_code/plotting/plot_deriv.py

@@ -39,9 +39,9 @@ from matplotlib.ticker import MaxNLocator  # needed for tick
 
 # READ DATA
 
-data_type ='varying+Yp'
+data_type ='varying_all'
 run_idx = 1
-
+print data_type
 
 # READ PARAMS
 dats = np.genfromtxt('../data/{}/run{}/fiducial_lenspotentialcls.dat'.format(data_type,run_idx))
@@ -167,22 +167,22 @@ plt.rcParams['axes.linewidth'] = 1.0
 
 # plot2 = plt.semilogx(data_fid[:,0] , 10.*np.nan_to_num((dats[:,1,]- dats[:,1,] )/data_fid[:,1]),linewidth=1, color='b',label=r'$C^{T}$')
 
+
 label = {}
 
 label['massless_neutrinos'] = 'N_{eff}'
 label['hubble'] = 'H_{0}'
-label['mnu'] = 'M'
 label['scalar_amp(1)'] = 'A_{s}'
 label['scalar_spectral_index(1)'] = 'n_{s}'
-label['omnuh2'] = r'\Omega_{\nu}h^{2}'
-label['re_optical_depth'] = r'\tau'
-label['ombh2'] = '\Omega_{b}h^{2}'
-label['ombch2'] = '\Omega_{m}h^{2}'
-label['omch2'] = '\Omega_{c}h^{2}'
-label['helium_fraction'] = 'Y_{p}'
+label['omnuh2'] = r'\Omega_{\nu}'
+label['re_optical_depth'] = r'~\tau'
+label['ombh2'] = '\Omega_{b}'
+label['omch2'] = '\Omega_{c}'
 label['w'] = 'w'
-
-
+label['wa'] = 'w_{a}'
+label['helium_fraction'] = 'Y_{p}'
+label['scalar_nrun(1)'] = r'\alpha_{s}'
+label['omk'] = r'\Omega_{k}'
 # fg = plt.figure(figsize=fig_dims)
 
 

### 2015-10-12 n_priors_code/plotting/plot_one_ellipse_plot_prior.py

@@ -57,7 +57,7 @@ base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_lambda'
 run_idx = 2
 lmax = 4499
-lmin = 100
+lmin = 4
 # ======
 fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type, str(run_idx)), "rb"))
 values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
@@ -213,7 +213,7 @@ label['ombh2'] = '\Omega_{b}'
 label['omch2'] = '\Omega_{c}'
 label['w'] = 'w'
 
-parameters =['omnuh2','omch2']
+parameters =['omnuh2','hubble']
 # parameters = ['w', 'hubble']
 
 fg = plt.figure(figsize=(10, 10))
@@ -274,6 +274,6 @@ plt.rcParams['legend.handletextpad'] = 0.3
 # ============================================
 # plt.savefig('../../images/trinagle.pdf', dpi=400, papertype='Letter',
 #             format='pdf', transparent=True)
-plt.savefig('ellipse_Omega_M_M_nu1002.pdf', dpi=400, papertype='Letter',
+plt.savefig('ellipse_H_M_nu1002.pdf', dpi=400, papertype='Letter',
             format='pdf', transparent=True)
 plt.close()

### 2015-10-12 n_priors_code/plotting/plot_prior_all.py

@@ -36,7 +36,7 @@ from palettable.colorbrewer.qualitative import Set1_9
 # ============================================
 
 no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
-plot_now = ['omnuh2','massless_neutrinos','w']
+plot_now = ['omnuh2','w']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
 
@@ -45,9 +45,9 @@ excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
 run_idx = 4
-lmax = 3000
-lmin = 50
-N_det = 10 ** 4
+lmax = 4499
+lmin = 4
+N_det = 10 ** 6
 # N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
 fsky = 0.75
 # ======
@@ -179,6 +179,7 @@ label['ombch2'] = '\Omega_{m}h^{2}'
 label['omch2'] = '\Omega_{c}h^{2}'
 label['helium_fraction'] = 'Y_{p}'
 label['w'] = 'w'
+label['wa'] = 'w_a'
 label['scalar_nrun(1)'] = r'\alpha_{s}'
 
 fisher_inv = np.linalg.inv(fisher_mat)
@@ -188,7 +189,7 @@ fisher_inplace = fisher_mat.copy()
 # CYCLE ON PARAMETERS (KEYS HERE)
 fg = plt.figure(figsize=fig_dims)
 
-for y, key_y in enumerate(plot_now):
+for y, key_y in enumerate(plot_param):
     print key_y
     ax1 = plt.subplot2grid((1, 1), (0, 0))
     ax1.set_color_cycle(Set1_9.mpl_colors)

### 2015-10-12 n_priors_code/utils.py

@@ -45,8 +45,10 @@ def load_data(data_folder,  values, lensed=False):
     dats = np.concatenate((cmb, lensing[:cmb.shape[0], :]), axis=1)
     # Load data for all parameters variations
     for key, value in values.iteritems():
-        for i in np.arange(0, 4):
-            print key, values[key][i]
+        if np.shape(values[values.keys()[0]])[0]!=4:
+            print "you are not using a 5 point formula are you sure?"
+        for i in np.arange(0, np.shape(values[values.keys()[0]])[0]):
+            print key, values[key][i],i
             filename = 'data/{}/'.format(data_folder)
             # filename_cmb = filename + key + '_{:.13f}'.format(values[key][i]) + '_lensedcls.dat'
             filename_cmb = filename + key + \
@@ -59,7 +61,7 @@ def load_data(data_folder,  values, lensed=False):
             newdat = np.concatenate((cmb, lensing[:dats.shape[0], :]), axis=1)
             # newdat = np.genfromtxt(filename)
             dats = np.dstack((dats, newdat))
-
+    np.save('data/{}/fisher_data.npy'.format(data_folder),dats)
     return dats
 
 
@@ -448,7 +450,6 @@ def save_cov_matrix(fisher_inv, filename='output_cmb/param_cov.txt'):
         for j in range(n_values):
             if i != j:
                 param_cov[i, j] = fisher_inv[i, j] / np.sqrt(fisher_inv[i, i] * fisher_inv[j, j])
-    # print param_cov
     np.savetxt(filename, param_cov)
 
 

### 2015-10-15 Notes/N_eff_prior_paper_v2.tex

@@
 We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$ unless the largest scale at reach for CMB S4 will be large enough to constraint $\tau$ internally thought the reionization bump.
-An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
+To imp a $3\%$ constraint on
 \end{abstract}
@@
 Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
-They have the potential to maintain this important role in the near future thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation.
+In the near future, they have the potential to maintain this important role  thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation.
 Indeed, the next generation of CMB experiments (S4), now in its planning stage, will measure the E-mode polarization with almost cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-This new sensitivity, in particular if supported with the essential synergy with other cosmological probes, will be a crucial step in our understanding of dark matter, inflation, Dark Energy and neutrinos.
+This new sensitivity, in particular if supported with the essential synergy with other cosmological probes, will be a a fundamental key in our understanding of dark matter, inflation, Dark Energy and neutrinos.
 
 Neutrinos properties can be effectively constrained with CMB measurements because they influence both the early history of the universe and the later stage of structure formation. 
-Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their effectivenumber $N_{\rm eff}$ affects the expansion rate.
+In the early Universe relativistic species, like neutrinos, are the main drivers of the cosmic expansion and the expansion rate depends on their effective number $N_{\rm eff}$.
 This can be powerfully tested using the CMB, by comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
-The future experiments, with their high resolution, will be able to measure the CMB power spectrum, at small scales, deeply into the damping tale, testing any possible deviations from $N_{\rm eff}=3.046$, the value predicted by the standard model.
-The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
-However massive neutrinos influence the growth of structure at scales shorter than their free streaming.
+S4 experiments, with their high resolution, will be able to measure the CMB power spectrum, at small scales, deeply into the damping tale, testing any possible deviations from $N_{\rm eff}=3.046$, the value predicted by the standard model.
+The total mass of neutrinos, on the other hand, has a modest effect on the primary CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
+However massive neutrinos influence the growth of structure at scales shorter than their free streaming one.
 Different neutrino masses consequently lead to different CMB lensing.
 Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction.
-This will turn into such a precise constraints of the sum of neutrino masses to get a possible hint of the hierarchies of the individual masses.
+This will turn into such a precise constraints of the sum of neutrino masses to get a possible hint of the hierarchical structure of the individual masses.
 Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $45\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
@@
 As previously mentioned, CMB is also a powerful probe of the dark energy properties (see \cite{2010MNRAS.405.2639J}).
-The challenge for future experiments is to go beyond the now well measured energy density and to determine other dark energy properties.
+The challenge for future experiments is to go beyond the, now well measured, energy density and to determine other dark energy properties.
 Among these, a crucial step would be to constraint the equation of state, the ratio of pressure and energy density, looking for possible deviations from the value $w=-1$ expected in a cosmological constant scenario.
-CMB is sensitive to Dark energy because this component alters the universe's expansion and it consequently changes the distance to the last scattering surface.
+CMB is sensitive to Dark energy because this component alters the late universe's expansion and it consequently changes the distance to the last scattering surface.
 Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing.
 However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
-As a consequence, similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
+As a consequence, similarly to the neutrino sector, CMB experiments will have to rely on external prior coming from BAO and supernova experiments to fully exploit their potential.
 
@@
 To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
-For all the parameters of the standard $\Lambda$CDM plus massive neutrinos ($\Omega_c h^2,\Omega_b h^2,A_s,n_s,\tau,H_0,M_{\nu}$) together with possibly testable non-standard extension ($N_{\rm eff},w,w_{a}$) we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
+For all the parameters of the standard $\Lambda$CDM plus massive neutrinos ($\Omega_c h^2,\Omega_b h^2,A_s,n_s,\tau,H_0,M_{\nu}$) together with possible non-standard extension ($N_{\rm eff},w,w_{a}$) we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
 
-This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
+This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we used to derive the effect of external priors on the CMB parameter constraints.
 In \refsec{results} we will describe our results and we then conclude with a discussion of them in \refsec{conclusions}.
@@
 To measure quantitatively the impact of external priors on the CMB ability to constrain cosmological parameters we use a Fisher matrix formalism.
-In this section we will quickly review  the technique and we will then present our fiducial cosmological model together with the experimental specifications of S4 CMB we used.
+In this section we will quickly review  the technique and we will then present our fiducial cosmological model together with the experimental specifications we used.
 
@@
 Furthermore it is easy to introduce external priors on cosmological parameters.
-Indeed we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
+Indeed we simply need to add to the Fisher matrixes elements the external priors (before we perform the matrix inversion of \refeq{cramer-rao}).
+For example, a $1\%$ prior on the parameter $i$ can be added by:
 \begin{equation}
-F_{\rm total}=F_{\rm CMB}+\sum F_{\rm external}.
-\end{equation}
-In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
-For example, a $1\%$ prior on $H_{0}$ can be added by:
-\begin{equation}
-F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,\text{fid}})^2}.
+F_{ii} \rightarrow F_{ii} + \frac{1}{(1\% \times  \theta_{i,\text{fid}})^2}.
 \end{equation}
@@
 As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing $\boldsymbol{C}_\ell$ negligible.
-We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
-%Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
-%Lastly we implement the same technique of \cite{2006astro.ph..9591A} to avoid possible numerical instability in the marginalizing procedure.
-This allow us not to invert the entire matrix when we want to marginalize over a set of parameters, in order to minimize numerical issues and conserve parameters degeneracies as much as possible.
+We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$.
+The constraint change by at most $10\%$ that should then be considered as a conservative numerical uncertainties upper limit.
 
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[scale=0.4]{PS_phi_with_noise.pdf}
-%\caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
-%In the figure: the deflection power spectrum for our fiducial cosmology together with the two examples of lensing reconstruction noise $N^{\phi}$ used in this work.}
-%\label{fig:phi-cl-noise}
-%\end{center}
-%\end{figure}
-%
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[scale=0.4]{PS_with_noise.pdf}
-%\caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
-%In the figure: CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
-%\label{fig:cmb-cl-noise}
-%\end{center}
-%\end{figure}
-
-We compute the power specctra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB.
+We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB.
 Our base cosmology is a flat $\nu \Lambda$CDM. 
@@
 $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc and a neutrino total mass $M_{\nu}$ $\simeq$ 85\ meV.
-Regarding the $\Lambda$CDM extensions, we chose, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1,w_{a}=0$.
+Regarding the $\Lambda$CDM extensions, we chose as fiducial, a standard $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1,w_{a}=0$.
 It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
@@
 This allow us to use all the polarization power spectrum multipoles all the way up to $\ell_{\rm E,max}=5000$.
-We deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
+We deal with the poisson noise from point sources in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
 The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
@@
 
-%Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from the CMB.
-The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons.
-This introduces non gaussianities that couple different, otherwise independent, CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
-The noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ is modeled assuming an iterative reconstruction technique.
-Indeed it has been shown \cite{seljak:2004} that the low level of noise expected in CMB S4 experiment will be in the regime  where this technique improve our constraints compared to the quadratic estimator introduced in \cite{okamoto:2003,hu:2002}.
+The noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.
+Indeed it has been shown that the low level of noise expected in CMB S4 experiment will be in the regime  where this technique improve our constraints compared to the quadratic estimator introduced in \cite{okamoto:2003,hu:2002}.
 
 \section{Results \label{sec:results}}
-In this section we present our main results. We will start from our base $\nu \Lambda$CDM and we then explore possible extensions.
+In this section we present our main results. We will start from our base $\nu \Lambda$CDM and we then explore extensions to this model ($N_{\rm eff},w$).
 
@@
 Since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$.
-This is particularly exciting because we now know that neutrinos are massive with a lower limit on $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
+This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
+
+In CMB experiment, neutrino masses constraint comes almost entirely from lensing.
+The CMB lensing power spectrum shape is not measured well enough to distinguish the effect of different neutrino masses.
+For this reason schematically the constraint comes from a comparison
+%==========
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
+\caption{$H_{0}$ and $\Omega_{c}h^{2}$ will be the more useful external prior to improve neutrino masses constraint.
+In the figure, the constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos} are shown.
+Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
+We chose a fiducial value $\sum m_{\nu}=85 $ meV.}
+\label{fig:prior_omeganuh2}
+\end{center}
+\end{figure}
+%==========
 
@@
 \begin{center}
-\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
-\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
-Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
-The lower limit on the sum of the neutrino masses is 50 meV.}
-\label{fig:prior_omeganuh2}
-\end{center}
-\end{figure}
-%==========
-%==========
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{prior_omnuh2_snow_mass_lmin=50_lmax=3000_ndet=10000_fsky=0.75}
-\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
+\includegraphics{prior_omnuh2_snow_mass_lmin=50_lmax=3000_ndet=1000000_fsky=0_75.pdf}
+\caption{Constraints on the sum of the neutrino masses for $\pmb{\ell_{\rm min}>50}$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
 Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
@@
 %==========
+As recently pointed out by \cite{allison:2015} there are two important regimes for neutrino masses constraint.
+In the first one,  large scale are assumed to be accessible by CMB experiments and, as a consequence the optical depth $\tau$ can be constrained internally by measuring the reionization bump.
+A second possibility is that CMB S4 will be limited to smaller scales and therefore it will have to rely on external priors to constrain $\tau$.
+This can be given by current stage CMB experiments like Planck as well as future 21 cm survey \cite{liu:2015}.
+This distinction is important because the optical depth and the neutrino mass are quite degenerate and this will be limiting the constraint in the context of very precise future CMB experiments (see \cite{allison:2015} for details).
 
@@
 Adding in external priors helps this significantly though, as discussed in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
-There, the prior was deescribed as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
-Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$, which indeed determines these low-redshift distances.
-As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior.
-Adding in other priors would help further.
+There, the prior was described as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
+Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed determines these low-redshift distances.
+Indeed it is important to keep in mind that a 
+As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior. Adding in other priors would help further.
+
+In \reffig{prior_omeganuh2_tau} we limit CMB accessible modes to $\ell_{\rm min}>50$.
 
@@
 However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably.
-{\bf We should try to understand why.
-Probably just a geometrical degeneracy in the way $w$ and $H_{0}$ enter the definition of the distance to the LSS $D(z_{\rm LSS})$}
+This primarily due to the fact that CMB constrained $w$ through a measurement of the last scattering surface.
+However that distance can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat).
+Indeed in a two dimensional ($H_{0},w$) the CMB constrained approximately lies on the constant large scattering surface line.
+Notice that an external prior on $H_{0}$ three times more accurate than CMB S4 can improve the CMB constraint almost by a factor of three from a $8\%$ error to a $3\%$ level.
 
@@
 \begin{description}
+\item[$M_{\nu}$] 
 \item[$N_{\rm eff}$] 
-\item[$M_{\nu}$] 
 \item[$w$,$w_{a}$] 
+\end{description}
+
 This results suggest...
-
 Comment on issue and future.
 This will be even more relevant when we open the parameter space.
-Will CMB + others be able to discover somethign CMB can't alone?
+Will CMB + others be able to discover something CMB can't alone?
 $N_{\rm eff}$ extensions, mass splitting etc
-
-\end{description}
 

### 2015-10-30 Notes/N_eff_prior_paper_v2.tex

@@
 \documentclass[aps,prd,reprint,superscriptaddress]{revtex4-1}
-%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
-%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
-%\usepackage{bm}
 \usepackage{amsmath}
@@
 %\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
-%\input{papers_packages_command._include_tex}
 \newcommand\refsec[1]{\S\ref{sec:#1}}
@@
 Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
-We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$ unless the largest scale at reach for CMB S4 will be large enough to constraint $\tau$ internally thought the reionization bump.
-To imp a $3\%$ constraint on
+We find that CMB S4 constraints on neutrino mass will be strongly enhanced by cold dark matter $\Omega_{c}h^{2}$ and Hubble constant $H_{0}$ external informations.
+If the largest scales ($\ell<50$) will not be measured an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important.
+A CMB constraint on $N_{\rm eff}$ will benefit from external prior on the spectral index $n_{s}$ and baryons energy density $\Omega_{b}h^{2}$.
+Finally, external prior on $H_{0}$ will be crucial to constraint a constant dark energy equation of state($w_{0}$).
 \end{abstract}
@@
 Indeed, the next generation of CMB experiments (S4), now in its planning stage, will measure the E-mode polarization with almost cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-This new sensitivity, in particular if supported with the essential synergy with other cosmological probes, will be a a fundamental key in our understanding of dark matter, inflation, Dark Energy and neutrinos.
+This new sensitivity, in particular if supported with the essential synergy with other cosmological probes, will be a a fundamental step forward in our understanding of dark matter, inflation, Dark Energy and neutrinos.
 
@@
 This can be powerfully tested using the CMB, by comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
-S4 experiments, with their high resolution, will be able to measure the CMB power spectrum, at small scales, deeply into the damping tale, testing any possible deviations from $N_{\rm eff}=3.046$, the value predicted by the standard model.
-The total mass of neutrinos, on the other hand, has a modest effect on the primary CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
+S4 experiments, with their high resolution, will be able to measure the CMB power spectrum, at small scales, deeply into the damping tale, testing  possible deviations from $N_{\rm eff}=3.046$, the value predicted by the standard model.
+The total mass of neutrinos, on the other hand, has a modest effect on the primary CMB because, for the range of masses allowed by recent constraints ($\sum m_\nu<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
 However massive neutrinos influence the growth of structure at scales shorter than their free streaming one.
@@
 This will turn into such a precise constraints of the sum of neutrino masses to get a possible hint of the hierarchical structure of the individual masses.
-Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $45\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
+Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $45\%$ with a factor of two improvement on $\sum m_\nu$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
@@
 As previously mentioned, CMB is also a powerful probe of the dark energy properties (see \cite{2010MNRAS.405.2639J}).
-The challenge for future experiments is to go beyond the, now well measured, energy density and to determine other dark energy properties.
+The challenge for future experiments is to go beyond the measurement of the energy density and to determine other dark energy properties.
 Among these, a crucial step would be to constraint the equation of state, the ratio of pressure and energy density, looking for possible deviations from the value $w=-1$ expected in a cosmological constant scenario.
@@
 To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
-For all the parameters of the standard $\Lambda$CDM plus massive neutrinos ($\Omega_c h^2,\Omega_b h^2,A_s,n_s,\tau,H_0,M_{\nu}$) together with possible non-standard extension ($N_{\rm eff},w,w_{a}$) we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
+For all the parameters of the standard $\Lambda$CDM plus massive neutrinos ($\Omega_c h^2,\Omega_b h^2,A_s,n_s,\tau,H_0,\sum m_\nu$) together with possible non-standard extension ($N_{\rm eff},w$) we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
 
@@
 where $\theta_{i,j}$ represents two of the cosmological parameters and $\boldsymbol{\theta_0}$ is the fiducial values array that, by definition, maximizes the likelihood. 
-It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian.
+%It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian.
 Even if the gaussian approximation has been shown to be appropriate most of the time, exceptions exist in the literature \cite{2012JCAP...09..009W}.
@@
 
-We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data and easily including external priors. 
-This technique however introduces also technical difficulties. 
-Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
-Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix lead to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
-Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard procedure used to forecast future constrains \cite{wu:2014}.
+%We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data and easily including external priors. 
+%This technique however introduces also technical difficulties. 
+%Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
+%Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix lead to a big change in its inverse.
+Because we use \refeq{cramer-rao} this can be a problem for error estimation.
+%Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard procedure used to forecast future constrains \cite{wu:2014}.
 We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
@@
 We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$.
-The constraint change by at most $10\%$ that should then be considered as a conservative numerical uncertainties upper limit.
+The constraints change by at most $10\%$ that should then be considered as a conservative numerical uncertainties.
 
@@
 Our base cosmology is a flat $\nu \Lambda$CDM. 
-We then extend the parameter space by introducing $N_{\rm eff}$ and $w,w_{a}$ as free parameters.
+We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters.
 We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
-$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc and a neutrino total mass $M_{\nu}$ $\simeq$ 85\ meV.
+$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc and a neutrino total mass $\sum m_\nu$ $\simeq$ 85\ meV.
 Regarding the $\Lambda$CDM extensions, we chose as fiducial, a standard $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1,w_{a}=0$.
@@
 where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
-We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin (PRELIMINARY).
-Note that $1.5$ $\mu$K-arcmin is the noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
+We decide to use a level of noise $s = 0.58$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin. This correspond to the $N_{\rm det}=10^{6}$ case of \cite{wu:2014}. 
+Note that the quoted noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
 
 The noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.
-Indeed it has been shown that the low level of noise expected in CMB S4 experiment will be in the regime  where this technique improve our constraints compared to the quadratic estimator introduced in \cite{okamoto:2003,hu:2002}.
 
@@
 
-In CMB experiment, neutrino masses constraint comes almost entirely from lensing.
-The CMB lensing power spectrum shape is not measured well enough to distinguish the effect of different neutrino masses.
-For this reason schematically the constraint comes from a comparison
+%In CMB experiment, neutrino masses constraint comes almost entirely from lensing.
+When the total mass of neutrinos the CMB lensing power spectrum amplitude decreases at scales smaller than the free-streeming scale of neutrinos.
+This effect is degenerate with an overall change of the power spectrum amplitude $A_{s}$.
+
 %==========
-
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
-\caption{$H_{0}$ and $\Omega_{c}h^{2}$ will be the more useful external prior to improve neutrino masses constraint.
-In the figure, the constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos} are shown.
-Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
-We chose a fiducial value $\sum m_{\nu}=85 $ meV.}
+\includegraphics{{prior_omnuh2_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{
+In the figure, the constraints on the sum of the neutrino masses as a function of priors on cosmological parameters are shown ($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ).
+Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm CMB}=0.6\%$.
+$H_{0}$ and $\Omega_{c}h^{2}$ will be the more useful external prior to improve neutrino masses constraint in this regime.
+We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-like prior.
+The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
 \label{fig:prior_omeganuh2}
@@
 \begin{center}
-\includegraphics{prior_omnuh2_snow_mass_lmin=50_lmax=3000_ndet=1000000_fsky=0_75.pdf}
-\caption{Constraints on the sum of the neutrino masses for $\pmb{\ell_{\rm min}>50}$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
-Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$.
-The lower limit on the sum of the neutrino masses is 50 meV.}
+\includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Some as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}. }
 \label{fig:prior_omeganuh2_tau}
@@
 %==========
-As recently pointed out by \cite{allison:2015} there are two important regimes for neutrino masses constraint.
+%As recently pointed out by \cite{allison:2015} there are two important regimes for neutrino masses constraint.
 In the first one,  large scale are assumed to be accessible by CMB experiments and, as a consequence the optical depth $\tau$ can be constrained internally by measuring the reionization bump.
@@
 
-\reffig{prior_omeganuh2} shows that, without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
+\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 experiment ($N_{det}=10^{6}$, $f_{sky}=0.75$,  $2<\ell<5000$) as a function of external priors.
+Without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
 Adding in external priors helps this significantly though, as discussed in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
@@
 Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed determines these low-redshift distances.
-Indeed it is important to keep in mind that a 
+Notice that external prior effect are not completely independent. An external prior on $H_{0}$ for example, improve the internal reconstruction of $\Omega_ch^2$ as well.
 As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior. Adding in other priors would help further.
@@
 In \reffig{prior_omeganuh2_tau} we limit CMB accessible modes to $\ell_{\rm min}>50$. 
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will be helpful.
+Indeed as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $\tau$.
+A measurement of the optical depth will help to constraint the primordial amplitude $A_{s}$ which is degenerate with the mass of neutrinos.
+Indeed, increasing the mass of neutrinos will decrease the CMB lensing power spectrum and this can be compensating by increasing $A_{s}$.
 
@@
 \begin{center}
-\includegraphics{prior_massless_neutrinos_snow_mass_lmin=4_lmax=4499.pdf}
+\includegraphics{{prior_massless_neutrinos_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model.
-Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm old}=4.1\%$.
 The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.
@@
 With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$.
-The sensitivity comes from the effect of extra species on the damping tail of the CMB anisotropies, both in temperature and polarization, so parameters that also affect the shape of the damping tail, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are most degenerate with $N_{\rm eff}$.
+The sensitivity comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also affect the shape of the damping tail, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are most degenerate with $N_{\rm eff}$.
 As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction.
@@
 
-While the primordial CMB has limited information about the late-time dark energy equation of state, CMB lensing probes the growth of structure at late times and therefore is sensitive to $w$.
-Here we are using only the information from the lensing power spectrum as inferred from the CMB, but there is even more information not considered here contained in the many galaxy clusters that upcoming missions will detect via the Sunyaev-Zel'dovich effect.
+The CMB constrained the late-time dark energy equation of state in two ways. The primordial CMB is very sensitive to the distance to the last scattering surface which depends on $w$. 
+Furthermore the CMB lensing probes the growth of structure at late times which is different for different dark matter equations of state.
+%==========
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. An external prior on $H_{0}$ will be crucial to improve the constraint.} 
+\label{fig:prior_w}
+\end{center}
+\end{figure}
+%==========
 
@@
 \begin{center}
-\includegraphics{prior_w_snow_mass_lmin=4_lmax=4499.pdf}
-\caption{Constraints on the dark energy equation of state $w$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.
-An external prior on $H_{0}$ will be crucial to improve the constraint.}
-\label{fig:prior_w}
+\includegraphics{{ellipse_H0_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0_75}.pdf}
+\caption{Two dimensional constraint 1$\sigma$ ellipse for $w,H_{0}$ space. The dashed (red) line correspond to values in the plane corresponding to the same distance to the last scattering surface.} 
+\label{fig:ellipse_w_H0}
 \end{center}
@@
 However that distance can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat).
-Indeed in a two dimensional ($H_{0},w$) the CMB constrained approximately lies on the constant large scattering surface line.
+Indeed in a two dimensional ($H_{0},w$) the CMB constrained approximately lies on the region of constant large scattering surface as shown in \reffig{ellipse_w_H0}.
 Notice that an external prior on $H_{0}$ three times more accurate than CMB S4 can improve the CMB constraint almost by a factor of three from a $8\%$ error to a $3\%$ level.
@@
 \section{Conclusions  \label{sec:conclusions}}
-The planning of the next CMB Stage IV experiments represents an exciting challenge.
+The design of the next CMB Stage IV experiments represents an exciting challenge.
 The possible reward is remarkable.
-We can finally put a tight constraint on the total mass of neutrinos with deeply implications on particle physics like, for example, the ability to distinguish between normal and inverted hierarchy.
-We will measure with great accuracy the number of relativistic species, testing both standard model prediction and, eventually, probing new physics.
+We can finally put a tight constraint on the total mass of neutrinos with deeply implications on particle physics like, for example, the solution of the mass hierarchy problem.
+We will measure with great accuracy the number of relativistic species, both testing standard model prediction and, eventually, probing new physics.
 The same is true for dark energy where we will tighten the constraint on $w$,$w_{a}$ getting closer to test the nature of the accelerated expansion of the universe.
-These however will not be possible without an optimal synergy with other experiments observing a variety of others cosmological probes.
-Indeed this should be taken into account in this CMB Stage IV planning phase.
-Understanding CMB weaknesses and the impact of external data can have a deep influence in planning the experiment specifications.
-For this reason in this work we explored the impact of external priors on CMB cosmological constraints and in particular on the neutrino sector and dark energy parameters.
+These results, however, will not be possible without an optimal synergy with the other experiments observing a variety of others cosmological probes.
+Indeed this should be taken into account, especially during the CMB Stage IV planning phase.
+Understanding CMB weaknesses and the impact of external data can have a deep influence in the needed experiment specifications.
+For this reason in this work we explored the impact of external priors on CMB S$4$ cosmological constraints and in particular on the neutrino sector and dark energy parameters.
 We find:
 \begin{description}
-\item[$M_{\nu}$] 
-\item[$N_{\rm eff}$] 
-\item[$w$,$w_{a}$] 
+\item[$\sum m_\nu$] To improve the CMB constraint by a factor of two, low-redshift measurements will be needed...
+If the final design of CMB S4 will not include measurement at large scale ($\ell<50$) external constraint on the primordial amplitude will be needed.
+Moreover, an external prior on $\tau$ will allow the CMB to constraint $A_{s}$ internally thus strongly alleviating the problem.
+\item[$N_{\rm eff}$] The effective neutrino numbers will benefit from external prior on parameters that affect the shape of the damping tail.
+\item[$w$] In this case the improvement will mainly come by external prior on $H_{0}$.
 \end{description}
 
-This results suggest...
-Comment on issue and future.
+\textit{Comment on issue and future.
 This will be even more relevant when we open the parameter space.
 Will CMB + others be able to discover something CMB can't alone?
-$N_{\rm eff}$ extensions, mass splitting etc
-
-% tables should appear as floats within the text
-%
-% Here is an example of the general form of a table:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
-% \begin{tabular}{} command.
-% The ruledtabular enviroment adds doubled rules to table and sets a
-% reasonable default table settings.
-% Use the table* environment to get a full-width table in two-column
-% Add \usepackage{longtable} and the longtable (or longtable*}
-% environment for nicely formatted long tables. Or use the the [H]
-% placement option to break a long table (with less control than 
-% in longtable).
-% \begin{table}%[H] add [H] placement to break table across pages
-% \caption{\label{}}
-% \begin{ruledtabular}
-% \begin{tabular}{}
-% Lines of table here ending with \\
-% \end{tabular}
-% \end{ruledtabular}
-% \end{table}
-
-% Surround table environment with turnpage environment for landscape
-% table
-% \begin{turnpage}
-% \begin{table}
-% \caption{\label{}}
-% \begin{ruledtabular}
-% \begin{tabular}{}
-% \end{tabular}
-% \end{ruledtabular}
-% \end{table}
-% \end{turnpage}
-
-% Specify following sections are appendices. Use \appendix* if there
-% only one appendix.
-
-%\appendix
+$N_{\rm eff}$ extensions, mass splitting etc}
 
 \begin{acknowledgments}
-We thank Youngsoo Park for his contribution in the early stage of this work.
+We thank Youngsoo Park for his important contribution in the early stage of this work.
 AM wants to thank Zhen Pan who allowed a careful cross-check of our results.

### 2015-10-30 n_priors_code/fisher_joint_test.py
change plot, test D_LSS

@@ -199,7 +199,7 @@ def C(iell, ell, parbin, data):
 # =============================
 l_t_max = 3000  # this is the multipole you want to cut the temperature Cl at, to simulate the effect of foregrounds
 lmax = 4499
-lmin = 5
+lmin = 4
 N_det = 10 ** 6
 N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_6.txt')
 data_folder = 'varying_all/run4'
@@ -348,7 +348,7 @@ planck_fisher = np.loadtxt('/home/manzotti/n_eff-dependence-on-prior/n_priors_co
 
 # print ''
 # print ''
-fisher +=planck_fisher
+# fisher +=planck_fisher
 
 print 'lmax =', ell
 # print fisher_inv

### 2015-10-30 n_priors_code/plotting/plot_prior_paper_M_nu_ell4.py

@@ -55,7 +55,7 @@ fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type,
 values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
 par_gaps = pickle.load(open(base_dir + 'data/{}/run{}/par_gaps.p'.format(data_type, str(run_idx)), "rb"))
 fisher_mat = np.loadtxt(
-    base_dir + 'data/{}/run{}/output/fisher_mat_joint_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(data_type, str(run_idx), lmin, lmax, N_det, fsky))
+    base_dir + 'data/{}/run{}/output/fisher_mat_CMB_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(data_type, str(run_idx), lmin, lmax, N_det, fsky))
 
 par_gaps, values, fid, fisher_mat = utils.exclude_parameters_from_fisher(
     excluded_parameters, par_gaps, values, fid, fisher_mat)
@@ -253,7 +253,7 @@ for key_y in ['omnuh2']:
     # ============================================
 
     # ============================================
-    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/CMB_prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                 format='pdf', bbox_inches='tight')
     plt.clf()
 

### 2015-10-30 n_priors_code/plotting/plot_prior_paper_M_nu_ell50.py

@@ -55,7 +55,7 @@ fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type,
 values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
 par_gaps = pickle.load(open(base_dir + 'data/{}/run{}/par_gaps.p'.format(data_type, str(run_idx)), "rb"))
 fisher_mat = np.loadtxt(
-    base_dir + 'data/{}/run{}/output/fisher_mat_joint_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(data_type, str(run_idx), lmin, lmax, N_det, fsky))
+    base_dir + 'data/{}/run{}/output/fisher_mat_CMB_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(data_type, str(run_idx), lmin, lmax, N_det, fsky))
 
 par_gaps, values, fid, fisher_mat = utils.exclude_parameters_from_fisher(
     excluded_parameters, par_gaps, values, fid, fisher_mat)
@@ -246,7 +246,7 @@ for key_y in ['omnuh2']:
     # ============================================
 
     # ============================================
-    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/CMB_prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                 format='pdf', bbox_inches='tight')
     plt.clf()
 

### 2015-10-30 n_priors_code/plotting/plot_prior_paper_w_wa.py

@@ -1,253 +1,0 @@
-'''
-This plotting script is used to plot the effect of external priors on all the parameters involved.
-
-To put all the parameters in the same plot the prior is defined as the improvement from the CMB S4 results we get.
-
-the same is true for the improvement on the y axis parameters
-
-
-
-'''
-
-
-import numpy as np
-import pickle as pickle
-import utils
-import matplotlib
-matplotlib.use('Agg')
-import matplotlib.pyplot as plt
-from scipy import integrate
-import scipy.special
-import scipy.interpolate
-from itertools import cycle
-#from matplotlib.colors import colorConverter
-from matplotlib.ticker import MaxNLocator  # needed for tick
-import sys
-from matplotlib.patches import Ellipse
-import n_priors_code.utils as utils
-from palettable.colorbrewer.qualitative import Set1_9
-# ============================================
-# ============================================
-# ============================================
-# TO DO
-#
-# ============================================
-# ============================================
-# ============================================
-
-no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
-plot_now = ['omnuh2','w']
-excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
-# omnuh2
-
-# READ DATA
-# DEFINE YOUR FOLDER HERE
-base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
-data_type = 'varying_all'
-run_idx = 2
-lmax = 4499
-lmin = 4
-N_det = 10 ** 6
-# N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
-fsky = 0.75
-# ======
-fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type, str(run_idx)), "rb"))
-values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
-par_gaps = pickle.load(open(base_dir + 'data/{}/run{}/par_gaps.p'.format(data_type, str(run_idx)), "rb"))
-fisher_mat = np.loadtxt(
[…]

### 2015-10-30 n_priors_code/plotting/plot_triangle_ellipses_plot_2.py

@@ -48,29 +48,32 @@ from palettable.colorbrewer.qualitative import Set1_9
 # READ DATA
 
 
-no_lcdm_parameters =  ['massless_neutrinos', 'w', 'omnuh2']
+no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
 plot_now = ['omnuh2']
-excluded_parameters = list(set(no_lcdm_parameters) -set(plot_now))
-excluded_parameters = None
+excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
 
 # READ DATA
 # DEFINE YOUR FOLDER HERE
 base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
-data_type = 'varying_lambda'
-run_idx = 2
+data_type = 'varying_all'
+run_idx = 4
 lmax = 4499
-lmin = 4
+lmin = 50
+N_det = 10 ** 6
+# N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
+fsky = 0.75
 # ======
 fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type, str(run_idx)), "rb"))
 values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
 par_gaps = pickle.load(open(base_dir + 'data/{}/run{}/par_gaps.p'.format(data_type, str(run_idx)), "rb"))
 fisher_mat = np.loadtxt(
-    base_dir + 'data/{}/run{}/output/fisher_mat_joint_lmin={}_lmax={}.txt'.format(data_type, str(run_idx), lmin, lmax))
+    base_dir + 'data/{}/run{}/output/fisher_mat_joint_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(data_type, str(run_idx), lmin, lmax, N_det, fsky))
 
 par_gaps, values, fid, fisher_mat = utils.exclude_parameters_from_fisher(
     excluded_parameters, par_gaps, values, fid, fisher_mat)
-# plot_param = list(set(fid.keys()) -set(excluded_parameters))
+plot_param = list(set(fid.keys()) - set(excluded_parameters))
+
 
 
 my_fisher_inv = np.linalg.inv(fisher_mat)
@@ -268,6 +271,5 @@ plt.rcParams['legend.handletextpad'] = 0.3
[…]

### 2015-11-02 Notes/N_eff_prior_paper_v2.tex
addressing Scott comments on the draft

@@
 \title{External priors for the next generation of CMB experiments.}
-
-\author{Alessandro Manzotti}
-\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
-\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-\author{Scott Dodelson}
-
-\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+%
+%\author{Alessandro Manzotti}
+%\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+%\author{Scott Dodelson}
+%
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
 
@@
 \end{eqnarray}
+The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be described at the end of this section.
 Note that we are neglecting the term $C_\ell^{E\phi}$.
 As also noticed in previous literature \cite{wu:2014,2013PhRvD..87h3008H} this term contains very little information while adding possible numerical issues.
-The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be described at the end of this section.
+Furthermore, as in \cite{wu:2014}, we decided to use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}.
+An eventual imperfect reconstruction of the unlensed spectra (delensing) and non gaussian effects described in \cite{benoit-levy:2012} will degrade the parameters constrained derived in this work.
 
-A lower bound on the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, can be easily obtained as:
+A lower bound on the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, can be obtained as:
 \begin{equation}
@@
 \end{equation}
-Furthermore it is easy to introduce external priors on cosmological parameters.
-Indeed we simply need to add to the Fisher matrixes elements the external priors (before we perform the matrix inversion of \refeq{cramer-rao}).
+Furthermore we can introduce external priors on cosmological parameters.
+Indeed we simply need to add to the Fisher matrix elements the external priors (before we perform the matrix inversion of \refeq{cramer-rao}).
 For example, a $1\%$ prior on the parameter $i$ can be added by:
@@
 Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
-This effect, probed by CMB lensing, will allow us to determine the contribution of neutrinos to the total matter density.
-Since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$.
+Indeed, below the current free-streaming scale, the matter power power spectrum is suppressed in the presence of
+massive neutrinos by a factor -8$f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
+This effect is probed by CMB lensing and since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$.
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
@@
 \includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Some as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}. }
+\caption{Some as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
 \label{fig:prior_omeganuh2_tau}
@@
 There, the prior was described as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
-Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed determines these low-redshift distances.
-Notice that external prior effect are not completely independent. An external prior on $H_{0}$ for example, improve the internal reconstruction of $\Omega_ch^2$ as well.
-As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior. Adding in other priors would help further.
+Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
+The improvement given by a prior on $\Omega_ch^2$ is a consequence of the fact that the CMB lensing is sensitive to $f_\nu$, therefore in order to constrain $\Omega_{\nu}h^2$ from this ratio we need to determine the matter energy density.
+A prior on  $H_{0}$ would help as well since the effects of external priors are not completely independent.
+The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^3$ and for this reason an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
+As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with these priors. Adding in other priors would help further.
 
 In \reffig{prior_omeganuh2_tau} we limit CMB accessible modes to $\ell_{\rm min}>50$. 
-Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will be helpful.
-Indeed as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $\tau$.
-A measurement of the optical depth will help to constraint the primordial amplitude $A_{s}$ which is degenerate with the mass of neutrinos.
-Indeed, increasing the mass of neutrinos will decrease the CMB lensing power spectrum and this can be compensating by increasing $A_{s}$.
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will also be helpful.
+Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and as a consequence between $\sum m_\nu$ and $\tau$. This can be understood as follow.
+The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig.5 of \cite{2013arXiv1309.5383A}).
+For this reason it is hard to measured the unsuppressed lensing CMB power spectra that is needed to constrain the values of $f_{\nu}$.
+The unsuppressed amplitude is actually derived from the primordial amplitude $A_{s}$.
+%Indeed, increasing the mass of neutrinos will decrease the CMB lensing power spectrum and this can be compensating by increasing $A_{s}$.
+A measurement of the optical depth will help to constraint the primordial amplitude $A_{s}$ internally since the CMB is sensitive to $A_{s}\exp^{-2\tau}$.
+If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} will be needed to fully exploit the CMB constraining power.
 
@@
 
-\reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other 6 parameters in the standard cosmological model ($H_0, \Omega_mh^2,\Omega_bh^2, n_s, A_s,$ and $\tau$).
+\reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other six parameters.
 With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$.
-The sensitivity comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also affect the shape of the damping tail, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are most degenerate with $N_{\rm eff}$.
+The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
+
+The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most degenerate with $N_{\rm eff}$.
 As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction.
-The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
+Also as noticed in \cite{baumann:2015} tight constraints on $N_{\rm eff}$ will allow to rule out early thermalized particles that decoupled early in the thermal history.
 
@@
 However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably.
-This primarily due to the fact that CMB constrained $w$ through a measurement of the last scattering surface.
-However that distance can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat).
-Indeed in a two dimensional ($H_{0},w$) the CMB constrained approximately lies on the region of constant large scattering surface as shown in \reffig{ellipse_w_H0}.
+This primarily due to the fact that CMB constrained $w$ through a precise measurement of the comoving distance to the last scattering surface.
+However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat).
+Indeed in a two dimensional ($H_{0},w$) the CMB constrained approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
 Notice that an external prior on $H_{0}$ three times more accurate than CMB S4 can improve the CMB constraint almost by a factor of three from a $8\%$ error to a $3\%$ level.
@@
 The same is true for dark energy where we will tighten the constraint on $w$,$w_{a}$ getting closer to test the nature of the accelerated expansion of the universe.
-These results, however, will not be possible without an optimal synergy with the other experiments observing a variety of others cosmological probes.
+These results, however, will not be possible without an optimal synergy with external experiments observing a variety of others cosmological probes.
 Indeed this should be taken into account, especially during the CMB Stage IV planning phase.
 Understanding CMB weaknesses and the impact of external data can have a deep influence in the needed experiment specifications.
-For this reason in this work we explored the impact of external priors on CMB S$4$ cosmological constraints and in particular on the neutrino sector and dark energy parameters.
+For this reason in this work we explored the impact of external priors on CMB S4 cosmological constraints and in particular on the neutrino sector and dark energy parameters.
 We find:

### 2015-11-04 Notes/N_eff_prior_paper_v2.tex
working on v2 of draft

@@
 %\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+
+%\author{Scott Dodelson}
+%
+%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+
+%\author{Youngsoo Park}
+%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
 
@@
 To measure quantitatively the impact of external priors on the CMB ability to constrain cosmological parameters we use a Fisher matrix formalism.
-In this section we will quickly review the technique and we will then present our fiducial cosmological model together with the experimental specifications we used.
+In this section we will quickly review the technique and we will then present the chosen fiducial cosmological model together with the experimental specifications.
 
@@
 The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be described at the end of this section.
-Note that we are neglecting the term $C_\ell^{E\phi}$.
-As also noticed in previous literature \cite{wu:2014,2013PhRvD..87h3008H} this term contains very little information while adding possible numerical issues.
+Note that we are neglecting the term $C_\ell^{E\phi}$. As also noticed in  \cite{wu:2014,2013PhRvD..87h3008H} this term contains very little information while adding possible numerical issues.
 Furthermore, as in \cite{wu:2014}, we decided to use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}.
-An eventual imperfect reconstruction of the unlensed spectra (delensing) and non gaussian effects described in \cite{benoit-levy:2012} will degrade the parameters constrained derived in this work.
+Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
 
@@
 
-The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the expected level for the next generation of CMB experiments (S4).
+The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the optimistic level for the next generation of CMB experiments (S4).
 For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
@@
 massive neutrinos by a factor -8$f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
-This effect is probed by CMB lensing and since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$.
+This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses $\sum m_\nu$.
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
@@
 
-\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 experiment ($N_{det}=10^{6}$, $f_{sky}=0.75$,  $2<\ell<5000$) as a function of external priors.
+\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment ($N_{det}=10^{6}$, $f_{sky}=0.75$,  $2<\ell<5000$) as a function of external priors.
 Without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
-Adding in external priors helps this significantly though, as discussed in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
+Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
 There, the prior was described as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
@@
 The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^3$ and for this reason an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
-As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with these priors. Adding in other priors would help further.
+As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors.
 
 In \reffig{prior_omeganuh2_tau} we limit CMB accessible modes to $\ell_{\rm min}>50$. 
-Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will also be helpful.
-Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and as a consequence between $\sum m_\nu$ and $\tau$. This can be understood as follow.
+Together with the importance of $H_{0}$ and $\Omega_ch^2$, that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will also be helpful.
+Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follow.
 The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig.5 of \cite{2013arXiv1309.5383A}).
-For this reason it is hard to measured the unsuppressed lensing CMB power spectra that is needed to constrain the values of $f_{\nu}$.
-The unsuppressed amplitude is actually derived from the primordial amplitude $A_{s}$.
-%Indeed, increasing the mass of neutrinos will decrease the CMB lensing power spectrum and this can be compensating by increasing $A_{s}$.
-A measurement of the optical depth will help to constraint the primordial amplitude $A_{s}$ internally since the CMB is sensitive to $A_{s}\exp^{-2\tau}$.
+For this reason it is hard to measure the unsuppressed lensing CMB power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$.
+The unsuppressed amplitude is actually better constrained by the primordial amplitude $A_{s}$.
+A measurement of the optical depth will help to constraint the primordial amplitude $A_{s}$ internally since the CMB is sensitive to the combination $A_{s}\exp^{-2\tau}$.
 If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} will be needed to fully exploit the CMB constraining power.
@@
 \includegraphics{{prior_massless_neutrinos_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model.
+\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of external priors.
 The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.
-The bottom-most (grey) curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
+The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
 \label{fig:prior_massless_neutrinos}
@@
 \includegraphics{{ellipse_H0_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0_75}.pdf}
-\caption{Two dimensional constraint 1$\sigma$ ellipse for $w,H_{0}$ space. The dashed (red) line correspond to values in the plane corresponding to the same distance to the last scattering surface.} 
+\caption{Two dimensional constraint (CMB S4 + Planck) 1$\sigma$ ellipse for $w,H_{0}$ space.
+The dashed (red) line correspond to values in the plane corresponding to the same distance to the last scattering surface.}
 \label{fig:ellipse_w_H0}
@@
 However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably.
-This primarily due to the fact that CMB constrained $w$ through a precise measurement of the comoving distance to the last scattering surface.
+This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface.
 However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat).
-Indeed in a two dimensional ($H_{0},w$) the CMB constrained approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
-Notice that an external prior on $H_{0}$ three times more accurate than CMB S4 can improve the CMB constraint almost by a factor of three from a $8\%$ error to a $3\%$ level.
+Indeed in a two dimensional ($H_{0},w$) space the CMB constrained approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
+Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint almost by a factor of 2 from a $8\%$ error to a $4\%$ level.
 
@@
 We will measure with great accuracy the number of relativistic species, both testing standard model prediction and, eventually, probing new physics.
-The same is true for dark energy where we will tighten the constraint on $w$,$w_{a}$ getting closer to test the nature of the accelerated expansion of the universe.
+The same is true for dark energy.
+In that context, we will tighten the constraint on $w$,$w_{a}$ getting closer to test the nature of the accelerated expansion of the universe.
 These results, however, will not be possible without an optimal synergy with external experiments observing a variety of others cosmological probes.
 Indeed this should be taken into account, especially during the CMB Stage IV planning phase.
-Understanding CMB weaknesses and the impact of external data can have a deep influence in the needed experiment specifications.
-For this reason in this work we explored the impact of external priors on CMB S4 cosmological constraints and in particular on the neutrino sector and dark energy parameters.
+Understanding CMB weaknesses and the impact of external data can have a deep influence in determining the needed experimental specifications.
+For this reason in this work we explored the impact of external priors on CMB S4 cosmological constraints and, in particular, on the neutrino sector and dark energy parameters.
 We find:
 \begin{description}
-\item[$\sum m_\nu$] To improve the CMB constraint by a factor of two, low-redshift measurements will be needed...
-If the final design of CMB S4 will not include measurement at large scale ($\ell<50$) external constraint on the primordial amplitude will be needed.
+\item[$\sum m_\nu$] To improve the CMB constraint by a factor of two, low-redshift measurements will be needed.
+Indeed external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level CMB S4 will reach alone can bring the error on the total mass of neutrinos from $30$meV to below 20meV.
+Furterhemore, if the final design of CMB S4 will not include measurement at large scale ($\ell<50$) external constraint on the primordial amplitude will be needed.
 Moreover, an external prior on $\tau$ will allow the CMB to constraint $A_{s}$ internally thus strongly alleviating the problem.
-\item[$N_{\rm eff}$] The effective neutrino numbers will benefit from external prior on parameters that affect the shape of the damping tail.
+\item[$N_{\rm eff}$] The effective neutrino numbers will benefit from external prior on parameters that affect the shape of the damping tail like the spectral index and the baryons energy density.
 \item[$w$] In this case the improvement will mainly come by external prior on $H_{0}$.
+For example a $2\%$ prior from external experiment will bring the reduce the error on w by a factor of two from 8$\%$ to $4\%$.
 \end{description}
@@
 \begin{acknowledgments}
-We thank Youngsoo Park for his important contribution in the early stage of this work.
 AM wants to thank Zhen Pan who allowed a careful cross-check of our results.

### 2015-11-10 Notes/N_eff_prior_paper.tex
edited v2

@@
 
+\newcommand\sd[1]{{\bf SD: #1}}
 %\newcommand\refsec[1]{\S\ref{sec:#1}}
@@
 
+\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
 \affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
 \affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
 
+\author{Youngsoo Park}
+\affiliation{Department of Physics, University of Arizona, Tucson, AZ 85721, USA}
 \date{\today}
 \begin{abstract}
-The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
+The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
 Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
-Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
-We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$  
+Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
+We find that CMB S4 constraints on neutrino masses will be limited by a degeneracy with the optical depth $\tau$  
 if the largest scale at reach is not enough to constraint $\tau$ internally.
@@
 \section{Introduction}\label{sec:intro}
-Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe and they will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation.
-The recent past is indeed reassuring. 
+Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe.
+They will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution expected in next generation, dubbed Stage IV (S4)~\cite{2013arXiv1309.5383A}.
+The recent past is  reassuring.
 Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
-This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck.
-With the improved sensitivity we extend our understanding of the universe scientific from a proof of its almost flat geometry to a well tested model $\Lambda$CDM with a percent level constraint on its parameters.
-The same success characterized ground CMB experiments, where the evolution from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
-These series of successes is far from its end.
-Indeed the next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As it happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.
+These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck.
+With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
+The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
 
-The cosmological dependence to neutrinos is primarily due to their contribution to the number of relativistic species $N_{\rm eff}$ in the early phase of the universe  together with the role they play, given their non zero mass $M_{\nu}$, into the late growth of cosmic structures.
+This pattern of successes is far from over.
+The next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy and neutrinos.
+
+The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $M_{\nu}$ affects the late-time growth of cosmic structures.
 Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$.
 This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
-The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale of the CMB power spectrum and will unequivocally test the value $N_{\rm eff}=3.046$ predicted by the standard model .
+The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$.
 The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
-As other low redshift effects however, massive neutrinos modify the CMB by altering the growth of the large scale structure responsible for the the lensing of the CMB photons.
-Different neutrino masses consequently lead to different CMB lensing.
-Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction.
+However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons.
+Different neutrino masses consequently lead to different CMB lensing spectra.
+A Stage IV experiment will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction.
 This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
-Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
+These signatures can be mimicked with variations in other cosmic parameters, so an important lingering question is: What external information can be used to improve projected constraints further by breaking degeneracies?
+%Quantitatively, the CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-As previously mentioned, CMB is also sensitive to the properties of dark energy (see \cite{2010MNRAS.405.2639J}).
+The CMB is sensitive also to the properties of dark energy (see \cite{2010MNRAS.405.2639J}).
 Thanks to the current generation of experiments we know with extraordinary accuracy its energy density.
 The challenge for the next generation is to reveal the nature of this mysterious component.
-For example, a crucial step to rule out part of the proposed models will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant.
+For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant.
 Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface.
-Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing.
-Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
-However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
-Similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.
+Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing.
+Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, the CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
+However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
+will be improved by external experiments.
 
 The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector.
-The future of cosmology will hinge on the ability of combining different probes.
-In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
-In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
-To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study in detail the dependance of CMB future constrains on the external priors assumed.
-For all the parameters of the standard $\Lambda$CDM with massive neutrinos and $w\neq-1$ dark energy model we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
+A general thrust in even the current generation of surveys is to combine information from different observations.
+%In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
+%In particular CMB will benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
+Here we study the dependance of projected constraints from CMB on the external priors assumed.
+For all the parameters of the standard $\Lambda$CDM with massive neutrinos and $w\neq-1$ dark energy model we quantify the CMB constraints as a function of external priors.
 

### 2015-11-10 Notes/N_eff_prior_paper_v2.tex
edited v2

@@
 \usepackage{graphicx}
+\usepackage{color}
+
+%\newcommand\sd[1]{{\bf SD: #1}}
+\newcommand{\sd}[1]{\textcolor{red}{[{\bf SD}: #1]}}
 
@@
 \title{External priors for the next generation of CMB experiments.}
+\author{Alessandro Manzotti}
+\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
+\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+\author{Scott Dodelson}
+
+\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
+\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
+\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
+
+\author{Youngsoo Park}
+\affiliation{Department of Physics, University of Arizona, Tucson, AZ 85721, USA}
 %
@@
 \begin{abstract}
-The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
-Indeed the low level of noise, together with improved angular resolution, will allow a precise measurement of the CMB polarization anisotropies as well as an improved reconstruction of the lensing potential of high redshift large scale structure.
+The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
+Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
-Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
-We find that CMB S4 constraints on neutrino mass will be strongly enhanced by cold dark matter $\Omega_{c}h^{2}$ and Hubble constant $H_{0}$ external informations.
-If the largest scales ($\ell<50$) will not be measured an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important.
-A CMB constraint on $N_{\rm eff}$ will benefit from external prior on the spectral index $n_{s}$ and baryons energy density $\Omega_{b}h^{2}$.
-Finally, external prior on $H_{0}$ will be crucial to constraint a constant dark energy equation of state($w_{0}$).
+Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
+$n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
+We find that CMB S4 constraints on neutrino mass will be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$.
+If the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important.
+A CMB constraint on $N_{\rm eff}$ will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$.
+Finally, an external prior on $H_{0}$ will be crucial to constrain a constant dark energy equation of state($w_{0}$).
 \end{abstract}
@@
 \section{Introduction}\label{sec:intro}
-Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe.
-In the near future, they have the potential to maintain this important role  thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation.
-Indeed, the next generation of CMB experiments (S4), now in its planning stage, will measure the E-mode polarization with almost cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-This new sensitivity, in particular if supported with the essential synergy with other cosmological probes, will be a a fundamental step forward in our understanding of dark matter, inflation, Dark Energy and neutrinos.
+Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe.
+They will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution expected in next generation, dubbed Stage IV (S4)~\cite{2013arXiv1309.5383A}.
+The recent past is  reassuring.
+Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
+These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck.
+With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
+The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
 
-Neutrinos properties can be effectively constrained with CMB measurements because they influence both the early history of the universe and the later stage of structure formation. 
-In the early Universe relativistic species, like neutrinos, are the main drivers of the cosmic expansion and the expansion rate depends on their effective number $N_{\rm eff}$.
-This can be powerfully tested using the CMB, by comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
-S4 experiments, with their high resolution, will be able to measure the CMB power spectrum, at small scales, deeply into the damping tale, testing  possible deviations from $N_{\rm eff}=3.046$, the value predicted by the standard model.
-The total mass of neutrinos, on the other hand, has a modest effect on the primary CMB because, for the range of masses allowed by recent constraints ($\sum m_\nu<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
-However massive neutrinos influence the growth of structure at scales shorter than their free streaming one.
-Different neutrino masses consequently lead to different CMB lensing.
-Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction.
-This will turn into such a precise constraints of the sum of neutrino masses to get a possible hint of the hierarchical structure of the individual masses.
-Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $45\%$ with a factor of two improvement on $\sum m_\nu$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
+This pattern of successes is far from over.
+The next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy and neutrinos.
+
+The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $M_{\nu}$ affects the late-time growth of cosmic structures.
+Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$.
+This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
+The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$.
+The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
+However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons.
+Different neutrino masses consequently lead to different CMB lensing spectra.
+A Stage IV experiment will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction.
+This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
+These signatures can be mimicked with variations in other cosmic parameters, so an important lingering question is: What external information can be used to improve projected constraints further by breaking degeneracies?
+%Quantitatively, the CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
 The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
-As previously mentioned, CMB is also a powerful probe of the dark energy properties (see \cite{2010MNRAS.405.2639J}).
-The challenge for future experiments is to go beyond the measurement of the energy density and to determine other dark energy properties.
-Among these, a crucial step would be to constraint the equation of state, the ratio of pressure and energy density, looking for possible deviations from the value $w=-1$ expected in a cosmological constant scenario.
-CMB is sensitive to Dark energy because this component alters the late universe's expansion and it consequently changes the distance to the last scattering surface.
-Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing.
-However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$.
-As a consequence, similarly to the neutrino sector, CMB experiments will have to rely on external prior coming from BAO and supernova experiments to fully exploit their potential.
+The CMB is sensitive also to the properties of dark energy (see \cite{2010MNRAS.405.2639J}).
+Thanks to the current generation of experiments we know with extraordinary accuracy its energy density.
+The challenge for the next generation is to reveal the nature of this mysterious component.
+For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant.
+Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface.
+Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing.
+Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, the CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
+However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
+will be improved by external experiments.
 
-The crucial importance of the feedback from external experiments is true not only for the dark energy and the neutrino sector. 
-In the future, cosmology will hinge on the ability of combining different probes.
-In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
-In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
-To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
-For all the parameters of the standard $\Lambda$CDM plus massive neutrinos ($\Omega_c h^2,\Omega_b h^2,A_s,n_s,\tau,H_0,\sum m_\nu$) together with possible non-standard extension ($N_{\rm eff},w$) we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.
+The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector.
+A general thrust in even the current generation of surveys is to combine information from different observations.
+%In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
+%In particular CMB will benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
+Here we study the dependance of projected constraints from CMB on the external priors assumed.
+For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$ ) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors.
+This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.
 
-This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we used to derive the effect of external priors on the CMB parameter constraints.
+This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
 In \refsec{results} we will describe our results and we then conclude with a discussion of them in \refsec{conclusions}.
@@
 To measure quantitatively the impact of external priors on the CMB ability to constrain cosmological parameters we use a Fisher matrix formalism.
-In this section we will quickly review the technique and we will then present the chosen fiducial cosmological model together with the experimental specifications.
+In this section we will quickly review the technique and then present the chosen fiducial cosmological model together with the experimental specifications.
 
@@
 \begin{equation}
- F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
+ F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right)
  \label{eqn:Fij_def2}
  \end{equation}
+ where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout.
 In this work we constrain cosmological parameters with CMB temperature and E mode polarization together with the reconstructed lensing potential of large scale structure.
-For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
+Therefore, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
  \begin{eqnarray}
@@
 \end{eqnarray}
-The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be described at the end of this section.
-Note that we are neglecting the term $C_\ell^{E\phi}$. As also noticed in  \cite{wu:2014,2013PhRvD..87h3008H} this term contains very little information while adding possible numerical issues.
-Furthermore, as in \cite{wu:2014}, we decided to use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}.
-Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
+The terms $N_\ell^{X}$ represent the instrumental noise power of the specific experiment and will be described at the end of this section.
+Note that we are neglecting the term $C_\ell^{E\phi}$ since it  contains very little information while adding possible numerical issues~\cite{wu:2014,2013PhRvD..87h3008H}.
+Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}.
+\sd{Can we say something more positive about these effects?
+Why are we allowing ourselves to make this approximation?} Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
 
-A lower bound on the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, can be obtained as:
+The projected error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, is then:
 \begin{equation}
@@
 \end{equation}
-Furthermore we can introduce external priors on cosmological parameters.
+We can introduce external priors on cosmological parameters.
 Indeed we simply need to add to the Fisher matrix elements the external priors (before we perform the matrix inversion of \refeq{cramer-rao}).
@@
 %Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard procedure used to forecast future constrains \cite{wu:2014}.
-We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
-We compute the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
+%We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
+We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB and
+the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
 %\begin{equation}
@@
 %\end{equation}
-this high order approach allows us to use bigger steps around the fiducial parameters to compute derivatives.
-As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing $\boldsymbol{C}_\ell$ negligible.
+this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives.
+As a consequence the differences of power spectra corresponding to different values of the parameters are big enough to ensure numerical accuracy.
+% issues in computing $\boldsymbol{C}_\ell$.
 We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$.
-The constraints change by at most $10\%$ that should then be considered as a conservative numerical uncertainties.
+The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties.
 
-We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB.
-Our base cosmology is a flat $\nu \Lambda$CDM. 
-We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters.
-We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
-$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc and a neutrino total mass $\sum m_\nu$ $\simeq$ 85\ meV.
-Regarding the $\Lambda$CDM extensions, we chose as fiducial, a standard $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1,w_{a}=0$.
-It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
+Our philosophy is to consider as few extensions as possible.
+Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models.
+As such, the key question is: What information is needed to reliably conclude that an additional parameter is required.
+The natural first baby step is to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist.
+So our fiducial cosmology is flat $\nu \Lambda$CDM, with assumed
+parameters from Table 2 of the \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
+$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc, supplemented by an arbitrary choice of $\sum m_\nu$ $\simeq$ 85\ meV.
+We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters, in each case keeping the other parameter fixed.
+The fiducial values of these are $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1$.
+%It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
 
-The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the optimistic level for the next generation of CMB experiments (S4).
+The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the optimistic level for the next generation of CMB experiments (S4) assumed in ~\cite{2013arXiv1309.5383A,wu:2014,2013PhRvD..87h3008H}.
 For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
@@
 where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
-We decide to use a level of noise $s = 0.58$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin. This correspond to the $N_{\rm det}=10^{6}$ case of \cite{wu:2014}. 
+We use a level of noise $s = 0.58$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin. This corresponds to the $N_{\rm det}=10^{6}$ case of \cite{wu:2014}. 
 Note that the quoted noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
-
-The noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.
+The noise $N_\ell^{\phi\phi}$ associated with the reconstructed $\phi$ spectrum is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.
 
@@
 Indeed, below the current free-streaming scale, the matter power power spectrum is suppressed in the presence of
-massive neutrinos by a factor -8$f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
+massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
 This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses $\sum m_\nu$.
-This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.
+This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>50$ meV that emerges from oscillation experiments.
 
@@
 \caption{
-In the figure, the constraints on the sum of the neutrino masses as a function of priors on cosmological parameters are shown ($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ).
-Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm CMB}=0.6\%$.
-$H_{0}$ and $\Omega_{c}h^{2}$ will be the more useful external prior to improve neutrino masses constraint in this regime.
-We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-like prior.
+Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. %($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ). 
+Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
+$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints.
+We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-like prior\sd{What does Planck prior mean here?}.
 The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
@@
 
-\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment ($N_{det}=10^{6}$, $f_{sky}=0.75$,  $2<\ell<5000$) as a function of external priors.
-Without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
+\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment\sd{What does Planck prior mean?} %($N_{det}=10^{6}$, $f_{sky}=0.75$,  $2<\ell<5000$) 
+as a function of external priors.
+Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
 Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
@@
 Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
-The improvement given by a prior on $\Omega_ch^2$ is a consequence of the fact that the CMB lensing is sensitive to $f_\nu$, therefore in order to constrain $\Omega_{\nu}h^2$ from this ratio we need to determine the matter energy density.
-A prior on  $H_{0}$ would help as well since the effects of external priors are not completely independent.
-The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^3$ and for this reason an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
+The improvement given by a prior on $\Omega_ch^2$ is also a consequence of the fact that the CMB lensing is sensitive to $f_\nu$; therefore in order to constrain $\Omega_{\nu}h^2$ from this ratio we need to determine the matter energy density.
+A prior on  $H_{0}$ helps with this as well since the effects of external priors are not completely independent.
+The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^3$\sd{get the right power from the Planck paper and cite that} and for this reason too an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
 As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors.
 
-In \reffig{prior_omeganuh2_tau} we limit CMB accessible modes to $\ell_{\rm min}>50$. 
-Together with the importance of $H_{0}$ and $\Omega_ch^2$, that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will also be helpful.
-Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follow.
+It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in \reffig{prior_omeganuh2_tau} we limit the modes to $\ell_{\rm min}>50$. 
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will also be helpful.
+Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
 The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig.5 of \cite{2013arXiv1309.5383A}).
 For this reason it is hard to measure the unsuppressed lensing CMB power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$.
-The unsuppressed amplitude is actually better constrained by the primordial amplitude $A_{s}$.
-A measurement of the optical depth will help to constraint the primordial amplitude $A_{s}$ internally since the CMB is sensitive to the combination $A_{s}\exp^{-2\tau}$.
+The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB.
+Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so
+a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$.
 If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} will be needed to fully exploit the CMB constraining power.
@@
 Furthermore any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model.
+As pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above ~100 GeV.
+%decoupled early in the thermal history.
+
 %==========
@@
 
-\reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other six parameters.
+\reffig{prior_massless_neutrinos} shows projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on the other seven parameters (the sum of the neutrino masses must be included as a free parameter).
 With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$.
@@
 As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction.
-Also as noticed in \cite{baumann:2015} tight constraints on $N_{\rm eff}$ will allow to rule out early thermalized particles that decoupled early in the thermal history.
 
@@
 
-The CMB constrained the late-time dark energy equation of state in two ways. The primordial CMB is very sensitive to the distance to the last scattering surface which depends on $w$. 
-Furthermore the CMB lensing probes the growth of structure at late times which is different for different dark matter equations of state.
+The CMB constrains the late-time dark energy equation of state in two ways. The observed CMB spectra are very sensitive to the distance to the last scattering surface which depends on $w$. 
+Furthermore CMB lensing probes the growth of structure at late times which is different for different dark energy equations of state.
 %==========

### 2015-11-18 Notes/N_eff_prior_paper_v2.tex
add omch2 right scaling in CMB, specified Planck prior (the same of what Dunkley used) and comment on Scott's comment

@@
 \newcommand{\sd}[1]{\textcolor{red}{[{\bf SD}: #1]}}
+\newcommand{\am}[1]{\textcolor{blue}{[{\bf AM}: #1]}}
 
@@
 With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
-The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
+The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} \am{it is ok for me to skip the list entirely otherwise feel free to add names I can add references later} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
 
@@
 Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
+\am{My understanding is that people do that: unlensed to avoid double counting in the informations with $C^{\phi}$, Aurelien and Wayne showed that unlensed + gaussian is similar to lensed + non-gaussian how big is the effect i do not know.
+Aurelien is writing a paper about that}
 \sd{Can we say something more positive about these effects?
@@
 $H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints.
-We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-like prior\sd{What does Planck prior mean here?}.
+We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-pol prior .
 The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
@@
 
-\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment\sd{What does Planck prior mean?} %($N_{det}=10^{6}$, $f_{sky}=0.75$,  $2<\ell<5000$) 
-as a function of external priors.
+\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment as a function of external priors. 
+Our Planck prior corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
 Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
@@
 A prior on  $H_{0}$ helps with this as well since the effects of external priors are not completely independent.
-The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^3$\sd{get the right power from the Planck paper and cite that} and for this reason too an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
+The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} and for this reason too an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
 As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors.
@@
 \begin{acknowledgments}
-AM wants to thank Zhen Pan who allowed a careful cross-check of our results.
+We thank Wayne Hu for useful discussions. AM wants to thank Zhen Pan who allowed a careful cross-check of our results.
 This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.

### 2015-11-24 n_priors_code/fisher_TT_test2.py
new images

@@ -1,309 +1,0 @@
-#!/usr/bin/env python
-
-
-'''
-
-SECOND VERSION TO SUBSTITUTE THE first one (AM)
-
-Simple fisher code to test prior effect on N effective neutrino estimates
-
-TODO:
-HIGH: lensing noise (almost there: Komatsu and quicklens.)
-
-low importance: ini file. Think about PCA to understand what is more important for N_eff
-
-
-create classes or at least external utilities. Maybe a better way to interface with cosmosis
-
-# DATA FROM CAMB
-
-we use the _lenspotentialcls so
-
- l CTT CEE CBB CTE Cdd CdT CdE
- 0  1  2    3   4   5   6  7
-
-  CX are l(l+1)Cl/2pi and Cdd=[l(l+1)]^2 Clphi/2pi, CdT=[l(l+1)]^(3/2) ClphiT/2pi, CdE=[l(l+1)]^(3/2)ClphiE/2pi
-
-
-The matrix is going to run an 4 parameters:
-
-CONVENTIONS:
-alphabetical in CAMB description. Use the ordered dict tools to explore it.
-hubble,massless_neutrinos,omnuh2,re_optical_depth,scalar_amp(1),scalar_spectral_index(1)
-
-
-GOAL
-
-1) reproduce Wu paper and beyond/ DONE. Cross checked with Zhen Pan UC Davis
-
-'''
-__author__ = "A.Manzotti"
-__license__ = "GPL"
-__version__ = "2.0"
-__maintainer__ = "A.Manzotti"
-__email__ = "manzotti.alessandro@gmail.com"
-__status__ = "Production"
-
-import numpy as np
-import math
-import matplotlib.pyplot as plt
-import utils
-import pickle
-import sys
-
-# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
-
-
-def years2sec(years):
-    ''' years to sec '''
-    return years * 365 * 24. * 60. * 60.
-
-
-def fsky2arcmin(fsky):
-    '''convert fsky in fraction of unity to arcmin^2'''
-    # 41253 square degrees in all sky
-    return 41253. * fsky * 60. * 60.
-
-
-def calc_deriv_vectorial(fisher_index, dats, pargaps, values):
-    '''
-    Given CMB data dats it normalize them anf create a 3x3 matrix
-
-    ell is the multiple
-    iell is the index in the data ell corresponds to
-
-    remember the order from CAMB
-     l CTT CEE CBB CTE Cdd CdT CdE
[…]

### 2015-11-24 n_priors_code/fisher_joint_test.py

@@ -51,6 +51,7 @@ import utils
 import pickle
 import sys
 import collections
+import statsmodels.api as sm
 
 # util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
 
@@ -68,15 +69,16 @@ def fsky2arcmin(fsky):
 
 def calc_deriv_vectorial(fisher_index, dats, pargaps, values, order=5):
     '''
-    Given CMB data dats it normalize them anf create a 3x3 matrix
-
-    ell is the multiple
-    iell is the index in the data ell corresponds to
-
-    remember the order from CAMB
-     l CTT CEE CBB CTE Cdd CdT CdE
+    Vectorize derivative computation. Each -calc_c_general gives you the matrix C for all ell and here you
+    compute the derivative in a vectorize fashion
     '''
-    if order == 5 and np.shape(values[values.keys()[0]])[0] == 4:
+
+    if order == 3 and np.shape(values[values.keys()[0]])[0] == 2:
+
+        ci = (-calc_c_general(dats, fisher_index * 2 + 2) - calc_c_general(dats,
+                                                                           fisher_index * 2 + 1)) / (2. * pargaps[values.keys()[fisher_index]])
+
+    elif order == 5 and np.shape(values[values.keys()[0]])[0] == 4:
         ci = (-calc_c_general(dats, fisher_index * 4 + 4) + 8. * calc_c_general(dats, fisher_index * 4 + 3) - 8. *
               calc_c_general(dats, fisher_index * 4 + 2) + calc_c_general(dats, fisher_index * 4 + 1)) / (12. * pargaps[values.keys()[fisher_index]])
 
@@ -88,11 +90,8 @@ def calc_deriv_vectorial(fisher_index, dats, pargaps, values, order=5):
         ci = (-3. * calc_c_general(dats, fisher_index * 8 + 8) + 32. * calc_c_general(dats, fisher_index * 8 + 7) - 168. *
               calc_c_general(dats, fisher_index * 8 + 6) + 672. * calc_c_general(dats, fisher_index * 8 + 5) - 672. * calc_c_general(dats, fisher_index * 8 + 4) + 168. * calc_c_general(dats, fisher_index * 8 + 3) - 32. * calc_c_general(dats, fisher_index * 8 + 2) + 3. * calc_c_general(dats, fisher_index * 8 + 1)) / (840. * pargaps[values.keys()[fisher_index]])
     else:
[…]

### 2015-11-24 n_priors_code/fisher_produce_BAO.py

@@ -3,39 +3,7 @@
 
 
 '''
-
-FAST VERSION
-Fisher code to test prior effect on N effective neutrino estimates
-
-TODO:
-HIGH: lensing noise (almost there: Komatsu and quicklens.)
-
-low importance: ini file. Think about PCA to understand what is more important for N_eff
-
-
-create classes or at least external utilities. Maybe a better way to interface with cosmosis
-
-# DATA FROM CAMB
-
-we use the _lenspotentialcls so
-
- l CTT CEE CBB CTE Cdd CdT CdE
- 0  1  2    3   4   5   6  7
-
-  CX are l(l+1)Cl/2pi and Cdd=[l(l+1)]^2 Clphi/2pi, CdT=[l(l+1)]^(3/2) ClphiT/2pi, CdE=[l(l+1)]^(3/2)ClphiE/2pi
-
-
-The matrix is going to run an 4 parameters:
-
-CONVENTIONS:
-alphabetical in CAMB description. Use the ordered dict tools to explore it.
-hubble,massless_neutrinos,omnuh2,re_optical_depth,scalar_amp(1),scalar_spectral_index(1)
-
-
-GOAL
-
-1) reproduce Wu paper and beyond/ DONE. Cross checked with Zhen Pan UC Davis
-
+produce Fisher matrix for BAO experiments
 '''
 __author__ = "A.Manzotti"
 __license__ = "GPL"
@@ -52,231 +20,155 @@ import pickle
 import sys
 import collections
 from scipy.integrate import quad
+from astropy.constants import hbar, k_B, G, c
+import NuDensity as nd
 
 
-# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
+# # ROUTINES TO COMPUTE WHAT BAO CONSTRAIN, rs/DV(z)
+
+# # following EISENSTEIN & HU
 
 
-def years2sec(years):
-    ''' years to sec '''
-    return years * 365 * 24. * 60. * 60.
+# def nu_plus_r_density():
+#     factor = (8.* np.pi**3 * k_B**4 * G ) / (45. * hbar**3 * c**5 *h*100)
+
+#  from http://arxiv.org/pdf/1411.1074v3.pdf
+# def integral_I(z,omnuh2,T_CMB):
+#     T_nu = T_CMB *(4./11.)**(1/3.) * (3.046/3)**(1/4.)
+#     m_nu =
+#     r_variable= (m_nu  * c**2 /k_B/T_nu)
[…]

### 2015-11-24 n_priors_code/fisher_produce_planckpol.py

@@ -118,16 +118,31 @@ def calc_c_fiducial(data):
     ell = data[lmin_index:lmax_index, 0, 0]
     # s = 350. * np.sqrt(arcmin_from_fsky) / np.sqrt(N_det * Y * sec_of_obs)  # half sky in arcmin^2
     s = {}
+    s_pol = {}
     t = {}
     N = {}
+    N_pol = {}
     fac = (ell * (ell + 1.) / 2. / np.pi) / (7.4311 * 10 ** 12)
     N_tot = np.zeros_like(ell)
-    s['143'] = 43 # this is supposed to be in muk arcmin to fit in the later used formulas
-    s['217'] = 66
-    t['217'] = 5. / 60. / 180. * np.pi  # 2arcmin to rads beam
-    t['143'] = 7. / 60. / 180. * np.pi  # 2arcmin to rads beam
+    N_pol_tot = np.zeros_like(ell)
 
-    for nu in ['143', '217']:
+    # 30,44,70,100,143,217,353
+    s['30'] = 145  # this is supposed to be in muk arcmin to fit in the later used formulas
+    s['44'] = 149
+    s['70'] = 137
+    s['100'] = 65
+    s['143'] = 43
+    s['217'] = 66
+    s['353'] = 200
+
+    t['30'] = 33. / 60. / 180. * np.pi  # 2arcmin to rads beam
+    t['44'] = 23. / 60. / 180. * np.pi  # 2arcmin to rads beam
+    t['70'] = 14. / 60. / 180. * np.pi  # 2arcmin to rads beam
+    t['100'] = 10. / 60. / 180. * np.pi  # 2arcmin to rads beam
+    t['143'] = 7. / 60. / 180. * np.pi  # 2arcmin to rads beam
+    t['217'] = 5. / 60. / 180. * np.pi  # 2arcmin to rads beam
+    t['353'] = 5. / 60. / 180. * np.pi  # 2arcmin to rads bea
+    for nu in ['30', '44', '70', '100', '143', '217', '353']:
 
         # Final CMB noise definition
         N[nu] = (s[nu] * np.pi / 180. / 60.) ** 2 * np.exp(ell * (ell + 1.) * t[nu] ** 2 / 8. / np.log(2))
@@ -135,7 +150,26 @@ def calc_c_fiducial(data):
         N_tot += 1 / N[nu]
 
     N_tot = 1. / N_tot
-    np.save('noise.npy',N_tot*fac)
+
+    # 30,44,70,100,143,217,353
[…]

### 2015-11-24 n_priors_code/generate_data.py

@@ -37,7 +37,7 @@ import os
 import warnings
 
 output_folder = 'varying_all'
-output_folder_2 = 'run2'
+output_folder_2 = 'run7'
 camb_location = '/home/manzotti/local/camb2013/camb'
 
 config = configparser.ConfigParser()
@@ -70,16 +70,16 @@ omch2 = config.getfloat('camb', 'omch2')
 fid = {}
 fid['hubble'] = config.getfloat('camb', 'hubble') / 100.
 fid['scalar_spectral_index(1)'] = config.getfloat('camb', 'scalar_spectral_index(1)')
-fid['helium_fraction'] = config.getfloat('camb', 'helium_fraction')
+# fid['helium_fraction'] = config.getfloat('camb', 'helium_fraction')
 fid['scalar_amp(1)'] = 10 ** 9 * config.getfloat('camb', 'scalar_amp(1)')
 fid['massless_neutrinos'] = config.getfloat('camb', 'massless_neutrinos') + 1.
 fid['re_optical_depth'] = config.getfloat('camb', 're_optical_depth')
 fid['w'] = config.getfloat('camb', 'w')  # DE W parameters
-fid['wa'] = config.getfloat('camb', 'wa')  # DE W parameters
+# fid['wa'] = config.getfloat('camb', 'wa')  # DE W parameters
 fid['ombh2'] = config.getfloat('camb', 'ombh2')
 fid['omch2'] = config.getfloat('camb', 'omch2')
-fid['omk'] = config.getfloat('camb', 'omk')
-fid['scalar_nrun(1)'] = config.getfloat('camb', 'scalar_nrun(1)')
+# fid['omk'] = config.getfloat('camb', 'omk')
+# fid['scalar_nrun(1)'] = config.getfloat('camb', 'scalar_nrun(1)')
 fid['omnuh2'] = config.getfloat('camb', 'omnuh2')
 fid = collections.OrderedDict(sorted(fid.items(), key=lambda t: t[0]))
 
@@ -95,26 +95,26 @@ with open("./data/{}/{}/fid_values.p".format(output_folder, output_folder_2), "w
 # ================================================
 
 pargaps_dict = {}
-pargaps_dict['hubble'] = fid['hubble'] * 0.02
-pargaps_dict['helium_fraction'] = fid['helium_fraction'] * 0.02
+pargaps_dict['hubble'] = fid['hubble'] * 0.05
[…]

### 2015-11-24 n_priors_code/plotting/plot_one_ellipse_different_probe.py

@@ -210,7 +210,7 @@ for i,probes in enumerate(['CMB', 'joint']):
 
 
     # parameters =['mnu','ombch2']
-    parameters = ['omnuh2', 'hubble']
+    parameters = ['omnuh2', 'omch2']
 
 
     sigmax_squared = my_fisher_inv[fid.keys().index(parameters[0]), fid.keys().index(parameters[0])]
@@ -235,16 +235,17 @@ for i,probes in enumerate(['CMB', 'joint']):
     # two_sigma = Ellipse((fid[key_j], fid[key_i]), np.sqrt(a_fisher)*2.48, np.sqrt(b_fisher)*2.48,
     #              angle=angle_fisher, linewidth=0.5, fill=False)
     ax1.add_patch(one_sigma)
+    one_sigma.set(clip_box=ax1.bbox,
+          label=probes)
     # plt.axhline(fid[parameters[1]] - 1.52 * np.sqrt(sigmay_squared), linewidth=1)
     # plt.axhline(fid[parameters[1]] + 1.52 * np.sqrt(sigmay_squared), linewidth=1)
 
-    xl = fid[parameters[0]] - 2.52 * np.sqrt(sigmax_squared)
-    xu = fid[parameters[0]] + 2.52 * np.sqrt(sigmax_squared)
-    yl = fid[parameters[1]] - 2.52 * np.sqrt(sigmay_squared)
-    yu = fid[parameters[1]] + 2.52 * np.sqrt(sigmay_squared)
+    xl = fid[parameters[0]] - 0.8 * fid[parameters[0]]
+    xu = fid[parameters[0]] + 0.8 * fid[parameters[0]]
+    yl = fid[parameters[1]]  - 0.03 * fid[parameters[1]]
+    yu = fid[parameters[1]]  + 0.03 * fid[parameters[1]]
     ax1.set_ylim(yl, yu)
     ax1.set_xlim(xl, xu)
-    ax1.plot(omnuh2_array, H_array,'-.' ,alpha=0.6 , linewidth=font_size / 10.)
 
     fg.tight_layout()
 
@@ -263,8 +264,9 @@ for i,probes in enumerate(['CMB', 'joint']):
     # ============================================
     # plt.savefig('../../images/trinagle.pdf', dpi=400, papertype='Letter',
     #             format='pdf', transparent=True)
[…]

### 2015-11-24 n_priors_code/plotting/plot_one_ellipse_paper_w_H0.py

@@ -46,7 +46,7 @@ from scipy.integrate import quad
 # READ DATA
 
 
-no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
+no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
 plot_now = ['omnuh2','w']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
@@ -55,7 +55,7 @@ excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # DEFINE YOUR FOLDER HERE
 base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
-run_idx = 2
+run_idx = 7
 lmax = 4499
 lmin = 4
 N_det = 10 ** 6

### 2015-11-24 n_priors_code/plotting/plot_prior_all.py

@@ -36,7 +36,7 @@ from palettable.colorbrewer.qualitative import Set1_9
 # ============================================
 
 no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
-plot_now = ['omnuh2']
+plot_now = ['omnuh2','w']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
 
@@ -247,7 +247,7 @@ for y, key_y in enumerate(plot_param):
     # ============================================
 
     # ============================================
-    plt.savefig(base_dir + 'data/{}/run{}/output/prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}.png'.format(data_type, str(run_idx), str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+    plt.savefig(base_dir + 'data/{}/run{}/output/prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}_planck_w.png'.format(data_type, str(run_idx), str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                 format='png', bbox_inches='tight')
     plt.clf()
 

### 2015-11-24 n_priors_code/plotting/plot_prior_paper_M_nu_ell4.py

@@ -35,7 +35,7 @@ from palettable.colorbrewer.qualitative import Set1_9
 # ============================================
 # ============================================
 
-no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
+no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
 plot_now = ['omnuh2']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
@@ -44,9 +44,9 @@ excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # DEFINE YOUR FOLDER HERE
 base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
-run_idx = 4
+run_idx = 7
 lmax = 4499
-lmin = 4
+lmin = 50
 N_det = 10 ** 6
 # N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
 fsky = 0.75
@@ -230,7 +230,7 @@ for key_y in ['omnuh2']:
     ax1.set_xlim((0.1 , 3.1))
     # ax1.set_title(r'$\sigma({0})={1:.1f}\%$'.format(str(label[key_y]), np.abs(sigma_just_CMB_y / fid[key_y] * 100.)))
     ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
-    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck}$')
+    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm \rm S4~ +~ Planck ~ Pol + ~BAO15}$')
 
     y1, y2 = ax1.get_ylim()
     ax2 = ax1.twinx()

### 2015-11-24 n_priors_code/plotting/plot_prior_paper_M_nu_ell50.py

@@ -35,7 +35,7 @@ from palettable.colorbrewer.qualitative import Set1_9
 # ============================================
 # ============================================
 
-no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
+no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
 plot_now = ['omnuh2']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
@@ -44,7 +44,7 @@ excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # DEFINE YOUR FOLDER HERE
 base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
-run_idx = 4
+run_idx = 7
 lmax = 4499
 lmin = 50
 N_det = 10 ** 6
@@ -230,7 +230,7 @@ for key_y in ['omnuh2']:
     ax1.set_xlim((0.1, 3.1))
     # ax1.set_title(r'$\sigma({0})={1:.1f}\%$'.format(str(label[key_y]), np.abs(sigma_just_CMB_y / fid[key_y] * 100.)))
     ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
-    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck }$')
+    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck ~ Pol + ~BAO15}$')
     y1, y2 = ax1.get_ylim()
     ax2 = ax1.twinx()
     minor_loc = ax1.yaxis.get_minor_locator()

### 2015-11-24 n_priors_code/plotting/plot_prior_paper_N_eff.py

@@ -35,7 +35,7 @@ from palettable.colorbrewer.qualitative import Set1_9
 # ============================================
 # ============================================
 
-no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2', 'helium_fraction','wa','omk','scalar_nrun(1)']
+no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
 plot_now = ['omnuh2','massless_neutrinos']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
@@ -44,7 +44,7 @@ excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # DEFINE YOUR FOLDER HERE
 base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
-run_idx = 2
+run_idx = 7
 lmax = 4499
 lmin = 4
 N_det = 10 ** 6
@@ -230,7 +230,7 @@ for  key_y in ['massless_neutrinos']:
     ax1.set_xlim((0.1, 3.1))
     # ax1.set_title(r'$\sigma({0})={1:.1f}\%$'.format(str(label[key_y]), np.abs(sigma_just_CMB_y / fid[key_y] * 100.)))
     ax1.set_ylabel(r'$\sigma(' + label[key_y] + r')$')
-    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck }$')
+    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck ~ Pol + ~BAO15 }$')
     y1, y2 = ax1.get_ylim()
     ax2 = ax1.twinx()
     minor_loc = ax1.yaxis.get_minor_locator()

### 2015-11-24 n_priors_code/plotting/plot_triangle_ellipses_plot_2.py

@@ -59,7 +59,7 @@ base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
 run_idx = 4
 lmax = 4499
-lmin = 50
+lmin = 4
 N_det = 10 ** 6
 # N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
 fsky = 0.75
@@ -271,5 +271,5 @@ plt.rcParams['legend.handletextpad'] = 0.3
 # ============================================
 # plt.savefig('../../images/trinagle.pdf', dpi=400, papertype='Letter',
 #             format='pdf', transparent=True)
-plt.savefig('triangle_all_prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str('joint_real'), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',format='pdf', transparent=True)
+plt.savefig('triangle_all_prior_{}_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}_no_planck.pdf'.format(str('joint_real'), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',format='pdf', transparent=True)
 plt.close()

### 2015-11-24 Notes/N_eff_prior_paper_v2_ypedit.tex

--- 2015-11-18 Notes/N_eff_prior_paper_v2.tex
@@
+	
 % Group addresses by affiliation; use superscriptaddress for long
@@
 \newcommand{\am}[1]{\textcolor{blue}{[{\bf AM}: #1]}}
+\newcommand{\yp}[1]{\textcolor{green}{[{\bf YP}: #1]}}
 
@@
 With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
-The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} \am{it is ok for me to skip the list entirely otherwise feel free to add names I can add references later} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
+The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} \am{it is ok for me to skip the list entirely otherwise feel free to add names I can add references later} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tail of the CMB spectrum with increasing accuracy.
 
@@
 The next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy and neutrinos.
+As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy, and neutrinos.
 
@@
 Here we study the dependance of projected constraints from CMB on the external priors assumed.
-For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$ ) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors.
+For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors.
 This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.
@@
 
-For CMB experiments the Fisher matrix can be related to the power spectrum $\boldsymbol{C}_\ell$ by:
+For CMB experiments the Fisher matrix can be related to the power spectrum $\boldsymbol{C}_\ell$ by \yp{I believe if you refer to $\boldsymbol{C}_\ell$ as the power spectrum, the trace in the following equation must be removed; the sum over ell is the trace taking.
+This equation seems to be in line with Eq.
+3 though, so perhaps you can just call $\boldsymbol{C}_\ell$ the covariance matrix?} :
 \begin{equation}
@@
  where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout.
-In this work we constrain cosmological parameters with CMB temperature and E mode polarization together with the reconstructed lensing potential of large scale structure.
+In this work we constrain cosmological parameters with CMB temperature and E-mode polarization together with the reconstructed lensing potential of large scale structure.
 Therefore, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
@@
 %We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
-We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB and
+We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \yp{cite CAMB?} and
 the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
@@
 The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties.
+\yp{Can the number 10\% be justified a bit more here?}
 
@@
 For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
-This allow us to use all the polarization power spectrum multipoles all the way up to $\ell_{\rm E,max}=5000$.
-We deal with the poisson noise from point sources in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
-The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
+This allows us to use all the polarization power spectrum multipoles all the way up to $\ell_{\rm E,max}=5000$.
+We deal with the Poisson noise from point sources in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
+The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way \yp{cite?}:
  \begin{equation}
@@
 Note that the quoted noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
+\yp{Sentence here to explain the factor of $\sqrt{2}$ would be helpful.}
 The noise $N_\ell^{\phi\phi}$ associated with the reconstructed $\phi$ spectrum is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.
@@
 \section{Results \label{sec:results}}
-In this section we present our main results. We will start from our base $\nu \Lambda$CDM and we then explore extensions to this model ($N_{\rm eff},w$).
+In this section we present our main results. We will start from our base $\nu \Lambda$CDM, and then will explore extensions to this model by considering ($N_{\rm eff},w$).
 
@@
 massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
-This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses $\sum m_\nu$.
+This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the the sum of the neutrino masses $\sum m_\nu$.
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>50$ meV that emerges from oscillation experiments.
-
+\yp{Consider changing $H_0 = 0.6\%$ to $\sigma(H_0) = 0.6\%$ in the legends to match the caption.}
 %In CMB experiment, neutrino masses constraint comes almost entirely from lensing.
@@
 \includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Some as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
+\caption{Same as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
 \label{fig:prior_omeganuh2_tau}
@@
 Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
-There, the prior was described as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
-Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
+There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
+In this study, we find the same improvement by adding a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
 The improvement given by a prior on $\Omega_ch^2$ is also a consequence of the fact that the CMB lensing is sensitive to $f_\nu$; therefore in order to constrain $\Omega_{\nu}h^2$ from this ratio we need to determine the matter energy density.
@@
 As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors.
+\yp{Alessandro, are you planning on including the $\Omega_b h^2$ discussion here?}
 
 It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in \reffig{prior_omeganuh2_tau} we limit the modes to $\ell_{\rm min}>50$. 
-Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we also observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ will also be helpful.
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we previously observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful.
 Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
-The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig.5 of \cite{2013arXiv1309.5383A}).
-For this reason it is hard to measure the unsuppressed lensing CMB power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$.
+The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig.
+5 of \cite{2013arXiv1309.5383A}).
+For this reason it is hard to measure the unsuppressed CMB lensing power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$.
 The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB.
@@
 
-In the standard cosmology, three active neutrinos are thermally produced in the early universe.
+In the standard model of cosmology, three active neutrinos are thermally produced in the early universe.
 Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation.
-However decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$.
-The additional fraction depends not only on well-known neutrino scattering rates but also finite temperature quantum corrections.
-Upcoming experiment have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
-Furthermore any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model.
-As pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above ~100 GeV.
+However, decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$.
+The additional fraction depends not only on well-known neutrino scattering rates but also on finite temperature quantum corrections.
+Upcoming experiments have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
+Furthermore, any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model.
+As pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above $\sim$100 GeV.
 %decoupled early in the thermal history.
@@
 \reffig{prior_massless_neutrinos} shows projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on the other seven parameters (the sum of the neutrino masses must be included as a free parameter).
-With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$.
+With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the expected deviation from $N_{\rm eff}=3$.
 The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
+\yp{The last sentence sounds a bit vague.}
 
@@
 The possible reward is remarkable.
-We can finally put a tight constraint on the total mass of neutrinos with deeply implications on particle physics like, for example, the solution of the mass hierarchy problem.
-We will measure with great accuracy the number of relativistic species, both testing standard model prediction and, eventually, probing new physics.
-The same is true for dark energy.
-In that context, we will tighten the constraint on $w$,$w_{a}$ getting closer to test the nature of the accelerated expansion of the universe.
+We can finally put a tight constraint on the total mass of neutrinos with deep implications on particle physics like, for example, the solution of the mass hierarchy problem.
+We can also measure with great accuracy the number of relativistic species, thereby testing the standard model prediction and possibly probing new physics.
+The same is true for dark energy: we can tighten the constraint on $w$,~$w_{a}$ and test the nature of the accelerated expansion of the universe.
+\yp{$w_a$ was never mentioned in this paper; perhaps it is better not to discuss it entirely.}
+
 These results, however, will not be possible without an optimal synergy with external experiments observing a variety of others cosmological probes.
-Indeed this should be taken into account, especially during the CMB Stage IV planning phase.
+Understanding the impact of such external information is crucial in determining the experimental specifications and requirements needed for CMB-S4, and ultimately in maximizing the scientific impact of the future CMB experiments.
+%Indeed this should be taken into account, especially during the CMB Stage IV planning phase.
 Understanding CMB weaknesses and the impact of external data can have a deep influence in determining the needed experimental specifications.
-For this reason in this work we explored the impact of external priors on CMB S4 cosmological constraints and, in particular, on the neutrino sector and dark energy parameters.
+In this work we contribute to that effort by exploring the impact of external priors on CMB S4 cosmological constraints, focusing on the neutrino sector and dark energy parameters.
 We find:
 \begin{description}
-\item[$\sum m_\nu$] To improve the CMB constraint by a factor of two, low-redshift measurements will be needed.
-Indeed external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level CMB S4 will reach alone can bring the error on the total mass of neutrinos from $30$meV to below 20meV.
-Furterhemore, if the final design of CMB S4 will not include measurement at large scale ($\ell<50$) external constraint on the primordial amplitude will be needed.
-Moreover, an external prior on $\tau$ will allow the CMB to constraint $A_{s}$ internally thus strongly alleviating the problem.
-\item[$N_{\rm eff}$] The effective neutrino numbers will benefit from external prior on parameters that affect the shape of the damping tail like the spectral index and the baryons energy density.
+\item[$\sum m_\nu$] To improve the CMB constraint on sum of neutrino masses by a factor of two, low-redshift measurements will be needed.
+Indeed external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level CMB S4 will reach alone can bring the 1-$\sigma$ uncertainty on $\sum m_\nu$ from $30$meV to below 20meV.
+Furthermore, if the final design of CMB S4 will not include measurements on large scales ($\ell<50$), external constraints on the primordial amplitude $A_{s}$ will be needed.
+Moreover, an external prior on $\tau$ will allow the CMB to constraint $A_{s}$ internally, thus strongly alleviating this problem.
+\item[$N_{\rm eff}$] The constraint on the effective neutrino number will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$ and the baryon energy density $\Omega_b h^2$.
 \item[$w$] In this case the improvement will mainly come by external prior on $H_{0}$.
@@
 \end{description}
+\yp{Neff and w sections feel like they need a bit more flesh.}
 

### 2015-11-24 n_priors_code/fisher_joint_test.py
change images new labels

@@ -352,13 +352,14 @@ np.save('data/{}/full_fisher_mat_joint_lmin={}_lmax={}_ndet={}_fsky={}.npy'.form
 
 np.savetxt('data/{}/ell_indeces_joint_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(output_folder, lmin, lmax, N_det, fsky),
            dats[lmin_index:lmax_index, 0, 0], header=header)
+fisher_single = fisher.copy()
 
 # utils.study_prior_tau_on_N_eff(fid, fisher, 'data/' + output_folder, header)
-
+np.savetxt('data/{}/fisher_mat_joint_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(output_folder,
+                                                                                 lmin, lmax, N_det, fsky), fisher_single, header=header)
 
 print 'finally how much constraint on parameters without external prior?'
 print ''
-fisher_single = fisher.copy()
 
 fisher_inv = np.linalg.inv(fisher_single)
 
@@ -368,8 +369,6 @@ utils.save_cov_matrix(
 
 np.savetxt('data/{}/invetered_sqrt_fisher_joint_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(output_folder,
                                                                                             lmin, lmax, N_det, fsky), np.sqrt(fisher_inv), header=header)
-np.savetxt('data/{}/fisher_mat_joint_lmin={}_lmax={}_ndet={}_fsky={}.txt'.format(output_folder,
-                                                                                 lmin, lmax, N_det, fsky), fisher_single, header=header)
 
 print 'fisher=', fisher
 

### 2015-11-24 n_priors_code/plotting/plot_one_ellipse_paper_w_H0.py

@@ -235,7 +235,7 @@ print 'rho', np.sqrt(sigmay_squared) / fid[parameters[1]] * 100.
 
 a_fisher = (sigmax_squared + sigmay_squared) / 2. + np.sqrt((sigmax_squared - sigmay_squared) ** 2 / 4. + sigmaxy ** 2)
 b_fisher = (sigmax_squared + sigmay_squared) / 2. - np.sqrt((sigmax_squared - sigmay_squared) ** 2 / 4. + sigmaxy ** 2)
-angle_fisher = -np.degrees(np.arctan(2. * sigmaxy / (sigmax_squared - sigmay_squared))) / 2.
+angle_fisher = -np.degrees(np.arctan(-2. * sigmaxy / (sigmax_squared - sigmay_squared))) / 2.
 print a_fisher,b_fisher,angle_fisher
 if sigmax_squared > sigmay_squared:
 
@@ -262,7 +262,6 @@ yu = fid[parameters[1]] + 2.52 * np.sqrt(sigmay_squared)
 ax1.set_ylim(yl, yu)
 ax1.set_xlim(xl, xu)
 ax1.plot(w_array, H_array,'-.' ,alpha=0.6 , linewidth=font_size / 10.)
-# ax1.invert_xaxis()
 
 fg.tight_layout()
 
@@ -283,6 +282,6 @@ plt.rcParams['legend.handletextpad'] = 0.3
 #             format='pdf', transparent=True)
 plt.savefig('ellipse_H_M_nu1002.pdf', dpi=400, papertype='Letter',
             format='pdf', transparent=True)
-plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/ellipse_H0_w_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/ellipse_H0_w_snow_mass_lmin={}_lmax={}_ndet={}_fsky={}_NOBAO.pdf'.format(lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                 format='pdf', bbox_inches='tight')
 plt.close()

### 2015-11-24 n_priors_code/plotting/plot_prior_paper_M_nu_ell4.py

@@ -215,7 +215,7 @@ for key_y in ['omnuh2']:
         normalize_y = new_sigma / sigma_just_CMB_y  # make the new sigma y relative.
         # plot
         # *93.14 * 1000. to go to M_nu
-        line_plot = ax1.plot(normalize_x, new_sigma*94. * 1000., label=r'${0}={1:.1f}\%$'.format(
+        line_plot = ax1.plot(normalize_x, new_sigma*94. * 1000., label=r'$\sigma({0})={1:.1f}\%$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     new_sigma_all = utils.return_simgax_all_prior(fid, fisher_mat, key_y)

### 2015-11-24 n_priors_code/plotting/plot_prior_paper_M_nu_ell50.py

@@ -215,7 +215,7 @@ for key_y in ['omnuh2']:
         normalize_y = new_sigma / sigma_just_CMB_y  # make the new sigma y relative.
         # plot
 
-        line_plot = ax1.plot(normalize_x, new_sigma *94. * 1000., label=r'${0}={1:.1f}\%$'.format(
+        line_plot = ax1.plot(normalize_x, new_sigma *94. * 1000., label=r'$\sigma({0})={1:.1f}\%$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     new_sigma_all = utils.return_simgax_all_prior(fid, fisher_mat, key_y)

### 2015-11-24 n_priors_code/plotting/plot_prior_paper_N_eff.py

@@ -215,7 +215,7 @@ for  key_y in ['massless_neutrinos']:
         normalize_y = new_sigma / sigma_just_CMB_y  # make the new sigma y relative.
         # plot
 
-        line_plot = ax1.plot(normalize_x, new_sigma, label=r'${0}={1:.1f}\%$'.format(
+        line_plot = ax1.plot(normalize_x, new_sigma, label=r'$\sigma({0})={1:.1f}\%$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     new_sigma_all = utils.return_simgax_all_prior(fid, fisher_mat, key_y)

### 2015-11-25 n_priors_code/plotting/plot_prior_paper_M_nu_ell4.py
fixing previous commit

@@ -46,7 +46,7 @@ base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
 run_idx = 7
 lmax = 4499
-lmin = 50
+lmin = 4
 N_det = 10 ** 6
 # N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
 fsky = 0.75

### 2015-11-25 Notes/N_eff_prior_paper_v2_ypedit.tex

@@
 
-For CMB experiments the Fisher matrix can be related to the power spectrum $\boldsymbol{C}_\ell$ by \yp{I believe if you refer to $\boldsymbol{C}_\ell$ as the power spectrum, the trace in the following equation must be removed; the sum over ell is the trace taking.
-This equation seems to be in line with Eq.
-3 though, so perhaps you can just call $\boldsymbol{C}_\ell$ the covariance matrix?} :
+For CMB experiments the Fisher matrix can be related to the covariance matrix $\boldsymbol{C}_\ell$ by:
 \begin{equation}
@@
 Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
-\am{My understanding is that people do that: unlensed to avoid double counting in the informations with $C^{\phi}$, Aurelien and Wayne showed that unlensed + gaussian is similar to lensed + non-gaussian how big is the effect i do not know.
-Aurelien is writing a paper about that}
 \sd{Can we say something more positive about these effects?
 Why are we allowing ourselves to make this approximation?} Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
+However, we expect this approximations to have a small impact on the final results and, for this reason, a more careful modeling is beyond the scope of this work.
+For example, we checked that using lensed spectra will change the final constraint by less than 10$\%$.
 
@@
 %We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
-We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \yp{cite CAMB?} and
+We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
 the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
@@
 We deal with the Poisson noise from point sources in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
-The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way \yp{cite?}:
+The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way \cite{knox:1995}:
  \begin{equation}
@@
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>50$ meV that emerges from oscillation experiments.
-\yp{Consider changing $H_0 = 0.6\%$ to $\sigma(H_0) = 0.6\%$ in the legends to match the caption.}
+
 %In CMB experiment, neutrino masses constraint comes almost entirely from lensing.
@@
 We can also measure with great accuracy the number of relativistic species, thereby testing the standard model prediction and possibly probing new physics.
-The same is true for dark energy: we can tighten the constraint on $w$,~$w_{a}$ and test the nature of the accelerated expansion of the universe.
-\yp{$w_a$ was never mentioned in this paper; perhaps it is better not to discuss it entirely.}
+The same is true for dark energy: we can tighten the constraint on $w$ and test the nature of the accelerated expansion of the universe.
 

### 2015-12-04 Notes/N_eff_prior_paper_v2_ypedit.tex
minor changes on YP draft

@@
 \sd{Can we say something more positive about these effects?
-Why are we allowing ourselves to make this approximation?} Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
+Why are we allowing ourselves to make this approximation?} This assumptions neglect the possibility of imperfect reconstruction of the unlensed spectra (delensing) and the non-gaussian effects described in \cite{benoit-levy:2012}.
 However, we expect this approximations to have a small impact on the final results and, for this reason, a more careful modeling is beyond the scope of this work.
@@
 %We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
-We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
+We compute the power spectra $C_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
 the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
@@
 %\end{equation}
-this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives.
-As a consequence the differences of power spectra corresponding to different values of the parameters are big enough to ensure numerical accuracy.
-% issues in computing $\boldsymbol{C}_\ell$.
+this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives improving numerical accuracy.
 We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$.
@@
 Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models.
-As such, the key question is: What information is needed to reliably conclude that an additional parameter is required.
-The natural first baby step is to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist.
+As such, the key question is what information is needed to reliably conclude that an additional parameter is required.
+The natural first step is to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist.
 So our fiducial cosmology is flat $\nu \Lambda$CDM, with assumed
@@
 The fiducial values of these are $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1$.
-%It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
+It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
+Furthermore we vary the value of the relic helium abundance $Y_{P}$ consistently with the value of $N_{\rm eff}$ and $\Omega_b h^2$ in such a way to keep $N_{\rm eff}^{\rm BBN}$ the same as $N_{\rm eff}^{\rm CMB}$\cite{1475-7516-2008-06-016}.
+Practically we impose Eq.
+14 of \cite{wu:2014}.
+As pointed out in \cite{wu:2014} this goes in the direction of improving the constraint by $15-20\%$ compared with fixing $Y_{P}$ to its fiducial value $Y_{P}=0.248$.
 
@@
 Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. %($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ). 
-Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
+Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment together with the current BAO and Planck results; e.g., $\sigma(H_0)_{\rm CMB+Planck pol+BAO15}=0.5\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.5\%.
 $H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints.
-We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-pol prior .
-The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
+We chose a fiducial value $\sum m_{\nu}=85 $ meV.
+The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.
+The blue horizontal line correspond to the CMB S4+Planck pol+DESI constraint.}
 \label{fig:prior_omeganuh2}
@@
 
-\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment as a function of external priors. 
-Our Planck prior corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
+\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment and current BAO measurement (BAO15) as a function of external priors.
+\am{we should probably say why we add BAO and Planck here}
+Throughout this work our Planck prior corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
+Furthermore for BAO15, we follow table V of \cite{allison:2015}.
 Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
@@
 There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
+The accuracy given a ``DESI BAO'' prior is shown in \reffig{prior_omeganuh2} by the horizontal line.
 In this study, we find the same improvement by adding a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
@@
 \caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of external priors.
-The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.
-The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
+The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.}
 \label{fig:prior_massless_neutrinos}
@@
 Furthermore CMB lensing probes the growth of structure at late times which is different for different dark energy equations of state.
+\am{we should probably add a few lines on the physical constraint from BAO}
 %==========
@@
 \begin{center}
-\includegraphics{{ellipse_H0_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0_75}.pdf}
+\includegraphics{{ellipse_H0_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Two dimensional constraint (CMB S4 + Planck) 1$\sigma$ ellipse for $w,H_{0}$ space.
-The dashed (red) line correspond to values in the plane corresponding to the same distance to the last scattering surface.}
+The dashed (blue) line correspond to values in the plane corresponding to the same distance to the last scattering surface.}
 \label{fig:ellipse_w_H0}
@@
 
-\reffig{prior_w} shows that without any priors, CMB experiments will not do much better than current constraints, which hover around 10\%.
+\reffig{prior_w} shows that without any priors, CMB S4 experiments together with the current state of the art BAO measurement (BAO15) will not do much better than current constraints, which hover around 5\%.
 However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably.
@@
 Indeed in a two dimensional ($H_{0},w$) space the CMB constrained approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
-Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint almost by a factor of 2 from a $8\%$ error to a $4\%$ level.
+Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint almost by a factor of 2 from a $4\%$ error to a $2\%$ level.
 

### 2015-12-04 Notes/N_eff_prior_paper_v3.tex
add version 3 of draft with Scott revision

--- 2015-11-25 Notes/N_eff_prior_paper_v2_ypedit.tex
@@
 \begin{abstract}
-The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
-Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
+Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
+The low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
 Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
-$n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
-We find that CMB S4 constraints on neutrino mass will be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$.
+%$n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
+We find that CMB S4 constraints on neutrino mass could be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$.
 If the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important.
-A CMB constraint on $N_{\rm eff}$ will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$.
-Finally, an external prior on $H_{0}$ will be crucial to constrain a constant dark energy equation of state($w_{0}$).
+A CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$.
+Finally, an external prior on $H_{0}$ will help constrain the dark energy equation of state ($w$).
 \end{abstract}
@@
 Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe.
-They will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution expected in next generation, dubbed Stage IV (S4)~\cite{2013arXiv1309.5383A}.
+They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
 The recent past is  reassuring.
-Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
+For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P, planck-collaboration:2014, 2003ApJS..148..175S, 2000ApJ...545L...5H,2000Natur.404..955D}.
 These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck.
 With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
-The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} \am{it is ok for me to skip the list entirely otherwise feel free to add names I can add references later} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tail of the CMB spectrum with increasing accuracy.
+The same success has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
 
-This pattern of successes is far from over.
-The next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
+The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
 As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy, and neutrinos.
 
-The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $M_{\nu}$ affects the late-time growth of cosmic structures.
+The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
 Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$.
@@
 The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$.
-The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface.
+The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($\sum m_{\nu}<230$ meV from \cite{planck-collaboration:2014}), neutrinos are still relativistic at the last scattering surface.
 However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons.
 Different neutrino masses consequently lead to different CMB lensing spectra.
-A Stage IV experiment will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction.
+CMB-S4 will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction.
 This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
-These signatures can be mimicked with variations in other cosmic parameters, so an important lingering question is: What external information can be used to improve projected constraints further by breaking degeneracies?
 %Quantitatively, the CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
@@
 
-The CMB is sensitive also to the properties of dark energy (see \cite{2010MNRAS.405.2639J}).
+The CMB is sensitive also to the properties of dark energy (see, e.g., \cite{2010MNRAS.405.2639J}).
 Thanks to the current generation of experiments we know with extraordinary accuracy its energy density.
@@
 Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing.
-Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, the CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
+%Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, the CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
 However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
@@
 
-The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector.
+In the case of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
+%The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector.
 A general thrust in even the current generation of surveys is to combine information from different observations.
@@
 
+%Our philosophy is to first examine the simplest model $\Lambda$CDM to see how the constraints on neutrino masses can be improved.
+Then we turn to two extensions, one at a time: extra relativistic degrees of freedom and the dark energy equation of state.
+Our philosophy is: since we will be searching for a crack in $\Lambda$CDM, finding even one crack will be a revolutionary discovery.
+As such, it makes sense to us to probe one new piece of physics -- extra light species and evolving dark energy -- at a time.
+%
 This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
@@
 \end{equation}
-where $\theta_{i,j}$ represents two of the cosmological parameters and $\boldsymbol{\theta_0}$ is the fiducial values array that, by definition, maximizes the likelihood. 
+where $\theta_{i,j}$ represent the cosmological parameters and $\boldsymbol{\theta_0}$ is the set of fiducial values that, by definition, maximizes the likelihood. 
 %It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian.
@@
 
-For CMB experiments the Fisher matrix can be related to the covariance matrix $\boldsymbol{C}_\ell$ by:
+For CMB experiments the Fisher matrix is related to the angular power spectra $\boldsymbol{C}_\ell$ by:
 \begin{equation}
@@
 Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
-\sd{Can we say something more positive about these effects?
-Why are we allowing ourselves to make this approximation?} Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
+%\sd{Can we say something more positive about these effects? Why are we allowing ourselves to make this approximation?} 
+Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
 However, we expect this approximations to have a small impact on the final results and, for this reason, a more careful modeling is beyond the scope of this work.
@@
 
-The projected error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, is then:
+The projected error on the parameter $\theta_i$, marginalized over all the other parameters, $\sigma_i$, is then:
 \begin{equation}
@@
 We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
-the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
+the derivatives in \refeq{Fij_def2} using a 5 point finite difference formula:
 %\begin{equation}
@@
 % issues in computing $\boldsymbol{C}_\ell$.
-We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$.
+We also test the robustness of this calculation by varying the derivative steps in the range $2-7\%$ .
 The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties.
-\yp{Can the number 10\% be justified a bit more here?}
+%\yp{Can the number 10\% be justified a bit more here?}
 
@@
 The fiducial values of these are $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1$.
+In this same spirit, we keep the curvature fixed to zero, as the level predicted in the vast majority of inflationary models is of order $10^{-5}$, much too small to have an impact on the parameters we consider.
+Similarly, we fix the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends only on the baryon density.
 %It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
 
-The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the optimistic level for the next generation of CMB experiments (S4) assumed in ~\cite{2013arXiv1309.5383A,wu:2014,2013PhRvD..87h3008H}.
+The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the optimistic level for CMB-S4 assumed in ~\cite{2013arXiv1309.5383A,wu:2014,2013PhRvD..87h3008H}.
 For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
-This allows us to use all the polarization power spectrum multipoles all the way up to $\ell_{\rm E,max}=5000$.
-We deal with the Poisson noise from point sources in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
-The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way \cite{knox:1995}:
+This allows us to use all the polarization power spectrum multipoles out to $\ell_{\rm E,max}=5000$.
+We deal with the Poisson noise from point sources in the temperature signal by simply discarding all  small scale modes with $\ell>\ell_{\rm T,max}=3000$.
+The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way: %\cite{knox:1995}:
  \begin{equation}
@@
 Note that the quoted noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.
-\yp{Sentence here to explain the factor of $\sqrt{2}$ would be helpful.}
+%\yp{Sentence here to explain the factor of $\sqrt{2}$ would be helpful.}
 The noise $N_\ell^{\phi\phi}$ associated with the reconstructed $\phi$ spectrum is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.
 
+We also include information from current Baryon Acoustic Oscillation experiments\sd{reference} and projected constraints from low-$l$ Planck polarization.
+The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
+
 \section{Results \label{sec:results}}
-In this section we present our main results. We will start from our base $\nu \Lambda$CDM, and then will explore extensions to this model by considering ($N_{\rm eff},w$).
+In this section we present our main results. We will start from $\nu \Lambda$CDM and then explore extensions to this model by considering constraints on $N_{\rm eff}$ and then $w$.
 
@@
 Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. %($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ). 
-Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
+Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.5\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
 $H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints.
@@
 %==========
+
+%==========
+%As recently pointed out by \cite{allison:2015} there are two important regimes for neutrino masses constraint.
+In the first one,  large scale are assumed to be accessible by CMB experiments and, as a consequence the optical depth $\tau$ can be constrained internally by measuring the reionization bump.
+A second possibility is that CMB S4 will be  limited to smaller scales and therefore it will have to rely on external priors to constrain $\tau$.
+This can be given by current stage CMB experiments like Planck as well as future 21 cm survey \cite{liu:2015}.
+This distinction is important because the optical depth and the neutrino mass are quite degenerate and this will be limiting the constraint in the context of very precise future CMB experiments (see \cite{allison:2015} for details).
+
+\reffig{prior_omeganuh2} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors. 
+Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario.
+Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
+There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
+This is represented by the blue band in \reffig{prior_omeganuh2}.
+In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2} generalizes the extent to which external information will improve the neutrino mass constraints.
+For example, the red curve shows the improvement as additional information on the Hubble constant is applied.
+The projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.5\%$ (and similarly the values for the other parameters are given in the inset).
+So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.005$ corresponds to 1 on the $x$-axis.
+This would improve the neutrino mass constraint to 17 meV.
+Similarly, introducing an external constraint on the dark matter density equal to its internal constraint of  0.3\% (so again 1 on the $x$-axis) improves the projected neutrino mass constraint to 14 meV.
+It is not surprising that this is equal to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
+
+The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density.
+In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density.
+A prior on  $H_{0}$ helps for a related, but subtle reason.
+%with this as well since the effects of external priors are not completely independent.
+The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} so an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
+As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
+The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
+%\yp{Alessandro, are you planning on including the $\Omega_b h^2$ discussion here?}
 
@@
 \end{figure}
-%==========
-%As recently pointed out by \cite{allison:2015} there are two important regimes for neutrino masses constraint.
-In the first one,  large scale are assumed to be accessible by CMB experiments and, as a consequence the optical depth $\tau$ can be constrained internally by measuring the reionization bump.
-A second possibility is that CMB S4 will be  limited to smaller scales and therefore it will have to rely on external priors to constrain $\tau$.
-This can be given by current stage CMB experiments like Planck as well as future 21 cm survey \cite{liu:2015}.
-This distinction is important because the optical depth and the neutrino mass are quite degenerate and this will be limiting the constraint in the context of very precise future CMB experiments (see \cite{allison:2015} for details).
 
-\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment as a function of external priors. 
-Our Planck prior corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
-Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario.
-Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
-There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
-In this study, we find the same improvement by adding a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
-The improvement given by a prior on $\Omega_ch^2$ is also a consequence of the fact that the CMB lensing is sensitive to $f_\nu$; therefore in order to constrain $\Omega_{\nu}h^2$ from this ratio we need to determine the matter energy density.
-A prior on  $H_{0}$ helps with this as well since the effects of external priors are not completely independent.
-The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} and for this reason too an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
-As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors.
-\yp{Alessandro, are you planning on including the $\Omega_b h^2$ discussion here?}
-
-It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in \reffig{prior_omeganuh2_tau} we limit the modes to $\ell_{\rm min}>50$. 
-Together with the importance of $H_{0}$ and $\Omega_ch^2$ that we previously observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful.
+It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in \reffig{prior_omeganuh2_tau} we limit the modes to $\ell>\ell_{\rm min}=50$. 
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful.\sd{this effect seems pretty small.}
 Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
@@
 The additional fraction depends not only on well-known neutrino scattering rates but also on finite temperature quantum corrections.
-Upcoming experiments have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
-Furthermore, any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model.
-As pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above $\sim$100 GeV.
+Upcoming experiments have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3.
+This will be not only a precision test of our understanding of the Universe when it was about a second old.
+It also will constrain Beyond the Standard Model physics.
+For example, as pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above $\sim$100 GeV.
 %decoupled early in the thermal history.
@@
 \reffig{prior_massless_neutrinos} shows projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on the other seven parameters (the sum of the neutrino masses must be included as a free parameter).
-With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the expected deviation from $N_{\rm eff}=3$.
-The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
-\yp{The last sentence sounds a bit vague.}
+With no priors on the other parameters, the projected $1$-sigma error is $0.016$, close to a 3-sigma detection of the expected deviation from $N_{\rm eff}=3$.
 
@@
 As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction.
+The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
+%\yp{The last sentence sounds a bit vague.}
 
@@
 \includegraphics{{prior_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. An external prior on $H_{0}$ will be crucial to improve the constraint.} 
+\caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. 
+An external prior on $H_{0}$ will be crucial to improve the constraint.} 
 \label{fig:prior_w}
@@
 \begin{center}
-\includegraphics{{ellipse_H0_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0_75}.pdf}
+\includegraphics{{ellipse_H0_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Two dimensional constraint (CMB S4 + Planck) 1$\sigma$ ellipse for $w,H_{0}$ space.
@@
 
-\reffig{prior_w} shows that without any priors, CMB experiments will not do much better than current constraints, which hover around 10\%.
-However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably.
+\reffig{prior_w} shows that without any priors, CMB experiments will do better than current constraints, which hover around 10\%.
+However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ by a factor of 2.
 This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface.
 However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat).
-Indeed in a two dimensional ($H_{0},w$) space the CMB constrained approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
-Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint almost by a factor of 2 from a $8\%$ error to a $4\%$ level.
+Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
+Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint almost by a factor of 2 from a $4\%$ error to a $2\%$ level.
 
 \section{Conclusions  \label{sec:conclusions}}
-The design of the next CMB Stage IV experiments represents an exciting challenge.
-The possible reward is remarkable.
-We can finally put a tight constraint on the total mass of neutrinos with deep implications on particle physics like, for example, the solution of the mass hierarchy problem.
-We can also measure with great accuracy the number of relativistic species, thereby testing the standard model prediction and possibly probing new physics.
-The same is true for dark energy: we can tighten the constraint on $w$ and test the nature of the accelerated expansion of the universe.
-
-These results, however, will not be possible without an optimal synergy with external experiments observing a variety of others cosmological probes.
+A CMB-S4 experiment will measure the sum of the neutrino masses, constrain extra relativistic degrees of freedom predicted in many extensions of the Standard Model, and measure the dark energy equation of state at the few percent level.
+These results can be improved even further by combining them with information from external experiments.
 Understanding the impact of such external information is crucial in determining the experimental specifications and requirements needed for CMB-S4, and ultimately in maximizing the scientific impact of the future CMB experiments.
@@
 Understanding CMB weaknesses and the impact of external data can have a deep influence in determining the needed experimental specifications.
-In this work we contribute to that effort by exploring the impact of external priors on CMB S4 cosmological constraints, focusing on the neutrino sector and dark energy parameters.
-We find:
+Toward that goal, we investigated here how these constraints depend on external priors:
 \begin{description}
-\item[$\sum m_\nu$] To improve the CMB constraint on sum of neutrino masses by a factor of two, low-redshift measurements will be needed.
-Indeed external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level CMB S4 will reach alone can bring the 1-$\sigma$ uncertainty on $\sum m_\nu$ from $30$meV to below 20meV.
-Furthermore, if the final design of CMB S4 will not include measurements on large scales ($\ell<50$), external constraints on the primordial amplitude $A_{s}$ will be needed.
-Moreover, an external prior on $\tau$ will allow the CMB to constraint $A_{s}$ internally, thus strongly alleviating this problem.
-\item[$N_{\rm eff}$] The constraint on the effective neutrino number will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$ and the baryon energy density $\Omega_b h^2$.
-\item[$w$] In this case the improvement will mainly come by external prior on $H_{0}$.
-For example a $2\%$ prior from external experiment will bring the reduce the error on w by a factor of two from 8$\%$ to $4\%$.
-\end{description}
-\yp{Neff and w sections feel like they need a bit more flesh.}
+\item[$\sum m_\nu$] To improve the CMB constraint on the sum of neutrino masses by a factor of two, low-redshift measurements will be needed.
+Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub=percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
+Furthermore, if the final design of CMB S4 will not include measurements on large scales ($\ell<50$), external constraints on the primordial amplitude $A_{s}$ will also be useful.
+%Moreover, an external prior on $\tau$ will allow the CMB to constraint $A_{s}$ internally, thus strongly alleviating this problem.
+\item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$ and the baryon energy density $\Omega_b h^2$.
+\item[$w$] In this case the improvement will mainly come from external priors on $H_{0}$.
+For example a $1\%$ prior from an external experiment will reduce the error on $w$ by a factor of two.\end{description}
+%\yp{Neff and w sections feel like they need a bit more flesh.}
 
-\textit{Comment on issue and future.
+Much of the improvement that would come on neutrino masses and $w$ from $H_0$ or $\Omega_ch^2$ priors is anticipated to be provided by distance measurements made by DESI.
+But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2} and \reffig{ellipse_w_H0}.
+Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST that would tighten the constraints on, say, the baryon density and $n_s$ \sd{could you add to fig 3 a curve that has priors on both $n_s$ and $\omega_b$?} could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
+Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.
+%\textit{Comment on issue and future.
 This will be even more relevant when we open the parameter space.

### 2015-12-04 n_priors_code/plotting/plot_prior_paper_N_eff.py
added ombh2+ns line to N_eff figure and change code to do that

@@ -219,6 +219,12 @@ for  key_y in ['massless_neutrinos']:
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     new_sigma_all = utils.return_simgax_all_prior(fid, fisher_mat, key_y)
+    new_sigma_ns_omb  =  utils.return_simgax_list_prior(fid, fisher_mat, ['scalar_spectral_index(1)', 'ombh2'] ,key_y)
+
+    plt.plot(normalize_x, new_sigma_ns_omb, label=r'${}+{}$'.format(
+            str(label['ombh2']),str(label['scalar_spectral_index(1)'])),
+             linestyle=next(linecycler), linewidth=font_size / 10., alpha=0.6)
+
 
     plt.plot(normalize_x, new_sigma_all, label=r'All',
              linestyle=next(linecycler), linewidth=font_size / 10., alpha=0.6)

### 2015-12-06 Notes/N_eff_prior_paper_v3.tex

@@
 \affiliation{Department of Physics, University of Arizona, Tucson, AZ 85721, USA}
-%
-%\author{Alessandro Manzotti}
-%\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
-%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-%\author{Scott Dodelson}
-%
-%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
-
-%\author{Scott Dodelson}
-%
-%\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
-%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
-%\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
-
-%\author{Youngsoo Park}
-%\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
 
@@
 Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
-%$n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
 We find that CMB S4 constraints on neutrino mass could be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$.
@@
 
-%Our philosophy is to first examine the simplest model $\Lambda$CDM to see how the constraints on neutrino masses can be improved.
-Then we turn to two extensions, one at a time: extra relativistic degrees of freedom and the dark energy equation of state.
-Our philosophy is: since we will be searching for a crack in $\Lambda$CDM, finding even one crack will be a revolutionary discovery.
-As such, it makes sense to us to probe one new piece of physics -- extra light species and evolving dark energy -- at a time.
-%
 This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
@@
 where $\theta_{i,j}$ represent the cosmological parameters and $\boldsymbol{\theta_0}$ is the set of fiducial values that, by definition, maximizes the likelihood.
-%It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian.
-Even if the gaussian approximation has been shown to be appropriate most of the time, exceptions exist in the literature \cite{2012JCAP...09..009W}.
-However, the level of accuracy needed in this work does not require a more careful modeling of the likelihoods.
-Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.
 
@@
 Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
-%\sd{Can we say something more positive about these effects? Why are we allowing ourselves to make this approximation?} 
 Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
@@
 
-%We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data and easily including external priors. 
-%This technique however introduces also technical difficulties. 
-%Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
-%Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix lead to a big change in its inverse.
-Because we use \refeq{cramer-rao} this can be a problem for error estimation.
-%Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard procedure used to forecast future constrains \cite{wu:2014}.
-%We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
 We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
 the derivatives in \refeq{Fij_def2} using a 5 point finite difference formula:
-%\begin{equation}
-%\begin{split}
-%\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim & \frac{1}{12 h} [ -C(\theta_{0}+2h)+ 8C(\theta_{0}+h) \\ &-8C(\theta_{0}-h)+C(\theta_{0}-2h)].
-%\end{split}
-%\label{eqn:deriv}
-%\end{equation}
 this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives.
 As a consequence the differences of power spectra corresponding to different values of the parameters are big enough to ensure numerical accuracy.
-% issues in computing $\boldsymbol{C}_\ell$.
 We also test the robustness of this calculation by varying the derivative steps in the range $2-7\%$ .
 The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties.
-%\yp{Can the number 10\% be justified a bit more here?}
-
 Our philosophy is to consider as few extensions as possible.
@@
 We deal with the Poisson noise from point sources in the temperature signal by simply discarding all  small scale modes with $\ell>\ell_{\rm T,max}=3000$.
-The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way: %\cite{knox:1995}:
+The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
  \begin{equation}
@@
 Note that the quoted noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$. 
-%\yp{Sentence here to explain the factor of $\sqrt{2}$ would be helpful.}
 The noise $N_\ell^{\phi\phi}$ associated with the reconstructed $\phi$ spectrum is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.
 
-We also include information from current Baryon Acoustic Oscillation experiments\sd{reference} and projected constraints from low-$l$ Planck polarization.
+We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
+As current BAO data, we follow \cite{allison:2015} and we take into account 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
 The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
@@
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit  $\sum m_\nu>50$ meV that emerges from oscillation experiments.
-
-%In CMB experiment, neutrino masses constraint comes almost entirely from lensing.
-When the total mass of neutrinos the CMB lensing power spectrum amplitude decreases at scales smaller than the free-streeming scale of neutrinos.
-This effect is degenerate with an overall change of the power spectrum amplitude $A_{s}$.
 
@@
 %==========
-%As recently pointed out by \cite{allison:2015} there are two important regimes for neutrino masses constraint.
-In the first one,  large scale are assumed to be accessible by CMB experiments and, as a consequence the optical depth $\tau$ can be constrained internally by measuring the reionization bump.
-A second possibility is that CMB S4 will be  limited to smaller scales and therefore it will have to rely on external priors to constrain $\tau$.
-This can be given by current stage CMB experiments like Planck as well as future 21 cm survey \cite{liu:2015}.
-This distinction is important because the optical depth and the neutrino mass are quite degenerate and this will be limiting the constraint in the context of very precise future CMB experiments (see \cite{allison:2015} for details).
 
@@
 It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in \reffig{prior_omeganuh2_tau} we limit the modes to $\ell>\ell_{\rm min}=50$. 
-Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful.\sd{this effect seems pretty small.}
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful.
+
 Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
@@
 For example a $1\%$ prior from an external experiment will reduce the error on $w$ by a factor of two.\end{description}
-%\yp{Neff and w sections feel like they need a bit more flesh.}
 
@@
 Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.
-%\textit{Comment on issue and future.
-This will be even more relevant when we open the parameter space.
-Will CMB + others be able to discover something CMB can't alone?
-$N_{\rm eff}$ extensions, mass splitting etc}
 

### 2015-12-06 n_priors_code/fisher_joint_test.py
re-organize comment code and another full run of th pipeline

@@ -5,15 +5,12 @@
 '''
 
 FAST VERSION
-Fisher code to test prior effect on N effective neutrino estimates
+Fisher code for CMB T,E,+ lensing phi.  used for the priors paper but pretty versatile.
+
 
 TODO:
-HIGH: lensing noise (almost there: Komatsu and quicklens.)
-
-low importance: ini file. Think about PCA to understand what is more important for N_eff
-
-
-create classes or at least external utilities. Maybe a better way to interface with cosmosis
+low importance: ini file. or sys argv passing parameters
+create classes or at least external utilities.
 
 # DATA FROM CAMB
 
@@ -24,17 +21,12 @@ we use the _lenspotentialcls so
 
   CX are l(l+1)Cl/2pi and Cdd=[l(l+1)]^2 Clphi/2pi, CdT=[l(l+1)]^(3/2) ClphiT/2pi, CdE=[l(l+1)]^(3/2)ClphiE/2pi
 
-
-The matrix is going to run an 4 parameters:
-
 CONVENTIONS:
 alphabetical in CAMB description. Use the ordered dict tools to explore it.
 hubble,massless_neutrinos,omnuh2,re_optical_depth,scalar_amp(1),scalar_spectral_index(1)
 
 
 GOAL
-
-1) reproduce Wu paper and beyond/ DONE. Cross checked with Zhen Pan UC Davis
 
 '''
 __author__ = "A.Manzotti"
@@ -45,16 +37,13 @@ __email__ = "manzotti.alessandro@gmail.com"
 __status__ = "Production"
 
 import numpy as np
-import math
-import matplotlib.pyplot as plt
 import utils
 import pickle
 import sys
 import collections
-import statsmodels.api as sm
 
-# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
 
+# Auxiliary functions.
 
 def years2sec(years):
     ''' years to sec '''
@@ -70,7 +59,9 @@ def fsky2arcmin(fsky):
 def calc_deriv_vectorial(fisher_index, dats, pargaps, values, order=5):
     '''
     Vectorize derivative computation. Each -calc_c_general gives you the matrix C for all ell and here you
-    compute the derivative in a vectorize fashion
+    compute the derivative in a vectorize fashion.
+
+    We accept 4 different 2 sides formulas. You pass the data so be careful you are doing i in a coherent way.
     '''
 
[…]

### 2015-12-06 n_priors_code/fisher_produce_planckpol.py

@@ -10,12 +10,7 @@ Here we produce the Planck Fisher matrix we will lately add to our fisher.
 
 
 TODO:
-HIGH: lensing noise (almost there: Komatsu and quicklens.)
-
 low importance: ini file. Think about PCA to understand what is more important for N_eff
-
-
-create classes or at least external utilities. Maybe a better way to interface with cosmosis
 
 # DATA FROM CAMB
 
@@ -47,15 +42,14 @@ __email__ = "manzotti.alessandro@gmail.com"
 __status__ = "Production"
 
 import numpy as np
-import math
-import matplotlib.pyplot as plt
 import utils
 import pickle
 import sys
 import collections
 
-# util.nl(noise_uK_arcmin, fwhm_arcmin, lmax)
 
+# Auxiliary functions.
+# ===================
 
 def years2sec(years):
     ''' years to sec '''
@@ -66,6 +60,7 @@ def fsky2arcmin(fsky):
     '''convert fsky in fraction of unity to arcmin^2'''
     # 41253 square degrees in all sky
     return 41253. * fsky * 60. * 60.
+# ===================
 
 
 def calc_deriv_vectorial(fisher_index, dats, pargaps, values, order=5):
@@ -102,8 +97,7 @@ def calc_c_fiducial(data):
     '''
     Given CMB data dats it normalize them anf create a 3x3 matrix
 
-    ell is the multiple
-    iell is the index in the data ell corresponds to
+    THE LEVEL OF NOISE IN THE CHANNEL IS TAKEN FROM ALLISON ET AL (DUNKLEY)
 
     remember the order from CAMB
      l CTT CEE CBB CTE Cdd CdT CdE
@@ -116,7 +110,6 @@ def calc_c_fiducial(data):
 
     global Y, sec_of_obs, arcmin_from_fsky, lmax_index
     ell = data[lmin_index:lmax_index, 0, 0]
-    # s = 350. * np.sqrt(arcmin_from_fsky) / np.sqrt(N_det * Y * sec_of_obs)  # half sky in arcmin^2
     s = {}
     s_pol = {}
     t = {}
@@ -142,6 +135,9 @@ def calc_c_fiducial(data):
     t['143'] = 7. / 60. / 180. * np.pi  # 2arcmin to rads beam
     t['217'] = 5. / 60. / 180. * np.pi  # 2arcmin to rads beam
     t['353'] = 5. / 60. / 180. * np.pi  # 2arcmin to rads bea
[…]

### 2015-12-07 Notes/N_eff_prior_paper_v3.tex

@@
 Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
-The low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
+Indeed the low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
@@
 If the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important.
-A CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$.
+Furthermore a CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$.
 Finally, an external prior on $H_{0}$ will help constrain the dark energy equation of state ($w$).
@@
 The same success has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
-
 The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
@@
 The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
-Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$.
+Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe, their number affects the expansion rate $H(z)$.
 This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
@@
 This will turn into a precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
-%Quantitatively, the CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}.
-The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.
 
@@
 Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing.
-%Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, the CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
 However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
@@
 In the case of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
-%The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector.
-A general thrust in even the current generation of surveys is to combine information from different observations.
-%In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters.
-%In particular CMB will benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
 Here we study the dependance of projected constraints from CMB on the external priors assumed.
@@
 Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
-Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
+This means that we neglect eventual imperfect reconstruction of the unlensed spectra (delensing) and the non-gaussian effects described in \cite{benoit-levy:2012} .
 However, we expect this approximations to have a small impact on the final results and, for this reason, a more careful modeling is beyond the scope of this work.
@@
 
-We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
+We compute the power spectra $C_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
 the derivatives in \refeq{Fij_def2} using a 5 point finite difference formula:
@@
 parameters from Table 2 of the \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e.
-$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc, supplemented by an arbitrary choice of $\sum m_\nu$ $\simeq$ 85\ meV.
+$\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc, supplemented by an choice of $\sum m_\nu$ $\simeq$ 85\ meV to be consistent with \cite{wu:2014}.
 We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters, in each case keeping the other parameter fixed.
@@
 Similarly, we fix the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends only on the baryon density.
-%It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).
 
@@
 For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible.
-This allows us to use all the polarization power spectrum multipoles out to $\ell_{\rm E,max}=5000$.
+This allows us to use all the polarization power spectrum multipoles out to $\ell_{\rm E,max}=4500$.
 We deal with the Poisson noise from point sources in the temperature signal by simply discarding all small scale modes with $\ell>\ell_{\rm T,max}=3000$.
@@
 We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
-As current BAO data, we follow \cite{allison:2015} and we take into account 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
+As current BAO data (labeled BAO15), we follow \cite{allison:2015} and we take into account 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
 The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
@@
 \caption{
-Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. %($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ). 
-Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.5\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
+Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. 
+Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment (+Planck+current BAO); e.g., $\sigma(H_0)_{\rm S4 + Planck Pol + BAO15}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
+The horizonal line correspond to CMB S4 experiment + Planck Pol + DESI.
 $H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints.
@@
 A prior on $H_{0}$ helps for a related, but subtle reason.
-%with this as well since the effects of external priors are not completely independent.
-The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} so an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
+The position of the CMB peaks, well constrained by measurement, in a flat universe is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
+. For this reason an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
 As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
 The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
-%\yp{Alessandro, are you planning on including the $\Omega_b h^2$ discussion here?}
 
@@
 a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$.
-If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} will be needed to fully exploit the CMB constraining power.
+If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.
 
@@
 Upcoming experiments have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3.
-This will be not only a precision test of our understanding of the Universe when it was about a second old.
+This will be a precision test of our understanding of the Universe when it was about a second old.
 It also will constrain Beyond the Standard Model physics.
 For example, as pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above $\sim$100 GeV.
-%decoupled early in the thermal history.
 
@@
 
-The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most degenerate with $N_{\rm eff}$.
+The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improce $N_{\rm eff}$ constraint.
 As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction.
 The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
-%\yp{The last sentence sounds a bit vague.}
 
@@
 Understanding the impact of such external information is crucial in determining the experimental specifications and requirements needed for CMB-S4, and ultimately in maximizing the scientific impact of the future CMB experiments.
-%Indeed this should be taken into account, especially during the CMB Stage IV planning phase.
-Understanding CMB weaknesses and the impact of external data can have a deep influence in determining the needed experimental specifications.
+
 Toward that goal, we investigated here how these constraints depend on external priors:
@@
 \item[$\sum m_\nu$] To improve the CMB constraint on the sum of neutrino masses by a factor of two, low-redshift measurements will be needed.
-Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub=percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
+Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub-percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
 Furthermore, if the final design of CMB S4 will not include measurements on large scales ($\ell<50$), external constraints on the primordial amplitude $A_{s}$ will also be useful. 
-%Moreover, an external prior on $\tau$ will allow the CMB to constraint $A_{s}$ internally, thus strongly alleviating this problem.
 \item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$ and the baryon energy density $\Omega_b h^2$.
@@
 But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2} and \reffig{ellipse_w_H0}.
-Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST that would tighten the constraints on, say, the baryon density and $n_s$ \sd{could you add to fig 3 a curve that has priors on both $n_s$ and $\omega_b$?} could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
+Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST that would tighten the constraints on, say, the baryon density and $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
 Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.

### 2015-12-07 Notes/N_eff_prior_paper_v3.tex
small edits

@@
 Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
-Indeed the low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
+The low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
 Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
-Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
+Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage IV (S4) experiment on neutrino and dark energy properties.
 We find that CMB S4 constraints on neutrino mass could be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$.
-If the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important.
-Furthermore a CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$.
+If polarization on the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important.
+A CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$.
 Finally, an external prior on $H_{0}$ will help constrain the dark energy equation of state ($w$).
@@
 The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
-As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy, and neutrinos.
+As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of dark matter, inflation, dark energy, neutrinos, and other  Beyond the Standard Model physics.
 
@@
 Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing.
-However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
+However, dark energy properties are strongly degenerate with other geometrical parameters such as $H_{0}$. As with the neutrino sector, dark energy constraints from the CMB 
 will be improved by external experiments.
 
-In the case of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
+In the cases of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
 Here we study the dependance of projected constraints from CMB on the external priors assumed.
@@
 
-This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints.
+This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions used to derive the effect of external priors on the CMB parameter constraints.
 In \refsec{results} we will describe our results and we then conclude with a discussion of them in \refsec{conclusions}.
@@
 We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
-As current BAO data (labeled BAO15), we follow \cite{allison:2015} and we take into account 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
+For current BAO data (labeled BAO15), we follow \cite{allison:2015} and include 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
 The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
@@
 For example, the red curve shows the improvement as additional information on the Hubble constant is applied.
-The projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.5\%$ (and similarly the values for the other parameters are given in the inset).
-So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.005$ corresponds to 1 on the $x$-axis.
+The projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.6\%$ (and similarly the values for the other parameters are given below the figure).
+So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.006$ corresponds to 1 on the $x$-axis.
 This would improve the neutrino mass constraint to 17 meV.
-Similarly, introducing an external constraint on the dark matter density equal to its internal constraint of 0.3\% (so again 1 on the $x$-axis) improves the projected neutrino mass constraint to 14 meV.
-It is not surprising that this is equal to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
+Similarly, introducing an external constraint on the dark matter density equal to its internal constraint of 0.4\% (so again 1 on the $x$-axis) improves the projected neutrino mass constraint to 20 meV.
+It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
 
@@
 A prior on $H_{0}$ helps for a related, but subtle reason.
-The position of the CMB peaks, well constrained by measurement, in a flat universe is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
-. For this reason an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
+The position of the CMB peaks, well constrained by measurement, in a flat universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
+. Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
 As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
@@
 \includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Same as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
+\caption{Same as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors also help because of their degeneracy with $\sum m_{\nu}$.}
 \label{fig:prior_omeganuh2_tau}
@@
 a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$.
-If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.
+If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon-borne experiments like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.
 
@@
 
-The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improce $N_{\rm eff}$ constraint.
-As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction.
+The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improve the $N_{\rm eff}$ constraint.
+As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction.
 The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
@@
 Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
-Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint almost by a factor of 2 from a $4\%$ error to a $2\%$ level.
+Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint on $w$ by a factor of two.
 

### 2015-12-07 n_priors_code/fisher_joint_test.py
switching to Planck Pol just for l<50

@@ -310,7 +310,7 @@ for iell, ell in enumerate(dats[lmin_index:lmax_index, 0, 0]):
 
 # LOADING EXTERNAL DATA TO BE ADDED.
 planck_fisher = np.loadtxt(
-    '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/data/fisher_mat_joint_lmin=2_lmax=2500_ndet=Planck_fsky=0.2.txt')
+    '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/data/fisher_mat_joint_lmin=2_lmax=50_ndet=Planck_fsky=0.2.txt')
 BAO_fisher = np.loadtxt('/home/manzotti/n_eff-dependence-on-prior/n_priors_code/data/fisher_mat_BAO.txt')
 BAO_fisher_DESI = np.loadtxt('/home/manzotti/n_eff-dependence-on-prior/n_priors_code/data/fisher_mat_BAO_DESI.txt')
 
@@ -324,7 +324,7 @@ print 'ADDING PLANCK POL'
 fisher += planck_fisher
 
 print 'ADDING BAO'
-fisher += BAO_fisher
+fisher += BAO_fisher_DESI
 
 print 'lmax =', ell
 # print fisher_inv
@@ -363,7 +363,7 @@ np.savetxt('data/{}/invetered_sqrt_fisher_joint_lmin={}_lmax={}_ndet={}_fsky={}.
 
 print 'fisher=', fisher
 no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
-plot_now = ['omnuh2', 'w']
+plot_now = ['omnuh2']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 
 par_gaps, values, fid, fisher_single = utils.exclude_parameters_from_fisher(

### 2015-12-07 n_priors_code/fisher_produce_planckpol.py

@@ -245,7 +245,7 @@ def C(iell, ell, parbin, data):
 
 
 l_t_max = 2500  # this is the multipole you want to cut the temperature Cl at, to simulate the effect of foreground
-lmax = 2500
+lmax = 50
 lmin = 2
 N_det = 'Planck'
 # No lensing
@@ -440,7 +440,7 @@ print 'lmax =', ell
 # lmin, lmax, N_det, fsky), np.sqrt(fisher_inv), header=header)
 
 print np.shape(fisher)
-np.savetxt('/home/manzotti/n_eff-dependence-on-prior/n_priors_code/data/fisher_mat_joint_lmin=2_lmax=2500_ndet=Planck_fsky=0.2.txt', fisher, header=header)
+np.savetxt('/home/manzotti/n_eff-dependence-on-prior/n_priors_code/data/fisher_mat_joint_lmin=2_lmax=50_ndet=Planck_fsky=0.2.txt', fisher, header=header)
 
 print 'fisher=', fisher
 

### 2015-12-07 n_priors_code/plotting/plot_prior_paper_M_nu_ell4.py

@@ -227,13 +227,13 @@ for key_y in ['omnuh2']:
     ax1.set_ylim((0.8 * np.amin(new_sigma_all) * 94. * 1000., 1.1 * np.amax(new_sigma) * 94. * 1000.))
     ax1.set_xlim((0.1, 3.1))
 
-    ax1.axhline(16.9 / 100. * fid['omnuh2'] * 94. * 1000., xmin=0., xmax=0.35, alpha=0.4, linewidth=2, label='DESI')
+    ax1.axhline(16.94 / 100. * fid['omnuh2'] * 94. * 1000., xmin=0., xmax=0.35, alpha=0.4, linewidth=2, label='DESI')
 
     ax1.legend(loc=0)
 
     # ax1.set_title(r'$\sigma({0})={1:.1f}\%$'.format(str(label[key_y]), np.abs(sigma_just_CMB_y / fid[key_y] * 100.)))
     ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
-    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm \rm S4~ +~ Planck ~ Pol + ~BAO15}$')
+    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm \rm S4~ +~ Planck ~ Pol  (\ell<50) + ~BAO15}$')
 
     y1, y2 = ax1.get_ylim()
     ax2 = ax1.twinx()

### 2015-12-07 n_priors_code/plotting/plot_prior_paper_M_nu_ell50.py

@@ -227,13 +227,13 @@ for key_y in ['omnuh2']:
     ax1.minorticks_on()
     ax1.set_ylim((0.8 * np.amin(new_sigma_all)*94. * 1000., 1.1 * np.amax(new_sigma)*94. * 1000.))
     ax1.set_xlim((0.1, 3.1))
-    ax1.axhline(23.2/100.*fid['omnuh2']*94. * 1000.,xmin=0.,xmax=0.35,alpha=0.4,linewidth=2,label='DESI')
+    ax1.axhline(23.24/100.*fid['omnuh2']*94. * 1000.,xmin=0.,xmax=0.35,alpha=0.4,linewidth=2,label='DESI')
 
 
     ax1.legend(loc=0)
     # ax1.set_title(r'$\sigma({0})={1:.1f}\%$'.format(str(label[key_y]), np.abs(sigma_just_CMB_y / fid[key_y] * 100.)))
     ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
-    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck ~ Pol + ~BAO15}$')
+    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck ~ Pol  (\ell<50) + ~BAO15}$')
     y1, y2 = ax1.get_ylim()
     ax2 = ax1.twinx()
     minor_loc = ax1.yaxis.get_minor_locator()

### 2015-12-07 n_priors_code/plotting/plot_prior_paper_N_eff.py

@@ -238,7 +238,7 @@ for  key_y in ['massless_neutrinos']:
     ax1.legend(loc=0)
     # ax1.set_title(r'$\sigma({0})={1:.1f}\%$'.format(str(label[key_y]), np.abs(sigma_just_CMB_y / fid[key_y] * 100.)))
     ax1.set_ylabel(r'$\sigma(' + label[key_y] + r')$')
-    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck ~ Pol + ~BAO15 }$')
+    ax1.set_xlabel(r'$\rm{prior}/\sigma(x)_{\rm S4~ +~ Planck ~ Pol (\ell<50) + ~BAO15 }$')
     y1, y2 = ax1.get_ylim()
     ax2 = ax1.twinx()
     minor_loc = ax1.yaxis.get_minor_locator()

### 2015-12-07 Notes/N_eff_prior_paper_v3.tex
minor changes on YP draft

@@
 \includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Same as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors also help because of their degeneracy with $\sum m_{\nu}$.}
+\caption{Same as \reffig{prior_omeganuh2} but with $\ell_{\rm min}>50$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
+
 \label{fig:prior_omeganuh2_tau}

### 2015-12-07 n_priors_code/fisher_joint_test.py
fixing previous commit with DESI in it

@@ -324,7 +324,7 @@ print 'ADDING PLANCK POL'
 fisher += planck_fisher
 
 print 'ADDING BAO'
-fisher += BAO_fisher_DESI
+fisher += BAO_fisher
 
 print 'lmax =', ell
 # print fisher_inv

### 2015-12-08 Notes/CMBS4_priors_paper_.tex
final changes and note CMBS4_priors_paper_.tex is the version on the arxiv ans that will be sent to the journal

--- 2015-12-07 Notes/N_eff_prior_paper_v3.tex
@@
 \documentclass[aps,prd,reprint,superscriptaddress]{revtex4-1}
+\usepackage{hyperref}
 \usepackage{amsmath}
@@
 \begin{document}
-\graphicspath{{images/}}
+\graphicspath{{./images/}}
 
@@
 \author{Alessandro Manzotti}
-\email{\href{mailto:manzotti.alessandro@gmail.com}{manzotti.alessandro@gmail.com}}
+\email{mailto:manzotti.alessandro@gmail.com}
 \affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
@@
 Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment (+Planck+current BAO); e.g., $\sigma(H_0)_{\rm S4 + Planck Pol + BAO15}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
-The horizonal line correspond to CMB S4 experiment + Planck Pol + DESI.
-$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints.
-We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-pol prior .
-The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI. The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.
+$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV.}
 \label{fig:prior_omeganuh2}
@@
 Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario.
-Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}.
+Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}.
 There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
@@
 . Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
-As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
+As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
 The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
@@
 \includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Same as \reffig{prior_omeganuh2} but with $\ell_{\rm min}>50$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
+\caption{Same as \reffig{prior_omeganuh2} but with $\ell^{\rm CMB-S4}_{\rm min}>50$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
 
@@
 
-\reffig{prior_w} shows that without any priors, CMB experiments will do better than current constraints, which hover around 10\%.
+\reffig{prior_w} shows that without any priors, CMB experiments (+BAO15) will do better than current constraints, which hover around 10\%.
 However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ by a factor of 2.

### 2015-12-10 n_priors_code/plotting/plot_prior_paper_M_nu_multiell.py
minor changes + Riess plots + multiell for movies

@@ -1,0 +1,273 @@
+'''
+This plotting script is used to plot the effect of external priors on all the parameters involved.
+
+To put all the parameters in the same plot the prior is defined as the improvement from the CMB S4 results we get.
+
+the same is true for the improvement on the y axis parameters
+
+
+
+'''
+
+
+import numpy as np
+import pickle as pickle
+import utils
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+from scipy import integrate
+import scipy.special
+import scipy.interpolate
+from itertools import cycle
+#from matplotlib.colors import colorConverter
+from matplotlib.ticker import MaxNLocator  # needed for tick
+import sys
+from matplotlib.patches import Ellipse
+import n_priors_code.utils as utils
+from palettable.colorbrewer.qualitative import Set1_9
+
+# ============================================
+# PLOTTING DEFINITION SKIP TO ~155
+# ============================================
+
+
+# ============================================
+# SIZE OF THE PICTURE
+# ============================================
+WIDTH = 246.0  # the number latex spits out
+# the fraction of the width you'd like the figure to occupy, 0.5 if double
+# column cause
+FACTOR = 0.9
+fig_width_pt = WIDTH * FACTOR
+
+inches_per_pt = 1.0 / 72.27
+# golden_ratio  = (np.sqrt(5) - 1.0) / 2.0  # because it looks good
+ratio = 0.8
+fig_width_in = fig_width_pt * inches_per_pt  # figure width in inches
+fig_height_in = fig_width_in * ratio   # figure height in inches
+fig_dims = [fig_width_in, fig_height_in]  # fig dims as a list
+# ============================================
+# ============================================
+
+# ============================================
+# SET LATEX
+plt.rc('text', usetex=True)
+plt.rc('font', **{'family': 'serif', 'serif': ['Computer Modern Roman']})
+plt.rcParams['font.weight'] = 400
+
+# ============================================
+
+# ============================================
+# FONT SIZES
+font_size = 14
+plt.rcParams['font.size'] = font_size
+plt.rcParams['axes.labelsize'] = font_size / 1.3
[…]

### 2015-12-10 n_priors_code/plotting/plot_riess_paper_M_nu_with_w.py

@@ -1,0 +1,264 @@
+'''
+This plotting script is used to plot the effect of external priors on all the parameters involved.
+
+To put all the parameters in the same plot the prior is defined as the improvement from the CMB S4 results we get.
+
+the same is true for the improvement on the y axis parameters
+
+
+
+'''
+
+
+import numpy as np
+import pickle as pickle
+import utils
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+from scipy import integrate
+import scipy.special
+import scipy.interpolate
+from itertools import cycle
+#from matplotlib.colors import colorConverter
+from matplotlib.ticker import MaxNLocator  # needed for tick
+import sys
+from matplotlib.patches import Ellipse
+import n_priors_code.utils as utils
+from palettable.colorbrewer.qualitative import Set1_9
+# ============================================
+# ============================================
+# ============================================
+# TO DO
+# DESI level = -3.21327730556 %
+# ============================================
+# ============================================
+# ============================================
+
+no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
+plot_now = ['omnuh2','w']
+excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
+# omnuh2
+
+# READ DATA
+# DEFINE YOUR FOLDER HERE
+base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
+data_type = 'varying_all'
+run_idx = 7
+lmax = 4499
+lmin = 4
+N_det = 10 ** 6
+# N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
+fsky = 0.75
+# ======
+fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type, str(run_idx)), "rb"))
+values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
+par_gaps = pickle.load(open(base_dir + 'data/{}/run{}/par_gaps.p'.format(data_type, str(run_idx)), "rb"))
+fisher_mat = np.loadtxt(
[…]

### 2015-12-10 n_priors_code/plotting/plot_riess_paper_w.py

@@ -1,0 +1,264 @@
+'''
+This plotting script is used to plot the effect of external priors on all the parameters involved.
+
+To put all the parameters in the same plot the prior is defined as the improvement from the CMB S4 results we get.
+
+the same is true for the improvement on the y axis parameters
+
+
+
+'''
+
+
+import numpy as np
+import pickle as pickle
+import utils
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+from scipy import integrate
+import scipy.special
+import scipy.interpolate
+from itertools import cycle
+#from matplotlib.colors import colorConverter
+from matplotlib.ticker import MaxNLocator  # needed for tick
+import sys
+from matplotlib.patches import Ellipse
+import n_priors_code.utils as utils
+from palettable.colorbrewer.qualitative import Set1_9
+# ============================================
+# ============================================
+# ============================================
+# TO DO
+# DESI level = -3.21327730556 %
+# ============================================
+# ============================================
+# ============================================
+
+no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
+plot_now = ['omnuh2','w']
+excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
+# omnuh2
+
+# READ DATA
+# DEFINE YOUR FOLDER HERE
+base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
+data_type = 'varying_all'
+run_idx = 7
+lmax = 4499
+lmin = 4
+N_det = 10 ** 6
+# N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
+fsky = 0.75
+# ======
+fid = pickle.load(open(base_dir + 'data/{}/run{}/fid_values.p'.format(data_type, str(run_idx)), "rb"))
+values = pickle.load(open(base_dir + 'data/{}/run{}/grid_values.p'.format(data_type, str(run_idx)), "rb"))
+par_gaps = pickle.load(open(base_dir + 'data/{}/run{}/par_gaps.p'.format(data_type, str(run_idx)), "rb"))
+fisher_mat = np.loadtxt(
[…]

### 2015-12-11 Notes/CMBS4_priors_paper_.tex
add legit DiValentino ref
journal version

@@
 \begin{center}
-\includegraphics{{prior_omnuh2_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_omnuh2.pdf}
 \caption{
@@
 Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario.
-Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}.
+Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}.
 There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
@@
 \begin{center}
-\includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_omnuh2_50.pdf}
 \caption{Same as \reffig{prior_omeganuh2} but with $\ell^{\rm CMB-S4}_{\rm min}>50$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
@@
 \begin{center}
-\includegraphics{{prior_massless_neutrinos_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_massless_neutrinos.pdf}
 \caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of external priors.
@@
 \begin{center}
-\includegraphics{{prior_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_w.pdf}
 \caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. 
@@
 \begin{center}
-\includegraphics{{ellipse_H0_w_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{ellipse_H0_w.pdf}
 \caption{Two dimensional constraint (CMB S4 + Planck) 1$\sigma$ ellipse for $w,H_{0}$ space.

### 2016-01-09 Notes/CMBS4_priors_paper.tex
working on the draft and addressing minor comments

--- 2015-12-11 Notes/CMBS4_priors_paper_.tex
@@
 \begin{abstract}
-Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
-The low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure.
-Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments.
+Planned cosmic microwave background (CMB) experiments will improve what we know about neutrino physics, inflation, and dark energy. 
+The low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized data as well as the reconstructed lensing potential of large scale structure.
+Projected constraints on cosmological parameters are tight, but these can be improved even further with information from external experiments.
 Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage IV (S4) experiment on neutrino and dark energy properties.
@@
 \section{Introduction}\label{sec:intro}
-Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe.
+Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the Universe.
 They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
 The recent past is reassuring.
-For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P, planck-collaboration:2014, 2003ApJS..148..175S, 2000ApJ...545L...5H,2000Natur.404..955D}.
+For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{fixsen:1996,BlueBook2005, bennett:2003}.
 These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck.
-With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
-The same success has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
+With the improved sensitivity we have extended our understanding of the Universe so that we now have solid evidence for a flat geometry and a well tested $\Lambda$CDM model  with percent level constraints on its parameters.
+The same improvement has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
 The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
@@
 
-The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
+The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the Universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
 Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe, their number affects the expansion rate $H(z)$.
@@
 Different neutrino masses consequently lead to different CMB lensing spectra.
-CMB-S4 will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction.
+CMB-S4 will measure small scale temperature and polarization anisotropies with low noise,  improving by at least one order of magnitude the current lensing reconstruction.
 This will turn into a precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.
@@
 The CMB is sensitive also to the properties of dark energy (see, e.g., \cite{2010MNRAS.405.2639J}).
-Thanks to the current generation of experiments we know with extraordinary accuracy its energy density.
+Thanks to the current generation of experiments we know with high accuracy its energy density.
 The challenge for the next generation is to reveal the nature of this mysterious component.
 For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant.
-Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface.
+Dark energy affects the CMB because it alters the Universe's expansion and it consequently changes the distance to the last scattering surface.
 Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing.
@@
 Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models.
-As such, the key question is: What information is needed to reliably conclude that an additional parameter is required.
-The natural first baby step is to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist.
+As such, the key question is: What information is needed to reliably conclude that an additional parameter is required?.
+In our baseline model we decide to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist.
 So our fiducial cosmology is flat $\nu \Lambda$CDM, with assumed
@@
 In this same spirit, we keep the curvature fixed to zero, as the level predicted in the vast majority of inflationary models is of order $10^{-5}$, much too small to have an impact on the parameters we consider.
-Similarly, we fix the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends only on the baryon density.
+Similarly, when we vary $N_{\rm eff}$ we adjust the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends also on the baryon density.
 
@@
 \end{equation}
-where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
+where $\theta^{\ 2}_{\textsc{fwhm}}$ is the so called full width half maximum (FWHM) of the experiment's beam and $s$ represents the instrumental white noise.
 We use a level of noise $s = 0.58$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin. This corresponds to the $N_{\rm det}=10^{6}$ case of \cite{wu:2014}. 
@@
 Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
-Indeed, below the current free-streaming scale, the matter power power spectrum is suppressed in the presence of
+Indeed, below the current free-streaming scale, the matter power spectrum is suppressed in the presence of
 massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
-This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the the sum of the neutrino masses $\sum m_\nu$.
+This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments.
@@
 Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}.
-There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI.
+There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}.
 This is represented by the blue band in \reffig{prior_omeganuh2}.
@@
 A prior on $H_{0}$ helps for a related, but subtle reason.
-The position of the CMB peaks, well constrained by measurement, in a flat universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
+The position of the CMB peaks, well constrained by measurement, in a flat Universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
 . Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
@@
 a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$.
-If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon-borne experiments like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.
+If CMB S4 will not be designed to measure scales $\ell<50$ to avoid difficulties in dealing with atmospheric noise, an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon-borne experiments like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.
 
@@
 
-In the standard model of cosmology, three active neutrinos are thermally produced in the early universe.
+In the standard model of cosmology, three active neutrinos are thermally produced in the early Universe.
 Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation.
@@
 \caption{Two dimensional constraint (CMB S4 + Planck) 1$\sigma$ ellipse for $w,H_{0}$ space.
-The dashed (red) line correspond to values in the plane corresponding to the same distance to the last scattering surface.}
+The dashed (blue) line correspond to values in the plane corresponding to the same distance to the last scattering surface.}
 \label{fig:ellipse_w_H0}
@@
 This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface.
-However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat).
+However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the Universe flat).
 Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
@@
 But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2} and \reffig{ellipse_w_H0}.
-Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST that would tighten the constraints on, say, the baryon density and $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
+Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, the baryon density and $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
 Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.

### 2016-02-05 Notes/response.txt
started response

Major comments:

- In my opinion the authors are over-discussing well known
degeneracies between cosmological parameters adding no new
information. They formulate the key question in the second-to-last
paragraph of the introduction but there is no answer at the end of the
reading. While acknowledging the importance of having priors on some
parameters to better constrain neutrinos and dark energy, I think the
authors should explain or highlight how those priors will be obtained,
i.e., which external experiments are needed and how they will help CMB
Stage-IV. Moreover, additional datasets like galaxies surveys have
already been proved to be useful by a number of papers in the
literature, the authors should highlight what is new in their results.
Following on these concerns, it is not clear to me if the authors have
an idea on how to get those improvements, for example how to get a
better prior on n_s with non-CMB data. I find the discussion simply
speculative. This impacts in particular the abstract, the results and
the conclusion sections.

The key issue here seems to be: How best to address the question: What external information can be used to improve projected constraints by breaking degeneracies? We think the referee is saying there are 2 parts to this answer: (i) quantify how much the added information on other parameters breaks degeneracies and (ii) give a sense of how we could get that information. The referee points out that we did not do a good job explaining the possibilities for (ii) and that at least qualitatively other papers have discussed degeneracies.

- The baseline dataset in all the results is using CMB+BAO which means
already CMB + external information. This is confusing, we are not
improving on CMB then but rather on a combination of CMB and some
galaxy data.

Right; we were not clear that CMB+BAO includes current BAO constraints, from e.g., BOSS. It doesn't make sense to us to exclude things we currently know, so we think
it important to include these constraints as our baseline.

- I cannot understand Figures 1,2,3,4. If the far right is the
expected CMB S4+Planck-Pol+BAO15 then the lines starting point should
correspond to prior/sigma = 1. How is it possible than adding a prior
on a parameter 3 times bigger than the sigma from CMB
S4+Planck-Pol+BAO15 (e.g., on Ho) gives a better constrain? If I think
about it in the likelihood space I am multiplying a distribution with
another one having 3 times its width and I do not expect to have any
impact on the combined result. Also, I do not understand why we should
consider adding a prior which is bigger than the baseline sigma from
CMB S4+Planck-Pol+BAO. What is shown on the right axis?

We agree that this was unecessarily complicated so have changed all the figures to absolute priors on the parameters. The referee's question still remains: why does a 3% prior on H_0 say help if the data itself constrain it to 1%. We stand by our result that it does help. Imagine a 1-D likelihood with one experiment giving a 1% constraint and another [here called "external"] a similar 1% constraint. When combining the 2, the constraint will be tighter, with the new error given by the inverse square root of the sum of the two inverse variances: [1/\sigma_1^2 + 1/\sigma_2^2]^{-1/2}. In the simple example above the external prior is equal to the internal constraint, so the total constraint is reduced by \sqrt{2}. In the example given by the referee [external constraint 3 times looser], there will still be improvement, albeit only by a factor of \sqrt{1 + 1/9}, ~5%. In multi-dimensional space, things are not that simple, but the plots show how things improve. To come back to our first response, we believe this information is extremely important for future planning. While it is true that people have studied degeneracies before, their impact has not been presented in this way. The very fact that the plots are not familiar indicates that we are introducing a new way, a productive way, of thinking about planning. We have added a paragraph in the text that attempts to explain this point more clearly and -- as noted -- have changed the x-axis on all plots to make things less confusing.

- I think that the claim of dramatically improving dark energy is
wrong. That authors themselves show that in the best case scenario
with lots of external information we will improve on w no more than a
factor of 2. More generally I think the discussion on dark energy need
to say more explicitly that the CMB is not the best probe of DE and it
will tell us something only in combination with other data.

Good point. We've toned down the language especially in the abstract and added the sentence: "For this reason CMB provides limited information about dark energy on its own. However CMB plays an important role in dark energy studies when combined with low redshift probes."'

- The references are not adequate in many cases, see minor comments
for details.

- I encourage the authors to review the text:
- - adjectives like “dramatically” and “extremely” are not useful in
giving an idea of an experiment constraining power;

good point; removed these

- - in the abstract the authors mention signal to noise of the CMB
polarization signal and I think they mean data;

got it; thanks

- - in the abstract high redshift large scale structure is wrong, the
CMB is a high redshift signal lensed by low redshift large scale
structure though;

right; we meant "high redshift" compared to other lensing data, but the wording is clearly bad. We've fixed by removing "high redshift"

"'"
- - in the introduction rephrase the 4th sentence. New experiments are
built to have a better sensitivity, it is not a success, it is meant
to happen;

not sure we understand this point

- - in the introduction rephrase the dark energy section following my
comment above.

done; as mentioned above

- Why do the authors use fsky=0.75? CMB S4 is expected to observe
roughly half the sky which, after galaxy removal, should correspond to
fsky ~ 0.4 .

Right; this is aggressive and the true value will be determined, but we chose this to be consistent with Ref. 11. Added a footnote: "The exact specifications of CMB-S4 are to be determined; we choose this value of fsky to afford easy comparison with the results of [11]."

- Keeping the helium fraction fixed is wrong, Yp varies with baryon
density AND Neff.

That's what we did, but the old version did not explain. We changed to:
"Similarly, when we vary $N_{\rm eff}$ we adjust the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends also on the baryon density."'

Minor comments:

- In the introduction: Ref. 4 is WMAP 1year results, in this context
the general WMAP experiment paper Bennett et al. 2003 or the final
mission results Bennet et al. 2012 would be more appropriate. Same
thing for Planck, cite the Blue Book and/or the 2015 Overview papers.
Ref. 5 is Maxima and Ref. 6 is Boomerang, they are not satellites and
are not mentioned in the text.
- ACT and SPT did not follow DASI, they have measured temperature with
ACTPol and SPTPol going for polarization. The first two should go into
the above temperature discussion with references to latest results or
general papers describing the experiments, and the last two should
contribute to the polarization discussion.
- In the introduction the cosmological dependence on neutrinos
properties jumps from number to masses to number again, it is
confusing please rephrase.
- Choose between Universe or universe.
- Define the cosmological parameters listed in the introduction.
- In the introduction “model LCDM” -> “LCDM model”
- Section II, “additional parameter is required” -> “additional
parameter is required?”
- Section II, the natural first “baby” step is not a step. The massive
neutrinos LCDM is the authors baseline model.
- Define FWHM.
- Section IIIA, cite DESI.
- Section IIIA, what do the authors mean with the “current

### 2016-02-05 Notes/CMBS4_priors_paper.tex
working on the draft and addressing minor comments
started response

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,prd,reprint,superscriptaddress]{revtex4-1}
+\documentclass[aps,prd,reprint,superscriptaddress,nofootinbib]{revtex4-1}
 \usepackage{hyperref}
@@
 Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing.
-However, dark energy properties are strongly degenerate with other geometrical parameters such as $H_{0}$. As with the neutrino sector, dark energy constraints from the CMB 
+However, dark energy properties are strongly degenerate with other geometrical parameters such as $H_{0}$.
+For this reason CMB provides limited information about dark energy on its own.
+However CMB plays an important role in dark energy studies when combined with low redshift probes.
+As with the neutrino sector, dark energy constraints from the CMB
 will be improved by external experiments.
@@
  \end{equation}
- where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout.
+ where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout\footnote{The exact specifications of CMB-S4 are to be determined; we choose this value of $f_{\rm sky}$ to afford easy comparison with the results of \cite{wu:2014}.}
 In this work we constrain cosmological parameters with CMB temperature and E-mode polarization together with the reconstructed lensing potential of large scale structure.

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
new images

@@ -222,9 +222,11 @@ for key_y in ['omnuh2']:
         # *93.14 * 1000. to go to M_nu
         line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
+        plt.title(r'$\rm S4~ + ~BAO15$')
 
 
         ax1.minorticks_on()
+        low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 16.94 / 100. * fid['omnuh2'])
         ax1.set_ylim((0.8 * np.amin(new_sigma) * 94. * 1000., 1.1 * np.amax(new_sigma) * 94. * 1000.))
         # ax1.set_xlim((0.1, 3.1))
         ax1.axhline(16.94 / 100. * fid['omnuh2'] * 94. * 1000., alpha=0.4, linewidth=2, label='DESI')

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_new.py

@@ -224,9 +224,14 @@ for key_y in ['omnuh2']:
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
         ax1.minorticks_on()
-        ax1.set_ylim((0.8 * np.amin(new_sigma) * 94. * 1000., 1.1 * np.amax(new_sigma) * 94. * 1000.))
+        # low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 23.24  / 100. * fid['omnuh2'])
+        # print 'lowlim', low_lim
+        print ax1.get_ylim()
+        ax1.set_ylim((0.8 * np.amin(new_sigma)*94. * 1000., 1.1 * np.amax(new_sigma)*94. * 1000.))
+        print ax1.get_ylim()
+        plt.title(r'$\rm S4~ +~ Planck ~ Pol  (\ell<50) + ~BAO15$')
         # ax1.set_xlim((0.1, 3.1))
-        ax1.axhline(16.94 / 100. * fid['omnuh2'] * 94. * 1000., alpha=0.4, linewidth=2, label='DESI')
+        ax1.axhline(23.24 / 100. * fid['omnuh2'] * 94. * 1000., alpha=0.4, linewidth=2, label='DESI')
         ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
         ax1.set_xlabel(r'$\rm{External ~ prior~on~'+ label[key] +'(\%)}$')
         vals = ax1.get_xticks()
@@ -243,6 +248,7 @@ for key_y in ['omnuh2']:
         for i, tick in enumerate(ax2.get_yticks().tolist()):
             new_ticks[i] = str(tick) + r'$\%$'
         ax2.set_yticklabels(new_ticks)
+        print ax1.get_ylim()
 
         # Put snow mass line
         # Put a legend below current axis

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_N_eff_new.py

@@ -241,6 +241,7 @@ for key_y in ['massless_neutrinos']:
         ax1.set_xticklabels([r'{:3.1f}$\%$'.format(x) for x in vals])
         xticks = ax1.xaxis.get_major_ticks()
         xticks[0].label1.set_visible(False)
+        plt.title(r'$\rm S4~ + ~BAO15$')
 
 
         ax1.minorticks_on()

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_w_new.py

@@ -230,6 +230,7 @@ for  key_y in ['w']:
         xticks = ax1.xaxis.get_major_ticks()
         xticks[0].label1.set_visible(False)
 
+        plt.title(r'$\rm S4~ + ~BAO15$')
 
 
 

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
new images

@@ -222,7 +222,6 @@ for key_y in ['omnuh2']:
         # *93.14 * 1000. to go to M_nu
         line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
-        plt.title(r'$\rm S4~ + ~BAO15$')
 
 
         ax1.minorticks_on()

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_new.py

@@ -227,7 +227,7 @@ for key_y in ['omnuh2']:
         # low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 23.24  / 100. * fid['omnuh2'])
         # print 'lowlim', low_lim
         print ax1.get_ylim()
-        ax1.set_ylim((0.8 * np.amin(new_sigma)*94. * 1000., 1.1 * np.amax(new_sigma)*94. * 1000.))
+        ax1.set_ylim((20.24 / 100. * fid['omnuh2'] * 94. * 1000., 1.1 * np.amax(new_sigma)*94. * 1000.))
         print ax1.get_ylim()
         plt.title(r'$\rm S4~ +~ Planck ~ Pol  (\ell<50) + ~BAO15$')
         # ax1.set_xlim((0.1, 3.1))

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
new images

@@ -228,7 +228,7 @@ for key_y in ['omnuh2']:
         low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 16.94 / 100. * fid['omnuh2'])
         ax1.set_ylim((0.8 * np.amin(new_sigma) * 94. * 1000., 1.1 * np.amax(new_sigma) * 94. * 1000.))
         # ax1.set_xlim((0.1, 3.1))
-        ax1.axhline(16.94 / 100. * fid['omnuh2'] * 94. * 1000., alpha=0.4, linewidth=2, label='DESI')
+        ax1.axhline(16.94 / 100. * fid['omnuh2'] * 94. * 1000., xmin=0., xmax=0.35, alpha=0.4, linewidth=2, label='DESI')
         ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
         ax1.set_xlabel(r'$\rm{External ~ prior~on~'+ label[key] +'(\%)}$')
 

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_new.py

@@ -231,12 +231,13 @@ for key_y in ['omnuh2']:
         print ax1.get_ylim()
         plt.title(r'$\rm S4~ +~ Planck ~ Pol  (\ell<50) + ~BAO15$')
         # ax1.set_xlim((0.1, 3.1))
-        ax1.axhline(23.24 / 100. * fid['omnuh2'] * 94. * 1000., alpha=0.4, linewidth=2, label='DESI')
+        ax1.axhline(23.24 / 100. * fid['omnuh2'] * 94. * 1000., xmin=0., xmax=0.35, alpha=0.4, linewidth=2, label='DESI')
         ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
         ax1.set_xlabel(r'$\rm{External ~ prior~on~'+ label[key] +'(\%)}$')
         vals = ax1.get_xticks()
         ax1.set_xticklabels([r'{:3.1f}$\%$'.format(x) for x in vals])
-
+        if key == 'ombh2':
+            ax1.set_xticklabels([r'{:3.2f}$\%$'.format(x) for x in vals])
         xticks = ax1.xaxis.get_major_ticks()
         xticks[0].label1.set_visible(False)
         y1, y2 = ax1.get_ylim()

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_N_eff_new.py

@@ -239,6 +239,8 @@ for key_y in ['massless_neutrinos']:
 
         vals = ax1.get_xticks()
         ax1.set_xticklabels([r'{:3.1f}$\%$'.format(x) for x in vals])
+        if key == 'ombh2':
+            ax1.set_xticklabels([r'{:3.2f}$\%$'.format(x) for x in vals])
         xticks = ax1.xaxis.get_major_ticks()
         xticks[0].label1.set_visible(False)
         plt.title(r'$\rm S4~ + ~BAO15$')

### 2016-02-09 Notes/response.txt
new images, working on the draft and response

@@
 
-Right; we were not clear that CMB+BAO includes current BAO constraints, from e.g., BOSS. It doesn't make sense to us to exclude things we currently know, so we think
-it important to include these constraints as our baseline.
+The referee is right; we were not clear that CMB+BAO includes current BAO constraints, from e.g., BOSS.
+It doesn't make sense to us to exclude things we currently know, so we think it important to include these constraints as our baseline.
+For example that will strongly bias the foracasted performance of CMB S4 when considered without future BAO experiments like DESI.
 
@@
 Good point.
-We've toned down the language especially in the abstract and added the sentence: "For this reason CMB provides limited information about dark energy on its own.
+We agree and we have toned down the language especially in the abstract and added the sentence: "For this reason CMB provides limited information about dark energy on its own.
 However CMB plays an important role in dark energy studies when combined with low redshift probes."'
@@
 
+see below
+
 - I encourage the authors to review the text:
@@
 
-good point; removed these
+good point; removed these.
 
@@
 
-That's what we did, but the old version did not explain. We changed to:
+That's indeed what we did, but the old version of the text did not explain. We changed to:
 "Similarly, when we vary $N_{\rm eff}$ we adjust the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends also on the baryon density."'
@@
 are not mentioned in the text.
+
+Thanks, Fixed with Bennett et al. 2003 and Blue book
+
 - ACT and SPT did not follow DASI, they have measured temperature with
@@
 contribute to the polarization discussion.
+
+We organized the discussion into ground base and satellite experiments. For this reason
+we mentioned ACT and SPT toghether with DASI. WE added references to general papers
+describing actpol and sptpol.
+
 - In the introduction the cosmological dependence on neutrinos
@@
 confusing please rephrase.
+-------------
+
 - Choose between Universe or universe.
+
+Done (Universe)
+
 - Define the cosmological parameters listed in the introduction.
+-------------
+
 - In the introduction “model LCDM” -> “LCDM model”
+
+Done
+
 - Section II, “additional parameter is required” -> “additional
 parameter is required?”
+
+Done
+
 - Section II, the natural first “baby” step is not a step. The massive
 neutrinos LCDM is the authors baseline model.
+
+Good point, we rephrased it.
+
 - Define FWHM.
+
+Done.
+
 - Section IIIA, cite DESI.
-- Section IIIA, what do the authors mean with the “current
+
+Done (http://arxiv.org/abs/1308.0847)
+
+- Section IIIA, what do the authors mean with the “current”
+free-streaming scale?
+
+We removed the word current
+
+- Section IIIA “power power spectrum is suppressed” -> “power spectrum
+is suppressed”
+
+Thanks, done.
+
+- Section IIIA, what is the meaning of “since the active neutrino
+number density are known in the standard model”?
+
+We meant that what is probed by the CMB is the energy density of neutrinos which can be
+turned into a total neutrino mass (in the massive case) by knowing the number density
+E= m n
+We rephrease the sentence cause the presence of the term standard model can cause confusion.
+
+- Section IIIA, “a constraint on the the” -> “a constraint on the”
+
+Done.
+
+- Section IIIA, add a reference to oscillation experiments result.
+
+Done.
+
+- Section IIIA, measuring ell<50 is not a design goal but rather an
+analysis one given that the problem is the atmospheric emission.
+
+We rephrase it --> If CMB S4 will not be designed to measure scales l < 50 to avoid difficulties in dealing with atmospheric noise
+
+- Section IIIA, with satellites need to cite LiteBird and Core/Prism,
+with the balloons Ebex and then Class for large modes from the ground.
+
+Done PRISM(http://arxiv.org/abs/1306.2259) LiteBIRD(http://arxiv.org/abs/1311.2847)
+EBEX (http://arxiv.org/abs/1007.3672) CLASS (http://arxiv.org/abs/1408.4788)
+
+- Fig. 5 caption, there is no red line in the plot
+We fixed it.
+
+- Last part of Section IV, cite LSST.
+
+Done (http://arxiv.org/abs/1211.0310)
+
+- Last part of Section IV, what do the authors mean with surveys
+beyoond DESI and LSST? LSST is beyond DESI and I am not aware of
+post-LSST surveys. Also, galaxy surveys will not help in constraining
+n_s.
+---------------------

### 2016-02-09 Notes/CMBS4_priors_paper.tex
new images, working on the draft and response

@@
 Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models.
-As such, the key question is: What information is needed to reliably conclude that an additional parameter is required?.
+As such, the key question is: What information is needed to reliably conclude that an additional parameter is required?
 In our baseline model we decide to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist.
@@
 This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
-This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments.
+This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).
 
 %==========
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{prior_omnuh2.pdf}
+%\caption{
+%Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. 
+%Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment (+Planck+current BAO); e.g., $\sigma(H_0)_{\rm S4 + Planck Pol + BAO15}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
+%The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI. The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.
+%$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV.}
+%\label{fig:prior_omeganuh2}
+%\end{center}
+%\end{figure}
+
+%==========
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics{prior_omnuh2.pdf}
-\caption{
-Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. 
-Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment (+Planck+current BAO); e.g., $\sigma(H_0)_{\rm S4 + Planck Pol + BAO15}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI. The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.
-$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV.}
-\label{fig:prior_omeganuh2}
+\includegraphics{{prior_omnuh2_hubble_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter, H$_{0}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_omeganuh2-h}
 \end{center}
@@
 
-\reffig{prior_omeganuh2} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors. 
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_omnuh2_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_omeganuh2-omch2}
+\end{center}
+\end{figure}
+%==========
+
+%==========
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_omnuh2_scalar_amp(1)_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $A_{s}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_omeganuh2-50-as}
+\end{center}
+\end{figure}
+%==========
+
+%==========
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_omnuh2_hubble_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $H_{0}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_omeganuh2-50-h}
+\end{center}
+\end{figure}
+%==========
+
+%==========
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_omnuh2_omch2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $\Omega_{c} h^{2}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_omeganuh2-50-omch2}
+\end{center}
+\end{figure}
+%==========
+
+%==========
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_omnuh2_scalar_spectral_index(1)_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $n_{s}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_omeganuh2-50-ns}
+\end{center}
+\end{figure}
+%==========
+
+%==========
+
+\reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} show the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors. 
 Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario.
@@
 There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}.
-This is represented by the blue band in \reffig{prior_omeganuh2}.
-In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2} generalizes the extent to which external information will improve the neutrino mass constraints.
-For example, the red curve shows the improvement as additional information on the Hubble constant is applied.
-The projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.6\%$ (and similarly the values for the other parameters are given below the figure).
-So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.006$ corresponds to 1 on the $x$-axis.
-This would improve the neutrino mass constraint to 17 meV.
-Similarly, introducing an external constraint on the dark matter density equal to its internal constraint of 0.4\% (so again 1 on the $x$-axis) improves the projected neutrino mass constraint to 20 meV.
+This is represented by the blue band in our figures.
+In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} generalize the extent to which external information will improve the neutrino mass constraints.
+For example, in \reffig{prior_omeganuh2-h} the red curve shows the improvement as additional information on the Hubble constant is applied.
+Note that the projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.6\%$ (and similarly the values for the other parameters are given below the figure).
+So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.6\%$ would improve the neutrino mass constraint to 17 meV.
+Similarly, in \reffig{prior_omeganuh2-omch2} introducing an external constraint on the dark matter density equal to its internal constraint of 0.4\% improves the projected neutrino mass constraint to 20 meV.
 It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
@@
 As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
-The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
+The left-hand side of \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
 
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{prior_omnuh2_50.pdf}
-\caption{Same as \reffig{prior_omeganuh2} but with $\ell^{\rm CMB-S4}_{\rm min}>50$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{prior_omnuh2_50.pdf}
+%\caption{Same as \reffig{prior_omeganuh2} but with $\ell^{\rm CMB-S4}_{\rm min}>50$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
+%\label{fig:prior_omeganuh2_tau}
+%\end{center}
+%\end{figure}
 
-\label{fig:prior_omeganuh2_tau}
-\end{center}
-\end{figure}
-
-It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in \reffig{prior_omeganuh2_tau} we limit the modes to $\ell>\ell_{\rm min}=50$. 
-Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed in \reffig{prior_omeganuh2}, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful.
+It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in Fig.\ref{fig:prior_omeganuh2-50-as}-\ref{fig:prior_omeganuh2-50-ns} we limit the modes to $\ell>\ell_{\rm min}=50$.
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).
 
@@
 a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$.
-If CMB S4 will not be designed to measure scales $\ell<50$ to avoid difficulties in dealing with atmospheric noise, an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon-borne experiments like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.
+If CMB S4 will not be designed to measure scales $\ell<50$ to avoid difficulties in dealing with atmospheric noise, an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like LiteBIRD \cite{matsumura:2014}, PIXIE\cite{kogut:2011}, and COrE/PRISM \cite{the-core-collaboration:2011,prism-collaboration:2013}, balloon-borne experiments like SPIDER \cite{filippini:2010}, EBEX \cite{reichborn-kjennerud:2010} or eventually ground experiments like CLASS \cite{essinger-hileman:2014}, will be needed to fully exploit the CMB constraining power.
 
@@
 
-%==========
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics{prior_massless_neutrinos.pdf}
-\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of external priors.
-The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.}
-\label{fig:prior_massless_neutrinos}
+\includegraphics{{prior_massless_neutrinos_ombh2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on the $\Omega _{b} h^{2}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_neff-ombh2}
 \end{center}
 \end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_massless_neutrinos_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_neff-omch2}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics{{prior_massless_neutrinos_scalar_spectral_index(1)_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $n_{s}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_neff-ns}
+\end{center}
+\end{figure}
+
 %==========
 
-\reffig{prior_massless_neutrinos} shows projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on the other seven parameters (the sum of the neutrino masses must be included as a free parameter).
+%==========
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{prior_massless_neutrinos.pdf}
+%\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of external priors.
+The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.}
+%\label{fig:prior_massless_neutrinos}
+%\end{center}
+%\end{figure}
+%==========
+
+\reffig{prior_neff-omch2}, \reffig{prior_neff-ns} and  \reffig{prior_neff-ombh2} show projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on $H_{0}$, $\Omega_{c}h^{2}$ and $\Omega_{b}h^{2}$(the sum of the neutrino masses must be included as a free parameter).
 With no priors on the other parameters, the projected $1$-sigma error is $0.016$, close to a 3-sigma detection of the expected deviation from $N_{\rm eff}=3$.
@@
 The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improve the $N_{\rm eff}$ constraint.
-As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction.
-The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
+As such, \reffig{prior_neff-ombh2} and \reffig{prior_neff-ns} show that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction.
 
@@
 Furthermore CMB lensing probes the growth of structure at late times which is different for different dark energy equations of state.
-%==========
 
@@
 \begin{center}
-\includegraphics{prior_w.pdf}
-\caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. 
-An external prior on $H_{0}$ will be crucial to improve the constraint.} 
-\label{fig:prior_w}
+\includegraphics{{prior_w_hubble_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on $w$ (absolute on the left and and relative to its value on the right) as a function of priors on $H_{0}$.
+The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
+\label{fig:prior_w-h}
 \end{center}
 \end{figure}
+
+%==========
+%
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{prior_w.pdf}
+%\caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. 
+%An external prior on $H_{0}$ will be crucial to improve the constraint.} 
+%\label{fig:prior_w}
+%\end{center}
+%\end{figure}
 %==========
@@
 
-\reffig{prior_w} shows that without any priors, CMB experiments (+BAO15) will do better than current constraints, which hover around 10\%.
+\reffig{prior_w-h} shows that without any priors, CMB experiments (+BAO15) will do better than current constraints, which hover around 10\%.
 However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ by a factor of 2.

### 2016-02-09 Notes/CMBS4_priors_paper.tex
new images, working on the draft and response

@@
 The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the Universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
-Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe, their number affects the expansion rate $H(z)$.
+The sensitivity to $N_{\rm eff}$ is due to its affects on the expansion rate $H(z)$ because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe.
 This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
 The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$.
-The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($\sum m_{\nu}<230$ meV from \cite{planck-collaboration:2014}), neutrinos are still relativistic at the last scattering surface.
+On the other hand, the total mass of neutrinos, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($\sum m_{\nu}<230$ meV from \cite{planck-collaboration:2014}), neutrinos are still relativistic at the last scattering surface.
 However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons.

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
new images

@@ -223,6 +223,7 @@ for key_y in ['omnuh2']:
         line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
+        plt.title(r'$\rm S4~ + ~BAO15$')
 
         ax1.minorticks_on()
         low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 16.94 / 100. * fid['omnuh2'])

### 2016-02-09 Notes/response.txt

@@
 confusing please rephrase.
--------------
+
+We introduce the two main topics of that paragraph (N_eff and M_nu) in the first sentence and we then
+describe one after the other in the remaining. We strengthen the connector hopefully reducing the confusion.
 
@@
 - Define the cosmological parameters listed in the introduction.
--------------
+
+The parameters used in our pipelines are quite standard in the field, and we think that pointing the
+reader who wants to know more to the Planck cosmology paper is probably enough.
 

### 2016-02-09 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_new.py
new images

@@ -195,7 +195,7 @@ for key_y in ['omnuh2']:
 
     sigma_just_CMB_y = (np.sqrt(fisher_inv[fid.keys().index(key_y), fid.keys().index(key_y)]))
 
-    for i, key in enumerate(['hubble','omch2','scalar_spectral_index(1)','scalar_amp(1)']):
+    for i, key in enumerate(['hubble','omch2','scalar_spectral_index(1)','scalar_amp(1)','re_optical_depth']):
         plt.clf()
         plt.close()
         fg = plt.figure(figsize=fig_dims)

### 2016-02-09 Notes/response.txt
typos fixed

@@
 We agree that this was unecessarily complicated so have changed all the figures to absolute priors on the parameters.
+We also explicitly describe the right axis that corresponds to the fractional error (obsolute error on the left axis)
 The referee's question still remains: why does a 3% prior on H_0 say help if the data itself constrain it to 1%.

### 2016-02-09 Notes/CMBS4_priors_paper.tex
typos fixed

@@
 \includegraphics{{prior_omnuh2_scalar_amp(1)_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $A_{s}$.
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $A_{s}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
@@
 \includegraphics{{prior_omnuh2_omch2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $\Omega_{c} h^{2}$.
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
@@
 \begin{center}
-\includegraphics{{prior_omnuh2_scalar_spectral_index(1)_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $n_{s}$.
+\includegraphics{{prior_omnuh2_re_optical_depth_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the optical depth $\tau$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_omeganuh2-50-ns}
+\label{fig:prior_omeganuh2-50-tau}
 \end{center}

### 2016-02-09 Notes/CMBS4_priors_paper.tex
typos fixed

@@
 Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
-Indeed, below the current free-streaming scale, the matter power spectrum is suppressed in the presence of
+Indeed, below the free-streaming scale, the matter power spectrum is suppressed in the presence of
 massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
@@
 %==========
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{prior_omnuh2.pdf}
-%\caption{
-%Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. 
-%Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment (+Planck+current BAO); e.g., $\sigma(H_0)_{\rm S4 + Planck Pol + BAO15}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
-%The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI. The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.
-%$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV.}
-%\label{fig:prior_omeganuh2}
-%\end{center}
-%\end{figure}
-
-%==========
-
 \begin{figure}[htbp]
@@
 %==========
-
 \begin{figure}[htbp]
@@
 
-\reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} show the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors. 
+\reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} show the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}$ and $\Omega_{c}h^{2}$.
 Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario.
@@
 For example, in \reffig{prior_omeganuh2-h} the red curve shows the improvement as additional information on the Hubble constant is applied.
-Note that the projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.6\%$ (and similarly the values for the other parameters are given below the figure).
-So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.6\%$ would improve the neutrino mass constraint to 17 meV.
-Similarly, in \reffig{prior_omeganuh2-omch2} introducing an external constraint on the dark matter density equal to its internal constraint of 0.4\% improves the projected neutrino mass constraint to 20 meV.
+We can see, for example that, imposing {\it external} information with at the level of $\sigma_{H_0}/H_0=0.6\%$ would improve the neutrino mass constraint to 17 meV.
+Similarly, in \reffig{prior_omeganuh2-omch2} adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV.
 It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
@@
 A prior on $H_{0}$ helps for a related, but subtle reason.
-The position of the CMB peaks, well constrained by measurement, in a flat Universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
-. Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
+The position of the CMB peaks, well constrained by measurement, in a flat Universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014}.
+Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
 As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
@@
 
-It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in Fig.\ref{fig:prior_omeganuh2-50-as}-\ref{fig:prior_omeganuh2-50-ns} we limit the modes to $\ell>\ell_{\rm min}=50$.
+Because of the presence of foregrounds, it may be difficult to get the very low-$\ell$ modes from the ground with CMB-S4, so in Fig.\ref{fig:prior_omeganuh2-50-as}-\ref{fig:prior_omeganuh2-50-tau} we limit the CMB S4 modes to $\ell>\ell_{\rm min}=50$.
+For the large scale modes $\ell<\ell_{\rm min}=50$ we use Planck at the level of accuracy forecasted by the Planck Blue Book.
 Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).
@@
 The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB.
-Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so
-a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$.
-If CMB S4 will not be designed to measure scales $\ell<50$ to avoid difficulties in dealing with atmospheric noise, an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like LiteBIRD \cite{matsumura:2014}, PIXIE\cite{kogut:2011}, and COrE/PRISM \cite{the-core-collaboration:2011,prism-collaboration:2013}, balloon-borne experiments like SPIDER \cite{filippini:2010}, EBEX \cite{reichborn-kjennerud:2010} or eventually ground experiments like CLASS \cite{essinger-hileman:2014}, will be needed to fully exploit the CMB constraining power.
+Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$.
+If CMB S4 will not be designed to measure scales $\ell<50$ to avoid difficulties in dealing with atmospheric emission, an external prior on the optical depth will be crucial.
+This can be obtained from future 21 cm surveys \cite{liu:2015} or other CMB experiments, satellite like LiteBIRD \cite{matsumura:2014}, PIXIE\cite{kogut:2011}, and COrE/PRISM \cite{the-core-collaboration:2011,prism-collaboration:2013}, balloon-borne experiments like SPIDER \cite{filippini:2010}, EBEX \cite{reichborn-kjennerud:2010} or eventually ground experiments like CLASS \cite{essinger-hileman:2014}.
 
@@
 
-%==========
-
-%==========
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{prior_massless_neutrinos.pdf}
-%\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of external priors.
-The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.}
-%\label{fig:prior_massless_neutrinos}
-%\end{center}
-%\end{figure}
-%==========
-
 \reffig{prior_neff-omch2}, \reffig{prior_neff-ns} and  \reffig{prior_neff-ombh2} show projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on $H_{0}$, $\Omega_{c}h^{2}$ and $\Omega_{b}h^{2}$(the sum of the neutrino masses must be included as a free parameter).
@@
 \end{figure}
-
-%==========
-%
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{prior_w.pdf}
-%\caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. 
-%An external prior on $H_{0}$ will be crucial to improve the constraint.} 
-%\label{fig:prior_w}
-%\end{center}
-%\end{figure}
-%==========
 

### 2016-02-09 Notes/response.txt
typos fixed

@@
 
-Thanks, Fixed with Bennett et al. 2003 and Blue book
+Thanks, Fixed adding Bennett et al. 2003 and Blue book
 
@@
 We organized the discussion into ground base and satellite experiments. For this reason
-we mentioned ACT and SPT toghether with DASI. WE added references to general papers
+we mentioned ACT and SPT toghether with DASI. We added references to general papers
 describing actpol and sptpol.
@@
 We introduce the two main topics of that paragraph (N_eff and M_nu) in the first sentence and we then
-describe one after the other in the remaining. We strengthen the connector hopefully reducing the confusion.
+describe one after the other in the remaining. We strengthen the connector, hopefully reducing the confusion.
 
@@
 
-Done (Universe)
+Done
 
@@
 The parameters used in our pipelines are quite standard in the field, and we think that pointing the
-reader who wants to know more to the Planck cosmology paper is probably enough.
+reader to the Planck cosmology paper is probably enough.
 
@@
 
-We removed the word current
+Thanks, We removed the word current.
 
@@
 
-We rephrase it --> If CMB S4 will not be designed to measure scales l < 50 to avoid difficulties in dealing with atmospheric noise
+We rephrase it --> If CMB S4 will not be designed to measure scales l < 50 to avoid difficulties in dealing with atmospheric emission
 

### 2016-02-09 Notes/CMBS4_priors_paper.tex
typos fixed

@@
 Much of the improvement that would come on neutrino masses and $w$ from $H_0$ or $\Omega_ch^2$ priors is anticipated to be provided by distance measurements made by DESI.
-But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2} and \reffig{ellipse_w_H0}.
+But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior_omeganuh2-50-h} and \reffig{ellipse_w_H0}.
 Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, the baryon density and $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.

### 2016-02-16 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
new images

@@ -194,15 +194,15 @@ for key_y in ['omnuh2']:
     print key_y
 
     sigma_just_CMB_y = (np.sqrt(fisher_inv[fid.keys().index(key_y), fid.keys().index(key_y)]))
+    fg = plt.figure(figsize=fig_dims)
+    ax1 = plt.subplot2grid((1, 1), (0, 0))
+    ax1.set_color_cycle(Set1_9.mpl_colors)
+    lines = ["-", "--", "-.", ":"]
+    linecycler = cycle(lines)
 
-    for i, key in enumerate(['hubble','omch2','scalar_spectral_index(1)','scalar_amp(1)']):
-        plt.clf()
-        plt.close()
-        fg = plt.figure(figsize=fig_dims)
-        ax1 = plt.subplot2grid((1, 1), (0, 0))
-        ax1.set_color_cycle(Set1_9.mpl_colors)
-        lines = ["-", "--", "-.", ":"]
-        linecycler = cycle(lines)
+    for i, key in enumerate(['hubble','omch2']):
+
+        print key
 
         if key == key_y:
             continue
@@ -211,7 +211,7 @@ for key_y in ['omnuh2']:
         # RELATIVE ERROR ON X IN THE CMB S4 we need relative cause this is how prior are used
         sigma_just_CMB_x = (np.sqrt(fisher_inv[fid.keys().index(key), fid.keys().index(key)]) / np.abs(fid[key]))
         # the abs value abbove is taken to deal with the negative fiducial value of w
-        prior_value = np.linspace(sigma_just_CMB_x / 10., sigma_just_CMB_x * 4.5, 900)
+        prior_value = np.linspace(0.0001, 0.03, 900)
         # print sigma_just_CMB_x
         # sigma_just_CMB_y_percent =sigma_just_CMB_y_percent /fid[key_y]
         # normalize_x = prior_value / sigma_just_CMB_x  # prior respect to actual error
@@ -223,44 +223,44 @@ for key_y in ['omnuh2']:
         line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
-        plt.title(r'$\rm S4~ + ~BAO15$')
+    plt.title(r'$\rm S4~ + ~BAO15$')
 
[…]

### 2016-02-16 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_tau.py

@@ -201,7 +201,7 @@ for key_y in ['omnuh2']:
     linecycler = cycle(lines)
 
 
-    for i, key in enumerate(['hubble','omch2','scalar_spectral_index(1)','scalar_amp(1)']):
+    for i, key in enumerate(['re_optical_depth']):
 
         if key == key_y:
             continue
@@ -210,7 +210,7 @@ for key_y in ['omnuh2']:
         # RELATIVE ERROR ON X IN THE CMB S4 we need relative cause this is how prior are used
         sigma_just_CMB_x = (np.sqrt(fisher_inv[fid.keys().index(key), fid.keys().index(key)]) / np.abs(fid[key]))
         # the abs value abbove is taken to deal with the negative fiducial value of w
-        prior_value = np.linspace(0.0001, 0.025, 900)
+        prior_value = np.linspace(0.001, 0.08, 900)
         # print sigma_just_CMB_x
         # sigma_just_CMB_y_percent =sigma_just_CMB_y_percent /fid[key_y]
         # normalize_x = prior_value / sigma_just_CMB_x  # prior respect to actual error
@@ -226,7 +226,7 @@ for key_y in ['omnuh2']:
     # low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 23.24  / 100. * fid['omnuh2'])
     # print 'lowlim', low_lim
     print ax1.get_ylim()
-    ax1.set_ylim((12, 35))
+    ax1.set_ylim((18, 32))
     # ax1.set_ylim(0,1)
     print ax1.get_ylim()
     plt.title(r'$\rm S4~ +~ Planck ~ Pol  (\ell<50) + ~BAO15$')
@@ -262,7 +262,7 @@ for key_y in ['omnuh2']:
     # ============================================
 
     # ============================================
-    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_{}_notau_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y),lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_{}_tau_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y),lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                 format='pdf', bbox_inches='tight')
     plt.clf()
 

### 2016-02-16 n_priors_code/plotting/plot_prior_paper_N_eff_2016.py

@@ -6,9 +6,7 @@ To put all the parameters in the same plot the prior is defined as the improveme
 the same is true for the improvement on the y axis parameters
 
 
-JAN 2016
 
-modify to absolute uncertaintes with the goal of improving clarity.
 '''
 
 
@@ -32,13 +30,13 @@ from palettable.colorbrewer.qualitative import Set1_9
 # ============================================
 # ============================================
 # TO DO
-# DESI LEVEL = 16.9%
+# DESI level 0.51 %
 # ============================================
 # ============================================
 # ============================================
 
 no_lcdm_parameters = ['massless_neutrinos', 'w', 'omnuh2']
-plot_now = ['omnuh2']
+plot_now = ['omnuh2', 'massless_neutrinos']
 excluded_parameters = list(set(no_lcdm_parameters) - set(plot_now))
 # omnuh2
 
@@ -48,7 +46,7 @@ base_dir = '/home/manzotti/n_eff-dependence-on-prior/n_priors_code/'
 data_type = 'varying_all'
 run_idx = 7
 lmax = 4499
-lmin = 50
+lmin = 4
 N_det = 10 ** 6
 # N_phi_l = np.loadtxt('data/noise/wu_cdd_noise_5.txt')
 fsky = 0.75
@@ -190,27 +188,28 @@ fisher_inplace = fisher_mat.copy()
 
 # CYCLE ON PARAMETERS (KEYS HERE)
 
-for key_y in ['omnuh2']:
+for key_y in ['massless_neutrinos']:
     print key_y
-
-    sigma_just_CMB_y = (np.sqrt(fisher_inv[fid.keys().index(key_y), fid.keys().index(key_y)]))
     fg = plt.figure(figsize=fig_dims)
     ax1 = plt.subplot2grid((1, 1), (0, 0))
     ax1.set_color_cycle(Set1_9.mpl_colors)
     lines = ["-", "--", "-.", ":"]
     linecycler = cycle(lines)
+    sigma_just_CMB_y = (np.sqrt(fisher_inv[fid.keys().index(key_y), fid.keys().index(key_y)]))
 
-
-    for i, key in enumerate(['hubble','omch2','scalar_spectral_index(1)','scalar_amp(1)']):
-
-        if key == key_y:
+    for i, key in enumerate(['omch2','ombh2','scalar_spectral_index(1)']):
[…]

### 2016-02-16 n_priors_code/utils.py

@@ -202,7 +202,7 @@ def return_simgax_y_prior(fid, fisher, x, y, prior_val):
 
     '''
 
-    print "CHECK"
+    # print "CHECK"
     if isinstance(prior_val, (int, long, float)):
         assert (fid.has_key(x))
         assert (fid.has_key(y))
@@ -228,7 +228,7 @@ def return_simgax_y_prior(fid, fisher, x, y, prior_val):
                 (prior * fid[y]) ** 2
 
             # print prior,y
-            print np.linalg.inv(fisher1).diagonal()/np.linalg.inv(fisher).diagonal()
+            # print np.linalg.inv(fisher1).diagonal()/np.linalg.inv(fisher).diagonal()
             # print ''
             sigma_x_prior[i] = np.sqrt(np.linalg.inv(fisher1)[fid.keys().index(x), fid.keys().index(x)])
             # print 'fisher', fisher[0,0]

### 2016-02-16 n_priors_code/plotting/plot_prior_paper_w_new.py
new images

@@ -193,7 +193,7 @@ for  key_y in ['w']:
 
     sigma_just_CMB_y = (np.sqrt(fisher_inv[fid.keys().index(key_y), fid.keys().index(key_y)]))
 
-    for i, key in enumerate(['omnuh2','hubble','omch2']):
+    for i, key in enumerate(['hubble']):
 
         if key == key_y:
             continue
@@ -210,7 +210,7 @@ for  key_y in ['w']:
         # RELATIVE ERROR ON X IN THE CMB S4 we need relative cause this is how prior are used
         sigma_just_CMB_x = (np.sqrt(fisher_inv[fid.keys().index(key), fid.keys().index(key)]) / np.abs(fid[key]))
         # the abs value abbove is taken to deal with the negative fiducial value of w
-        prior_value = np.linspace(sigma_just_CMB_x / 10., sigma_just_CMB_x * 4.5, 900)
+        prior_value = np.linspace(0.0001, 0.08, 900)
         # print sigma_just_CMB_x
         # sigma_just_CMB_y_percent =sigma_just_CMB_y_percent /fid[key_y]
         # compute the new sigma on y given the prior

### 2016-02-16 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
new images

@@ -200,7 +200,7 @@ for key_y in ['omnuh2']:
     lines = ["-", "--", "-.", ":"]
     linecycler = cycle(lines)
 
-    for i, key in enumerate(['hubble','omch2']):
+    for i, key in enumerate(['hubble','omch2','scalar_spectral_index(1)','scalar_amp(1)']):
 
         print key
 
@@ -211,7 +211,7 @@ for key_y in ['omnuh2']:
         # RELATIVE ERROR ON X IN THE CMB S4 we need relative cause this is how prior are used
         sigma_just_CMB_x = (np.sqrt(fisher_inv[fid.keys().index(key), fid.keys().index(key)]) / np.abs(fid[key]))
         # the abs value abbove is taken to deal with the negative fiducial value of w
-        prior_value = np.linspace(0.0001, 0.03, 900)
+        prior_value = np.linspace(0.0001, 0.025, 900)
         # print sigma_just_CMB_x
         # sigma_just_CMB_y_percent =sigma_just_CMB_y_percent /fid[key_y]
         # normalize_x = prior_value / sigma_just_CMB_x  # prior respect to actual error
@@ -228,10 +228,12 @@ for key_y in ['omnuh2']:
     ax1.minorticks_on()
     low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 16.94 / 100. * fid['omnuh2'])
     ax1.set_ylim((0.8 * np.amin(new_sigma) * 94. * 1000., 1.1 * np.amax(new_sigma) * 94. * 1000.))
+    ax1.set_ylim((12, 28))
+
     # ax1.set_xlim((0.1, 3.1))
     ax1.axhline(16.94 / 100. * fid['omnuh2'] * 94. * 1000., xmin=0., xmax=0.35, alpha=0.4, linewidth=2, label='DESI')
     ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
-    ax1.set_xlabel(r'$\rm{External ~ prior~on~'+ label[key] +'(\%)}$')
+    ax1.set_xlabel(r'$\rm{External ~ prior~on~(\%)}$')
 
     vals = ax1.get_xticks()
     ax1.set_xticklabels([r'{:3.1f}$\%$'.format(x) for x in vals])

### 2016-02-16 Notes/CMBS4_priors_paper.tex

@@
 \begin{center}
-\includegraphics{{prior_omnuh2_hubble_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{{prior_omnuh2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter, H$_{0}$.
@@
 \begin{center}
-\includegraphics{{prior_omnuh2_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
@@
 \begin{center}
-\includegraphics{{prior_omnuh2_scalar_amp(1)_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{{prior_omnuh2_tau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $A_{s}$.
@@
 %==========
-
-%==========
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{{prior_omnuh2_hubble_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $H_{0}$.
+%
+%%==========
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{{prior_omnuh2_hubble_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $H_{0}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_omeganuh2-50-h}
-\end{center}
-\end{figure}
-%==========
-
-%==========
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{{prior_omnuh2_omch2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
+%\label{fig:prior_omeganuh2-50-h}
+%\end{center}
+%\end{figure}
+%%==========
+%
+%
+%%==========
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{{prior_omnuh2_omch2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_omeganuh2-50-omch2}
-\end{center}
-\end{figure}
-%==========
-
-%==========
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{{prior_omnuh2_re_optical_depth_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the optical depth $\tau$.
+%\label{fig:prior_omeganuh2-50-omch2}
+%\end{center}
+%\end{figure}
+%%==========
+%
+%%==========
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{{prior_omnuh2_re_optical_depth_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the optical depth $\tau$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_omeganuh2-50-tau}
-\end{center}
-\end{figure}
-%==========
+%\label{fig:prior_omeganuh2-50-tau}
+%\end{center}
+%\end{figure}
+%%==========
+%
 
@@
 \begin{center}
-\includegraphics{{prior_massless_neutrinos_ombh2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{{prior_massless_neutrinos_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on the $\Omega _{b} h^{2}$.
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{{prior_massless_neutrinos_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{{prior_massless_neutrinos_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+%\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_neff-omch2}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{{prior_massless_neutrinos_scalar_spectral_index(1)_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $n_{s}$.
+%\label{fig:prior_neff-omch2}
+%\end{center}
+%\end{figure}
+%
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{{prior_massless_neutrinos_scalar_spectral_index(1)_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+%\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $n_{s}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_neff-ns}
-\end{center}
-\end{figure}
+%\label{fig:prior_neff-ns}
+%\end{center}
+%\end{figure}
 

### 2016-02-22 Notes/response.txt
working on response

@@
 
-We agree that this was unecessarily complicated so have changed all the figures to absolute priors on the parameters.
-We also explicitly describe the right axis that corresponds to the fractional error (obsolute error on the left axis)
-The referee's question still remains: why does a 3% prior on H_0 say help if the data itself constrain it to 1%.
+We agree that this was unnecessarily complicated so have changed all the figures to absolute priors on the parameters.
+We also explicitly include the the fractional error on right axis in addition to the absolute error on the left axis.
+The referee's question still remains: why does a 3% prior on H_0, say, help if the data itself constrain it to 1%.
 We stand by our result that it does help.
@@
 While it is true that people have studied degeneracies before, their impact has not been presented in this way.
-The very fact that the plots are not familiar indicates that we are introducing a new way, a productive way, of thinking about planning.
+The very fact that the plots are not familiar indicates that we are introducing a new way, we think a productive way, of thinking about planning.
 We have added a paragraph in the text that attempts to explain this point more clearly and -- as noted -- have changed the x-axis on all plots to make things less confusing.
@@
 
-"'"
 - - in the introduction rephrase the 4th sentence. New experiments are
@@
 
-We organized the discussion into ground base and satellite experiments. For this reason
-we mentioned ACT and SPT toghether with DASI. We added references to general papers
-describing actpol and sptpol.
+We were not focused so much on the experiments but the achievements that feed into the projections in this paper.
+First, there was detection of polarization; then measurement of small scale anisotropies that yield the damping tail and the lensing maps.
+Reworded taking out explicit mention of the experiments to be clearer.
 
@@
 
-We introduce the two main topics of that paragraph (N_eff and M_nu) in the first sentence and we then
-describe one after the other in the remaining. We strengthen the connector, hopefully reducing the confusion.
+We introduce the two main topics of that paragraph (N_eff and M_nu) in the first sentence and we then describe one after the other in the remaining.
+We strengthen the connector, hopefully reducing the confusion.
 
@@
 
-The parameters used in our pipelines are quite standard in the field, and we think that pointing the
-reader to the Planck cosmology paper is probably enough.
+Done
 
@@
 
-We meant that what is probed by the CMB is the energy density of neutrinos which can be
-turned into a total neutrino mass (in the massive case) by knowing the number density
-E= m n
-We rephrease the sentence cause the presence of the term standard model can cause confusion.
+We meant that the number density of active neutrinos is a robust prediction emerging from the Standard Model of particle physics. Capitalized to make a little clearer
 
@@
 - Last part of Section IV, what do the authors mean with surveys
-beyoond DESI and LSST? LSST is beyond DESI and I am not aware of
+beyond DESI and LSST? LSST is beyond DESI and I am not aware of
 post-LSST surveys. Also, galaxy surveys will not help in constraining
 n_s.
+
+Part of the point here is to motivate surveys that extract more of the information on the sky than will DESI and LSST.
+We agree with the referee that such surveys [galaxy or otherwise] are not currently planned, but we think the arguments here can help motivate people to start planning them.
+Already, several of the plots in this paper were shown at the NRAO 2020 Futures meeting [https://science.nrao.edu/science/meetings/2015/2020futures/program]
+
 ---------------------

### 2016-02-22 Notes/CMBS4_priors_paper.tex
working on response

@@
 With the improved sensitivity we have extended our understanding of the Universe so that we now have solid evidence for a flat geometry and a well tested $\Lambda$CDM model  with percent level constraints on its parameters.
-The same improvement has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
+Similar progress has characterized ground-based CMB experiments, where the first detection of polarization anisotropy \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail~\cite{2011ApJ...739...52D,2011ApJ...743...28K} and of the projected gravitational potential~\cite{2011PhRvL.107b1301D,2012ApJ...756..142V}.
+
 The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
@@
 The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the Universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
-The sensitivity to $N_{\rm eff}$ is due to its affects on the expansion rate $H(z)$ because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe.
+The sensitivity to $N_{\rm eff}$ is due to its effects on the expansion rate $H(z)$ because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe.
 This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein).
@@
 Here we study the dependance of projected constraints from CMB on the external priors assumed.
-For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors.
+The canonical set of parameters is: cold dark matter density $\Omega_ch^2$; baryon density $\Omega_bh^2$; amplitude of primordial perturbations $A_s$; slope of primordial spectrum  $n_s$; optical depth to the last scattering surface $\tau$; Hubble constant $H_0$; and sum of the neutrino masses $\sum m_\nu$.
+Together with extensions (the number of relativistic degrees of freedom $N_{\rm eff}$ and dark energy equation of state $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors.
 This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.
@@
 massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
-This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
+This effect is probed by CMB lensing and, since the active neutrino number densities are known in the Standard Model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).
@@
 
-\reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} show the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}$ and $\Omega_{c}h^{2}$.
+\reffig{prior_omeganuh2-h} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}, \Omega_{c}h^{2}, n_s$ and $A_s$. 
 Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario.
@@
 This is represented by the blue band in our figures.
-In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} generalize the extent to which external information will improve the neutrino mass constraints.
-For example, in \reffig{prior_omeganuh2-h} the red curve shows the improvement as additional information on the Hubble constant is applied.
-We can see, for example that, imposing {\it external} information with at the level of $\sigma_{H_0}/H_0=0.6\%$ would improve the neutrino mass constraint to 17 meV.
+In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} generalizes the extent to which external information will improve the neutrino mass constraints.
+For example, in \reffig{prior_omeganuh2-h} the red solid curve shows the improvement as additional information on the Hubble constant is applied.
+We can see, for example that, imposing {\it external} information at the level of $\sigma_{H_0}/H_0=0.5\%$ would improve the neutrino mass constraint to 17 meV.
 Similarly, in \reffig{prior_omeganuh2-omch2} adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV.

### 2016-02-22 Notes/response.txt
SD semi-final response

@@
 
-The key issue here seems to be: How best to address the question: What external information can be used to improve projected constraints by breaking degeneracies?
+The key issue raised here seems to be: How best to address the question: What external information can be used to improve projected constraints by breaking degeneracies?
 We think the referee is saying there are 2 parts to this answer: (i) quantify how much the added information on other parameters breaks degeneracies and (ii) give a sense of how we could get that information.
-The referee points out that we did not do a good job explaining the possibilities for (ii) and that at least qualitatively other papers have discussed degeneracies.
+The referee points out that we did not do a good job explaining the possibilities for (ii) and that for (i) at least qualitatively other papers have discussed degeneracies.
+We think this paper does advance the discussion of how degeneracies impact future constraints.
+The plots in Figs.
+1-4 are a new way of answering this question and we think quite useful.
+We thought it important to explain the degeneracies in the text for completeness, but we agree with the referee that these explanations are not new.
+It is the quantitative information in the plots that is new.
+
+The second issue is explaining how to get this information.
+In the case of mnu, we have added text about the importance of getting \tau from either 21 cm reionization measurements or low-l cmb polarization experiments.
+For N_eff, we point out that high-z surveys obtain more Fourier modes in the linear regime, and therefore the range of modes [k_min,k_max] that are measured precisely can be widened.
+It is this widening that will improve the constraints on n_s [and therefore N_eff].
+We now emphasize that z>1 surveys are crucial for this.
+H_0 is clearer: there are already a number of efforts underway to improve direct measurements of H_0, so the plots here are useful in that they lend further importance to those measurements.
 
@@
 It doesn't make sense to us to exclude things we currently know, so we think it important to include these constraints as our baseline.
-For example that will strongly bias the foracasted performance of CMB S4 when considered without future BAO experiments like DESI.
+For example that will strongly bias the forecasted performance of CMB S4 when considered without future BAO experiments like DESI.
+We’ve added a line to make clear why we are including current BAO.
 

### 2016-02-22 Notes/CMBS4_priors_paper.tex
SD semi-final response

@@
 
-We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
+We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization, as this is or will shortly be available.
 For current BAO data (labeled BAO15), we follow \cite{allison:2015} and include 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
@@
 \includegraphics{{prior_omnuh2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter, H$_{0}$.
+\caption{Projected constraints on the sum of the neutrino masses (absolute on the left and relative on the right) from CMB-S4 and current BAO measurements as a function of priors on the other cosmological parameters.
+The {\it internal} (marginalized constraints from CMB-S4 and current BAO) constraints on each parameter are given as $\sigma_{\rm pipeline}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
@@
 
-%==========
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
+%%==========
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_omeganuh2-omch2}
-\end{center}
-\end{figure}
-%==========
+%\label{fig:prior_omeganuh2-omch2}
+%\end{center}
+%\end{figure}
+%%==========
 
@@
 \begin{center}
-\includegraphics{{prior_omnuh2_tau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $A_{s}$.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_omeganuh2-50-as}
+\includegraphics{{prior_omnuh2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75_test}.pdf}
+\caption{Same as Fig.~1, but with $l<50$ modes excluded from CMB-S4, instead coming from Planck.
+Note the importance of obtaining information on the optical depth $\tau$ if the low-$l$ modes are not measured accurately.}
+\label{fig:prior-50}
 \end{center}
@@
 We can see, for example that, imposing {\it external} information at the level of $\sigma_{H_0}/H_0=0.5\%$ would improve the neutrino mass constraint to 17 meV.
-Similarly, in \reffig{prior_omeganuh2-omch2} adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV.
 It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
+Similarly, adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV.
+These are multi-dimensional examples of the familiar fact that adding two sets of equal constraints (in this case internal to S4+BAO15 and external) reduces the overall error by a factor of $\sqrt{2}$.
 
-The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density.
+The physical reason underlying the improvement from $\Omega_ch^2$ is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density.
 In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density.
@@
 As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
-The left-hand side of \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
+The left-hand side of \reffig{prior_omeganuh2-h} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
 
@@
 
-Because of the presence of foregrounds, it may be difficult to get the very low-$\ell$ modes from the ground with CMB-S4, so in Fig.\ref{fig:prior_omeganuh2-50-as}-\ref{fig:prior_omeganuh2-50-tau} we limit the CMB S4 modes to $\ell>\ell_{\rm min}=50$.
+Because of the presence of foregrounds, it may be difficult to get the very low-$\ell$ modes from the ground with CMB-S4, so in \reffig{prior-50}, we limit the CMB S4 modes to $\ell>\ell_{\rm min}=50$.
 For the large scale modes $\ell<\ell_{\rm min}=50$ we use Planck at the level of accuracy forecasted by the Planck Blue Book.
-Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).
+Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be important.
+% (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).
 
@@
 As such, \reffig{prior_neff-ombh2} and \reffig{prior_neff-ns} show that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction.
+Besides futuristic high redshift 21 cm ideas, the best bet to improve the constraints on $n_s$ are probes that push to redshift $z>1$ such as the Lyman alpha forest.
+There are many more linear Fourier modes available at high $z$ that can be used to obtain the larger level arm from large to small scales that is essential to extract high precision measurements of $n_s$.
 
@@
 Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub-percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
-Furthermore, if the final design of CMB S4 will not include measurements on large scales ($\ell<50$), external constraints on the primordial amplitude $A_{s}$ will also be useful. 
-\item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$ and the baryon energy density $\Omega_b h^2$.
+Furthermore, if the final design of CMB S4 does not include measurements on large scales ($\ell<50$), external constraints from 21 cm experiments or all-sky CMB experiments on the optical depth $\tau$ will become very important.% powerful.
+\item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$.
+These are most likely to come from surveys that measure a large fraction of the $z>1$ linear Fourier modes.
 \item[$w$] In this case the improvement will mainly come from external priors on $H_{0}$.
@@
 But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior_omeganuh2-50-h} and \reffig{ellipse_w_H0}.
-Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, the baryon density and $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
+Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
 Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.

### 2016-02-24 Notes/response.txt
very minor changes

@@
 1-4 are a new way of answering this question and we think quite useful.
-We thought it important to explain the degeneracies in the text for completeness, but we agree with the referee that these explanations are not new.
+We thought it was important to explain the degeneracies in the text for completeness, but we agree with the referee that these explanations are not new.
 It is the quantitative information in the plots that is new.

### 2016-02-24 Notes/CMBS4_priors_paper.tex
very minor changes

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,prd,reprint,superscriptaddress,nofootinbib]{revtex4-1}
+\documentclass[aps,prd,reprint,superscriptaddress,nofootinbib,floatfix]{revtex4-1}
 \usepackage{hyperref}
@@
 
-%\newcommand\sd[1]{{\bf SD: #1}}
 \newcommand{\sd}[1]{\textcolor{red}{[{\bf SD}: #1]}}
@@
 
-%\newcommand\refsec[1]{\S\ref{sec:#1}}
-%\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
 \newcommand\refsec[1]{\S\ref{sec:#1}}
@@
 \newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
-%\newcommand{\apj}{ApJ}
 \newcommand{\physrep}{Physics Reports}
@@
 \newcommand{\mnras}{MNRAS}
-%\newcommand{\prd}{Phys. Rev. D}
 \newcommand{\physrev}{Phys. Rev.}
@@
 \section{Results \label{sec:results}}
-In this section we present our main results. We will start from $\nu \Lambda$CDM and then explore extensions to this model by considering constraints on $N_{\rm eff}$ and then $w$.
+In this section we present our main results. Starting from $\nu \Lambda$CDM we will then explore extensions to this model by considering constraints on $N_{\rm eff}$ and $w$.
 
@@
 massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
-This effect is probed by CMB lensing and, since the active neutrino number densities are known in the Standard Model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
+This effect is probed by CMB lensing and, since the active neutrino number densities are known in the Standard Model, it can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).
@@
 %==========
-
-%%==========
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-%\label{fig:prior_omeganuh2-omch2}
-%\end{center}
-%\end{figure}
-%%==========
-
 %==========
@@
 \begin{center}
-\includegraphics{{prior_omnuh2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75_test}.pdf}
+\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
 \caption{Same as Fig.~1, but with $l<50$ modes excluded from CMB-S4, instead coming from Planck.
@@
 %==========
-%
-%%==========
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{{prior_omnuh2_hubble_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter $H_{0}$.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-%\label{fig:prior_omeganuh2-50-h}
-%\end{center}
-%\end{figure}
-%%==========
-%
-%
-%%==========
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{{prior_omnuh2_omch2_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-%\label{fig:prior_omeganuh2-50-omch2}
-%\end{center}
-%\end{figure}
-%%==========
-%
-%%==========
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{{prior_omnuh2_re_optical_depth_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-%\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the optical depth $\tau$.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-%\label{fig:prior_omeganuh2-50-tau}
-%\end{center}
-%\end{figure}
-%%==========
-%
 
@@
 There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}.
-This is represented by the blue band in our figures.
+This is represented by the light blue band in our figures.
 In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} generalizes the extent to which external information will improve the neutrino mass constraints.
@@
 Similarly, adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV.
-These are multi-dimensional examples of the familiar fact that adding two sets of equal constraints (in this case internal to S4+BAO15 and external) reduces the overall error by a factor of $\sqrt{2}$.
+These are multi-dimensional examples of the familiar fact that adding two sets of equal and indipendent constraints (in this case internal to S4+BAO15 and external) reduces the overall error by a factor of $\sqrt{2}$.
 
@@
 
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{prior_omnuh2_50.pdf}
-%\caption{Same as \reffig{prior_omeganuh2} but with $\ell^{\rm CMB-S4}_{\rm min}>50$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
-%\label{fig:prior_omeganuh2_tau}
-%\end{center}
-%\end{figure}
-
 Because of the presence of foregrounds, it may be difficult to get the very low-$\ell$ modes from the ground with CMB-S4, so in \reffig{prior-50}, we limit the CMB S4 modes to $\ell>\ell_{\rm min}=50$.
@@
 Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be important. 
-% (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).
-
 Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
@@
 \includegraphics{{prior_massless_neutrinos_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on the $\Omega _{b} h^{2}$.
+\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega _{b} h^{2}$, $\Omega _{c} h^{2}$ and $n_{s}$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-\label{fig:prior_neff-ombh2}
+\label{fig:prior_neff}
 \end{center}
@@
 
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{{prior_massless_neutrinos_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-%\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-%\label{fig:prior_neff-omch2}
-%\end{center}
-%\end{figure}
-%
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics{{prior_massless_neutrinos_scalar_spectral_index(1)_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
-%\caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $n_{s}$.
-The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
-%\label{fig:prior_neff-ns}
-%\end{center}
-%\end{figure}
-
-\reffig{prior_neff-omch2}, \reffig{prior_neff-ns} and  \reffig{prior_neff-ombh2} show projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on $H_{0}$, $\Omega_{c}h^{2}$ and $\Omega_{b}h^{2}$(the sum of the neutrino masses must be included as a free parameter).
+\reffig{prior_neff} shows projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on $H_{0}$, $\Omega_{c}h^{2}$ and $\Omega_{b}h^{2}$ (the sum of the neutrino masses must be included as a free parameter).
 With no priors on the other parameters, the projected $1$-sigma error is $0.016$, close to a 3-sigma detection of the expected deviation from $N_{\rm eff}=3$.
@@
 The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improve the $N_{\rm eff}$ constraint.
-As such, \reffig{prior_neff-ombh2} and \reffig{prior_neff-ns} show that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction.
+As such, the blue and green dashed lines in \reffig{prior_neff} show that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction.
 Besides futuristic high redshift 21 cm ideas, the best bet to improve the constraints on $n_s$ are probes that push to redshift $z>1$ such as the Lyman alpha forest.
@@
 \reffig{prior_w-h} shows that without any priors, CMB experiments (+BAO15) will do better than current constraints, which hover around 10\%.
-However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ by a factor of 2.
+However, an external an external prior on $H_{0}$ two times more accurate than CMB S4 (i.e.
+$1\%$) can improve the CMB constraint on $w$ by a factor of two.
 This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface.
@@
 Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
-Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint on $w$ by a factor of two.
 
@@
 Much of the improvement that would come on neutrino masses and $w$ from $H_0$ or $\Omega_ch^2$ priors is anticipated to be provided by distance measurements made by DESI.
-But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior_omeganuh2-50-h} and \reffig{ellipse_w_H0}.
+But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior-50}, \reffig{prior_neff} and \reffig{prior_w-h}.
 Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.

### 2016-02-24 Notes/CMBS4_priors_paper.tex
very minor changes

@@
 
-\reffig{prior_neff} shows projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on $H_{0}$, $\Omega_{c}h^{2}$ and $\Omega_{b}h^{2}$ (the sum of the neutrino masses must be included as a free parameter).
+\reffig{prior_neff} shows projections for how well CMB-S4 will do at measuring $N_{\rm eff}$ as a function of priors on $H_{0}$, $\Omega_{b}h^{2}$ and $\Omega_{c}h^{2}$ (the sum of the neutrino masses must be included as a free parameter).
 With no priors on the other parameters, the projected $1$-sigma error is $0.016$, close to a 3-sigma detection of the expected deviation from $N_{\rm eff}=3$.
@@
 Much of the improvement that would come on neutrino masses and $w$ from $H_0$ or $\Omega_ch^2$ priors is anticipated to be provided by distance measurements made by DESI.
-But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior-50}, \reffig{prior_neff} and \reffig{prior_w-h}.
-Indeed, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
+But, additional $H_0$ measurements beyond DESI would help, as evidenced by the sharply falling curves in the left-most regions of \reffig{prior_omeganuh2-h}, \reffig{prior-50}, and \reffig{prior_w-h}.
+Furthermore, perhaps the most tantalizing finding is that future surveys beyond DESI and LSST \cite{lsst-dark-energy-science-collaboration:2012} that would tighten the constraints on, say, $n_s$ could indirectly lead to constraints in $N_{\rm eff}$ much less than a percent.
 Given the power of these constraints to probe physics beyond the Standard Model, such surveys seem particularly interesting to pursue.

### 2016-02-25 Notes/CMBS4_priors_paper.tex
ready for submission

@@
 \begin{center}
-\includegraphics{{prior_omnuh2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_omnuh2_lmin4}
 \caption{Projected constraints on the sum of the neutrino masses (absolute on the left and relative on the right) from CMB-S4 and current BAO measurements as a function of priors on the other cosmological parameters.
@@
 \begin{center}
-\includegraphics{{prior_omnuh2_notau_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_omnuh2_l50}
 \caption{Same as Fig.~1, but with $l<50$ modes excluded from CMB-S4, instead coming from Planck.
@@
 \begin{center}
-\includegraphics{{prior_massless_neutrinos_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_massless_neutrinos}
 \caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega _{b} h^{2}$, $\Omega _{c} h^{2}$ and $n_{s}$.
@@
 \begin{center}
-\includegraphics{{prior_w_hubble_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
+\includegraphics{prior_w_hubble}
 \caption{Constraints on $w$ (absolute on the left and and relative to its value on the right) as a function of priors on $H_{0}$.

### 2016-03-12 Notes/CMBS4_priors_paper.tex

@@
 In this same spirit, we keep the curvature fixed to zero, as the level predicted in the vast majority of inflationary models is of order $10^{-5}$, much too small to have an impact on the parameters we consider.
-Similarly, when we vary $N_{\rm eff}$ we adjust the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends also on the baryon density.
+When we vary $N_{\rm eff}$ we adjust the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends also on the baryon density.
 
@@
 massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
-This effect is probed by CMB lensing and, since the active neutrino number densities are known in the Standard Model, it can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
+This effect is probed by CMB lensing and, considering the robust prediction from particle physics Standard Model for the number density of active neutrinos, it can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$.
 This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).
@@
 \reffig{prior_w-h} shows that without any priors, CMB experiments (+BAO15) will do better than current constraints, which hover around 10\%.
-However, an external an external prior on $H_{0}$ two times more accurate than CMB S4 (i.e.
+However, an external prior on $H_{0}$ two times more accurate than CMB S4 (i.e.
 $1\%$) can improve the CMB constraint on $w$ by a factor of two.
@@
 Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub-percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
-Furthermore, if the final design of CMB S4 does not include measurements on large scales ($\ell<50$), external constraints from 21 cm experiments or all-sky CMB experiments on the optical depth $\tau$ will become very important.% powerful.
+Furthermore, if the final design of CMB S4 does not include measurements on large scales ($\ell<50$), external constraints on the optical depth $\tau$, from 21 cm experiments or CMB experiments measuring the large
+scales, will become very important.
 \item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$.

### 2016-03-12 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
new images

@@ -187,7 +187,7 @@ label['scalar_nrun(1)'] = r'\alpha_{s}'
 fisher_inv = np.linalg.inv(fisher_mat)
 fisher_inplace = fisher_mat.copy()
 
-
+utils.print_resume_stat(fisher_mat,fid)
 # CYCLE ON PARAMETERS (KEYS HERE)
 
 for key_y in ['omnuh2']:
@@ -220,7 +220,10 @@ for key_y in ['omnuh2']:
         # normalize_y = new_sigma / sigma_just_CMB_y  # make the new sigma y relative.
         # plot
         # *93.14 * 1000. to go to M_nu
-        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        # line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
+
+        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     plt.title(r'$\rm S4~ + ~BAO15$')
@@ -233,7 +236,7 @@ for key_y in ['omnuh2']:
     # ax1.set_xlim((0.1, 3.1))
     ax1.axhline(16.94 / 100. * fid['omnuh2'] * 94. * 1000., xmin=0., xmax=0.35, alpha=0.4, linewidth=2, label='DESI')
     ax1.set_ylabel(r'$\sigma(\sum m_\nu) $ meV')
-    ax1.set_xlabel(r'$\rm{External ~ prior~on~(\%)}$')
+    ax1.set_xlabel(r'$\rm{External ~ prior~ (\%)}$')
 
     vals = ax1.get_xticks()
     ax1.set_xticklabels([r'{:3.1f}$\%$'.format(x) for x in vals])
@@ -261,7 +264,7 @@ for key_y in ['omnuh2']:
     # ============================================
 
     # ============================================
[…]

### 2016-03-12 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_new.py

@@ -219,7 +219,9 @@ for key_y in ['omnuh2']:
         # normalize_y = new_sigma / sigma_just_CMB_y  # make the new sigma y relative.
         # plot
         # *93.14 * 1000. to go to M_nu
-        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        # line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
+        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     ax1.minorticks_on()
@@ -264,7 +266,7 @@ for key_y in ['omnuh2']:
     # ============================================
 
     # ============================================
-    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_{}_notau_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y),lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_omnuh2_l50.pdf'.format(str(key_y),lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                 format='pdf', bbox_inches='tight')
     plt.clf()
 

### 2016-03-12 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_tau.py

@@ -219,7 +219,9 @@ for key_y in ['omnuh2']:
         # normalize_y = new_sigma / sigma_just_CMB_y  # make the new sigma y relative.
         # plot
         # *93.14 * 1000. to go to M_nu
-        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        # line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
+        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})/{0}$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     ax1.minorticks_on()

### 2016-03-12 n_priors_code/plotting/plot_prior_paper_N_eff_2016.py

@@ -218,7 +218,10 @@ for key_y in ['massless_neutrinos']:
         # normalize_y = new_sigma / sigma_just_CMB_y  # make the new sigma y relative.
         # plot
 
-        line_plot = ax1.plot(prior_value*100., new_sigma, label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        # line_plot = ax1.plot(prior_value*100., new_sigma, label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
+        #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
+
+        line_plot = ax1.plot(prior_value*100., new_sigma , label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
         # new_sigma_all = utils.return_simgax_all_prior(fid, fisher_mat, key_y)
@@ -268,7 +271,7 @@ for key_y in ['massless_neutrinos']:
     # ============================================
 
     # ============================================
-    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_{}_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+    plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_massless_neutrinos.pdf'.format(str(key_y), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                 format='pdf', bbox_inches='tight')
     plt.clf()
 

### 2016-03-12 n_priors_code/plotting/plot_prior_paper_w_new.py

@@ -220,9 +220,9 @@ for  key_y in ['w']:
         # line_plot = ax1.plot(prior_value*100., new_sigma, label=r'$\sigma({0})={1:.1f}\%$'.format(
         #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
-        line_plot = ax1.plot(prior_value*100., new_sigma, label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
-            str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
+        line_plot = ax1.plot(prior_value*100., new_sigma , label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
+            str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
 
         vals = ax1.get_xticks()
@@ -259,7 +259,7 @@ for  key_y in ['w']:
         # ============================================
 
         # ============================================
-        plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_{}_{}_lmin={}_lmax={}_ndet={}_fsky={}.pdf'.format(str(key_y),str(key), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
+        plt.savefig('/home/manzotti/n_eff-dependence-on-prior/Notes/images/prior_w_hubble.pdf'.format(str(key_y),str(key), lmin, lmax, N_det, fsky), dpi=400, papertype='Letter',
                     format='pdf', bbox_inches='tight')
         plt.clf()
 

### 2016-03-12 Notes/CMBS4_priors_paper.tex

@@
 \caption{Projected constraints on the sum of the neutrino masses (absolute on the left and relative on the right) from CMB-S4 and current BAO measurements as a function of priors on the other cosmological parameters.
-The {\it internal} (marginalized constraints from CMB-S4 and current BAO) constraints on each parameter are given as $\sigma_{\rm pipeline}$.
+For comparison, the {\it internal} (marginalized constraints from CMB-S4 and current BAO) constraints on each parameter are:
+$\sigma_{\rm pipeline}(H_{0})/H_{0}=0.6\%$, 
+$\sigma_{\rm pipeline}(\Omega_{c}h^{2})/\Omega_{c}h^{2}=0.4\%$,
+$\sigma_{\rm pipeline}(n_{s})/n_{s}=0.1\%$,
+$\sigma_{\rm pipeline}(A_{s})/A_{s}=0.4\%$. 
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
@@
 \caption{Same as Fig.~1, but with $l<50$ modes excluded from CMB-S4, instead coming from Planck.
-Note the importance of obtaining information on the optical depth $\tau$ if the low-$l$ modes are not measured accurately.}
+Note the importance of obtaining information on the optical depth $\tau$ if the low-$l$ modes are not measured accurately.
+For comparison, the {\it internal} (marginalized constraints from CMB-S4 and current BAO) constraints on each parameter are:  
+$\sigma_{\rm pipeline}(H_{0})/H_{0}=0.6\%$, 
+$\sigma_{\rm pipeline}(\Omega_{c}h^{2})/\Omega_{c}h^{2}=0.4\%$,
+$\sigma_{\rm pipeline}(n_{s})/n_{s}=0.1\%$,
+$\sigma_{\rm pipeline}(A_{s})/A_{s}=0.4\%$,
+$\sigma_{\rm pipeline}(\tau)/\tau=5\%$.}
 \label{fig:prior-50}
@@
 \caption{Constraints on $N_{\rm{eff}}$ (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega _{b} h^{2}$, $\Omega _{c} h^{2}$ and $n_{s}$. 
+For comparison, the {\it internal} (marginalized constraints from CMB-S4 and current BAO) constraints on each parameter are:  
+$\sigma_{\rm pipeline}(\Omega_{c}h^{2})/\Omega_{c}h^{2}=0.4\%$,
+$\sigma_{\rm pipeline}(n_{s})/n_{s}=0.1\%$,
+$\sigma_{\rm pipeline}(\Omega_{b}h^{b})/\Omega_{c}h^{2}=0.1$,
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
@@
 \caption{Constraints on $w$ (absolute on the left and and relative to its value on the right) as a function of priors on $H_{0}$. 
+For comparison, the {\it internal} (marginalized constraints from CMB-S4 and current BAO) constraints on $H_{0}$ when w is free to vary is  
+$\sigma_{\rm pipeline}(H_{0})/H_{0}=1.9\%$.
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}

### 2016-03-14 n_priors_code/plotting/plot_prior_paper_M_nu_ell4_new.py
still undecided about new images

@@ -187,7 +187,7 @@ label['scalar_nrun(1)'] = r'\alpha_{s}'
 fisher_inv = np.linalg.inv(fisher_mat)
 fisher_inplace = fisher_mat.copy()
 
-utils.print_resume_stat(fisher_mat,fid)
+utils.print_resume_stat(fisher_mat, fid)
 # CYCLE ON PARAMETERS (KEYS HERE)
 
 for key_y in ['omnuh2']:
@@ -200,7 +200,7 @@ for key_y in ['omnuh2']:
     lines = ["-", "--", "-.", ":"]
     linecycler = cycle(lines)
 
-    for i, key in enumerate(['hubble','omch2','scalar_spectral_index(1)','scalar_amp(1)']):
+    for i, key in enumerate(['hubble', 'omch2', 'scalar_spectral_index(1)', 'scalar_amp(1)']):
 
         print key
 
@@ -222,14 +222,18 @@ for key_y in ['omnuh2']:
         # *93.14 * 1000. to go to M_nu
         # line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
         #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
+        linestyle = next(linecycler)
+        color = next(colorcycler)
 
-        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
-            str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
+        # ax1.axvline(x=np.abs(sigma_just_CMB_x * 100.), ymin=0.9, linestyle=linestyle, color=color,alpha=0.8)
+
+        line_plot = ax1.plot(prior_value * 100., new_sigma * 94. * 1000., label=r'${0}$'.format(
+            str(label[key]), np.abs(sigma_just_CMB_x * 100.)), color=color, linestyle=linestyle)
 
     plt.title(r'$\rm S4~ + ~BAO15$')
 
     ax1.minorticks_on()
-    low_lim = np.min(0.8 * np.amin(new_sigma),0.8 * 16.94 / 100. * fid['omnuh2'])
+    low_lim = np.min(0.8 * np.amin(new_sigma), 0.8 * 16.94 / 100. * fid['omnuh2'])
[…]

### 2016-03-14 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_new.py

@@ -221,7 +221,7 @@ for key_y in ['omnuh2']:
         # *93.14 * 1000. to go to M_nu
         # line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
         #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
-        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
+        line_plot = ax1.plot(prior_value*100., new_sigma * 94. * 1000., label=r'${0}$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
     ax1.minorticks_on()

### 2016-03-14 n_priors_code/plotting/plot_prior_paper_N_eff_2016.py

@@ -221,7 +221,7 @@ for key_y in ['massless_neutrinos']:
         # line_plot = ax1.plot(prior_value*100., new_sigma, label=r'$\sigma_{{\rm{{pipeline}}}}({0})={1:.1f}\%$'.format(
         #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
-        line_plot = ax1.plot(prior_value*100., new_sigma , label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
+        line_plot = ax1.plot(prior_value*100., new_sigma , label=r'${0}$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
         # new_sigma_all = utils.return_simgax_all_prior(fid, fisher_mat, key_y)

### 2016-03-14 n_priors_code/plotting/plot_prior_paper_w_new.py

@@ -221,7 +221,7 @@ for  key_y in ['w']:
         #     str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
 
-        line_plot = ax1.plot(prior_value*100., new_sigma , label=r'$\sigma_{{\rm{{pipeline}}}}({0})$'.format(
+        line_plot = ax1.plot(prior_value*100., new_sigma , label=r'${0}$'.format(
             str(label[key]), np.abs(sigma_just_CMB_x * 100.)), linestyle=next(linecycler))
 
 
@@ -241,7 +241,7 @@ for  key_y in ['w']:
         ax1.legend(loc=0)
         # ax1.set_title(r'$\sigma({0})={1:.1f}\%$'.format(str(label[key_y]), np.abs(sigma_just_CMB_y / fid[key_y] * 100.)))
         ax1.set_ylabel(r'$\sigma(' + label[key_y] + r')$')
-        ax1.set_xlabel(r'$\rm{External ~ prior~on~'+ label[key] +'(\%)}$')
+        ax1.set_xlabel(r'$\rm{External ~ prior~ (\%)}$')
         y1, y2 = ax1.get_ylim()
         ax2 = ax1.twinx()
         minor_loc = ax1.yaxis.get_minor_locator()

### 2016-03-14 n_priors_code/plotting/plot_prior_paper_M_nu_ell50_new.py
still undecided about new images

@@ -255,10 +255,10 @@ for key_y in ['omnuh2']:
 
     # Put snow mass line
     # Put a legend below current axis
-    # ax1.legend(loc=0,fancybox=True)
+    ax1.legend(loc=0,fancybox=True)
         # Put a legend below current axis
-    ax1.legend(loc='upper center', bbox_to_anchor=(0.5, -0.25),
-              fancybox=True, shadow=True, ncol=2)
+    # ax1.legend(loc='upper center', bbox_to_anchor=(0.5, -0.25),
+    #           fancybox=True, shadow=True, ncol=2)
     ax1.grid(True, alpha=0.4, linewidth=0.01)
     # ============================================
     # FINALLY SAVE

### 2016-03-14 Notes/CMBS4_priors_paper.tex
answer to referee

@@
 $\sigma_{\rm pipeline}(n_{s})/n_{s}=0.1\%$,
-$\sigma_{\rm pipeline}(A_{s})/A_{s}=0.4\%$,
+$\sigma_{\rm pipeline}(A_{s})/A_{s}=0.9\%$,
 $\sigma_{\rm pipeline}(\tau)/\tau=5\%$.}
@@
 $\sigma_{\rm pipeline}(n_{s})/n_{s}=0.1\%$,
-$\sigma_{\rm pipeline}(\Omega_{b}h^{b})/\Omega_{c}h^{2}=0.1$,
+$\sigma_{\rm pipeline}(\Omega_{b}h^{b})/\Omega_{c}h^{2}=0.1\%$,
 The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
